### 2019-11-08 paper/paper.tex
PER working, also included a new update of qblade where pitch works

\documentclass[hyperref,german,beleg]{cgvpub}

\usepackage{acro}
\input{abbreviations.tex}

%weitere Optionen zum Ergänzen (in eckigen Klammern):
% 
% female	weibliche Titelbezeichnung bei Diplom
% bibnum	numerische Literaturschlüssel
% final 	für Abgabe	
% lof			Abbildungsverzeichis
% lot			Tabellenverzeichnis
% noproblem	keine Aufgabenstellung
% notoc			kein Inhaltsverzeichnis
% twoside		zweiseitig
\author{Nico Westerbeck}
\title{Reinforcement-Learning Windturbine Controller}
\birthday{10. Januar 1994}
\placeofbirth{Bielefeld}
\matno{3951488}

\betreuer{Dr. Dmitrij Schlesinger}
\bibfiles{literatur}
\problem{
  \section{Background:}
  Here at the HFI Experimental fluid mechanics group, we have developed an
  open source project called QBlade. QBlade is a simulation tool used for
  testing wind turbines in the hostile environment that they normally operate.
  We normally tackle problems of aerodynamic or structural optimization but
  we have also a research focus on the development of the control systems of
  the wind turbines. We currently have a research effort looking at developing
  cluster-based controllers building on the work of Professor Bernd Noack who
  is a guest professor at our group. In the last year or so (Nair, A. G., Yeh, C.-
  A., Kaiser, E., Noack, B. R., Brunton, S. L., \& Taira, K. (2018). Cluster-based
  feedback control of turbulent post-stall separated flows. Journal of Physics
  Fluid Dynamics, (M), 1-32. Retrieved from http://arxiv.org/abs/1809.07220).
  AI projects such as openAI have enabled he rapid development of neural
  network in the field of control using reinforcement learning. The goal of this
  project is to use QBlade as a wind turbine simulator and attempt to control
  the pitch and rotor speed in a way that doesn’t cause the wind turbine to
  shatter but instead to yield energy, i.e. reward and death condition. This first
  stage of work should be considered as exploratory but will hopefully open up
  avenues of controlling active flow control elements such as flaps.

  \section{Tasks}
  The major tasks of the project are as follows:
  \begin{itemize}
    \item Build up and interface between QBlade and python the model code so that an external code can run as a controller within a QBlade simulation.
    \item Gain a rough understanding of the mechanics of wind turbines and their controllers.
    \item Research reinforcement learning methods suitable for use as a windturbine controller and perform a literature review on these approaches.
    \item Create a reinforcement learning agent which uses the Qblade interface for controllers to control a windturbine. 
      \begin{itemize}
        \item Inputs to the agent could be defined by the standartized controller input format to Nordex turbines, which consists of 39 real-valued sensor-inputs. However, initial tests can be conducted with whichever inputs are easiest to tackle. If required, further hidden state from the simulation can be exported to enrich data quality. If aiming for industrial quality, more inputs and also sensor faults could be optionally incorporated.
        \item Outputs are in a minimum version pitch angles for the 3 blades and turbine torque. Optionally the agent should be able to control active element such as flaps on the blades.
      \end{itemize}
    \item Optimize the agent to deliver maximum energy yield.
    \item Optimize under respect of certain boundary conditions (maximum pitch acceleration, maximum power, maximum blade load, blade touching the tower) and optionally other boundary conditions like long term turbine wear.
    \item If necessary for the training process, scale the simulation to run at a larger scale.
    \item Implement and attempt to get the agent to perform something close to sensible control of the wind turbine. Optionally evaluate the results against existing controllers and try to outperform them.
    \item Optionally, create a conference paper, poster or blog post etc.. on the results. 
  \end{itemize}
}

\copyrighterklaerung{Hier soll jeder Autor die von ihm eingeholten
Zustimmungen der Copyright-Besitzer angeben bzw. die in Web Press
Rooms angegebenen generellen Konditionen seiner Text- und
Bild"ubernahmen zitieren.}
\acknowledgments{Die Danksagung...}

\begin{document}
\chapter{Intro}

Recent advancements in reinforcment learning have managed to tackle more and more complex problems, like StarCraft or Go. The range of topics, RL is applicable to, increases. However, so far, there hasn't been an effort to control a windturbine with a reinforcement learning algorithm. Current state of the art linear controllers are performing well in maintaining turbine control, and the optimization margin for these is small, as they achieve the theoretical maximum energy output up to rated windspeed and manage to keep the turbine intact above rated windspeeds.

Trying it nontheless can still provide for some foundational knowledge for future developments. Park control becomes a bigger topic in wind-energy, where controlling all windturbines of a park together can increase overall energy yield or protect windturbines further down the windstream from wakes by turbines up front. C

\chapter{Background}

This section aims at facilitating the background knowledge necessary for understanding this paper. We fill first introduce windturbines and common terms around that

\section{Windturbine control}
% Quote wind-energy handbook, talk about how currently wind turbines are controlled.
There are two mayor types of wind turbine designs, \ac{HAWT} and \ac{VAWT}, which differ by their rotation axis. In this work, we will only look at \ac{HAWT} and all mentions of windturbines mean \ac{HAWT} type turbines. Such a turbine is made up by 3 big components, a tower on which a nacelle is rested which in turn has a rotor in front. 
The joint between tower and nacelle allows for rotation to turn the rotor into the wind. Except for recent ideas \quote{Howland_Lele_Dabiri_2019}, the best value for the rotational direction was to always turn the rotor directly into the wind, and for the sake of simplicity we will omit this control parameter from our simulation.
In the nacelle there is, most importantly, an electrical generator, which can be given a specific torque. It will then slow down the rotor with that torque and at the same time generate energy proportional to rotor speed and torque.

\section{QBlade}
% Quickly introduce qblade as a simulation tool
Our data source in this work is the open-source simulation tool QBlade \quote{Marten_Wendler_Pechlivanoglou_Nayeri_Paschereit_2013} developed at TU Berlin. QBlade is an accessible and performant tool with the primary purpose of designing and simulating wind turbines in a graphical user interface. Its simulation results are on par with current state-of-the-art proprietary simulation tools and it yields good computational efficiency. To be used in reinforcement learning, we designed an interface, over which the QBlade simulation can be embedded into an external environment. Most machine-learning frameworks are written in python, so we decided to compile QBlade into library format and expose the most fundamental functions to allow it to communicate with any programming language that can load libraries. As python has a ctypes interface to load c-code, we could link a python agent to the QBlade C++ environment.

QBlade allows different simulation scenarios, for our testing we decided to only use the NREL 5MW \quote{Jonkman_Butterfield_Musial_Scott_2009} turbine with the default structural model and using all implemented correction mechanisms to achieve the most realistic data possible. As reference data to this turbine is easily available and publicly published by NREL, this allows us for good cross-validation.

\section{Reinforcement learning}
%

\section{RL on windturbines}

\end{document}

### 2019-11-08 python/ddpg/ddpg.py

@@ -173,28 +173,27 @@ class DDPG:
     def get_action(self, obs, add_noise=False):
         action = None
         # Do some random exploration at the beginning
-        if(self.time < self.hparams.random_exploration_steps and add_noise):
-            action = self.denormalize_action(self.random_exploration_noise.get_noise(self.time), False)
-        else:
-            # Send the observation to the device, normalize it
-            # Then calculate the action and denormalize it
-            state = torch.FloatTensor(obs).unsqueeze(0).to(self.device)
 
-            if (self.hparams.normalize_observations):
-                state = self.normalize_state(state, True)
-            
-            # Run the network
-            action = self.actor.forward(state)
+        # Send the observation to the device, normalize it
+        # Then calculate the action and denormalize it
+        state = torch.FloatTensor(obs).unsqueeze(0).to(self.device)
 
-            # Add noise
+        if (self.hparams.normalize_observations):
+            state = self.normalize_state(state, True)
+        
+        # Run the network
+        action = self.actor.forward(state)
+
+        # Add noise
+        if(add_noise):
             noise = torch.FloatTensor(self.action_noise.get_noise(self.time)).to(self.device)
             action = action + noise
-            
-            if(self.hparams.normalize_actions):
-                action = self.denormalize_action(action, True)
-            
-            # Get it to the cpu
-            action = action.squeeze(0).cpu().detach().numpy()
+        
+        if(self.hparams.normalize_actions):
+            action = self.denormalize_action(action, True)
+        
+        # Get it to the cpu
+        action = action.squeeze(0).cpu().detach().numpy()
 
         action = np.clip(action, self.act_low, self.act_high)
 
@@ -345,21 +344,40 @@ class DDPG:
             self.replay_buffer.update_priorities(indices, td_error)
 
         if(self.hparams.log_net_insights and time % self.hparams.log_net_insights == 0):
-            critic_state = dict(self.critic.cpu().named_parameters())
-            self.logger.add_histogram_nofilter('critic-weights/linear1', critic_state['linear1.weight'], time)
-            self.logger.add_histogram_nofilter('critic-weights/linear2', critic_state['linear2.weight'], time)
[…]

### 2019-11-08 python/ddpg/models.py

@@ -13,11 +13,17 @@ class Critic(nn.Module):
         if(self.simple):
             self.linear1 = nn.Linear(self.obs_dim + self.action_dim, hidden_size_1)
             self.linear2 = nn.Linear(hidden_size_1, 1)
+            print('Initialized Critic with %d, %d, %d' % (self.obs_dim + self.action_dim, hidden_size_1, 1))
         else:
             self.linear1 = nn.Linear(self.obs_dim, hidden_size_1)
             self.linear2 = nn.Linear(hidden_size_1 + self.action_dim, hidden_size_2)
             self.linear3 = nn.Linear(hidden_size_2, hidden_size_3)
             self.linear4 = nn.Linear(hidden_size_3, 1)
+            print('Initialized Critic with %d, %d, %d, %d, %d' % (self.obs_dim + self.action_dim,
+                                                              hidden_size_1,
+                                                              hidden_size_2,
+                                                              hidden_size_3,
+                                                              1))
 
         self.init_weights(init_w)
 
@@ -28,6 +34,12 @@ class Critic(nn.Module):
         if(not self.simple):
             self.linear3.weight.data.uniform_(-init_w, init_w)
             self.linear4.weight.data.uniform_(-init_w, init_w)
+
+    def first_layer(self, x, a):
+        if(self.simple):
+            return F.relu(self.linear1(torch.cat([x, a], 1)))
+        else:
+            return F.relu(self.linear1(x))
 
     def forward_complex(self, x, a):
         xa = F.relu(self.linear1(x))
@@ -62,10 +74,12 @@ class Actor(nn.Module):
         if(self.simple):
             self.linear1 = nn.Linear(self.obs_dim, hidden_size_1)
             self.linear2 = nn.Linear(hidden_size_1, self.action_dim)
+            print("Initialized actor with %d, %d, %d" % (self.obs_dim, hidden_size_1, self.action_dim))
         else:
             self.linear1 = nn.Linear(self.obs_dim, hidden_size_1)
             self.linear2 = nn.Linear(hidden_size_1, hidden_size_2)
             self.linear3 = nn.Linear(hidden_size_2, self.action_dim)
+            print("Initialized actor with %d, %d, %d, %d" % (self.obs_dim, hidden_size_1, hidden_size_2, self.action_dim))
 
         self.init_weights(init_w)
 
@@ -74,6 +88,9 @@ class Actor(nn.Module):
[…]

### 2019-11-08 python/ddpg/utils.py

@@ -330,11 +330,6 @@ class BasicBuffer:
             print("Could not load memory: %s" % str(e))
 
 
-x = BasicBuffer(10)
-x.push(1,1,1,1,1, 1)
-x.push(2,2,2,2,2, 0.5)
-x.push(3,3,3,3,3, 0.1)
-
 class QBladeLogger:
 
     def __init__(self, logdir, log_steps):
@@ -343,29 +338,34 @@ class QBladeLogger:
         self.writer = SummaryWriter(logdir + '/' + str(datetime.now()))
 
     def logObservation(self, step, observation, prefix='obs'):
-        self.add_scalar('%s/rotational speed [rad/s]' % prefix, observation[0], step)
-        self.add_scalar('%s/power [W]' % prefix, observation[1], step)
-        self.add_scalar('%s/HH wind velocity [m/s]' % prefix, observation[2], step)
-        self.add_scalar('%s/yaw angle [deg]' % prefix, observation[3], step)
-        self.add_scalar('%s/pitch blade 1 [deg]' % prefix, observation[4], step)
-        self.add_scalar('%s/pitch blade 2 [deg]' % prefix, observation[5], step)
-        self.add_scalar('%s/pitch blade 3 [deg]' % prefix, observation[6], step)
-        self.add_scalar('%s/tower top bending local x [Nm]' % prefix, observation[7], step)
-        self.add_scalar('%s/tower top bending local y [Nm]' % prefix, observation[8], step)
-        self.add_scalar('%s/tower top bending local z [Nm]' % prefix, observation[9], step)
-        self.add_scalar('%s/oop bending blade 1 [Nm]' % prefix, observation[10], step)
-        self.add_scalar('%s/oop bending blade 2 [Nm]' % prefix, observation[11], step)
-        self.add_scalar('%s/oop bending blade 3 [Nm]' % prefix, observation[12], step)
-        self.add_scalar('%s/ip bending blade 1 [Nm]' % prefix, observation[13], step)
-        self.add_scalar('%s/ip bending blade 2 [Nm]' % prefix, observation[14], step)
-        self.add_scalar('%s/ip bending blade 3 [Nm]' % prefix, observation[15], step)
-        self.add_scalar('%s/oop tip deflection blade 1 [m]' % prefix, observation[16], step)
-        self.add_scalar('%s/oop tip deflection blade 2 [m]' % prefix, observation[17], step)
[…]

### 2019-11-08 python/env/qbladeadapter.py

@@ -35,7 +35,8 @@ class QBladeAdapter:
 
     print("creating qblade instance")
     self._qbladeLib._Z14createInstancev()
-    x = ctypes.c_char_p(b"../sample_projects/NREL_5MW_STR.wpa")
+
+    x = ctypes.c_char_p(b"../sample_projects/NREL_5MW_STR_NEW.wpa")
     print("loading qblade project")
     self._loadProject(x)
 
@@ -67,9 +68,10 @@ class QBladeAdapter:
   def reset(self):
     print('Resetting simulation')
     self._initializeSimulation(ctypes.c_int(0))
+    self.lastAction = np.zeros(5)
     print('reset done')
 
-    return self.extractObservation()
+    return self.maskObservation(self.extractObservation())
 
   def get_obs_dim(self):
     #return 23
@@ -90,24 +92,24 @@ class QBladeAdapter:
     return [0, 0]
 
   def calc_reward(self, observation, action):
-    rated_power = 1e3
-    rated_speed = 1.1
+    rated_power = 3200
+    rated_speed = 0.8
     #return -np.abs(observation[1]-rated_power) - 1e3*(np.abs(observation[16]) + np.abs(observation[17]) + np.abs(observation[18]))
     #return np.clip(observation[1], 0, None) - 1e4*(np.abs(observation[16]) + np.abs(observation[17]) + np.abs(observation[18]))
-    return np.clip(5 
-      - np.abs((observation[1]-rated_power)/rated_power)
-      - 5e-2*(np.abs(observation[16]) + np.abs(observation[17]) + np.abs(observation[18])), -50, 50)
-    #return -np.abs(observation[0]-rated_speed)
+    #return np.clip(5 
+    #  - np.abs((observation[1]-rated_power)/rated_power)
+    #  - 5e-2*(np.abs(observation[16]) + np.abs(observation[17]) + np.abs(observation[18])), -10, 10)
+    return -np.abs(observation[0]-rated_speed)
 
   def calc_death(self, observation):
 
     # 63 is the rotor size, so if anything bends further than that, it's broken off.
     broken_state = 20
-    broken_rpm = 3
     return np.abs(observation[16]) > broken_state or \
            np.abs(observation[17]) > broken_state or \
            np.abs(observation[18]) > broken_state or \
-           np.abs(observation[0]) > broken_rpm
+           observation[0] > 4 or \
+           observation[0] < -0.5
 
   def padAction(self, action):
     action = [action[0], 0, action[1], action[1], action[1]]
[…]

### 2019-11-08 python/automated_hparam_search.py
Implemented automated hyperparameter searching

@@ -1,0 +1,45 @@
+from multiprocessing import Pool
+import subprocess
+import json
+import os
+
+experiments = [
+  {"id": 0, "batch_size": 64},
+  {"id": 1, "batch_size": 100},
+  {"id": 2, "batch_size": 128},
+  {"id": 3, "batch_size": 150},
+  {"id": 4, "batch_size": 200},
+  {"id": 5, "batch_size": 256},
+]
+
+
+def run_one(experiment):
+  id = experiment['id']
+  del experiment['id']
+
+  test_output = 'output_%d' % id
+  checkpoint_dir = 'checkpoints/autorun_%d' % id
+
+  experiment['test_output'] = test_output
+  experiment['checkpoint_dir'] = checkpoint_dir
+
+  os.makedirs(checkpoint_dir, exist_ok=True)
+
+  hparams = ','.join([ ('%s=%s' % (key, val)) for key, val in experiment.items() ])
+
+  print('Launching experiment %d with hparams %s' % (id, hparams))
+
+  handle = subprocess.run('xvfb-run -a python main.py --hparams=%s' % hparams, shell=True)
+  handle.wait()
+
+  with open(test_output, 'r') as f:
+    data = json.loads(f.read())
+
+  return (id, data['total_reward'], data['total_deaths'])
+
+# Set the number of parallel runs here
+pool = Pool(3)
+results = pool.map(run_one, experiments)
+
+for (id, reward, deaths) in results:
+  print("Run %d, reward: %f, deaths: %d" % (id, reward, deaths))

### 2019-11-08 python/ddpg/ddpg.py

@@ -600,7 +600,10 @@ class DDPG:
             epochs = 2000,
 
             # Number of steps to run after the training, testing the policy
-            test_steps = 1000,
+            test_steps = 50000,
+
+            # Whether, and where to write test results to
+            test_output = 'test_output',
 
             # Batch size for experience replay
             # The bigger the less it will overfit to specific samples
@@ -625,20 +628,20 @@ class DDPG:
             critic_lr = 1e-4,
 
             # Neural network sizes of the critic
-            critic_sizes = [512, 128, 64],
+            critic_sizes = [64, 32, 16],
 
             # Whether to use a 4-layer or a 2-layer structure
             # for the critic
-            critic_simple = False,
+            critic_simple = True,
 
             # Learning rate of the policy
             actor_lr = 1e-4,
 
             # Network sizes of the policy
-            actor_sizes = [128, 32],
+            actor_sizes = [32, 8],
 
             # Whether to use a 2-layer or a 3-layer structure for the actor
-            actor_simple = False,
+            actor_simple = True,
 
             # Network parameters of both the actor and critic will be initialized to
             # a uniform random value between [-init_weight_limit, init_weight_limit]

### 2019-11-08 python/main.py

@@ -82,6 +82,36 @@ def main():
       o = env.reset()
       a = agent.reset_finalize(o)
 
+  # Run a test run after training if wanted
+  if(hparams.test_steps):
+    print("Starting test run")
+    o = env.reset()
+    a = agent.prepare(o)
+    total_reward = 0
+    total_deaths = 0
+
+    for t in range(0, hparams.test_steps):
+      o, r, d = env.step(a)
+
+      total_reward += r
+
+      if(d):
+        total_deaths += 1
+        o = env.reset()
+
+      a = agent.get_action(o)
+
+    print("Test results:")
+    print("Total reward: %d" % total_reward)
+    print("Total deaths: %d" % total_deaths)
+
+    if(hparams.test_output):
+      with open(hparams.test_output, 'w') as f:
+        f.write(json.dumps({
+          "total_reward": total_reward,
+          "total_deaths": total_deaths,
+          "hparams": hparams.values()
+        }))
 
 
   env.storeProject("checkpoints/sampleproject.wpa")

### 2019-11-11 paper/paper.tex
Continued on the paper, improved automated hparam search

@@
 \usepackage{acro}
+\usepackage{siunitx}
 \input{abbreviations.tex}
@@
 \begin{document}
+\chapter{Abstract}
+Small
+
 \chapter{Intro}
@@
 In this work, we will only look at \ac{HAWT} and all mentions of windturbines mean \ac{HAWT} type turbines.
-Such a turbine is made up by 3 big components, a tower on which a nacelle is rested which in turn has a rotor in front.
+Such a turbine is made up by 3 big components, a tower on which a nacelle is rested which itself has a rotor in front.
 The joint between tower and nacelle allows for rotation to turn the rotor into the wind.
-Except for recent ideas \quote{Howland_Lele_Dabiri_2019}, the best value for the rotational direction was to always turn the rotor directly into the wind, and for the sake of simplicity we will omit this control parameter from our simulation.
+Except for recent ideas \cite{Howland_Lele_Dabiri_2019}, the best value for the rotational direction was to always yaw the rotor directly into the wind, and for the sake of simplicity we will omit this control parameter from our simulation and leave it at 0 degrees with wind facing straight onto the rotor.
 In the nacelle there is, most importantly, an electrical generator, which can be given a specific torque.
 It will then slow down the rotor with that torque and at the same time generate energy proportional to rotor speed and torque.
+This torque is one of the two more important control parameters, together with blade pitch.
+The blades are designed to operate on maximal aerodynamic efficiency when they are pitched at 0 degrees, increasing the pitch angle will (except for possible slight improvements in the first few degrees) reduce aerodynamic efficiency of the rotor.
+
+As described in \cite[sec 8.3]{Burton_Jenkins_Sharpe_Bossanyi_2011}, the normal controller design for windturbines divides the operation of a turbine into two parts, below rated and above rated.
+Rated is a windspeed, in which the turbine gives maximum energy on highest aerodynamic effiency.
+At this point, the generator runs on its maximum torque and the blades are pitched to the optimum angle of 0 degrees.
+At this point, the tip-speed-ratio, which is a good indicator for loads on the blades is at maximum too.
+Below that windspeed, the rotor generates less rotational torque than the maximum generator torque, thus the generator torque needs to be reduced to not slow the rotor down too much.
+Slowing the rotor down complicates the energy generation process, and though variable-speed wind turbines allow for some play, all windturbines are limited to a certain operational range.
+Fixed-speed wind turbines can only run on one single rotational speed and thus need to change the torque more agressively.
+Above rated windspeed, the blades need to be pitched out to stop the rotor from spinning at faster speeds than what the turbine was designed for.
+The generator torque can be kept at maximum in this area, but to protect the turbine from damage, aerodynamic efficiency of the rotor is reduced by turning the blades along their long axis and thus moving their angle of attack against the incoming air into a less optimal range.
+
+Additionally to this trivial control, optimizations to reduce vibrations and stress on the structures are implemented.
+The blades can be pitched individually and quickly enough to allow for different blade pitches during a single rotation of the rotor, which can be used to reduce the turbulence that hits the tower or to account for different windspeeds closer to the ground and further up in the air.
+Also, both generator torque and blade pitch can be used to counteract natural resonant frequencies of the structure, reducing material stress through extensive swinging of structural parts.
+
+Usually this control is implemented by two PID controllers, which are hand-designed as described in \cite[sec 8.4]{Burton_Jenkins_Sharpe_Bossanyi_2011}.
+Below rated windspeed, the controller for torque is active, above rated the one for pitch.
+The parameters to those controllers can be calculated according to the laws of control theory, and the resulting controllers work reasonably close to theoretical maximum.
+Still, in this work, we are trying to replace these hand-designed controllers with a reinforcement-learning algorithm.
 
@@
 % Quickly introduce qblade as a simulation tool
-Our data source in this work is the open-source simulation tool QBlade \quote{Marten_Wendler_Pechlivanoglou_Nayeri_Paschereit_2013} developed at TU Berlin.
+Our data source in this work is the open-source simulation tool QBlade \cite{Marten_Wendler_Pechlivanoglou_Nayeri_Paschereit_2013} developed at TU Berlin.
 QBlade is an accessible and performant tool with the primary purpose of designing and simulating wind turbines in a graphical user interface.
 Its simulation results are on par with current state-of-the-art proprietary simulation tools and it yields good computational efficiency.
+It uses an algorithm which approximates the 3D blade structures with 2D structures and corrects the results through several error terms.
+Though a full CFD simulation would yield higher accuracy, our computational resources don't allow for that.
 To be used in reinforcement learning, we designed an interface, over which the QBlade simulation can be embedded into an external environment.
@@
 
-QBlade allows different simulation scenarios, for our testing we decided to only use the NREL 5MW \quote{Jonkman_Butterfield_Musial_Scott_2009} turbine with the default structural model and using all implemented correction mechanisms to achieve the most realistic data possible.
+QBlade allows different simulation scenarios, for our testing we decided to only use the NREL 5MW \cite{Jonkman_Butterfield_Musial_Scott_2009} turbine with the default structural model and using all implemented correction mechanisms to achieve the most realistic data possible.
 As reference data to this turbine is easily available and publicly published by NREL, this allows us for good cross-validation.
 
+The interface is made of 7 functions, which allow for loading a project, resetting the simulation, setting controller inputs and getting environment measurements and advancing the simulation a timestep.
+The observations returned in getControlVars match those that are visible to standart industry controllers, which in turn are modelled after what can be measured on a real windturbine.
+Concretely, we measure:
+\begin{code}
+* rotational speed [rad/s]
+* power [kW]
+* wind velocity [m/s]
+* yaw angle [deg]
+* pitch blade 1 [deg]
+* pitch blade 2 [deg]
+* pitch blade 3 [deg]
+* tower top bending local x [Nm]
+* tower top bending local y [Nm]
+* tower top bending local z [Nm]
+* out-of-plane bending blade 1 [Nm]
+* out-of-plane bending blade 2 [Nm]
+* out-of-plane bending blade 3 [Nm]
+* in-plane bending blade 1 [Nm]
+* in-plane bending blade 2 [Nm]
+* in-plane bending blade 3 [Nm]
+* out-of-plane tip deflection blade 1 [m]
+* out-of-plane tip deflection blade 2 [m]
+* out-of-plane tip deflection blade 3 [m]
+* in-plane tip deflection blade 1 [m]
+* in-plane tip deflection blade 2 [m]
+* in-plane tip deflection blade 3 [m]
+* current time [s]
+\end{code}
+
+Because it has no semantic meaning, we mask away current time from our reinforcement learning agent. The inputs we can give to the turbine are:
+\begin{code}
+* generator torque [Nm]
+* yaw angle [deg]
+* pitch blade 1 [deg]
+* pitch blade 2 [deg]
+* pitch blade 3 [deg]
+\end{code}
+
+As mentioned before, we set the yaw angle to always 0 degrees, as in reality controlling the orientation of the nacelle is trivial.
+The generator torque on our NREL 5MW turbine has a sensible range of \num{0} to \num{4e6} Nm% TODO insert formula here
+Because changing torque from zero to maximum in one timestep is not realistic for a real turbine, we restrict it to move a maximum of \num{3e5} Nm per one timestep of 0.1 seconds.
+If our agent chooses a value outside of that area, we set it to the closest value inside of that area.
+Blades can be pitched between 0 and 90 degrees, 0 being not pitched out at all and operating at maximum efficiency and 90 resulting in no aerodynamic torque from the rotor whatsoever.
+A sensible pitch motor can only turn the blades at a limited speed, we assume a limit of 5 degrees per second and again set controller inputs to a sensible value inside that.
+The simulation theoretically accepts a full pitch change in one timestep, however as inertia on the blade is so high for such a change, the blades break instantly and the rest of the simulation needs to be reset.
+
+QBlade simulates blade damage, also high rotational speeds cause blades to fall off.
+QBlade is not validated to realistically simulate these extreme conditions and thus posed a problem to our controllers.
+In the course of our experiments, we decided to reset the simulation at an rotational speed of 4 rad/s, as our turbine operates at ca 0.8 rad/s.
+Also, high generator torque inputs cause the simulation to rotate the rotor backwards with a negative energy yield, effectively creating a multimillion-dollar leafblower.
+As this is neither a realistic scenario, we also reset the simulation at -0.5 rad/s.
+
 \section{Reinforcement learning}
-%
 
@@
 
+\chapter{Early experimentation}
+%Early experiments, taking a reference implementation and openai gym training. Then finding out qblade is screwed...
+
+\chapter{Experimentation}
+
+\section{Designing reward functions}
+
+\section{Action Noise}
+
+\section{Random exploration}
+
+\section{Parameter noise}
+
+\section{Replay noise}
+
+\section{Simplifying the architecture}
+
+\section{Normalization}
+
+\section{Gradient actionspace}
+
+\section{Feeding past timesteps}
+
+\section{Prioritized experience replay}
+
+\section{Automatized hyperparameter search}
+
+\chapter{Evaluation}
+
+% What should I put here?
+
+\chapter{Future work}
+
+% Reworking reward functions
+% Trying different algorithms for better stability?
+% Pretraining the policy with a reference controller
+% Whatever comes to our mind
+
+\chapter{Conclusion}
+
+% It's pretty hard to create a working controller for a windturbine, apparently RL needs some more foundational research.
+
 \end{document}

### 2019-11-11 python/automated_hparam_search.py

@@ -3,13 +3,24 @@ import subprocess
 import json
 import os
 
-experiments = [
+'''experiments = [
   {"id": 0, "batch_size": 64},
   {"id": 1, "batch_size": 100},
   {"id": 2, "batch_size": 128},
   {"id": 3, "batch_size": 150},
   {"id": 4, "batch_size": 200},
   {"id": 5, "batch_size": 256},
+]'''
+
+experiments = [
+  {"id": 6, "critic_lr": 1e-2},
+  {"id": 7, "critic_lr": 1e-3},
+  {"id": 8, "critic_lr": 1e-5},
+  {"id": 9, "critic_lr": 1e-6},
+  {"id": 10, "actor_lr": 1e-2},
+  {"id": 10, "actor_lr": 1e-3},
+  {"id": 10, "actor_lr": 1e-5},
+  {"id": 10, "actor_lr": 1e-6},
 ]
 
 
@@ -29,13 +40,16 @@ def run_one(experiment):
 
   print('Launching experiment %d with hparams %s' % (id, hparams))
 
-  handle = subprocess.run('xvfb-run -a python main.py --hparams=%s' % hparams, shell=True)
-  handle.wait()
+  try:
+    handle = subprocess.run('xvfb-run -a python main.py --hparams=%s' % hparams, shell=True)
+    handle.wait()
 
-  with open(test_output, 'r') as f:
-    data = json.loads(f.read())
+    with open(test_output, 'r') as f:
+      data = json.loads(f.read())
 
-  return (id, data['total_reward'], data['total_deaths'])
+    return (id, data['total_reward'], data['total_deaths'])
+  except:
+    return (id, -1, -1)
 
 # Set the number of parallel runs here
 pool = Pool(3)

### 2019-11-11 python/ddpg/ddpg.py

@@ -331,8 +331,10 @@ class DDPG:
         expected_Q = reward_batch + self.gamma * next_Q
         
         # update critic
-        # Idea: huber loss?
-        q_loss = F.mse_loss(curr_Q, expected_Q.detach())
+        if(self.hparams.critic_loss == 'huber'):
+            q_loss = F.smooth_l1_loss(curr_Q, expected_Q.detach())
+        else:
+            q_loss = F.mse_loss(curr_Q, expected_Q.detach())
 
         q_loss.backward() 
         self.critic_optimizer.step()
@@ -633,6 +635,11 @@ class DDPG:
             # Whether to use a 4-layer or a 2-layer structure
             # for the critic
             critic_simple = True,
+
+            # Loss function for the critic
+            # default is mean square error
+            # Other values: "mse", "huber"
+            critic_loss = 'huber',
 
             # Learning rate of the policy
             actor_lr = 1e-4,

### 2019-11-11 python/ddpg/utils.py

@@ -361,7 +361,7 @@ class QBladeLogger:
           19: 'ip tip deflection blade 1 [m]',
           20: 'ip tip deflection blade 2 [m]',
           21: 'ip tip deflection blade 3 [m]',
-          22: 'current time'
+          22: 'current time [s]'
         }
 
         for i in range(0, len(observation)):

### 2019-11-11 python/env/qbladeadapter.py

@@ -52,7 +52,7 @@ class QBladeAdapter:
     self._setControlVars = qbladeLib._Z14setControlVarsPd
     self._setControlVars.argtypes = [ctypes.POINTER(ctypes.c_double)]
     self._getControlVars = qbladeLib._Z14getControlVarsPd
-    self._setControlVars.argtypes = [ctypes.POINTER(ctypes.c_double)]
+    self._getControlVars.argtypes = [ctypes.POINTER(ctypes.c_double)]
     self._advanceSingleTimestep = qbladeLib._Z21advanceSingleTimestepv
 
   def del_functions(self):

### 2019-11-12 paper/paper.tex
Wrote several chapters in experimentation and a rl background subsection

@@
 \usepackage{siunitx}
-\input{abbreviations.tex}
+\usepackage{mathtools}
+\usepackage{commath}
+\input{abbreviations}
 
@@
 The joint between tower and nacelle allows for rotation to turn the rotor into the wind.
-Except for recent ideas \cite{Howland_Lele_Dabiri_2019}, the best value for the rotational direction was to always yaw the rotor directly into the wind, and for the sake of simplicity we will omit this control parameter from our simulation and leave it at 0 degrees with wind facing straight onto the rotor.
+Except for recent ideas \cite{howlandWindFarmPower2019}, the best value for the rotational direction was to always yaw the rotor directly into the wind, and for the sake of simplicity we will omit this control parameter from our simulation and leave it at 0 degrees with wind facing straight onto the rotor.
 In the nacelle there is, most importantly, an electrical generator, which can be given a specific torque.
@@
 
-As described in \cite[sec 8.3]{Burton_Jenkins_Sharpe_Bossanyi_2011}, the normal controller design for windturbines divides the operation of a turbine into two parts, below rated and above rated.
+As described in \cite[sec 8.3]{burtonWindEnergyHandbook2011}, the normal controller design for windturbines divides the operation of a turbine into two parts, below rated and above rated.
 Rated is a windspeed, in which the turbine gives maximum energy on highest aerodynamic effiency.
@@
 
-Usually this control is implemented by two PID controllers, which are hand-designed as described in \cite[sec 8.4]{Burton_Jenkins_Sharpe_Bossanyi_2011}.
+Usually this control is implemented by two PID controllers, which are hand-designed as described in \cite[sec 8.4]{burtonWindEnergyHandbook2011}.
 Below rated windspeed, the controller for torque is active, above rated the one for pitch.
 The parameters to those controllers can be calculated according to the laws of control theory, and the resulting controllers work reasonably close to theoretical maximum.
+% TODO find reference
  Still, in this work, we are trying to replace these hand-designed controllers with a reinforcement-learning algorithm.
@@
 % Quickly introduce qblade as a simulation tool
-Our data source in this work is the open-source simulation tool QBlade \cite{Marten_Wendler_Pechlivanoglou_Nayeri_Paschereit_2013} developed at TU Berlin.
+Our data source in this work is the open-source simulation tool QBlade \cite{martenQBLADEOPENSOURCE2013} developed at TU Berlin.
 QBlade is an accessible and performant tool with the primary purpose of designing and simulating wind turbines in a graphical user interface.
@@
 
-QBlade allows different simulation scenarios, for our testing we decided to only use the NREL 5MW \cite{Jonkman_Butterfield_Musial_Scott_2009} turbine with the default structural model and using all implemented correction mechanisms to achieve the most realistic data possible.
+QBlade allows different simulation scenarios, for our testing we decided to only use the NREL 5MW \cite{jonkmanDefinition5MWReference2009} turbine with the default structural model and using all implemented correction mechanisms to achieve the most realistic data possible.
 As reference data to this turbine is easily available and publicly published by NREL, this allows us for good cross-validation.
@@
 Concretely, we measure:
-\begin{code}
+\begin{verbatim}
 * rotational speed [rad/s]
@@
 * current time [s]
-\end{code}
+\end{verbatim}
 
 Because it has no semantic meaning, we mask away current time from our reinforcement learning agent. The inputs we can give to the turbine are:
-\begin{code}
+\begin{verbatim}
 * generator torque [Nm]
@@
 * pitch blade 3 [deg]
-\end{code}
+\end{verbatim}
 
@@
 
+This section should aim to give a brief overview into Reinforcement Learning assumptions and derive briefly a couple of methods until in the end, we will describe \ac{DDPG}, the method we utilized to control our virtual windturbine.
+
+\subsection{Environment assumptions}
+
+At first, we need to do some assumptions on our environment.
+The concept environment encapsulates the reality or simulation that the reinforcement learning algorithm works in.
+This could be an actual cyber-physical system or a simulation.
+The first assumption to this environment is that this system operates over time and we can discretize time into timesteps.
+Our environment here is the simulated wind turbine, but it could be a computer game \cite{lillicrapCONTINUOUSCONTROLDEEP2015}, a physics simulation\cite{brockmanOpenAIGym2016} or even an actual existing wind turbine \cite{kolterDesignAnalysisLearning2012}.
+In every timestep, the environment supplies us with a state, also sometimes called observation, and a scalar reward.
+The observed state doesn't have to represent the entirety of the state of the simulation, in fact it usually is only a small subset.
+For our wind turbine example, the observed state is limited to what can be measured with sensors on a wind turbine, in an arcade game it could be the pixel output or in a physical simulation it could be some key values in the simulation.
+The reward is a scalar value which judges the current state of the simulation.
+A higher reward means the current state is better than that of a lower reward.
+Sometimes, the reward function is obvious from the system, as in an arcade game it would surely be the score or in chess it could be 1 for win, -1 for lose and 0 for not decided yet.
+However in our paper, we will dedicate a section on how we designed our reward function.
+In general reinforcement learning theory, it is assumed to exist and be supplied by the environment.
+In every timestep, the agent supplies an action in the form of a tensor.
+This action could be button presses in an arcade game or blade pitch in a windturbine simulation.
+The mission of the agent is to pick actions which maximize cumulative reward.
+In other words, it should steer the environment to achieve best performance.
+Additionally to the state and reward, an environment can optionally send out a done signal, which indicates a state in which further simulation is not possible and the environment wants to be reset to an initial state.
+In a chess game, this would be a loss or a win, and with a windturbine that could for example be fatal structural damage.
+
+The algorithms we looked at further assume that the state progressions can be modelled as a Markov-Chain, as in that what happens at timestep t+1 is only dependent on the hidden state at timestep t, not also on for example t-10.
+As the hidden state could for example be the full RAM of the simulation, we can safely assume that a simulation only calculates the next timestep on what is in the RAM at the current timestep.
+Also, Markov-Chains allow for several states as an outcome of any state, and that we can model the transition between these state with a probablistic model.
+
+\subsection{Q-Learning}
+
+At first we will present a very basic algorithm for a machine learning agent, which is based on the basic Bellman-Equation \cite{bellmanTheoryDynamicProgramming1954}
+
+\begin{equation}
+Q(s, a) = r + \max_{a'}Q(s', a')
+\end{equation}
+
+% Bellman equation
+
+\subsection{PG}
+
+\subsection{DDPG}
+% Also problems with ddpg
+
 \section{RL on windturbines}
@@
 
+At the start of our experimentation, we took off with a well-known problem and a proven implementation.
+We used the OpenAI-Pendulum task, in which a pendulum in gravity needs to be held upright by applying a force to it.
+OpenAI Gyms generally serve as benchmarks for reinforcement learning algorithms and most foundational papers and articles evaluate their solutions based on how well they perform in OpenAI Gyms.
+Starting with a reference-implementation in pytorch by \cite{DeepDeterministicPolicy}, we experimented with the pendulum environment.
+We observed that the algorithm got stuck in a local optimum, where it rotates the pendulum instead of holding it upright, for long periods of the training process.
+Only with high noise, we could get it to hold a pendulum upright.
+We also tried to run the HalfCheetah task, where a two-dimensional dog-like being should be brought to running and moving its joints, but the results were more a crawling and shaking towards the goal.
+However, we hoped that the implementation might yield results on our windturbine environment already and exchanged the OpenAI gym for qblade.
+
+The first runs looked marvellous, perfectly holding rated speed, until we found out that those perfect values were prescribed by a setting in the qblade project file, and disabling this setting marked the beginning of our actual experimentation phase.
+
 \chapter{Experimentation}
+
+On our first experimentation steps, we did not yet have experience with how the qblade simulation behaves with a random or not well designed controller.
+Our observation was, that the simulation was resting between two different states, which we want to call \ac{FMM} and \ac{BMM}.
+
+\ac{FMM} is characterized by a rotational speed of ca 5 times rated speed, at 4.3 rad/s.
+At this speed, for some reason, the rotor will not speed up anymore.
+However, severe vibrations and sometimes the case of a disjointed blade happen.
+Especially in the case of a blade flying off, deflection values skyrocket up to several orders of magnitude higher than values seen in normal operation, and with one or more missing blades, the rotor will quickly stop rotating.
+This state is usually reached by a too low generator torque while not pitching out blades and sometimes stays stable for several thousand steps until a blade falls off.
+
+\ac{BMM} is characterized by a negative rotational speed, up to -3.8 rad/s.
+This can be reached by setting a higher generator torque than the rotor torque which is produced by aerodynamic lift on the blades.
+Also in this state, extreme vibrations and sometimes blades flying off can be observed, and like \ac{FMM} this state is stable until a blade falls off.
 
@@
 
-\section{Action Noise}
+As qblade doesn't provide a built-in reward function, we needed to craft our own reward function based on the observed state.
+We considered several possible variants.
+We will first introduce all of them and give a more detailed explanation later
 
-\section{Random exploration}
+\begin{itemize}
 
-\section{Parameter noise}
+\item \textbf{Rated speed} Hold a rated speed.
+Observe current rotational speed \(s_{rot}\) and with given rated speed \(s^*_{rot}\) calculate reward \(r(s) = -\abs{\frac{s_{rot} - s^*_{rot}}{s^*_{rot}}}\)
 
-\section{Replay noise}
+\item \textbf{Rated power} Hold a rated power. Observe current power \(s_{pow}\) and with given rated power \(s^*_{pow}\) calculate reward \(r(s) = -\abs{\frac{s_{pow} - s^*_{pow}}{s^*_{pow}}}\)
 
-\section{Simplifying the architecture}
+\item \textbf{Maximal power} Generate the maximum power possible. Observe current power \(s_{pow}\) and with given normalization constant \(c\) calculate \(r(s) = s_{pow} * c\)
 
-\section{Normalization}
+\item \textbf{Penalize current stress} Penalize high blade bending.
+Observe current bending \(s_{bend}\) and with given normalization constant \(c\) and another reward function \(r(s)\) calculate new reward \(r'(s) = r(s) - c * s_{bend} \)
 
-\section{Gradient actionspace}
+\item \textbf{Penalize accumulated stress} Penalize accumulated structural stress.
+With \(x = rainflow(s_0...s_t\) being the rainflow table and \(p(x)\) being a function that maps from rainflow outputs to a scalar penalty, calculate \(r'(s_0...s_t) = r(s_t) - p(rainflow(s_0...s_t))\)
 
-\section{Feeding past timesteps}
+\end{itemize}
 
-\section{Prioritized experience replay}
+\textbf{Rated speed} is the easiest of the rewards, as holding a rated speed can be achieved by either pitch or torque or a combination of the two.
+So both a high pitch and a high torque will stop it from running faster than it should.
+We used this reward for our first working version.
+\textbf{Rated power} is a bit more difficult.
+For reaching rated power, a certain aerodynamic torque is required, so pitching the blades out completely won't deliver rated power.
+However especially if setting an artificial rated power lower than what the turbine was built for, there is a certain play in how strongly to use blade pitch and how much needs to be done through torque.
+\textbf{Maximal power} is an intuitive reward, but might yield unrealistic results.
+This rewards high power regardless of the stress that is induced to the turbine, so a good policy might spin the turbine at high speeds beyond any turbine specification and then apply maximum torque.
+So this reward makes more sense when combined with a stress penalty.
+\textbf{Penalize current stress} is the easiest stress penalty, where a penalty proportional to the current bending of the structure can be used.
+As there are several bending modes on most of the components of the windturbine, a combination or selection of modes and components should be taken before.
+We decided to limit us to out-of-plane bending of the blade tips.
+This is a simple variant which leaves out bending in the middle of the blades, bending in the rotational plane, vibration or bending of the tower and torsional stress on the shaft.
+\textbf{Penalize accumulated stress} is a more complete, but time dependent variant.
+\ac{RFC} is a commonly used method for calculating structural fatigue in windturbines \cite{berglindFatigueEstimationMethods2014}.
+\ac{RFC} takes in a bending signal over time and returns a table with amplitudes and their frequency, as in how often the structure swung how far.
+It filters out some higher-frequency swinging in between two lower frequency swings and thus also works for structures that swing in more than one frequency.
+To use it as a penalty, we needed to create a function that maps from that table to a scalar penalty.
+This penalty accumulates past stress, so it increases over time.
+We suspect this penalty to be hard to learn, as it reacts slowly to changes and a once given penalty will never be lifted again.
+
+Most of our experimentation we did with rewarding rated speed.
+
+\section{Aiding exploration}
+
+As in the beginning we observed our algorithm to get stuck in either \ac{FMM} or \ac{BMM}, we wanted to improve exploration.
+Exploration is a common problem in reinforcement learning, as algorithms tend to stay in local maxima.
+
+\subsection{Action Noise}
+
+We first added a gaussian noise to our actions and after not seeing any improvemnts, switched to an Ornstein-Uhlenbeck process \cite{uhlenbeckTheoryBrownianMotion1930} as recommended in the DDPG paper \cite[formula (7)]{lillicrapCONTINUOUSCONTROLDEEP2015}.
+Additionally to the DDPG paper, we let the sigma parameter decay over time, so that the noise gets less when training has proceeded.
+
+\subsection{Random exploration}
+
+The TD3-paper \cite{fujimotoAddressingFunctionApproximation2018} utilizes a random exploration phase, in which a completely random policy \cite{mozerDiscoveringStructureReactive1990} initially explores the environment, before the actual agent switches in.
+We implemented the random walk at first through gaussian noise and then through an Ornstein-Uhlenbeck process.
+The later version creates less vibration through less excessive changes in control parameters.
+As we found that both random policies rarely explore a sensible operational range, we later added an expert PID controller, which implements torque control, and combined that with our Ornstein-Uhlenbeck process.
+We set the parameters to that controller by experimentation.
+
+\subsection{Parameter noise}
+
+As we still didn't see good results, we tried adding parameter-space noise \cite{plappertParameterSpaceNoise2017} to the actor function, as the paper promised better exploration.
+We could not observe any improvements and deactivated parameter space noise for the working version.
+
+\section{Zero Gradients}
+All our efforts to aid exploration by adding noise did not yield better exploration, so we had a look at our gradients and found that all our actor gradients are zero.
+The critic had gradients, but only in the later levels of the net.
+Eventually, we solved this problem by both simplyfing the architecture and utilizing normalization.
+
+\subsection{Simplifying the architecture}
+
+We tackled this problem by at first reducing the number and size of layers from the example implementation to one fully connected layer of 64 neurons in the critic and one fully connected layer of 32 neurons in the actor.
+The reference-implementation of \cite{DeepDeterministicPolicy} utilized 3 levels in the critic with 1024, 512 and 300 fully connected neurons and in the actor 2 levels of 512 and 128 fully connected neurons.
+Reducing complexity partially tackled the problem of zero gradients, we could then observe small gradients which however vanished over time.
+
+\subsection{Normalization}
+
+More effectively, we added data normalization.
+We normalized the data so states, actions and reward always stayed between 1 and -1.
+This finally fixed our problem of zero gradients and we could observe normal learning.
+We recommend this technique as a default technique for every reinforcement learning algorithm, as we can not imagine any drawbacks of normalizing input data.
+As in our simulation, where our state reflects measurements from very different parts of the simulation, some values in the state-array were 8 orders of magnitude different from others.
+As most of the high-magnitude values are vibrations, the net could not create a link between these and the state-action, while low-magnitude values like rotational speed were not considered due to their small absolute values.
+
+\section{High action gradients}
+
+We observed that our controller delivers high action gradients and reacts strongly between timesteps.
+This creates vibrations in the structure and especially high blade pitch changes lead to blades breaking or falling off.
+We tried different methods to circumvent this.
+
+\subsection{Gradient actionspace}
+
+At first, we implemented gradient actionspaces, where the controller output is not an absolute pitch or torque value but a difference to the last output, starting with everything set to 0.
+This way, we could clip the action gradients by reducing the gradient action space, so that in each timestep, only a fraction of the actual action space could be traversed.
+If the controller would output a positive change at maximum pitch/torque or a negative change at minimum pitch, this would be ignored.
+This did effectively limit action gradients, but also stopped us from getting sensible results.
+
+\subsection{Feeding past timesteps}
+
+Another idea was that we feed the last n timesteps concatenated to the current observations.
+We hoped that this way, the agent could derive its own gradients internally.
+The result of this were high and very noisy q-losses and no convergence whatsoever.
+Though we still had gradients, we suspect that our relu-activations might have drifted far off to the low end, and thus could not recover.
+
+\subsection{Clipping action gradients}
+
+We eventually solved the problem with the high action gradients by still letting the controller output absolute actions across the entire action spaces, but if gradients exceed a certain threshold, the actual action taken is set to the nearest value with sensible gradients.
+Though the actions returned from the policy are still wildly unstable, they at least didn't result in death of the turbine anymore and still we didn't hinder convergence through our gradient actionspace.
+
+\section{Improvements without sense}
+
+\subsection{Prioritized experience replay}
+
+\subsection{Data augmentation}
+
+We still had some parts of the code left where there wasn't any noise involved, so we decided to add some to the training process.
+Data-augmentation with noise is a common technique when using neural networks \cite[p.347]{bishopNeuralNetworksPattern1995} to reduce overfitting, so we expected it to also make our model generalize better.
+There have been some experiments with data augmentation in reinforcement learning \cite{cobbeQuantifyingGeneralizationReinforcement2019}, but it is not often mentioned on tutorials or blog posts.
+We implemented by adding a small noise term to the sampled states, actions and next states \((s, a, s')\) when performing experience replay.
+We didn't really see any indication of bad generalization, but as implementing it wasn't a big issue and normally doesn't bring any disadvantages, we still added this feature and set the noise level to a very small value.
+
+\section{Exploding Q-Loss}
+
+We observed that our Q-Loss, especially after a death, explodes into e15 magnitude values, while most of the times it stayed around e2 magnitude.
+Instable Q-Losses are a general problem with DQN and DDPG, and some literature \cite{fujimotoAddressingFunctionApproximation2018} proposes to use two Q networks, of which only the minimum is used for training.
+We could already greatly improve our results by using huber-loss \cite{huber1964} for the critic instead of mean squared error loss as described in DDPG paper.
+Also, we believe that adding double Q networks such as in \cite{fujimotoAddressingFunctionApproximation2018} would also solve this problem.
 
@@
 
+% Holds power, doesn't diverge, way to many actions, super noisy, we are at least somewhere close to sensible
+% Different reward function
+% Different wind speed
+
 \chapter{Future work}
@@
 % Whatever comes to our mind
+% Punishment for high action gradients
+% Switch reward functions during training?
 

### 2019-11-14 paper/paper.tex
Added twin delayed features

@@
 Instable Q-Losses are a general problem with DQN and DDPG, and some literature \cite{fujimotoAddressingFunctionApproximation2018} proposes to use two Q networks, of which only the minimum is used for training.
+
+\subsection{Huber loss}
 We could already greatly improve our results by using huber-loss \cite{huber1964} for the critic instead of mean squared error loss as described in DDPG paper.
-Also, we believe that adding double Q networks such as in \cite{fujimotoAddressingFunctionApproximation2018} would also solve this problem.
+However we still saw some loss explosions.
+
+\subsection{Double critic}
+Adding twin critic networks such as in \cite{fujimotoAddressingFunctionApproximation2018} eventually solved our problem with loss explosions completely.
 
@@
 
+We name our final variant of \ac{DDPG} \ac{DDDDDPPG}.
+Our variant is different to \cite{fujimotoAddressingFunctionApproximation2018} in that we don't use target policy smoothing.
+However, we just have more D.
 % What should I put here?

### 2019-11-14 python/ddpg/ddpg.py

@@ -55,7 +55,7 @@ class DDPG:
         
         self.actor = Actor(self.obs_dim, self.act_dim, hparams.actor_sizes[0], hparams.actor_sizes[1], hparams.init_weight_limit, hparams.actor_simple).to(self.device)
         self.actor_target = Actor(self.obs_dim, self.act_dim, hparams.actor_sizes[0], hparams.actor_sizes[1], hparams.init_weight_limit, hparams.actor_simple).to(self.device)
-    
+
         # Copy target parameters
         for target_param, param in zip(self.critic_target.parameters(), self.critic.parameters()):
             target_param.data.copy_(param.data)
@@ -65,7 +65,21 @@ class DDPG:
         # optimizers
         self.critic_optimizer = optim.Adam(self.critic.parameters(), lr=hparams.critic_lr)
         self.actor_optimizer  = optim.Adam(self.actor.parameters(), lr=hparams.actor_lr)
-    
+
+        # If in twin critic mode, add another critic
+        if(self.hparams.twin_critics):
+            self.critic2 = Critic(self.obs_dim, self.act_dim, hparams.critic_sizes[0], hparams.critic_sizes[1], hparams.critic_sizes[2], hparams.init_weight_limit, hparams.critic_simple).to(self.device)
+            self.critic2_target = Critic(self.obs_dim, self.act_dim, hparams.critic_sizes[0], hparams.critic_sizes[1], hparams.critic_sizes[2], hparams.init_weight_limit, hparams.critic_simple).to(self.device)
+            for target_param, param in zip(self.critic2_target.parameters(), self.critic2.parameters()):
+                target_param.data.copy_(param.data)
+            self.critic2_optimizer = optim.Adam(self.critic2.parameters(), lr=hparams.critic_lr)
+
+        if(self.hparams.critic_loss == 'huber'):
+            self.critic_loss_function = F.smooth_l1_loss
+        else:
+            self.critic_loss_function = F.mse_loss
+
+        # Buffor for experience replay
         self.replay_buffer = BasicBuffer(hparams.buffer_maxlen)
 
         # Noises
@@ -330,19 +344,40 @@ class DDPG:
         next_Q = self.critic_target.forward(next_state_batch, next_actions.detach())
         expected_Q = reward_batch + self.gamma * next_Q
         
+        # If using twin q, take minimum of the two next_Q estimations and replace that with the current one
+        if(self.hparams.twin_critics):
[…]

### 2019-11-14 python/ddpg/utils.py

@@ -216,6 +216,8 @@ class BasicBuffer:
         if(priority == 0):
             priority = 1e-15
 
+        done = 1 if done else 0
+
         experience = (state, action, np.array([reward]), next_state, done)
 
         self.max_priority = max(self.max_priority, priority)

### 2019-11-14 python/env/qbladeadapter.py

@@ -95,11 +95,12 @@ class QBladeAdapter:
     rated_power = 3200
     rated_speed = 0.8
     #return -np.abs(observation[1]-rated_power) - 1e3*(np.abs(observation[16]) + np.abs(observation[17]) + np.abs(observation[18]))
+    return 1-np.abs((observation[1]-rated_power)/rated_power)
     #return np.clip(observation[1], 0, None) - 1e4*(np.abs(observation[16]) + np.abs(observation[17]) + np.abs(observation[18]))
     #return np.clip(5 
     #  - np.abs((observation[1]-rated_power)/rated_power)
     #  - 5e-2*(np.abs(observation[16]) + np.abs(observation[17]) + np.abs(observation[18])), -10, 10)
-    return -np.abs(observation[0]-rated_speed)
+    #return -np.abs(observation[0]-rated_speed)
 
   def calc_death(self, observation):
 

### 2019-11-25 paper/paper.tex
Added batch-norm and todolist

@@
 
-As mentioned before, we set the yaw angle to always 0 degrees, as in reality controlling the orientation of the nacelle is trivial.
-The generator torque on our NREL 5MW turbine has a sensible range of \num{0} to \num{4e6} Nm% TODO insert formula here
+After a while of experimenting with the unrestricted simulation, we added some bounds to our control parameters. In our final version, we had the following restrictions:
+As mentioned before, we set the yaw angle to always 0 degrees, as in reality controlling the orientation of the nacelle is trivial and in our simulations, wind will always come from the front.
+The generator torque on our NREL 5MW turbine has a sensible range of \num{0} to \num{4e6} Nm.
 Because changing torque from zero to maximum in one timestep is not realistic for a real turbine, we restrict it to move a maximum of \num{3e5} Nm per one timestep of 0.1 seconds.
@@
 \section{Designing reward functions}
+\label{rewardfunctions}
 
@@
 With \(x = rainflow(s_0...s_t\) being the rainflow table and \(p(x)\) being a function that maps from rainflow outputs to a scalar penalty, calculate \(r'(s_0...s_t) = r(s_t) - p(rainflow(s_0...s_t))\)
+
+\item \textbf{Penalize action gradients} Penalize high action gradients.
+With \(a_t\) being the current action, \(a_{t-1}\) the last action, \(c\) a penalty constant and \(r(s)\) another reward function, calculate: \(r'(s, a) = r(s) - c * (a_t - a_{t-1})\)
 
@@
 We suspect this penalty to be hard to learn, as it reacts slowly to changes and a once given penalty will never be lifted again.
+\textbf{Penalize action gradients} is an option which is useful in contexts where high action gradients are impossible on the real system.
+Instead of clipping them, penalizing them could include action gradients into the learning process.
+We only expect sensible results with a stateful agent which knows at least the last action it took or one which has recurrent connections, as otherwise the last action taken is unknown to the agent.
 
-Most of our experimentation we did with rewarding rated speed.
+Most of our experimentation we did with rewarding rated speed, as we expected this reward function to be the easiest to learn.
 
@@
 As we still didn't see good results, we tried adding parameter-space noise \cite{plappertParameterSpaceNoise2017} to the actor function, as the paper promised better exploration.
-We could not observe any improvements and deactivated parameter space noise for the working version.
+We could not observe any improvements and deactivated parameter space noise.
 
@@
 As most of the high-magnitude values are vibrations, the net could not create a link between these and the state-action, while low-magnitude values like rotational speed were not considered due to their small absolute values.
+At first, we used observed states and rewards from the random exploration phase to calculate maximum/minimum values.
+As sometimes after random exploration, extreme values were observed, we started normalising on the replay data of a previous run, in which many simulation states were explored.
 
@@
 Though the actions returned from the policy are still wildly unstable, they at least didn't result in death of the turbine anymore and still we didn't hinder convergence through our gradient actionspace.
+We at first hid this clipping from the replay buffer, but then decided to store the actual values taken, as the models then don't have to also learn the clipping effect.
+We observed better learning when not hiding the clipping from the replay buffer.
 
@@
 \subsection{Double critic}
-Adding twin critic networks such as in \cite{fujimotoAddressingFunctionApproximation2018} eventually solved our problem with loss explosions completely.
+Adding twin critic networks such as in \cite{fujimotoAddressingFunctionApproximation2018} eventually solved our problem with loss explosions completely, however the thing didn't learn anymore.
+To use it in conjunction with \ac{PER}, we needed to decide for a priority formula and decided to take the minimum of the two q values, after the maximum showed biasing towards one of the networks
+
+\subsection{Batch normalization}
+As in deep learning, batch normalization is helpful for increasing lr and thus to skip over local maxima, we tried to add batch normalization between the layers.
+The authors suspected problems with the noise in RL, however it looked good.
 
 \section{Automatized hyperparameter search}
+
+To test all the above things, we tried simulations with different hyperparameters.
 
@@
 However, we just have more D.
+
+\section{First working version}
+How we achieved first successes, hparams, etc
+
+\section{HParam tuning}
+We tried different hyper parameters.
+To evaluate our controller, we measure the time the simulation stayed within a sensible range.
+We define sensible as rotational speeds between 0 and 1.2 rad/s.
+We let each simulation run 1 million steps.
+As during our random exploration, the simulation was always in sensible range, we remove the random exploration phase from our evaluation.
+
+\section{Comparison to industry controllers}
+We aren't even close
+
 % What should I put here?
@@
 \chapter{Future work}
-
 % Reworking reward functions
@@
 % Pretraining the policy with a reference controller
+% Expert policy in instable conditions
 % Whatever comes to our mind
@@
 % Switch reward functions during training?
+% Interest from TU Berlin from control elements guy
+
+\section{Reward functions}
+
+We did all our training under the simple hold-speed reward function.
+A realistic reward function however would incorporate power generated and damage incurred.
+As soon as a stable agent under the hold speed reward function was created, using different rewards could bring the performance of the controller closer to that of industry controllers.
+A complex reward function like what we proposed in section \ref{rewardfunctions} however also makes learning more difficult, and thus would need more work in refining the algorithm.
+
+\section{Switch reward functions during training}
+To combine the learning ease of a simple reward function with the performance of a complex reward, we could imagine fading over between a simple and complex reward function.
+In early stages of training, the network is conditioned to only output anything sane, while in later training it could optimize power or damage prevention more.
+
+\section{Expert policy training}
+
+We could imagine that first training the policy to output expert actions would improve learning.
+We imagine a phase at the beginning and after random exploration, where observations from the random exploration phase are fed into the expert controller and the policy is trained to estimate those values directly.
+This way, after random exploration, the agent would start off with a policy which behaves like the expert controller but can deviate from that in future training, by taking into account q values.
+A possible algorithm for this could look like:
+
+Alternatively, we could imagine leaving the expert policy training active the whole time, not just once after random exploration, but to reduce the learning rate gradually.
+This way, in initial training, predicting the expert policies actions would have a higher impact than what the Q-function would suggest, whereas in later training with the smaller learning rate, deviations from the expert policy would be punished less and the policy is allowed to deviate further from the expert policy.
+This would involve two backward passes per training iteration.
+
+\section{Expert policy in instable conditions}
+When the turbine is reaching a dangerous state (high/low rotational speed, high vibrations), we could fall back to our expert controller to save the situation.
+This would result in a safe controller, that upon insane policies would fall back to sane behavior.
+It would degrade exploration beyond what we doom safe, so a controller with this fallback method would never explore a death condition.
+This would not necessarily hinder learning, as a complete death condition might not be necessary to explore to see the decreasing loss gradient leading to it.
+However, seeing a death condition could result in more extreme Q-predictions and thus more incentive to stay clear of a death.
+
+\section{Active control elements}
+There is ongoing research into adding active control elements to windturbine blades such as flaps.
+A difficulty incurred in that research is the design of a controller, as the high-dimensional action space makes it difficult for a human to engineer a controller.
+Using our automatically learning agent could enable a first working version to that research, providing a first direction or even lead to performance increases over not using active control elements in a turbine.
 

### 2019-11-25 python/automated_hparam_search.py

@@ -12,7 +12,7 @@ import os
   {"id": 5, "batch_size": 256},
 ]'''
 
-experiments = [
+'''experiments = [
   {"id": 6, "critic_lr": 1e-2},
   {"id": 7, "critic_lr": 1e-3},
   {"id": 8, "critic_lr": 1e-5},
@@ -21,6 +21,26 @@ experiments = [
   {"id": 10, "actor_lr": 1e-3},
   {"id": 10, "actor_lr": 1e-5},
   {"id": 10, "actor_lr": 1e-6},
+]'''
+
+experiments = [
+  {"id": 11, "twin_critics": True},
+  {"id": 12, "twin_critics": True, "critic_loss": 'mse'},
+  {"id": 13, "critic_simple": False},
+  {"id": 14, "actor_simple": False},
+  {"id": 15, "actor_simple": False, "critic_simple": False},
+  {"id": 16, "replay_noise": 0},
+  {"id": 17, "replay_noise": 1e-4},
+  {"id": 18, "replay_noise": 1e-3},
+  {"id": 19, "replay_noise": 1e-2},
+  {"id": 20, "tau": 1e-3},
+  {"id": 21, "tau": 1e-4},
+  {"id": 22, "tau": 5e-2},
+  {"id": 23, "tau": 0.1},
+  {"id": 24, "gamma": 0.9},
+  {"id": 25, "gamma": 0.999},
+  {"id": 26, "gamma": 0.9999},
+  {"id": 27, "gamma": 0.8},
 ]
 
 
@@ -52,7 +72,7 @@ def run_one(experiment):
     return (id, -1, -1)
 
 # Set the number of parallel runs here
-pool = Pool(3)
+pool = Pool(1)
 results = pool.map(run_one, experiments)
 
 for (id, reward, deaths) in results:

### 2019-11-25 python/ddpg/ddpg.py

@@ -183,6 +183,21 @@ class DDPG:
         m, c = self.reward_normalizer_params_gpu if gpu else self.reward_normalizer_params
         return reward * m + c
 
+    def clip_action(self, action):
+        if(self.hparams.clip_action_gradients):
+            max_threshold = np.array(self.hparams.act_max_grad)
+            action_grads = action - self.last_action
+
+            # Limit higher values
+            indices = action_grads > max_threshold
+            action[indices] = self.last_action[indices] + max_threshold[indices]
+
+            # Limit smaller values
+            indices = action_grads < -max_threshold
+            action[indices] = self.last_action[indices] - max_threshold[indices]
+
+        action = np.clip(action, self.act_low, self.act_high)
+        return action
 
     def get_action(self, obs, add_noise=False):
         action = None
@@ -209,7 +224,7 @@ class DDPG:
         # Get it to the cpu
         action = action.squeeze(0).cpu().detach().numpy()
 
-        action = np.clip(action, self.act_low, self.act_high)
+        action = self.clip_action(action)
 
         return action
 
@@ -375,7 +390,7 @@ class DDPG:
             # Thus, if any of the two nets is performing badly on thet sample,
             # we'll sample it more often.
             if(self.hparams.twin_critics):
-                td_error = (expected_Q.detach() - torch.max(curr_Q.detach(), curr_Q2.detach())).abs().cpu().data.numpy()
+                td_error = (expected_Q.detach() - torch.min(curr_Q.detach(), curr_Q2.detach())).abs().cpu().data.numpy()
             else:
                 td_error = (expected_Q.detach() - curr_Q.detach()).abs().cpu().data.numpy()
             self.replay_buffer.update_priorities(indices, td_error)
@@ -486,6 +501,7 @@ class DDPG:
         if(done):
             self.killcount += 1
             self.epoch_killcount += 1
+            self.last_action = np.zeros(self.act_dim)
             if(self.hparams.feed_past):
                 self.past_obs.clear()
 
@@ -520,7 +536,7 @@ class DDPG:
         if(self.time < self.hparams.random_exploration_steps):
             action = [2e7*state[0]-12e6,0]
             action += self.denormalize_action(self.random_exploration_noise.get_noise(self.time), False)
-            action = np.clip(action, self.act_low, self.act_high)
+            action = self.clip_action(action)
         else:
             action = self.get_action(state, True)
 
@@ -540,8 +556,8 @@ class DDPG:
[…]

### 2019-11-25 python/ddpg/models.py

@@ -3,12 +3,23 @@ import torch.nn as nn
 import torch.nn.functional as F
 
 class Critic(nn.Module):
-    def __init__(self, obs_dim, action_dim, hidden_size_1 = 1024, hidden_size_2 = 512, hidden_size_3 = 300, init_w = 0.1, simple = False):
+    def __init__(self, obs_dim, action_dim, hidden_size_1 = 1024, hidden_size_2 = 512, hidden_size_3 = 300, init_w = 0.1, simple = False, batch_normalization = False):
         super(Critic, self).__init__()
 
         self.obs_dim = obs_dim
         self.action_dim = action_dim
         self.simple = simple
+
+        # If batch normalization is enabled, add batch-norm layers
+        # Otherwise just put the identity function
+        if(batch_normalization):
+            self.bn1 = nn.BatchNorm1d(hidden_size_1)
+            self.bn2 = nn.BatchNorm1d(hidden_size_2)
+            self.bn3 = nn.BatchNorm1d(hidden_size_3)
+        else:
+            self.bn1 = lambda x: x
+            self.bn2 = lambda x: x
+            self.bn3 = lambda x: x
 
         if(self.simple):
             self.linear1 = nn.Linear(self.obs_dim + self.action_dim, hidden_size_1)
@@ -37,22 +48,22 @@ class Critic(nn.Module):
 
     def first_layer(self, x, a):
         if(self.simple):
-            return F.relu(self.linear1(torch.cat([x, a], 1)))
+            return F.relu(self.bn1(self.linear1(torch.cat([x, a], 1))))
         else:
-            return F.relu(self.linear1(x))
+            return F.relu(self.bn1(self.linear1(x)))
 
     def forward_complex(self, x, a):
-        xa = F.relu(self.linear1(x))
+        xa = F.relu(self.bn1(self.linear1(x)))
         xa = torch.cat([xa,a], 1)
-        xa = F.relu(self.linear2(xa))
-        xa = F.relu(self.linear3(xa))
+        xa = F.relu(self.bn2(self.linear2(xa)))
+        xa = F.relu(self.bn3(self.linear3(xa)))
         qval = self.linear4(xa)
 
         return qval
 
     def forward_simple(self, x, a):
         xa_cat = torch.cat([x,a], 1)
-        xa = F.relu(self.linear1(xa_cat))
+        xa = F.relu(self.bn1(self.linear1(xa_cat)))
         qval = self.linear2(xa)
 
[…]

### 2019-11-25 python/env/qbladeadapter.py

@@ -36,7 +36,7 @@ class QBladeAdapter:
     print("creating qblade instance")
     self._qbladeLib._Z14createInstancev()
 
-    x = ctypes.c_char_p(b"../sample_projects/NREL_5MW_STR_NEW.wpa")
+    x = ctypes.c_char_p(b"../sample_projects/NREL_5MW_STR.wpa")
     print("loading qblade project")
     self._loadProject(x)
 
@@ -91,11 +91,20 @@ class QBladeAdapter:
     #return [0, 0, 0, 0, 0]
     return [0, 0]
 
+  def get_act_max_grad(self):
+    return [3e5, 0.5]
+
+
   def calc_reward(self, observation, action):
+    if(self.calc_death(observation)):
+      death_penalty = 3
+    else:
+      death_penalty = 0
+
     rated_power = 3200
     rated_speed = 0.8
     #return -np.abs(observation[1]-rated_power) - 1e3*(np.abs(observation[16]) + np.abs(observation[17]) + np.abs(observation[18]))
-    return 1-np.abs((observation[1]-rated_power)/rated_power)
+    return 1-np.abs((observation[1]-rated_power)/rated_power)-death_penalty
     #return np.clip(observation[1], 0, None) - 1e4*(np.abs(observation[16]) + np.abs(observation[17]) + np.abs(observation[18]))
     #return np.clip(5 
     #  - np.abs((observation[1]-rated_power)/rated_power)
@@ -110,7 +119,7 @@ class QBladeAdapter:
            np.abs(observation[17]) > broken_state or \
            np.abs(observation[18]) > broken_state or \
            observation[0] > 4 or \
-           observation[0] < -0.5
+           observation[0] < -1
 
   def padAction(self, action):
     action = [action[0], 0, action[1], action[1], action[1]]
@@ -121,17 +130,6 @@ class QBladeAdapter:
     action = np.clip(action, self.get_act_low(), self.get_act_high())
     action = np.nan_to_num(self.padAction(action))
 
-    # Make sure action gradients don't exceed a certain threshold
-    action_grads = action - self.lastAction
-    max_threshold = np.array([3e5, 1, 0.5, 0.5, 0.5])
-
-    # Limit higher values
-    indices = action_grads > max_threshold
-    action[indices] = self.lastAction[indices] + max_threshold[indices]
-
-    # Limit smaller values
-    indices = action_grads < -max_threshold
[…]

### 2019-11-25 python/main.py

@@ -45,6 +45,7 @@ def main():
   hparams.act_dim = env.get_act_dim()
   hparams.act_high = env.get_act_high()
   hparams.act_low = env.get_act_low()
+  hparams.act_max_grad = env.get_act_max_grad()
 
   checkpoint_steps = hparams.checkpoint_steps
   checkpoint_dir = hparams.checkpoint_dir

### 2019-11-27 paper/paper.tex
Added Q-Learning and PG section in paper, some DDPG rewrites for action grad clipping

@@
 Our environment here is the simulated wind turbine, but it could be a computer game \cite{lillicrapCONTINUOUSCONTROLDEEP2015}, a physics simulation\cite{brockmanOpenAIGym2016} or even an actual existing wind turbine \cite{kolterDesignAnalysisLearning2012}.
-In every timestep, the environment supplies us with a state, also sometimes called observation, and a scalar reward.
-The observed state doesn't have to represent the entirety of the state of the simulation, in fact it usually is only a small subset.
+In every timestep, the environment supplies us with a state \(x\), also sometimes called observation, and a scalar reward \(r\) for taking an action \(a\) in that state.
+In theory, the full trajectory \((x_0, a_0, ..., x_t)\) could be needed to describe the full state of the simulation at timestep t \(s_t\).
+However, the algorithms we looked at assume that the state progressions can be modelled as a Markov-Chain of observations, as in that what happens at timestep t+1 is only dependent on the observation at timestep t, not also on for example t-10.
+In Markov Chains, there is a state transition probability \(p(x_{t+1}|x_t, a_t)\) for each transition, which we aim to learn.
+Because we assume the observation \(x\) to be fully descriptive, we will use \(s_t = x_t\).
+In reality, the observed state doesn't have to represent the entirety of the state of the simulation, in fact it usually is only a small subset.
 For our wind turbine example, the observed state is limited to what can be measured with sensors on a wind turbine, in an arcade game it could be the pixel output or in a physical simulation it could be some key values in the simulation.
-The reward is a scalar value which judges the current state of the simulation.
+
+The reward \(r\) is a scalar value which judges the current state of the simulation.
 A higher reward means the current state is better than that of a lower reward.
 Sometimes, the reward function is obvious from the system, as in an arcade game it would surely be the score or in chess it could be 1 for win, -1 for lose and 0 for not decided yet.
-However in our paper, we will dedicate a section on how we designed our reward function.
-In general reinforcement learning theory, it is assumed to exist and be supplied by the environment.
-In every timestep, the agent supplies an action in the form of a tensor.
+In our paper, we will dedicate a section on how we designed our reward function, whereas in general reinforcement learning theory, it is assumed to exist and be supplied by the environment.
+
+In every timestep, the agent supplies an action \(a\).
 This action could be button presses in an arcade game or blade pitch in a windturbine simulation.
@@
 In other words, it should steer the environment to achieve best performance.
-Additionally to the state and reward, an environment can optionally send out a done signal, which indicates a state in which further simulation is not possible and the environment wants to be reset to an initial state.
+
+Additionally to the state and reward, an environment can optionally send out a done signal \(d\), which indicates a state in which further simulation is not possible and the environment wants to be reset to an initial state.
 In a chess game, this would be a loss or a win, and with a windturbine that could for example be fatal structural damage.
 
-The algorithms we looked at further assume that the state progressions can be modelled as a Markov-Chain, as in that what happens at timestep t+1 is only dependent on the hidden state at timestep t, not also on for example t-10.
-As the hidden state could for example be the full RAM of the simulation, we can safely assume that a simulation only calculates the next timestep on what is in the RAM at the current timestep.
-Also, Markov-Chains allow for several states as an outcome of any state, and that we can model the transition between these state with a probablistic model.
+\subsection{Definitions}
+
+TODO HUGE MESS FOLLOWING
+Before diving into the algorithm, we would define some parameters. From our environment, we have
+\begin{itemize}
+  \item \(s_t\) a single state at timestep t
+  \item \(r_t\) the reward at that timestep t
+  \item \(a_t\) an action taken at timestep t
+  \item \(\varepsilon\) the unknown transition probability matrix of the environment
+  \item \(d_t\) a flag (0 or 1) which indicates a terminal condition in the environment
+  \item \((s_t, a_t, r_t, d_t, s_{t+1})\) a state transition, going from state \(s_t\), taking action \(a_t\), getting a reward for that action \(r_t\) and a terminal flag \(d_t\) and reaching state \(s_{t+1}\)
+\end{itemize}
+
+We furthermore define \(\pi_{prob}: S \rightarrow P(A)\) a probablistic policy that returns a probability distribution over actions for given state and \(\pi: S \rightarrow A\) a deterministic policy that returns a single action for a given state.
+We will use deterministic policies in all of the paper except for the Q-Learning explanation, which is easier to grasp with probablistic policies.
+
+  \(R_{\pi}(t) = \sum^T_{t'=t}{\gamma^{t'-t}r_{t'}}\) A discounted return, summing up all returns from timestep t up to some point in the far future T while applying a discounting term \(\gamma \in [0, 1]\).
+The discounting gives less significance for rewards far in the future than rewards close to now.
+Also, as we need to assume state/action transitions in the future, the return is dependend upon a policy.
+If we wanted to compute this return, we would need to calculate the expectation over a construct made of all possible trajectories \( (s_0, a_0 \leftarrow \pi_{prob}(s_0), s_1, ...) \), combined with the initial state distribution \(p(s_0)\) and the environment Markov Chain \(p(s_{t}|s_{t-1}, a_{t-1}\).
+As we don't know the environment probability distributions, we need to find another way.
+
+\begin{itemize}
+  \item \((s_0, a_0, s_1, a_1, ... , a_{t-1}, s_t) \rightarrow \tau\) a trajectory of many states and their actions.
+  \item \(R(\tau) = \sum_t{r_t} \) an undiscounted return from a trajectory
+\end{itemize}
+
+\(V(s)\) is the value of a state defined as accumulating all rewards from this step infinitely into the future: \(V(s) = \sum_t{r_t}\).
+There are a couple of problems with this function, at first summing infinitely into the future might yield infinite or otherwise unusable results, which is why we introduce a discounting factor \(\gamma\) in \(V(s) = \sum_t{r_t * \gamma^t}\).
+And furthermore and more obviously, inputting a state to the function, we can not
+
+HUGE MESS OVER
+
+Some terms which we use
+\begin{itmeize}
+  \item \textbf{Return} A return with respect of a policy and a starting state s gives the accumulative reward of following that policy from state s until the future.
+As we have a stochastic environment, we have to add an expectation term, so it is the expected accumulative reward of following that policy.
+Furthermore a return can be discounted or undiscounted.
+Discounting means we multiply a discounting factor $\gamma \in [0,1]$ to rewards in the fashion of $r_0 \gamma^0 + r_1 \gamma^1 + r_2 \gamma^2 ...$, which results in distant future rewards being weighted less than closer ones.
+  \item \textbf{Optimal policy} The theoretical construct of a policy that always takes the best action with respect to accumulative reward.
+Though we rarely know this policy, there must always be at least one policy which gives the highest returns of all possible policies
+  \item \textbf{Value} The value of a state s is the return of that state following the optimal policy.
+\end{itemize}
 
@@
 
-% Bellman equation
+This equation uses the notion of a state-value function, also called Q function.
+In the Bellman Equation, it describes the total accumulated return until infinity taking an action a in state s and then following the optimal policy afterwards.
 
-\subsection{PG}
+The idea behind Q-Learning is that we want to learn the Q function from observations which we have.
+Assuming we already know the Q value for the next state \(Q(s_{t+1}, a_{t+1})\), we can then calculate the current Q value \( Q(s_t, a_t) = r_t + \gamma Q(s_{t+1}, a_{t+1}) \).
+Unfortunately, we don't have a definite Q for the next state, so we need to use our estimation which we have already.
+This estimation is most likely noisy or even plain wrong, so we want to only gradually update our Q.
+For this, we define our learning target for learning step i as
+\begin{equation}
+y_i = \mathbb{E}_{s' \sim \varepsilon}[r + \gamma Q(s', \pi(s');\theta_{i-1})]
+\end{equation}
+For the moment, let's assume the optimal policy $\pi$ which always chooses the action with the highest Q value.
+Our $y$ is then the current reward added to the highest possible reward in the future.
+If we had another policy, we would get the future accumulated reward under that policy.
+We use the parameters $\theta_{i-1}$ from the previous timestep here so that we don't have to derive y when we derive for $\theta_i$, which simplifies the learning term.
+
+Now, on our learning target y, we can define a loss as
+\begin{equation}
+L(\theta_i) = \mathbb{E}_{s, a}[(y_i - Q(s, a))^2]
+\end{equation}
+This already looks close to some squared error term.
+To calculate the expectation value over the environment $\varepsilon$, we will randomly draw samples from the replay buffer and average across them, doing something similar to a Monte-Carlo experiment.
+The result is equivalent to a normal mean square error calculation.
+
+Deriving for $\theta_i$ and applying the chain rule, we get the loss derivative
+\begin{equation}
+\Delta_{\theta_i} L(\theta_i) = \mathbb{E}[2(y_i - Q(s, a))\Delta_{\theta_i}Q(s, a)]
+\end{equation}
+Using our Monte-Carlo like batch from the replay buffer instead of calculating the full expectation term, we end up with stochastic gradient descent.
+
+Thus we just derived a method to train a Q function under the assumption of a policy.
+In \cite{mnihPlayingAtariDeep2013}, the team had a finite state space and could use the greedy policy $argmax_a Q(s', a)$ by searching the entire action space.
+We are presented with an infinite action space though, which is why we can't implement a greedy policy without high computational expense.
+So, for our work, we have to also learn a policy and not just a Q estimator.
+Luckily, our Q learning rule also works for other policies than the optimal policy, so let's try to derive a way to learn a policy.
+
+\subsection{Policy Gradients}
+
+We will at first give an intuitive explanation of a basic algorithm which forms the basis for many modern reinforcement leaarning algorithms, including the one we chose.
+The algorithm is called Policy Gradients or REINFORCE and was originally presented by \cite{williamsSimpleStatisticalGradientfollowing}.
+
+Let's assume we have a probablistic policy $\pi: S \rightarrow P(A)$ which is parametrized by $\theta$.
+An intuitive way of improving this policy could be to increase the gradient of the best possible action $a^*$ for a state $s$:
+\begin{equation}
+\theta_{i+1} = \theta_i + \alpha \Delta \pi_{\theta_i}(a^*|s)
+\end{equation}
+We could do this for all states we see and would end up with an iterative way of improving our policy.
+However, we might not know the best action for any state.
+Luckily, we derived Q-Learning in the section before and thus have an estimation which action will give which return in this step.
+We can use this and, for all actions and states, do
+\begin{equation}
+\theta_{i+1} = \theta_i + \alpha Q(s, a)\Delta \pi_{\theta_i}(a|s)
+\end{equation}
+This way, actions with higher Q values will receive a higher gradient step, and as Q is constant with respect to $\theta$ we don't have to derive along Q here.
+However there is still a problem to this.
+We assume that we use the same $\pi$ for exploration while doing training, so this policy is responsible for taking actions in the environment.
+This style is also called on-policy learning.
+So let's assume the policy is poorly initialized at the beginning of our training and gives action 1 a probability 4 times as high as action 2, though taking action 2 would yield a 10\% better return in our Q function.
+Now, as we are taking action 1 4 times as often as action 2, we will also perform gradient updates on action 1 4 times as often.
+Though each of the gradient steps is smaller due to the Q factor, the higher frequency of updating action 1 will lead to action 1 getting an advantage over action 2 in this update strategy.
+To correct for this phenomenon, we could divide the probability of each of the actions from the update term:
+\begin{equation}
+\theta_{i+1} = \theta_i + \alpha Q(s, a) \frac{ \Delta \pi_{\theta_i}(a|s)}{\pi_{\theta_i}(a|s)}
+\end{equation}
+We might remember from calculus that $\frac{\Delta x}{x} = \Delta \log x$ so we simplify to
+\begin{equation}
+\theta_{i+1} = \theta_i + \alpha Q(s, a) \Delta \log \pi_{\theta_i}(a|s)
+\end{equation}
+
+This update equation is our basic policy gradient equation.
 
@@
 Also in this state, extreme vibrations and sometimes blades flying off can be observed, and like \ac{FMM} this state is stable until a blade falls off.
+
+In neither of these modes, the simulation is within realistic bounds and the observations gotten from the simulation at this part are not credible.
+After we observed that the controller itself wasn't able to avoid these states on its own, we inhibited reaching these states by outputting a death condition when being in negative rotational areas or high positive.
+After a death, the simulation gets reset to initial state and the controller has more chances to learn.
 
@@
 As in the beginning we observed our algorithm to get stuck in either \ac{FMM} or \ac{BMM}, we wanted to improve exploration.
+The term exploration in reinforcement learning means how well the agent explores the state space of the environment.
 Exploration is a common problem in reinforcement learning, as algorithms tend to stay in local maxima.
+Intuitively, we are optimizing a policy to move towards an optimal state, while computing what is the optimal state based on what we observed.
+Thus, if a certain policy already gets locally good values, it will move towards that locally optimal state and never explore out of it, thus the Q network will not know about the good values beyond that.
+
+Additionally, as we discount future rewards, if the slope between the local optimum and the global optimum is too high and wide, the higher return of the global optimum will not propagate until the local optimum because of the discounting done close to it, and though the Q-network knows the global optimum, the gradients along the Q-function still lead into the local optimum.
+
+We suspected the first problem to happen, as neither of the two modes had good rewards, and going out of these modes continously improves rewards.
+To feed a network with new values closer to an optimum, a common practice in reinforcement learning is to add noise.
 
@@
 
-We first added a gaussian noise to our actions and after not seeing any improvemnts, switched to an Ornstein-Uhlenbeck process \cite{uhlenbeckTheoryBrownianMotion1930} as recommended in the DDPG paper \cite[formula (7)]{lillicrapCONTINUOUSCONTROLDEEP2015}.
-Additionally to the DDPG paper, we let the sigma parameter decay over time, so that the noise gets less when training has proceeded.
+In \cite{mnihPlayingAtariDeep2013} they use an epsilon-greedy strategy, where with a certain small probability epsilon, a random action is taken.
+As we have a continous action space, we can add a little bit of noise towards every action: \(a' = a + n \).
+We first tried a gaussian noise: \(n \leftarrow \mathcal{N}(\mu, \sigma^2) \).
+We didn't see any improvements over our old behavior, so we switched to an Ornstein-Uhlenbeck process \cite{uhlenbeckTheoryBrownianMotion1930} as recommended in the \ac{DDPG} paper \cite[formula (7)]{lillicrapCONTINUOUSCONTROLDEEP2015}.
+Additionally to the \ac{DDPG} paper, we let the sigma parameter decay over time, so that the noise gets less when training has proceeded.
 
@@
 As we found that both random policies rarely explore a sensible operational range, we later added an expert PID controller, which implements torque control, and combined that with our Ornstein-Uhlenbeck process.
-We set the parameters to that controller by experimentation.
+We set the parameters to that controller by experimentation as we did not need the actual control parameters of the controller.
+However it could be imagined to use an established controller here.
+The PID controller was easily able to keep the windturbine in a sensible operation range, and without noise it resulted in a smooth and constant operation.
+Though our PID controller is certainly inferior to industry-grade controllers, we will view it as an expert controller which is already achieving a certain degree of quality.
 
@@
 As we still didn't see good results, we tried adding parameter-space noise \cite{plappertParameterSpaceNoise2017} to the actor function, as the paper promised better exploration.
-We could not observe any improvements and deactivated parameter space noise.
+We could not observe any improvements and thus deactivated parameter space noise for the rest of the experimentation.
 
 \section{Zero Gradients}
-All our efforts to aid exploration by adding noise did not yield better exploration, so we had a look at our gradients and found that all our actor gradients are zero.
+All our efforts to aid exploration by adding noise did not yield better results, so we had a look at our gradients and found that all our actor gradients are zero.
 The critic had gradients, but only in the later levels of the net.
-Eventually, we solved this problem by both simplyfing the architecture and utilizing normalization.
+The problem of vanishing or exploding gradients is typical for deep learning, but not so common in flatter architectures.
 
@@
 This finally fixed our problem of zero gradients and we could observe normal learning.
-We recommend this technique as a default technique for every reinforcement learning algorithm, as we can not imagine any drawbacks of normalizing input data.
+We recommend this technique as a default technique for every reinforcement learning algorithm.
 As in our simulation, where our state reflects measurements from very different parts of the simulation, some values in the state-array were 8 orders of magnitude different from others.
-As most of the high-magnitude values are vibrations, the net could not create a link between these and the state-action, while low-magnitude values like rotational speed were not considered due to their small absolute values.
+As most of the high-magnitude values are vibrations, the net could not create a link between these and the state-action, while low-magnitude but insightful observations like rotational speed were not considered due to their small absolute values.
 At first, we used observed states and rewards from the random exploration phase to calculate maximum/minimum values.
-As sometimes after random exploration, extreme values were observed, we started normalising on the replay data of a previous run, in which many simulation states were explored.
+As sometimes after random exploration, extreme values were observed, we started normalising on the replay data of a previous run.
+We could also imagine manually setting normalization bounds on low-dimensional problems, as sometimes human knowledge about the possible state space is present.
 
@@
 
-We observed that our controller delivers high action gradients and reacts strongly between timesteps.
+We observed that our controller delivers high action gradients and reacts strongly between timesteps, sometimes jumping from no pitch/torque all the way to the maximum.
 This creates vibrations in the structure and especially high blade pitch changes lead to blades breaking or falling off.
@@
 At first, we implemented gradient actionspaces, where the controller output is not an absolute pitch or torque value but a difference to the last output, starting with everything set to 0.
+For aiding training, we added the actual actions taken to the observation space of the next step.
 This way, we could clip the action gradients by reducing the gradient action space, so that in each timestep, only a fraction of the actual action space could be traversed.
 If the controller would output a positive change at maximum pitch/torque or a negative change at minimum pitch, this would be ignored.
-This did effectively limit action gradients, but also stopped us from getting sensible results.
+This did effectively limit action gradients, but also did not lead to any sensible results, so we deactivated this again.
 
@@
 We hoped that this way, the agent could derive its own gradients internally.
-The result of this were high and very noisy q-losses and no convergence whatsoever.
-Though we still had gradients, we suspect that our relu-activations might have drifted far off to the low end, and thus could not recover.
+The result of this were high and very noisy q-losses and no sensible results still.
+Though gradients looked sensible, we suspect that the high Q-losses resulted in too high steps along the gradient and thus diverging behavior.
 
@@
 Though the actions returned from the policy are still wildly unstable, they at least didn't result in death of the turbine anymore and still we didn't hinder convergence through our gradient actionspace.
+This resulted in our first ever controller keeping the turbine within sensible operational range for at least a part of the training, though diverging quickly after.
+
 We at first hid this clipping from the replay buffer, but then decided to store the actual values taken, as the models then don't have to also learn the clipping effect.
 We observed better learning when not hiding the clipping from the replay buffer.
+
+\subsection{Pretraining the policy}
+
+We added a small period of direct pretraining of the policy on the actions taken by the expert policy during random exploration.
+We hoped for a more sensible default policy at the beginning and indeed observed reaching sensible behavior a little earlier.
 
@@
 \subsection{Double critic}
-Adding twin critic networks such as in \cite{fujimotoAddressingFunctionApproximation2018} eventually solved our problem with loss explosions completely, however the thing didn't learn anymore.
-To use it in conjunction with \ac{PER}, we needed to decide for a priority formula and decided to take the minimum of the two q values, after the maximum showed biasing towards one of the networks
+Adding twin critic networks such as in \cite{fujimotoAddressingFunctionApproximation2018} eventually solved our problem with loss explosions completely, however the thing didn't reach sensible states anymore.
+To use it in conjunction with \ac{PER}, we needed to decide for a priority formula and decided to take the minimum of the two q values, after the maximum showed biasing towards one of the networks.
 
@@
 
-We name our final variant of \ac{DDPG} \ac{DDDDDPPG}.
-Our variant is different to \cite{fujimotoAddressingFunctionApproximation2018} in that we don't use target policy smoothing.
-However, we just have more D.
+It must be stated that we did not achieve a single run which stayed within sensible operational range for longer than 30K steps.
+Even our handcrafted PID controller for which we took the first set of parameters that somehow works widely outperforms our learnt policies, an industry controller is even better.
+This result in itself is notable, because it needed a whole set of additions to plain \ac{DDPG} to get first sensible results on our windturbine environment.
+So apparently, our windturbine simulation is more complex to learn than the OpenAI gym examples, which are usually used as benchmarks for \ac{RL} performance.
+As the final product we developped is close to \cite{fujimotoAddressingFunctionApproximation2018} and has improvements from a wide variety of papers, we conclude that tackling wind turbine control with \ac{RL} will need more research.
+Though we were not able to outperform industry controllers, the speed in which the area of \ac{RL} is advancing makes us hopeful to achieve good results in the near future.
+With our work on integrating qblade, we provided an open-source, accessible framework to develop future reinforcement learning controllers for wind turbines.
 
@@
 
+We tried
+\begin{itemize} 
+  \item increasing/decreasing both learning rates
+  \item increasing/decreasing tau and gamma
+  \item increasing/decreasing batch size
+  \item increasing/decreasing replay noise
+\end{itemize}
+
 \section{Comparison to industry controllers}
-We aren't even close
 
@@
 
-We could imagine that first training the policy to output expert actions would improve learning.
-We imagine a phase at the beginning and after random exploration, where observations from the random exploration phase are fed into the expert controller and the policy is trained to estimate those values directly.
-This way, after random exploration, the agent would start off with a policy which behaves like the expert controller but can deviate from that in future training, by taking into account q values.
-A possible algorithm for this could look like:
-
-Alternatively, we could imagine leaving the expert policy training active the whole time, not just once after random exploration, but to reduce the learning rate gradually.
+We implemented expert policy pretraining.
+Alternatively, we could imagine leaving the expert policy training active the whole time, not just once after random exploration, and to reduce the learning rate of the expert policy training gradually.
 This way, in initial training, predicting the expert policies actions would have a higher impact than what the Q-function would suggest, whereas in later training with the smaller learning rate, deviations from the expert policy would be punished less and the policy is allowed to deviate further from the expert policy.
-This would involve two backward passes per training iteration.
+This would involve two backward passes per training iteration and would require prediction targets from the expert policy on new seen observations
 

### 2019-11-27 python/automated_hparam_search.py

@@ -28,16 +28,16 @@ experiments = [
   {"id": 12, "twin_critics": True, "critic_loss": 'mse'},
   {"id": 13, "critic_simple": False},
   {"id": 14, "actor_simple": False},
-  {"id": 15, "actor_simple": False, "critic_simple": False},
+  {"id": 15, "actor_simple": False, "critic_simple": False}, # Worked kinda
   {"id": 16, "replay_noise": 0},
   {"id": 17, "replay_noise": 1e-4},
   {"id": 18, "replay_noise": 1e-3},
   {"id": 19, "replay_noise": 1e-2},
-  {"id": 20, "tau": 1e-3},
+  {"id": 20, "tau": 1e-3}, # Worked kinda
   {"id": 21, "tau": 1e-4},
   {"id": 22, "tau": 5e-2},
   {"id": 23, "tau": 0.1},
-  {"id": 24, "gamma": 0.9},
+  {"id": 24, "gamma": 0.9}, # Worked kinda
   {"id": 25, "gamma": 0.999},
   {"id": 26, "gamma": 0.9999},
   {"id": 27, "gamma": 0.8},

### 2019-11-27 python/ddpg/ddpg.py

@@ -154,7 +154,6 @@ class DDPG:
         self.time = 0
         self.last_state = np.zeros(self.obs_dim)
         self.last_action = np.zeros(self.act_dim)
-        self.last_death = False
         self.epoch_reward = 0
         self.killcount = 0
         self.epoch_killcount = 0
@@ -191,7 +190,7 @@ class DDPG:
             # Limit higher values
             indices = action_grads > max_threshold
             action[indices] = self.last_action[indices] + max_threshold[indices]
-
+        
             # Limit smaller values
             indices = action_grads < -max_threshold
             action[indices] = self.last_action[indices] - max_threshold[indices]
@@ -199,10 +198,8 @@ class DDPG:
         action = np.clip(action, self.act_low, self.act_high)
         return action
 
-    def get_action(self, obs, add_noise=False):
-        action = None
-        # Do some random exploration at the beginning
 
+    def get_action(self, obs, add_noise=False):
         # Send the observation to the device, normalize it
         # Then calculate the action and denormalize it
         state = torch.FloatTensor(obs).unsqueeze(0).to(self.device)
@@ -226,20 +223,24 @@ class DDPG:
 
         action = self.clip_action(action)
 
+        assert(not np.any(np.isnan(action)))
+
         return action
 
 
     def prepare(self, state):
+        self.last_action = np.zeros(self.act_dim)
         state = self.feed_past_obs(state)
         if(self.hparams.action_gradients):
             return self.real_last_action
-        action = self.get_action(state, True)
+        action = self.get_action(state, False)
         self.last_action = action
         self.last_state = state
         return action
 
     def reset_finalize(self, obs):
-        return self.prepare(obs)
+        action = self.prepare(obs)
+        return action
 
     def calc_normalizations(self):
         # Load additional data if wanted
@@ -314,6 +315,50 @@ class DDPG:
 
         assert(len(state) == self.obs_dim)
         return state
+
+    def pretrain_policy(self):
+        optimizer = optim.Adam(self.actor.parameters(), lr=self.hparams.pretrain_policy_lr)
+        validation_size = int(len(self.replay_buffer) / 0.03)
+        validation_indices = np.random.choice(len(self.replay_buffer), validation_size)
+        self.replay_buffer.update_priorities(validation_indices, np.ones(validation_size)*1e-10)
+
+        validation_states, validation_actions, _, _, _ = zip(*self.replay_buffer.get_buffer()[validation_indices])
+
[…]

### 2019-11-27 python/ddpg/utils.py

@@ -374,18 +374,24 @@ class QBladeLogger:
         if(time%self.log_steps != 0):
             return
 
-        self.writer.add_scalar(name, val, time)
+        self.add_scalar_nofilter(name, val, time)
 
     def add_scalar_nofilter(self, name, val, time):
-        self.writer.add_scalar(name, val, time)
+        try:
+            self.writer.add_scalar(name, val, time)
+        except Exception as e:
+            print(e)
 
     def add_histogram(self, name, val, time):
         if(time%self.log_steps != 0):
             return
-        self.writer.add_histogram(name, val, time)
+        self.add_histogram_nofilter(name, val, time)
 
     def add_histogram_nofilter(self, name, val, time):
-        self.writer.add_histogram(name, val, time)
+        try:
+            self.writer.add_histogram(name, val, time)
+        except Exception as e:
+            print(e)
 
     def logAction(self, step, action, prefix='act'):
         self.add_scalar('%s/generator torque [Nm]' % prefix, action[0], step)

### 2019-11-27 python/env/qbladeadapter.py

@@ -36,9 +36,7 @@ class QBladeAdapter:
     print("creating qblade instance")
     self._qbladeLib._Z14createInstancev()
 
-    x = ctypes.c_char_p(b"../sample_projects/NREL_5MW_STR.wpa")
-    print("loading qblade project")
-    self._loadProject(x)
+    self.load_default_project()
 
     print("qblade init done")
 
@@ -64,11 +62,22 @@ class QBladeAdapter:
     del self._getControlVars
     del self._advanceSingleTimestep
 
+  def load_default_project(self):
+    x = ctypes.c_char_p(b"../sample_projects/NREL_5MW_STR.wpa")
+    print("loading qblade project")
+    self._loadProject(x)
 
   def reset(self):
     print('Resetting simulation')
     self._initializeSimulation(ctypes.c_int(0))
     self.lastAction = np.zeros(5)
+    self.storeAction(np.zeros(self.get_act_dim()))
+
+    # Do some steps with zero action
+    #action = np.zeros(self.get_act_dim())
+    #for i in range(0, 5):
+    #  self.step(action)
+
     print('reset done')
 
     return self.maskObservation(self.extractObservation())
@@ -95,9 +104,9 @@ class QBladeAdapter:
     return [3e5, 0.5]
 
 
-  def calc_reward(self, observation, action):
-    if(self.calc_death(observation)):
-      death_penalty = 3
+  def calc_reward(self, observation, action, death):
+    if(death):
+      death_penalty = 0.5
     else:
       death_penalty = 0
 
@@ -112,6 +121,11 @@ class QBladeAdapter:
     #return -np.abs(observation[0]-rated_speed)
 
   def calc_death(self, observation):
+    # If there are nan values, accumulate, after a few, die
+    if(np.any(np.isnan(observation))):
+      return True
+
+    observation = np.nan_to_num(observation)
 
     # 63 is the rotor size, so if anything bends further than that, it's broken off.
     broken_state = 20
@@ -119,7 +133,7 @@ class QBladeAdapter:
            np.abs(observation[17]) > broken_state or \
            np.abs(observation[18]) > broken_state or \
            observation[0] > 4 or \
-           observation[0] < -1
+           observation[0] < -0.2
 
   def padAction(self, action):
     action = [action[0], 0, action[1], action[1], action[1]]
@@ -129,7 +143,6 @@ class QBladeAdapter:
     # Copy action to control vars
     action = np.clip(action, self.get_act_low(), self.get_act_high())
[…]

### 2019-11-29 paper/paper.tex
PER improvements, not working yet

@@
 
-At first, we need to do some assumptions on our environment.
+At first, we need to do some assumptions on our environment $\varepsilon$.
 The concept environment encapsulates the reality or simulation that the reinforcement learning algorithm works in.
@@
 Our environment here is the simulated wind turbine, but it could be a computer game \cite{lillicrapCONTINUOUSCONTROLDEEP2015}, a physics simulation\cite{brockmanOpenAIGym2016} or even an actual existing wind turbine \cite{kolterDesignAnalysisLearning2012}.
-In every timestep, the environment supplies us with a state \(x\), also sometimes called observation, and a scalar reward \(r\) for taking an action \(a\) in that state.
+In every timestep, the environment supplies us with an observation \(x\) and a scalar reward \(r\) for taking an action \(a\) in that state.
 In theory, the full trajectory \((x_0, a_0, ..., x_t)\) could be needed to describe the full state of the simulation at timestep t \(s_t\).
-However, the algorithms we looked at assume that the state progressions can be modelled as a Markov-Chain of observations, as in that what happens at timestep t+1 is only dependent on the observation at timestep t, not also on for example t-10.
-In Markov Chains, there is a state transition probability \(p(x_{t+1}|x_t, a_t)\) for each transition, which we aim to learn.
-Because we assume the observation \(x\) to be fully descriptive, we will use \(s_t = x_t\).
+However, the algorithms we looked at assume that the state progressions can be modelled as a \ac{MDP} of observations, as in that what will happen at timestep t+1 is only dependent on the observation at timestep t, not also on for example t-10.
+In \ac{MDP}, the state transition probability \(p(x_{t+1}|x_t, a_t)\) is fully descriptive, as in it is equal to \(p(x_{t+1}|x_0, a_0, ..., x_t, a_t\).
+Because we assume the observation \(x\) to be fully descriptive, we will use the term state as equivalent to observation \(s_t = x_t\).
 In reality, the observed state doesn't have to represent the entirety of the state of the simulation, in fact it usually is only a small subset.
@@
 
-The reward \(r\) is a scalar value which judges the current state of the simulation.
-A higher reward means the current state is better than that of a lower reward.
+The reward \(r\) is a scalar value which judges the action taken in the current state of the simulation.
+A higher reward means the action is better than that of a lower reward.
 Sometimes, the reward function is obvious from the system, as in an arcade game it would surely be the score or in chess it could be 1 for win, -1 for lose and 0 for not decided yet.
+In both cases, sometimes a certain timespan passes before a reward for an action kicks in, which is also to be learned.
 In our paper, we will dedicate a section on how we designed our reward function, whereas in general reinforcement learning theory, it is assumed to exist and be supplied by the environment.
@@
 Some terms which we use
-\begin{itmeize}
+\begin{itemize}
   \item \textbf{Return} A return with respect of a policy and a starting state s gives the accumulative reward of following that policy from state s until the future.
@@
   \item \textbf{Value} The value of a state s is the return of that state following the optimal policy.
+  \item \textbf{Trajectory} A trajectory is a sequence of states and actions $(s_0, a_0, s_1, ..., a_{i-1}, s_i)$
 \end{itemize}
@@
 This equation uses the notion of a state-value function, also called Q function.
-In the Bellman Equation, it describes the total accumulated return until infinity taking an action a in state s and then following the optimal policy afterwards.
+In the Bellman Equation, it describes the total accumulated return until infinity taking an action $a$ in state $s$, receiving reward $r$, ending up in state $s'$ and then following the optimal policy afterwards.
+With the help of this function, we can compare different actions in our current state.
+And also, this small definition gives us a hint on how to create this function.
+We could initialize an array for all states and actions and then repeatedly loop through the array and fill it according to the above equation.
+However, this table is only computationally feasible when we have small finite state and action spaces.
+If any of the two spaces is big or even continuous, we have to replace the table by an estimator for Q.
 
@@
 
-We will at first give an intuitive explanation of a basic algorithm which forms the basis for many modern reinforcement leaarning algorithms, including the one we chose.
-The algorithm is called Policy Gradients or REINFORCE and was originally presented by \cite{williamsSimpleStatisticalGradientfollowing}.
+We will at first give an intuitive explanation of a basic algorithm which forms the basis for many modern reinforcement learning algorithms, including the one we chose.
+The algorithm is called REINFORCE and was originally presented by \cite{williamsSimpleStatisticalGradientfollowing}.
+Then we will add a Q term to it according to \cite{suttonPolicyGradientMethods} and is called \ac{PG}.
+\cite{degrisOffPolicyActorCritic2013} generalized that term and made an off-policy version of policy gradients.
+That version was then used by \cite{silverDeterministicPolicyGradient2013} and the stochastic policy was replaced by a deterministic one, the algorithm was called \ac{DPG}.
+This algorithm finally lays the basis for \ac{DDPG} by \cite{lillicrapCONTINUOUSCONTROLDEEP2015}, which is what we used until finally optimizing it towards \ac{TD3} \cite{fujimotoAddressingFunctionApproximation2018}.
 
@@
 Luckily, we derived Q-Learning in the section before and thus have an estimation which action will give which return in this step.
-We can use this and, for all actions and states, do
+\cite{suttonPolicyGradientMethods} include Q-learning into the REINFORCE algorithm and call the resulting algorithm \ac{PG}.
+So, with our newly gained Q, for all actions and states, do
 \begin{equation}
@@
 \subsection{DDPG}
-% Also problems with ddpg
 
 \section{RL on windturbines}
+We are aware of one paper \cite{kolterDesignAnalysisLearning2012} which tried RL on windturbines already, we want to dedicate a section to it.
+The team behind that paper built a miniature model of a windturbine and used a variation of the policy gradient algorithm to control it.
 
-\chapter{Early experimentation}
-%Early experiments, taking a reference implementation and openai gym training. Then finding out qblade is screwed...
+\chapter{Experimentation}
+
+Unfortunately, we are not able to provide a scientific comparison between different methods, as our learning proved too instable to derive sane metrics from it.
+Thus, this section will be structured as an assortment of problems and their solutions
+
+\section{Gym Experiments}
 
@@
 
-\chapter{Experimentation}
+\section{Starting with qblade}
 
@@
 
+\cite{schaulPrioritizedExperienceReplay2016} proposed a method to sample experiences from the replay buffer according to how much they benefit training, and not just uniformly.
+In their paper, they showed how especially cliffwalk problems, but effectively all problems, benefit from using prioritized experience replay.
+In their case, they use it for \ac{DQN}, but we will apply it to \ac{DDPG}, as our windturbine problem is also such a problem in which rewards are sparse.
+
+As knowing how much a sample benefits training is difficult, they propose to approximate this importance by the temporal difference error $\tau = y - Q(s, a)$.
+This difference, which in Q learning is comparable to the loss of the approximator, tells us how far off the prediction was in this step.
+So, alongside with the experiences, a sampling probability for each item $p_i$ is stored in the replay buffer.
+To be able to efficiently sample from it, we implemented it as proposed in their paper in the form of a sum-tree.
+
+When replaying, the sampling probability of a sample is set to $P(j) = \frac{p_j^{\alpha}}{\sum_i{p_i^{\alpha}}}$, $\alpha$ being a hyperparameter allowing to fade between uniform sampling ($\alpha=0$) and only priority sampling ($\alpha=1$).
+New samples are always stored with maximum priority, and to correct for oversampling bias, a correction term is applied to the temporal difference error: $w_j = \frac{(N*P(j))^{-\beta}}{\max_i{w_i}}$.
+It is simply multiplied to the real TD-error.
+$\beta$ is a parameter describing how strongly to correct oversampling, where $\beta=1$ is full correction and $\beta=0$ no correction at all.
+
+We decided to split policy and critic updates, as the oversampling effects could also happen to policy updates, and still sample uniformly for policy updates.
+
 \subsection{Data augmentation}
@@
 
-% It's pretty hard to create a working controller for a windturbine, apparently RL needs some more foundational research.
+% It's pretty hard to create a working controller for a windturbine, apparently RL needs some more foundational research or we are just not able to
 

### 2019-11-29 python/ddpg/ddpg.py

@@ -202,6 +202,7 @@ class DDPG:
     def get_action(self, obs, add_noise=False):
         # Send the observation to the device, normalize it
         # Then calculate the action and denormalize it
+        assert(not np.any(np.isnan(obs)))
         state = torch.FloatTensor(obs).unsqueeze(0).to(self.device)
 
         if (self.hparams.normalize_observations):
@@ -221,6 +222,8 @@ class DDPG:
         # Get it to the cpu
         action = action.squeeze(0).cpu().detach().numpy()
 
+        assert(not np.any(np.isnan(action)))
+
         action = self.clip_action(action)
 
         assert(not np.any(np.isnan(action)))
@@ -229,6 +232,8 @@ class DDPG:
 
 
     def prepare(self, state):
+        assert(not np.any(np.isnan(state)))
+
         self.last_action = np.zeros(self.act_dim)
         state = self.feed_past_obs(state)
         if(self.hparams.action_gradients):
@@ -249,6 +254,12 @@ class DDPG:
             tmpbuffer.load(self.hparams.normalization_extradata)
             extrastates = [state for state, _, _, _, _ in tmpbuffer.get_buffer()]
             extrarewards = [reward for _, _, reward, _, _ in tmpbuffer.get_buffer()]
+
+            # Length of our observations and the observations in the data must match
+            if(len(extrastates[0]) != self.obs_dim):
+                print('Normalization extradata incompatible')
+                extrastates = []
+
 
 
         # Calculate state max and min
@@ -397,6 +408,14 @@ class DDPG:
 
         if(self.hparams.normalize_actions):
             action_batch = self.normalize_action(action_batch, True)
+
+        # When using PER, calculate IS-weights
+        if(self.hparams.prioritized_experience_replay):
+            probabilities = self.replay_buffer.get_probabilities(indices)
+            probabilities = probabilities ** (-self.hparams.prioritized_experience_replay_beta)
+            probabilities = probabilities / np.max(probabilities)
+            weights = torch.FloatTensor(probabilities).to(self.device)
+
 
         self.critic_optimizer.zero_grad()
         curr_Q = self.critic.forward(state_batch, action_batch)
@@ -674,10 +693,10 @@ class DDPG:
 
             # Number of steps to sample random actions
             # before starting to utilize the policy
-            random_exploration_steps = 2000,
+            random_exploration_steps = 5000,
 
             # How many steps to train the agent after random exploration
[…]

### 2019-11-29 python/ddpg/utils.py

@@ -131,13 +131,16 @@ A tree structure, where every of the nodes is the sum of all the leaves below
 This can be used to do efficient sampling (without np.random.choice fucking me up)
 The weights buffer is a buffer with the tree weights stored like
 1  2 3  4 5 6 7  8 9 10 11 12 13 14 15...
+Note that alpha=0 is not supported!
 '''
 class SumTree:
-    def __init__(self, max_size):
+    def __init__(self, max_size, alpha=1):
         self.max_size = max_size
+        self.alpha = alpha
         self.tree_layers = int(np.log2(max_size)) + 1
         self.tree_size = np.sum([2**x for x in range(0, self.tree_layers)])
         self.weights = np.zeros(self.tree_size + self.max_size)
+        self.leaf_weights = np.zeros(self.max_size)
 
     # Get the weight trajectory for the leaf
     # All the weights from the leaf to the root node are in here
@@ -156,13 +159,43 @@ class SumTree:
 
         return trajectory
 
+    # Return the weights directly
     def get_leaf_weights(self):
+        return self.leaf_weights
+
+    def get_sum(self):
+        return np.sum(self.leaf_weights)
+
+    # Return the weights potentiated by alpha
+    def get_leaf_weights_alpha(self):
         return self.weights[self.tree_size:]
+
+    # We already have the sum of all items stored
+    def get_sum_alpha(self):
+        return self.weights[0]
+
+    # Sets a new alpha
+    # Warning, incurs complete recomputation
+    def set_alpha(self, alpha):
+        assert(alpha != 0)
+        self.alpha = alpha
+        self.recompute()
+
+    # Recompute all weights to correct for floating point imprecision
+    # Also necessary when changing alpha
+    def recompute(self):
+        self.weights = np.zeros(self.tree_size + self.max_size)
+        for i in range(0, self.max_size):
+            self.update_entry(i, self.leaf_weights[i])
+
+    # The root node holds the sum of all weights
+    def sum_weights(self):
+        return self.weights[0]
 
     # Update a leaf and propagate the sums back through the net
     def update_entry(self, leaf_idx, weight):
-        assert(weight >= 0)
-        weight = float(weight)
+        self.leaf_weights[leaf_idx] = float(weight)
+        weight = float(weight)**self.alpha
 
         trajectory = self.leaf_trajectory(leaf_idx)
         # Get the previous weight at this position
@@ -170,21 +203,26 @@ class SumTree:
         # Add the difference to all the nodes in the trajectory
[…]

### 2019-11-29 python/env/qbladeadapter.py

@@ -84,7 +84,7 @@ class QBladeAdapter:
 
   def get_obs_dim(self):
     #return 23
-    return 22
+    return 2
 
   def get_act_dim(self):
     #return 5
@@ -106,19 +106,19 @@ class QBladeAdapter:
 
   def calc_reward(self, observation, action, death):
     if(death):
-      death_penalty = 0.5
+      death_penalty = 0.5 if observation[0] > 0 else 3
     else:
       death_penalty = 0
 
     rated_power = 3200
     rated_speed = 0.8
     #return -np.abs(observation[1]-rated_power) - 1e3*(np.abs(observation[16]) + np.abs(observation[17]) + np.abs(observation[18]))
-    return 1-np.abs((observation[1]-rated_power)/rated_power)-death_penalty
+    #return 1-np.abs((observation[1]-rated_power)/rated_power)-death_penalty
     #return np.clip(observation[1], 0, None) - 1e4*(np.abs(observation[16]) + np.abs(observation[17]) + np.abs(observation[18]))
     #return np.clip(5 
     #  - np.abs((observation[1]-rated_power)/rated_power)
     #  - 5e-2*(np.abs(observation[16]) + np.abs(observation[17]) + np.abs(observation[18])), -10, 10)
-    #return -np.abs(observation[0]-rated_speed)
+    return 1-np.abs((observation[0]-rated_speed)/rated_speed)-death_penalty
 
   def calc_death(self, observation):
     # If there are nan values, accumulate, after a few, die
@@ -132,8 +132,8 @@ class QBladeAdapter:
     return np.abs(observation[16]) > broken_state or \
            np.abs(observation[17]) > broken_state or \
            np.abs(observation[18]) > broken_state or \
-           observation[0] > 4 or \
-           observation[0] < -0.2
+           observation[0] > 3 or \
+           observation[0] < -0.5
 
   def padAction(self, action):
     action = [action[0], 0, action[1], action[1], action[1]]
@@ -149,23 +149,26 @@ class QBladeAdapter:
 
     self.lastAction = action
 
-  def extractObservation(self):
+  def extractObservation(self, remove_nan=True):
     # 23 values are hardcoded in the library
     out_data = (ctypes.c_double * 23)()
     self._getControlVars(out_data)
     observation = [out_data[i] for i in range(0, 23)]
 
+    if(remove_nan):
+      observation = np.nan_to_num(observation)
+
     return observation
 
   def maskObservation(self, obs):
-    return np.array(obs[0:22])
[…]

### 2019-11-30 paper/paper.tex
Derived Q learning, REINFORCE, policy gradients and now doing deterministic pg

@@
 \begin{itemize}
+  \item \textbf{Policy} A function that maps from states to actions. There are deterministic policies $\mu: S -> A$ and stochastic policies $\pi: S -> P(A)$.
   \item \textbf{Return} A return with respect of a policy and a starting state s gives the accumulative reward of following that policy from state s until the future.
-As we have a stochastic environment, we have to add an expectation term, so it is the expected accumulative reward of following that policy.
+$R_{\pi}(s_t) = \mathbb{E}_{s_i \sim \varepsilon, a_i \sim \pi, i>t} r(s_i, a_i) $.
+As we have a stochastic environment and policy, we have to add an expectation term over actions taken and states resulting from those actions.
+It is the expected accumulative reward of following that policy.
 Furthermore a return can be discounted or undiscounted.
@@
 Though we rarely know this policy, there must always be at least one policy which gives the highest returns of all possible policies
-  \item \textbf{Value} The value of a state s is the return of that state following the optimal policy.
+  \item \textbf{Q-Value} The return of taking action a in state s plus the return following a policy after that $Q_{\pi}(s,a) = r(s, a) + \mathbb{E}_{s' \sim \varepsilon} R_{\pi}(s')$
   \item \textbf{Trajectory} A trajectory is a sequence of states and actions $(s_0, a_0, s_1, ..., a_{i-1}, s_i)$
@@
 
-At first we will present a very basic algorithm for a machine learning agent, which is based on the basic Bellman-Equation \cite{bellmanTheoryDynamicProgramming1954}
+At first we will present a very basic algorithm for a machine learning agent, which is based on the basic Bellman-Equation \cite{bellmanTheoryDynamicProgramming1954}. It is proven, that
 
 \begin{equation}
-Q(s, a) = r + \max_{a'}Q(s', a')
+Q*(s, a) = \mathbb{E}_{s' \sim \varepsilon | s,a}[r(s,a) + \gamma\max_{a'}Q*(s', a')]
 \end{equation}
 
+converges to the optimal solution when using iterative value updates.
+I will quickly explain this Bellman Equation.
 This equation uses the notion of a state-value function, also called Q function.
 In the Bellman Equation, it describes the total accumulated return until infinity taking an action $a$ in state $s$, receiving reward $r$, ending up in state $s'$ and then following the optimal policy afterwards.
+$\gamma$ acts as a discounting factor just as in our return function.
+It is even possible to implement an algorithm from this.
+We could create a lookup table for all s and a, initialize it all zero and whenever observing a state transition, we could update the table at that state and action after the above notation.
+
 With the help of this function, we can compare different actions in our current state.
-And also, this small definition gives us a hint on how to create this function.
-We could initialize an array for all states and actions and then repeatedly loop through the array and fill it according to the above equation.
 However, this table is only computationally feasible when we have small finite state and action spaces.
@@
 
-The idea behind Q-Learning is that we want to learn the Q function from observations which we have.
-Assuming we already know the Q value for the next state \(Q(s_{t+1}, a_{t+1})\), we can then calculate the current Q value \( Q(s_t, a_t) = r_t + \gamma Q(s_{t+1}, a_{t+1}) \).
-Unfortunately, we don't have a definite Q for the next state, so we need to use our estimation which we have already.
-This estimation is most likely noisy or even plain wrong, so we want to only gradually update our Q.
+Even with an estimated Q, though no longer guaranteed, we can still converge towards the optimal solution.
+For training an estimator, we need a loss which we can derive.
 For this, we define our learning target for learning step i as
 \begin{equation}
-y_i = \mathbb{E}_{s' \sim \varepsilon}[r + \gamma Q(s', \pi(s');\theta_{i-1})]
+y_i = \mathbb{E}_{s' \sim \varepsilon}[r + \gamma \max_{a'}Q(s', a';\theta_{i})]
 \end{equation}
-For the moment, let's assume the optimal policy $\pi$ which always chooses the action with the highest Q value.
-Our $y$ is then the current reward added to the highest possible reward in the future.
-If we had another policy, we would get the future accumulated reward under that policy.
-We use the parameters $\theta_{i-1}$ from the previous timestep here so that we don't have to derive y when we derive for $\theta_i$, which simplifies the learning term.
+For the moment, let's assume we can still calculate $\max_{a'}$.
+In fact, in a small finite action space we could just try all possible values of a.
+Our $y$ is then the current reward added to the highest possible rewards in the future, with other words exactly what we want to have as Q.
+We use the parameters $\theta_{i}$ in the target, so in theory, when deriving a later loss, we would also have to derive over our targets.
+However, in Q learning, this is commonly ignored.
 
@@
 \begin{equation}
-L(\theta_i) = \mathbb{E}_{s, a}[(y_i - Q(s, a))^2]
+L(\theta_i) = \mathbb{E}_{s \sim \varepsilon}[(y_i - Q(s, a;\theta_i))^2]
 \end{equation}
 This already looks close to some squared error term.
-To calculate the expectation value over the environment $\varepsilon$, we will randomly draw samples from the replay buffer and average across them, doing something similar to a Monte-Carlo experiment.
-The result is equivalent to a normal mean square error calculation.
+To calculate the expectation value over the environment $\varepsilon$, we need to do some form of Monte-Carlo experiment, in which we approximate that expectation value by repeatedly drawing samples from the underlying distribution.
+Luckily, we can easily draw samples from our environment, and we can even reuse old samples.
+Thus we store all samples in a replay buffer and uniformly draw from it to resemble a Monte-Carlo experiment.
 
@@
 \begin{equation}
-\Delta_{\theta_i} L(\theta_i) = \mathbb{E}[2(y_i - Q(s, a))\Delta_{\theta_i}Q(s, a)]
+\Delta_{\theta_i} L(\theta_i) = \mathbb{E}[(y_i - Q(s, a;\theta_i))\Delta_{\theta_i}Q(s, a;\theta_i)]
 \end{equation}
-Using our Monte-Carlo like batch from the replay buffer instead of calculating the full expectation term, we end up with stochastic gradient descent.
+Using our Monte-Carlo like batch from the replay buffer, we end up with stochastic gradient descent.
 
 Thus we just derived a method to train a Q function under the assumption of a policy.
-In \cite{mnihPlayingAtariDeep2013}, the team had a finite state space and could use the greedy policy $argmax_a Q(s', a)$ by searching the entire action space.
+In \cite{mnihPlayingAtariDeep2013}, the authors assumed a finite state space and could use the greedy policy $argmax_a Q(s', a)$ by searching the entire action space.
 We are presented with an infinite action space though, which is why we can't implement a greedy policy without high computational expense.
@@
 Luckily, we derived Q-Learning in the section before and thus have an estimation which action will give which return in this step.
-\cite{suttonPolicyGradientMethods} include Q-learning into the REINFORCE algorithm and call the resulting algorithm \ac{PG}.
 So, with our newly gained Q, for all actions and states, do
@@
 This way, actions with higher Q values will receive a higher gradient step, and as Q is constant with respect to $\theta$ we don't have to derive along Q here.
-However there is still a problem to this.
+Problematic is though, that the Q term we defined before, always took the best possible actions, which could be vastly different from the actions under our policy.
+Thus, we need to reformulate Q to account for any policy and not just the one which always takes the best action.
+
+\begin{equation}
+Q(s, a) = \mathbb{E}_{s' \sim \varepsilon, a' \sim \pi | s,a}[r + \gamma Q(s', a')]
+\end{equation}
+
+However there is still a problem if we integrate this new Q into our policy updates.
 We assume that we use the same $\pi$ for exploration while doing training, so this policy is responsible for taking actions in the environment.
-This style is also called on-policy learning.
 So let's assume the policy is poorly initialized at the beginning of our training and gives action 1 a probability 4 times as high as action 2, though taking action 2 would yield a 10\% better return in our Q function.
-Now, as we are taking action 1 4 times as often as action 2, we will also perform gradient updates on action 1 4 times as often.
+Now, as we are taking action 1 4 times as often as action 2, we will also perform gradient updates on action 1 4 times as often in our Monte-Carlo expectation draw.
 Though each of the gradient steps is smaller due to the Q factor, the higher frequency of updating action 1 will lead to action 1 getting an advantage over action 2 in this update strategy.
-To correct for this phenomenon, we could divide the probability of each of the actions from the update term:
+To correct for this oversampling bias, we could divide the probability of each of the actions from the update term:
 \begin{equation}
@@
 This update equation is our basic policy gradient equation.
+In reinforcement-learning, it has proven helpful to do off-policy learning, in which another policy than the one being trained can generate experiences.
+Thus, we can store past experiences in a replay buffer.
+And in fact, we were already able to use a replay buffer in the Q learning part.
+However we lost that again when we introduced our policy to Q learning.
+If we remember our expectation term, in which actions were sampled from the policy $\mathbb{E}_{a \sim \pi}$.
+We needed this expectation term, because we were dealing with stochastic policies.
+If instead learning deterministic policies, we can exclude the term from the equation.
+
+\begin{equation}
+Q^{\mu}(s,a) = \mathbb{E}_{s' \sim \varepsilon | s,a}[r + \gamma Q^{\mu}(s', \mu(s'))]
+\end{equation}
+
+Now, our Q is only dependent on the environment distribution, so when we do our Monte-Carlo experiments to update it, we can take our generated samples from the environment and assume they are modelling the underlying environment.
+However, by introducing our deterministic policy, we ruined our trick of dividing the probability value of an action from its gradient.
+
+TODO: deterministic policy gradients
+
+\subsection{Deterministic Policy Gradients}
 
 \subsection{DDPG}
+
+\ac{DDPG} is a combination of \ac{DPG} and \ac{DQN}.
+\cite{lillicrapCONTINUOUSCONTROLDEEP2015} tackle several problems in the two algorithms to combine them.
+At first, \ac{DQN} updates the Q network parameters based on a target calculated on the Q network parameters.
+To tackle this, \cite{mnihPlayingAtariDeep2013} introduced a second Q network which they call target network for which the parameters are held constant over a period of updates.
+After a number of steps, the parameters from the actual Q network are then copied over to the target.
+In \ac{DDPG}, they instead use soft target updates, where every step, the parameters of the actual network fade over to the target: $\theta_{target} = \tau \theta + (1 - \tau) \theta_{target}$.
+The fade-over parameter $\tau$ was recommended to be set to a small number like 0.01.
+Additionally to the target Q-network, they also found that a target policy
 
@@
 
-\section{Improvements without sense}
+\section{Other improvements}
+
+Because still generally not performing well, we added some techniques which we thought would aid general performance.
 
@@
 This result in itself is notable, because it needed a whole set of additions to plain \ac{DDPG} to get first sensible results on our windturbine environment.
-So apparently, our windturbine simulation is more complex to learn than the OpenAI gym examples, which are usually used as benchmarks for \ac{RL} performance.
+So apparently, our windturbine simulation is more complex to learn than the OpenAI gym examples, which are usually used as benchmarks for \ac{RL} performance, or we are having errors in our programme.
 As the final product we developped is close to \cite{fujimotoAddressingFunctionApproximation2018} and has improvements from a wide variety of papers, we conclude that tackling wind turbine control with \ac{RL} will need more research.
@@
 \section{First working version}
+
+Our model consists of the DDPG algorithm with following additions:
+We integrated prioritized experience replay to
+
 How we achieved first successes, hparams, etc

### 2019-11-30 python/automated_hparam_search.py

@@ -12,18 +12,18 @@ import os
   {"id": 5, "batch_size": 256},
 ]'''
 
-'''experiments = [
-  {"id": 6, "critic_lr": 1e-2},
-  {"id": 7, "critic_lr": 1e-3},
-  {"id": 8, "critic_lr": 1e-5},
-  {"id": 9, "critic_lr": 1e-6},
-  {"id": 10, "actor_lr": 1e-2},
-  {"id": 10, "actor_lr": 1e-3},
-  {"id": 10, "actor_lr": 1e-5},
-  {"id": 10, "actor_lr": 1e-6},
-]'''
-
 experiments = [
+  {"id": 30, "critic_lr": 1e-2},
+  {"id": 31, "critic_lr": 1e-3},
+  {"id": 32, "critic_lr": 1e-5},
+  {"id": 33, "critic_lr": 1e-6},
+  {"id": 34, "actor_lr": 1e-2},
+  {"id": 35, "actor_lr": 1e-3},
+  {"id": 36, "actor_lr": 1e-5},
+  {"id": 37, "actor_lr": 1e-6},
+]
+
+'''experiments = [
   {"id": 11, "twin_critics": True},
   {"id": 12, "twin_critics": True, "critic_loss": 'mse'},
   {"id": 13, "critic_simple": False},
@@ -41,7 +41,9 @@ experiments = [
   {"id": 25, "gamma": 0.999},
   {"id": 26, "gamma": 0.9999},
   {"id": 27, "gamma": 0.8},
-]
+]'''
+
+
 
 
 def run_one(experiment):
@@ -53,6 +55,7 @@ def run_one(experiment):
 
   experiment['test_output'] = test_output
   experiment['checkpoint_dir'] = checkpoint_dir
+  experiment['run_name'] = 'id%d' % id
 
   os.makedirs(checkpoint_dir, exist_ok=True)
 
@@ -72,7 +75,7 @@ def run_one(experiment):
     return (id, -1, -1)
 
 # Set the number of parallel runs here
-pool = Pool(1)
+pool = Pool(2)
 results = pool.map(run_one, experiments)
 
 for (id, reward, deaths) in results:

### 2019-11-30 python/ddpg/ddpg.py

@@ -62,9 +62,19 @@ class DDPG:
         for target_param, param in zip(self.actor_target.parameters(), self.actor.parameters()):
             target_param.data.copy_(param.data)
         
+        if(self.hparams.clip_gradients):
+            for p in self.actor.parameters():
+                p.register_hook(lambda grad: torch.clamp(grad, -self.hparams.clip_gradients, self.hparams.clip_gradients))
+            for p in self.critic.parameters():
+                p.register_hook(lambda grad: torch.clamp(grad, -self.hparams.clip_gradients, self.hparams.clip_gradients))
+
         # optimizers
-        self.critic_optimizer = optim.Adam(self.critic.parameters(), lr=hparams.critic_lr)
-        self.actor_optimizer  = optim.Adam(self.actor.parameters(), lr=hparams.actor_lr)
+        if(self.hparams.optimizer == 'adam'):
+            self.optimizer = optim.Adam
+        else:
+            self.optimizer = optim.SGD
+        self.critic_optimizer = self.optimizer(self.critic.parameters(), lr=hparams.critic_lr)
+        self.actor_optimizer  = self.optimizer(self.actor.parameters(), lr=hparams.actor_lr)
 
         # If in twin critic mode, add another critic
         if(self.hparams.twin_critics):
@@ -72,12 +82,25 @@ class DDPG:
             self.critic2_target = Critic(self.obs_dim, self.act_dim, hparams.critic_sizes[0], hparams.critic_sizes[1], hparams.critic_sizes[2], hparams.init_weight_limit, hparams.critic_simple).to(self.device)
             for target_param, param in zip(self.critic2_target.parameters(), self.critic2.parameters()):
                 target_param.data.copy_(param.data)
-            self.critic2_optimizer = optim.Adam(self.critic2.parameters(), lr=hparams.critic_lr)
+
+            if(self.hparams.clip_gradients):
+                for p in self.critic2.parameters():
+                    p.register_hook(lambda grad: torch.clamp(grad, -self.hparams.clip_gradients, self.hparams.clip_gradients))
+
+            self.critic2_optimizer = self.optimizer(self.critic2.parameters(), lr=hparams.critic_lr)
 
         if(self.hparams.critic_loss == 'huber'):
-            self.critic_loss_function = F.smooth_l1_loss
+            loss = torch.nn.SmoothL1Loss(reduction="none")
         else:
-            self.critic_loss_function = F.mse_loss
+            loss = torch.nn.MSELoss(reduction="none")
+
[…]

### 2019-11-30 python/ddpg/utils.py

@@ -290,13 +290,16 @@ class BasicBuffer:
         else:
             return self.buffer[0:self.iterator]
 
-    def sample(self, batch_size):
+    def sample(self, batch_size, uniform=False):
         if(len(self.buffer) == 0):
             return [], [], [], [], [], []
 
 
-        # Sample according to priorities
-        indices = [self.priorities.sample() for _ in range(0, batch_size)]
+        # Sample according to priorities or uniformly
+        if(not uniform):
+            indices = [self.priorities.sample() for _ in range(0, batch_size)]
+        else:
+            indices = np.random.choice(len(self), batch_size)
 
         # Convert indices to lists of individual batches
         batch = self.buffer[indices]
@@ -304,7 +307,6 @@ class BasicBuffer:
         state_batch, action_batch, reward_batch, next_state_batch, done_batch = zip(*batch)
 
         return (state_batch, action_batch, reward_batch, next_state_batch, done_batch, indices)
-
 
     def sample_sequence(self, batch_size):
         state_batch = []
@@ -379,10 +381,11 @@ class BasicBuffer:
 
 class QBladeLogger:
 
-    def __init__(self, logdir, log_steps):
+    def __init__(self, logdir, log_steps, run_name):
         self.logdir = logdir
         self.log_steps = log_steps
-        self.writer = SummaryWriter(logdir + '/' + str(datetime.now()))
+        run_name = run_name if run_name else str(datetime.now())
+        self.writer = SummaryWriter('%s/%s' % (logdir, run_name))
 
     def logObservation(self, step, observation, prefix='obs'):
         labels = {

### 2019-12-01 paper/paper.tex
Finished DPG section

@@
 Though we rarely know this policy, there must always be at least one policy which gives the highest returns of all possible policies
-  \item \textbf{Q-Value} The return of taking action a in state s plus the return following a policy after that $Q_{\pi}(s,a) = r(s, a) + \mathbb{E}_{s' \sim \varepsilon} R_{\pi}(s')$
+  \item \textbf{Q-Value} The return of taking action a in state s plus the return following a policy after that $Q_{\pi}(s,a) = r(s, a) + \mathbb{E}_{s' \sim \varepsilon} R_{\pi}(s')$.
+When following the optimal policy, we simply write $Q(s,a)$.
   \item \textbf{Trajectory} A trajectory is a sequence of states and actions $(s_0, a_0, s_1, ..., a_{i-1}, s_i)$
@@
 \begin{equation}
-Q*(s, a) = \mathbb{E}_{s' \sim \varepsilon | s,a}[r(s,a) + \gamma\max_{a'}Q*(s', a')]
+Q(s, a) = \mathbb{E}_{s' \sim \varepsilon | s,a}[r(s,a) + \gamma\max_{a'}Q(s', a')]
 \end{equation}
@@
 We will at first give an intuitive explanation of a basic algorithm which forms the basis for many modern reinforcement learning algorithms, including the one we chose.
-The algorithm is called REINFORCE and was originally presented by \cite{williamsSimpleStatisticalGradientfollowing}.
+The algorithm is called \ac{REINFORCE} (no joke) and was originally presented by \cite{williamsSimpleStatisticalGradientfollowing}.
 Then we will add a Q term to it according to \cite{suttonPolicyGradientMethods} and is called \ac{PG}.
@@
 Let's assume we have a probablistic policy $\pi: S \rightarrow P(A)$ which is parametrized by $\theta$.
-An intuitive way of improving this policy could be to increase the gradient of the best possible action $a^*$ for a state $s$:
+An intuitive way of improving this policy could be to increase the gradient proportional to the reward an action yielded:
 \begin{equation}
-\theta_{i+1} = \theta_i + \alpha \Delta \pi_{\theta_i}(a^*|s)
+\theta_{i+1} = \theta_i + \alpha r(s,a) \Delta \pi_{\theta_i}(a|s)
 \end{equation}
 We could do this for all states we see and would end up with an iterative way of improving our policy.
-However, we might not know the best action for any state.
+In fact, this is similar to the \ac{REINFORCE} algorithm (if we add our importance sampling trick from below).
+However, we might not know rewards for any action, state or have noisy rewards.
 Luckily, we derived Q-Learning in the section before and thus have an estimation which action will give which return in this step.
@@
 \end{equation}
-This way, actions with higher Q values will receive a higher gradient step, and as Q is constant with respect to $\theta$ we don't have to derive along Q here.
-Problematic is though, that the Q term we defined before, always took the best possible actions, which could be vastly different from the actions under our policy.
-Thus, we need to reformulate Q to account for any policy and not just the one which always takes the best action.
+This way, actions with higher Q values will receive a higher gradient step.
+As Q is constant with respect to $\theta$ we don't have to derive along Q here.
+Problematic is though, that the Q term we used, always took the best possible actions, which could be vastly different from the actions under our policy.
+Thus, we need to reformulate Q to account for our policy instead of the optimal policy
 
 \begin{equation}
-Q(s, a) = \mathbb{E}_{s' \sim \varepsilon, a' \sim \pi | s,a}[r + \gamma Q(s', a')]
+Q^{\pi}(s, a) = \mathbb{E}_{s' \sim \varepsilon, a' \sim \pi | s,a}[r + \gamma Q^{\pi}(s', a')]
 \end{equation}
 
+Now, in theory, we can not regard Q as constant with respect to $\theta$ anymore, because it is dependent on the policy.
+\cite{suttonPolicyGradientMethods} claim though that it can be ommitted and still regarded as constant.
 However there is still a problem if we integrate this new Q into our policy updates.
-We assume that we use the same $\pi$ for exploration while doing training, so this policy is responsible for taking actions in the environment.
+We assume that we use the same $\pi$ for exploration while doing training, so this policy is responsible for taking actions in the environment and also responsible for which experiences we see during our training.
 So let's assume the policy is poorly initialized at the beginning of our training and gives action 1 a probability 4 times as high as action 2, though taking action 2 would yield a 10\% better return in our Q function.
 Now, as we are taking action 1 4 times as often as action 2, we will also perform gradient updates on action 1 4 times as often in our Monte-Carlo expectation draw.
-Though each of the gradient steps is smaller due to the Q factor, the higher frequency of updating action 1 will lead to action 1 getting an advantage over action 2 in this update strategy.
-To correct for this oversampling bias, we could divide the probability of each of the actions from the update term:
+Though each of the gradient update steps is smaller due to the lower Q factor, the higher frequency of updating action 1 will lead to action 1 getting an advantage over action 2 in this update strategy.
+To correct for this oversampling bias, we could employ a trick which is called importance sampling.
+Because we know the bias from the probability distribution of our policy, we divide that probability from the update term:
 \begin{equation}
-\theta_{i+1} = \theta_i + \alpha Q(s, a) \frac{ \Delta \pi_{\theta_i}(a|s)}{\pi_{\theta_i}(a|s)}
+\theta_{i+1} = \theta_i + \alpha Q^{\pi}(s, a) \frac{ \Delta \pi_{\theta_i}(a|s)}{\pi_{\theta_i}(a|s)}
 \end{equation}
@@
 \begin{equation}
-\theta_{i+1} = \theta_i + \alpha Q(s, a) \Delta \log \pi_{\theta_i}(a|s)
+\theta_{i+1} = \theta_i + \alpha Q^{\pi}(s, a) \Delta \log \pi_{\theta_i}(a|s)
 \end{equation}
@@
 This update equation is our basic policy gradient equation.
+
+Before, we ignored for brevity that we are dealing with expectation calculations and that we are doing these updates under a stochastic environment, so to write it down formally correct, we need to introduce the density function of a state under a policy.
+According to \cite{suttonPolicyGradientMethods}, $d^{\pi}(s) = \lim_{t -> \infty} p(s_t = s | s_0, \pi)$ describes how likely it is to be in state s when following a policy $\pi$ forever, and also according to them this distribution is independent of the starting state $s_0$ and only dependent on the policy and environment.
+\cite{silverDeterministicPolicyGradient2013} used a discounted version of this, $\rho^{\pi}(s)$, which weights future probabilities less and thus can be described as the probability of being in state s \textit{soon} when following policy $\pi$.
+Using the discounted version approximates the Monte-Carlo experiments we are doing better than the undiscounted version, as effectively we do not train forever on one constant policy, but change it frequently.
+
+Having this $\rho$, let's write our objective function $J$ which we want to maximize.
+\begin{equation}
+J(\pi) = \int_S \rho^{\pi}(s) \int_A \pi(s, a)Q^{\pi}(s, a) dads = \mathbb{E}_{s \sim \rho^{\pi}, a \sim \pi}[Q^{\pi}(s,a)]
+\end{equation}
+
+We explained intuitively above, and \cite{suttonPolicyGradientMethods} proves that deriving this yields
+\begin{equation}
+\Delta J(\pi) = \mathbb{E}_{s \sim \rho^{\pi}, a \sim \pi}[\Delta \log \pi(a|s)Q^{\pi}(s, a)]
+\end{equation}
+
 In reinforcement-learning, it has proven helpful to do off-policy learning, in which another policy than the one being trained can generate experiences.
-Thus, we can store past experiences in a replay buffer.
-And in fact, we were already able to use a replay buffer in the Q learning part.
-However we lost that again when we introduced our policy to Q learning.
-If we remember our expectation term, in which actions were sampled from the policy $\mathbb{E}_{a \sim \pi}$.
+Thus, we can store past experiences in a replay buffer and learn from that.
+In fact, we were already able to use a replay buffer in the Q learning part.
+If we wanted to do that on our policy, we would need to correct for the bias of the other policy aswell with importance sampling.
+However we will not derive stochastic off-policy updates, because using the \ac{DPG} algorithm, this problem is solved.
+
+Additionally to this problem, another problem will be solved by \ac{DPG} as well: If we remember our Q expectation term, in which actions were sampled from the policy $\mathbb{E}_{a \sim \pi}$.
 We needed this expectation term, because we were dealing with stochastic policies.
-If instead learning deterministic policies, we can exclude the term from the equation.
+If instead learning deterministic policies, we can exclude policy expectation from the equation.
+But before going through all the advantages of deterministic policies, let's first explain it.
+
+\subsection{Deterministic Policy Gradients}
+
+Before \cite{silverDeterministicPolicyGradient2013}, stochastic policies were preferred because the stochastic nature aided exploration and because there was simply no theory on how to derive deterministic policy gradients.
+The idea behind deterministic policy gradients is to replace the stochastic policy $\pi: S -> P(A)$ with a deterministic one $\mu: S -> A$.
+Also, we are switching from small finite action spaces to continous action spaces.
+Now, updating this policy according to the equations above does not work anymore due to two problems.
+At first and most significantly, above we needed all $a$ or at least some form of maximization to calculate the full gradient of $\pi(a|s)$.
+This is difficult with a continous action space now.
+Furthermore our importance sampling trick, our correction for overestimation bias based on the bias of the policy, does not work anymore.
+We don't now the bias of that policy, because it is deterministic and will yield only a single action for any state.
+
+So, following the policy gradient has become difficult. However, let's have a look at our Q function. We can replace our stochastic policy with a deterministic one and get
 
@@
 
-Now, our Q is only dependent on the environment distribution, so when we do our Monte-Carlo experiments to update it, we can take our generated samples from the environment and assume they are modelling the underlying environment.
-However, by introducing our deterministic policy, we ruined our trick of dividing the probability value of an action from its gradient.
+Now, our Q is only dependent on the environment distribution, not anymore both on $s' \sim \varepsilon, a' \sim \pi$.
+This is an advantage, because computing only one expectation requires less samples and we do not have to worry for any bias from the policy in Q-updates.
+When we do our Monte-Carlo experiments to update it, we can take our generated samples from the environment and assume they are modelling the underlying environment.
 
-TODO: deterministic policy gradients
+If we are visualizing our Q as a function of state and action, we get a fully differentiable surface.
+If we derive this surface along the action dimension, we get a gradient which points in the direction of a better action for this state.
+We could direcly follow this gradient and calculate our policy update from it:
 
-\subsection{Deterministic Policy Gradients}
+\begin{equation}
+\theta_{i+1} = \theta_i + \alpha \mathbb{E}_{s \sim \rho^{\mu_{\theta_i}}}[\Delta Q^{\mu_{\theta_i}}(s, \mu_{\theta_i}(s))]
+\end{equation}
+
+\cite[Theorem 1]{silverDeterministicPolicyGradient2013} proves that this gradient always exists.
+We now managed to remove any expectation that is related to our action space from our gradient updates.
+This is beautiful because at first, we do not have to sample over many actions to learn this expectation.
+But more importantly, this term also works off policy \cite[Section 4.2]{silverDeterministicPolicyGradient2013} and with a Q function that approximates based on any other policy $Q^{\omega}$.
+So now we can use a replay buffer again and learn from past experiences, though we did not generate them with our policy.
 
@@
 
-\ac{DDPG} is a combination of \ac{DPG} and \ac{DQN}.
+\ac{DDPG} is a combination of \ac{DPG} and \ac{DQN} with several improvements.
 \cite{lillicrapCONTINUOUSCONTROLDEEP2015} tackle several problems in the two algorithms to combine them.
@@
 The fade-over parameter $\tau$ was recommended to be set to a small number like 0.01.
-Additionally to the target Q-network, they also found that a target policy
+Additionally to the target Q-network, they also found that a target policy improves stability, and they update it the same way as the main policy.
 
@@
 We are aware of one paper \cite{kolterDesignAnalysisLearning2012} which tried RL on windturbines already, we want to dedicate a section to it.
-The team behind that paper built a miniature model of a windturbine and used a variation of the policy gradient algorithm to control it.
+The team behind that paper built a miniature model of a windturbine and used a variation of the \ac{REINFORCE} algorithm to control it.
 

### 2019-12-02 paper/paper.tex
Finished DPG section, started with DDPG

@@
 Park control becomes a bigger topic in wind-energy, where controlling all windturbines of a park together can increase overall energy yield or protect windturbines further down the windstream from wakes by turbines up front.
-C
 
@@
 
-This section aims at facilitating the background knowledge necessary for understanding this paper. We fill first introduce windturbines and common terms around that
+This section aims at facilitating the background knowledge necessary for understanding this paper. We will first introduce windturbines and common terms around this area of research.
 
@@
 For our wind turbine example, the observed state is limited to what can be measured with sensors on a wind turbine, in an arcade game it could be the pixel output or in a physical simulation it could be some key values in the simulation.
+We denote the probability of a state transition to a specific $s'$ as $p(s'|s,a;\varepsilon)$ and the distribution over all states $s'$ as $P_{s' \sim \varepsilon | s,a}$ and when $s,a$ is clear, we omit it.
 
-The reward \(r\) is a scalar value which judges the action taken in the current state of the simulation.
+The reward \(r(s,a)\) is a scalar value which judges the action $a$ taken in the current state $s$ of the simulation.
 A higher reward means the action is better than that of a lower reward.
@@
 In a chess game, this would be a loss or a win, and with a windturbine that could for example be fatal structural damage.
+As this is commonly ignored in reinforcement learning, we will also ignore it.
 
@@
 
-TODO HUGE MESS FOLLOWING
-Before diving into the algorithm, we would define some parameters. From our environment, we have
+Additionally to the terms defined by the environment, we introduce some terms which are common to reinforcement learning
 \begin{itemize}
-  \item \(s_t\) a single state at timestep t
-  \item \(r_t\) the reward at that timestep t
-  \item \(a_t\) an action taken at timestep t
-  \item \(\varepsilon\) the unknown transition probability matrix of the environment
-  \item \(d_t\) a flag (0 or 1) which indicates a terminal condition in the environment
-  \item \((s_t, a_t, r_t, d_t, s_{t+1})\) a state transition, going from state \(s_t\), taking action \(a_t\), getting a reward for that action \(r_t\) and a terminal flag \(d_t\) and reaching state \(s_{t+1}\)
-\end{itemize}
-
-We furthermore define \(\pi_{prob}: S \rightarrow P(A)\) a probablistic policy that returns a probability distribution over actions for given state and \(\pi: S \rightarrow A\) a deterministic policy that returns a single action for a given state.
-We will use deterministic policies in all of the paper except for the Q-Learning explanation, which is easier to grasp with probablistic policies.
-
-  \(R_{\pi}(t) = \sum^T_{t'=t}{\gamma^{t'-t}r_{t'}}\) A discounted return, summing up all returns from timestep t up to some point in the far future T while applying a discounting term \(\gamma \in [0, 1]\).
-The discounting gives less significance for rewards far in the future than rewards close to now.
-Also, as we need to assume state/action transitions in the future, the return is dependend upon a policy.
-If we wanted to compute this return, we would need to calculate the expectation over a construct made of all possible trajectories \( (s_0, a_0 \leftarrow \pi_{prob}(s_0), s_1, ...) \), combined with the initial state distribution \(p(s_0)\) and the environment Markov Chain \(p(s_{t}|s_{t-1}, a_{t-1}\).
-As we don't know the environment probability distributions, we need to find another way.
-
-\begin{itemize}
-  \item \((s_0, a_0, s_1, a_1, ... , a_{t-1}, s_t) \rightarrow \tau\) a trajectory of many states and their actions.
-  \item \(R(\tau) = \sum_t{r_t} \) an undiscounted return from a trajectory
-\end{itemize}
-
-\(V(s)\) is the value of a state defined as accumulating all rewards from this step infinitely into the future: \(V(s) = \sum_t{r_t}\).
-There are a couple of problems with this function, at first summing infinitely into the future might yield infinite or otherwise unusable results, which is why we introduce a discounting factor \(\gamma\) in \(V(s) = \sum_t{r_t * \gamma^t}\).
-And furthermore and more obviously, inputting a state to the function, we can not
-
-HUGE MESS OVER
-
-Some terms which we use
-\begin{itemize}
-  \item \textbf{Policy} A function that maps from states to actions. There are deterministic policies $\mu: S -> A$ and stochastic policies $\pi: S -> P(A)$.
-  \item \textbf{Return} A return with respect of a policy and a starting state s gives the accumulative reward of following that policy from state s until the future.
-$R_{\pi}(s_t) = \mathbb{E}_{s_i \sim \varepsilon, a_i \sim \pi, i>t} r(s_i, a_i) $.
+  \item \textbf{Policy} A function that maps from states to actions.
+There are deterministic policies $\mu: S -> A$ and stochastic policies $\pi: S -> P(A)$.
+If we approximate a policy by a function approximator parametrized with $\theta$, we write $\pi(s;\theta)$ or omit the parameters when clear.
+  \item \textbf{Value} A value with respect to a policy and a starting state s gives the accumulative reward of following that policy from state s until infinitely in the future.
+$V^{\pi}(s_t) = \mathbb{E}_{s_i \sim \varepsilon, a_i \sim \pi, i>t} r(s_i, a_i) $.
 As we have a stochastic environment and policy, we have to add an expectation term over actions taken and states resulting from those actions.
 It is the expected accumulative reward of following that policy.
-Furthermore a return can be discounted or undiscounted.
+Furthermore a value can be discounted or undiscounted.
 Discounting means we multiply a discounting factor $\gamma \in [0,1]$ to rewards in the fashion of $r_0 \gamma^0 + r_1 \gamma^1 + r_2 \gamma^2 ...$, which results in distant future rewards being weighted less than closer ones.
+  \item \textbf{Return} If we measured a value function, that would be the return.
+Obviously we can't measure infinitely, so a return is only defined on a finite trajectory of state-transitions, but otherwise is equal to a value.
   \item \textbf{Optimal policy} The theoretical construct of a policy that always takes the best action with respect to accumulative reward.
-Though we rarely know this policy, there must always be at least one policy which gives the highest returns of all possible policies
-  \item \textbf{Q-Value} The return of taking action a in state s plus the return following a policy after that $Q_{\pi}(s,a) = r(s, a) + \mathbb{E}_{s' \sim \varepsilon} R_{\pi}(s')$.
-When following the optimal policy, we simply write $Q(s,a)$.
-  \item \textbf{Trajectory} A trajectory is a sequence of states and actions $(s_0, a_0, s_1, ..., a_{i-1}, s_i)$
+Though we rarely know this policy, there must always be at least one policy which gives the highest value of all possible policies in a finite \ac{MDP}
+  \item \textbf{Q-Value} The reward of taking action a in state s plus the value following a policy after that $Q^{\pi}(s,a) = r(s, a) + \mathbb{E}_{s' \sim \varepsilon} V^{\pi}(s')$.
+When following the optimal policy, we simply write $Q^*(s,a)$.
+Usually, our Q functions are discounted.
+Note that we can also define Values on Q-Values: $V^{\pi}(s) = \mathbb{E}_{a \sim \pi} Q(s, a)$.
+If we approximate Q by a function approximator parametrized with $\theta$, we write $Q(s, a;\theta)$ or omit it in case it is clear.
+  \item \textbf{Actor-Critic} Actor is just another name for a learnt policy and critic just another name for an estimated Q-Value function.
+These two frequently appear together, if that happens we are using an actor-critic algorithm.
 \end{itemize}
@@
 \begin{equation}
-Q(s, a) = \mathbb{E}_{s' \sim \varepsilon | s,a}[r(s,a) + \gamma\max_{a'}Q(s', a')]
+Q^*(s, a) = \mathbb{E}_{s' \sim \varepsilon | s,a}[r(s,a) + \gamma\max_{a'}Q^*(s', a')]
 \end{equation}
@@
 converges to the optimal solution when using iterative value updates.
-I will quickly explain this Bellman Equation.
+We will quickly explain this Bellman Equation.
 This equation uses the notion of a state-value function, also called Q function.
@@
 \begin{equation}
-y_i = \mathbb{E}_{s' \sim \varepsilon}[r + \gamma \max_{a'}Q(s', a';\theta_{i})]
+y_i = \mathbb{E}_{s' \sim \varepsilon}[r + \gamma \max_{a'}Q^*(s', a';\theta_{i})]
 \end{equation}
@@
 \begin{equation}
-L(\theta_i) = \mathbb{E}_{s \sim \varepsilon}[(y_i - Q(s, a;\theta_i))^2]
+L(\theta_i) = \mathbb{E}_{s \sim \varepsilon}[(y_i - Q^*(s, a;\theta_i))^2]
 \end{equation}
@@
 \begin{equation}
-\Delta_{\theta_i} L(\theta_i) = \mathbb{E}[(y_i - Q(s, a;\theta_i))\Delta_{\theta_i}Q(s, a;\theta_i)]
+\Delta_{\theta_i} L(\theta_i) = \mathbb{E}[(y_i - Q^*(s, a;\theta_i))\Delta_{\theta_i}Q^*(s, a;\theta_i)]
 \end{equation}
@@
 
-Thus we just derived a method to train a Q function under the assumption of a policy.
+Thus we just derived a method to train a Q function under the assumption of an optimal policy.
 In \cite{mnihPlayingAtariDeep2013}, the authors assumed a finite state space and could use the greedy policy $argmax_a Q(s', a)$ by searching the entire action space.
@@
 
-In reinforcement-learning, it has proven helpful to do off-policy learning, in which another policy than the one being trained can generate experiences.
-Thus, we can store past experiences in a replay buffer and learn from that.
+In reinforcement-learning, it has proven helpful to do off-policy learning, in which another policy $\beta$ than the one being trained can generate experiences.
+We would then reformulate our expectation to $\mathbb{E}_{s \sim \rho^{\beta}, a \sim \beta}$.
+Thus, we could store past experiences in a replay buffer and learn from that.
 In fact, we were already able to use a replay buffer in the Q learning part.
-If we wanted to do that on our policy, we would need to correct for the bias of the other policy aswell with importance sampling.
+If we wanted to do that on our policy however, we would need to correct for the bias of the other policy aswell with importance sampling.
 However we will not derive stochastic off-policy updates, because using the \ac{DPG} algorithm, this problem is solved.
+If you want to learn about \ac{Off-PAC} with stochastic policies, we recommend to read \cite{degrisOffPolicyActorCritic2013}
 
@@
 
-Now, our Q is only dependent on the environment distribution, not anymore both on $s' \sim \varepsilon, a' \sim \pi$.
+Now, our Q is only calculating an expectation on the environment distribution, not anymore both on $s' \sim \varepsilon, a' \sim \pi$.
 This is an advantage, because computing only one expectation requires less samples and we do not have to worry for any bias from the policy in Q-updates.
@@
 This is beautiful because at first, we do not have to sample over many actions to learn this expectation.
-But more importantly, this term also works off policy \cite[Section 4.2]{silverDeterministicPolicyGradient2013} and with a Q function that approximates based on any other policy $Q^{\omega}$.
-So now we can use a replay buffer again and learn from past experiences, though we did not generate them with our policy.
+But more importantly, this term also works off policy, and without importance sampling.
+We can use a different exploration policy $\beta$ and replace our expectation term $\mathbb{E}_{s \sim \rho^{\mu}}$ with $\mathbb{E}_{\rho^{\beta}}$ here completely unpenalized.
+Why is this so?
+\cite{degrisOffPolicyActorCritic2013} proved off-policy gradients possible, but had to use importance sampling to correct for the bias in $\beta$.
+In fact, generally, it is not possible to do this replacement.
+However, \cite[Section 4.3]{silverDeterministicPolicyGradient2013} formulated a compatibility theorem.
+They proved, that a linear function approximator $Q^{\omega}$ which minimizes the MSE between $Q^{\omega}$ and $Q^{\mu}$ will retain the gradient, even when trained on samples generated by $\beta$, and called this Q \textit{compatible}.
+However, in practice we are not minimizing that loss and neither are we using linear function approximators.
+\cite{bhatnagarConvergentTemporalDifferenceLearning2009} proved that using a non-linear function approximator to predict the value function $V(s)$ will converge towards a local optimum.
+As we can formulate our Q by using V, we transfer this to our scenario.
+However, \ac{DPG} is not definitely proven to converge for a non-linear function approximator, as this transfer was, to our knowledge, never fully formulated.
+Observations show that it still does converge.
+
+So let's restate our objective function for the actor.
+
+\begin{equation}
+J(\mu_{\theta}) = \int_S \rho^{\beta}(s) Q^{\omega}(s,\mu_{\theta}(s)) ds = \mathbb{E}_{s \sim \rho^{\beta}}[Q^{\omega}(s,\mu_{\theta}(s))]
+\end{equation}
 
@@
 
-\ac{DDPG} is a combination of \ac{DPG} and \ac{DQN} with several improvements.
+\begin{figure}
+  \centering
+  \includegraphics[width=0.6\linewidth]{DDPG_Setup.png}
+  \caption{DDPG Network}
+  \label{fig:ddpgsetup}
+\end{figure}
+
+\ac{DDPG} is a combination of \ac{DPG} and \ac{DQN}.
 \cite{lillicrapCONTINUOUSCONTROLDEEP2015} tackle several problems in the two algorithms to combine them.
-At first, \ac{DQN} updates the Q network parameters based on a target calculated on the Q network parameters.
-To tackle this, \cite{mnihPlayingAtariDeep2013} introduced a second Q network which they call target network for which the parameters are held constant over a period of updates.
+At first, traditional Q learning updates the Q network parameters based on a target $y$ calculated on the Q network parameters themselves.
+This introduces instability and to tackle this, \cite{mnihPlayingAtariDeep2013} introduced a second Q network $Q'$ which they call target network, for which the parameters are held constant over a period of updates.
+They calculate the $y$ based on the second network $Q'$.
 After a number of steps, the parameters from the actual Q network are then copied over to the target.
-In \ac{DDPG}, they instead use soft target updates, where every step, the parameters of the actual network fade over to the target: $\theta_{target} = \tau \theta + (1 - \tau) \theta_{target}$.
-The fade-over parameter $\tau$ was recommended to be set to a small number like 0.01.
-Additionally to the target Q-network, they also found that a target policy improves stability, and they update it the same way as the main policy.
+The stability gain in their paper depends greatly on the number of steps after which to perform an update.
+In \ac{DDPG}, they instead use smooth target updates, where every step, the parameters of the actual network fade over to the target: $\theta_{target} = \tau \theta + (1 - \tau) \theta_{target}$.
+The fade-over parameter $\tau$ was recommended to be set to a small number like 0.001.
+Additionally to the target Q-network, they also found that a target policy $\mu'$ improves stability, and they update it the same way as the main policy.
+But this is a lot at once.
+Let's have a look at the Q update functions
+
+\begin{equation}
+y = r(s, a) + \gamma Q'(s', \mu'(s';\theta^{\mu'});\theta^{Q'})
+\end{equation}
+\begin{equation}
+L(\theta^Q) = \mathbb{E}_{s \sim \rho^{\beta}, a \sim \beta, s' \sim \varepsilon}[(Q(s, a;\theta^Q) - y)^2]
+\end{equation}
+
+We sketched an overview in \ref{fig:ddpgsetup}.
+Remember that actor was another name for policy and critic another name for Q estimator.
+Notice how for calculating the Q targets, they used both the target actor $\mu'$ for generating actions which are then evaluated by the target critic $Q'$.
+This is their version of the improvement over \ac{DQN} for their instable Q function, as they also observed instabilities.
+Note how the Q function here does not really predict the value of following a specific policy anymore, as before we could always denote which policy Q was trained on.
+Also, this Q is approximated by a non-linear function approximator instead of the linear one from compatible \ac{DPG}.
+\cite{lillicrapCONTINUOUSCONTROLDEEP2015} do not provide a proof of convergence for this new, mixed Q update.
+Still, we have an off-policy way of improving the critic now.
+The actor is directly copied from \ac{DPG} even though it does not fulfil the \textit{compatible} property.
+
+\begin{equation}
+J(\theta^{\mu}) = \mathbb{E}_{s \sim \rho^{\beta}}[Q(s, \mu(s;\theta^{\mu});\theta^Q)]
+\end{equation}
+
+Let's derive these two equations. First the policy gradient:
+
+\begin{equation}
+\Delta_{\theta^{\mu}}J(\theta^{\mu}) = \mathbb{E}_{s \sim \rho^{\beta}}[\Delta_{\theta^{\mu}}Q(s, \mu(s;\theta^{\mu});\theta^Q)] = \mathbb{E}_{s \sim \rho^{\beta}}[\Delta_{\theta^{\mu}}Q(s, a;\theta^Q | a=\mu(s)) \Delta_{\theta^{\mu}}\mu(s;\theta^{\mu})]
+\end{equation}
+
+And the Q update:
+\begin{equation}
+\Delta_{\theta^Q} L(\theta^Q) = \mathbb{E}_{s \sim \rho^{\beta}, a \sim \beta, s' \sim \varepsilon}[(Q(s, a;\theta^Q) - y)\Delta_{\theta^Q}Q(s, a;\theta^Q)]
+\end{equation}
+
+Similar to other reinforcement learning algorithms, we will again not compute the full gradient but use stochastic gradient ascend with batch learning.
+To rephrase gradient ascend to gradient descend, they multiply the policy loss with $-1$.
+In fact, in the original \ac{DDPG} algorithm, they used Adam \cite{kingmaAdamMethodStochastic2017} optimization instead of stochastic gradient descend.
 

### 2019-12-03 paper/paper.tex
Added death condition handling, fixed some bugs, wrote PER section

@@
 Bild"ubernahmen zitieren.}
-\acknowledgments{Die Danksagung...}
+\acknowledgments{At first thanks to my advisors Mattew Lennie and Dmitrij Shlezinger for their support.
+Also, I guess, thanks David Silver - 7 of my most important papers were from you.
+Thanks to my laptop for almost dying, but then surviving.
+And big thanks to the mighty god of windturbines for letting me crush $\num{12345}$ windturbines for science - yes I counted.}
 
@@
 In fact, in the original \ac{DDPG} algorithm, they used Adam \cite{kingmaAdamMethodStochastic2017} optimization instead of stochastic gradient descend.
+As activations, they used relu \cite{agarapDeepLearningUsing2019}.
+Their network architecture for the actor consisted of two fully connected layers with 400 and 300 units and instead of relu they used tanh activations in the last layer to bound actions.
+For the critic, they had a version which learnt upon pixel outputs and used convolutional layers, while we only look at the low-dimensional version.
+For this, they also used a two layer variant with 400 and 300 units.
+State inputs were fed through the whole network, while actions were concatenated in onto the second layer, skipping the first.
+
+Furthermore, to aid exploration, they added noise in form of a Ornstein-Uhlenbeck process \cite{uhlenbeckTheoryBrownianMotion1930}
 
@@
 
-Unfortunately, we are not able to provide a scientific comparison between different methods, as our learning proved too instable to derive sane metrics from it.
-Thus, this section will be structured as an assortment of problems and their solutions
+For everyone who is interested in a recap of our design process, in this section we will provide such.
+We will again present our algorithm in a later section.
+To understand our design decisions, it might be interesting to read this passage, but you can also skip this passage and just read the rest.
 
@@
 OpenAI Gyms generally serve as benchmarks for reinforcement learning algorithms and most foundational papers and articles evaluate their solutions based on how well they perform in OpenAI Gyms.
-Starting with a reference-implementation in pytorch by \cite{DeepDeterministicPolicy}, we experimented with the pendulum environment.
-We observed that the algorithm got stuck in a local optimum, where it rotates the pendulum instead of holding it upright, for long periods of the training process.
-Only with high noise, we could get it to hold a pendulum upright.
-We also tried to run the HalfCheetah task, where a two-dimensional dog-like being should be brought to running and moving its joints, but the results were more a crawling and shaking towards the goal.
-However, we hoped that the implementation might yield results on our windturbine environment already and exchanged the OpenAI gym for qblade.
+We managed to find a solution with our DDPG implementation.
+We hoped that the implementation might yield results on our windturbine environment as well and exchanged the OpenAI gym for qblade.
 
@@
 
-In \cite{mnihPlayingAtariDeep2013} they use an epsilon-greedy strategy, where with a certain small probability epsilon, a random action is taken.
-As we have a continous action space, we can add a little bit of noise towards every action: \(a' = a + n \).
-We first tried a gaussian noise: \(n \leftarrow \mathcal{N}(\mu, \sigma^2) \).
-We didn't see any improvements over our old behavior, so we switched to an Ornstein-Uhlenbeck process \cite{uhlenbeckTheoryBrownianMotion1930} as recommended in the \ac{DDPG} paper \cite[formula (7)]{lillicrapCONTINUOUSCONTROLDEEP2015}.
+We first tried a gaussian noise instead of the recommended \ac{OU-Noise} \cite{uhlenbeckTheoryBrownianMotion1930} from \ac{DDPG}: \(n \leftarrow \mathcal{N}(\mu, \sigma^2) \).
+We didn't see any improvements over our old behavior, so we switched to \ac{OU-Noise}.
 Additionally to the \ac{DDPG} paper, we let the sigma parameter decay over time, so that the noise gets less when training has proceeded.
+We hope to only need this noise in the first stage of the training and then, later, see better results.
+Our decay is parameterized by a start sigma, end sigma and end step and then held constant at the end sigma.
 
@@
 As we found that both random policies rarely explore a sensible operational range, we later added an expert PID controller, which implements torque control, and combined that with our Ornstein-Uhlenbeck process.
-We set the parameters to that controller by experimentation as we did not need the actual control parameters of the controller.
-However it could be imagined to use an established controller here.
-The PID controller was easily able to keep the windturbine in a sensible operation range, and without noise it resulted in a smooth and constant operation.
-Though our PID controller is certainly inferior to industry-grade controllers, we will view it as an expert controller which is already achieving a certain degree of quality.
+We set the parameters to that controller by experimentation and came up with the following equation, observing rotational speed $rot$ from the environment: $(torque, pitch) = (\num{2e7}*rot-\num{12e6}, 128*rot-103)$.
+These actions were clipped to our action bounds.
+So effectively, we designed a P controller, and only using this, we could already achieve relatively stable control.
+As further work, it could be imagined to use an established controller here, but we did not need this for our purposes.
+The PID controller was easily able to keep the windturbine in a sensible operation range, and without adding noise it resulted in a smooth and constant operation.
 
@@
 We tackled this problem by at first reducing the number and size of layers from the example implementation to one fully connected layer of 64 neurons in the critic and one fully connected layer of 32 neurons in the actor.
-The reference-implementation of \cite{DeepDeterministicPolicy} utilized 3 levels in the critic with 1024, 512 and 300 fully connected neurons and in the actor 2 levels of 512 and 128 fully connected neurons.
 Reducing complexity partially tackled the problem of zero gradients, we could then observe small gradients which however vanished over time.
@@
 We normalized the data so states, actions and reward always stayed between 1 and -1.
+This is a normal technique in machine learning, and we are irritated by the fact that this is not part of the default algorithm.
+\cite{vanhasseltLearningValuesMany2016} propose an integration mechanism into \ac{RL}, but we decided to implement a more straight-forward variant.
+At first, we used observed states and rewards from the random exploration phase to calculate maximum/minimum values.
+As sometimes after random exploration, no extreme values were observed, we started normalising on the min/max of replay data of a previous run.
+We could also imagine manually setting normalization bounds on low-dimensional problems, as sometimes human knowledge about the possible state space is usually present.
 This finally fixed our problem of zero gradients and we could observe normal learning.
@@
 As most of the high-magnitude values are vibrations, the net could not create a link between these and the state-action, while low-magnitude but insightful observations like rotational speed were not considered due to their small absolute values.
-At first, we used observed states and rewards from the random exploration phase to calculate maximum/minimum values.
-As sometimes after random exploration, extreme values were observed, we started normalising on the replay data of a previous run.
-We could also imagine manually setting normalization bounds on low-dimensional problems, as sometimes human knowledge about the possible state space is present.
 
@@
 \cite{schaulPrioritizedExperienceReplay2016} proposed a method to sample experiences from the replay buffer according to how much they benefit training, and not just uniformly.
+This method is called \ac{PER} According to how much they benefit means that we add a priority to the experiences, which we somehow derive from the performance of the algorithm on that samples.
 In their paper, they showed how especially cliffwalk problems, but effectively all problems, benefit from using prioritized experience replay.
-In their case, they use it for \ac{DQN}, but we will apply it to \ac{DDPG}, as our windturbine problem is also such a problem in which rewards are sparse.
+We argue that our windturbine is similar to such a cliffwalk, as random exploration usually only shortly passes sensible operation range and then quickly destroys the turbine.
+Seeing this short high reward might not be enough for the controller to learn that being there is good.
+In the original case, they use it on \ac{DQN}, but we will apply it to \ac{DDPG}.
+There hasn't been much work on how to use \ac{PER} in conjunction with \ac{DDPG}, in fact we only found two papers: \cite{houImprovingDDPGPrioritized} and \cite{zhaExperienceReplayOptimization2019}.
+We will at first present the general prioritized experience replay algorithm, then we will discuss the papers and then present how we incorporated \ac{PER}.
 
-As knowing how much a sample benefits training is difficult, they propose to approximate this importance by the temporal difference error $\tau = y - Q(s, a)$.
+As knowing how much a sample benefits training is difficult, \cite{schaulPrioritizedExperienceReplay2016} propose to approximate this importance by the \ac{TD-Error} $\delta = y - Q(s, a)$.
 This difference, which in Q learning is comparable to the loss of the approximator, tells us how far off the prediction was in this step.
-So, alongside with the experiences, a sampling probability for each item $p_i$ is stored in the replay buffer.
-To be able to efficiently sample from it, we implemented it as proposed in their paper in the form of a sum-tree.
+So, alongside with the experiences, a sampling priority for each item $p_i$ is stored in the replay buffer.
+In theory, we would have to calculate this \ac{TD-Error} every time we update our critic.
+This would mean to feed our entire replay buffer through the critic on every update step.
+Clearly, this is a huge performance hit, which is why in the algorithm, they approximated by only updating the error on samples that are anyways being used for training.
+We calculate the \ac{TD-Error} anyways for the critic update, so we can store this without performance penalty.
+New samples are always stored with maximum priority to make sure they are sampled at least once.
 
-When replaying, the sampling probability of a sample is set to $P(j) = \frac{p_j^{\alpha}}{\sum_i{p_i^{\alpha}}}$, $\alpha$ being a hyperparameter allowing to fade between uniform sampling ($\alpha=0$) and only priority sampling ($\alpha=1$).
-New samples are always stored with maximum priority, and to correct for oversampling bias, a correction term is applied to the temporal difference error: $w_j = \frac{(N*P(j))^{-\beta}}{\max_i{w_i}}$.
-It is simply multiplied to the real TD-error.
+When replaying, the sampling probability of a sample is set to $P(j) = \frac{p_j^{\alpha}}{\sum_i{p_i^{\alpha}}}$, $\alpha$ being a hyperparameter allowing to fade between uniform sampling ($\alpha=0$) and pure priority based sampling ($\alpha=1$).
+Because we introduced oversampling bias by sampling some experiences more often than others, a correction term is applied to the \ac{TD-Error}: $w_j = \frac{(N*P(j))^{-\beta}}{\max_i{w_i}}$.
+It is simply multiplied to the real \ac{TD-Error}.
 $\beta$ is a parameter describing how strongly to correct oversampling, where $\beta=1$ is full correction and $\beta=0$ no correction at all.
+This importance sampling is a common technique in reinforcement learning and we have addressed it before in our background part.
 
-We decided to split policy and critic updates, as the oversampling effects could also happen to policy updates, and still sample uniformly for policy updates.
+In addition to sampling directly based on $\abs{\delta}$, \cite{schaulPrioritizedExperienceReplay2016} propose a variant where they sample according to the rank of an experience: $p_i = \frac{1}{\text{rank}(i)}$ where rank is defined as the position in the replay memory when sorted according to $\abs{\delta}$.
+This method is more robust to outliers and slightly increases performance, but because of the extra computational effort of having to calculate the rank every time, we decided to go for the first method.
+Also, \cite{zhaExperienceReplayOptimization2019} measured worse performance of the rank based version in contrast to even plain \ac{DDPG}.
+
+The importance sampling is also the reason why we deem it not so trivial to introduce \ac{PER} to \ac{DDPG}, as there is no specified way how to integrate this into policy updates.
+But first we will describe two other efforts.
+\cite{houImprovingDDPGPrioritized} implemented it straightforward on the Q-Learning part of \ac{DDPG} and ignored any effects on policy updates.
+They measured improvements tackling the OpenAI Pendulum task in contrast to vanilla \ac{DDPG}.
+However, they only evaluated this pendulum task, which we already suspect not to be representative of our windturbine problem.
+We prefer to rely on the work of \cite{zhaExperienceReplayOptimization2019} - the team has formulated experience replay as a learning problem and they trained another prediction network on which replay to choose.
+They evaluated their learnt experience replay against sampling based on rank or priority and found their algorithm to outperform both, while rank-based sampling performed far worse than even vanilla \ac{DDPG} and priority based sampling slightly better than vanilla \ac{DDPG}.
+However, the team completely disregarded importance sampling in their work.
+Also, we deem it as unnecessary to introduce yet another learning task, especially because we are calculating on weak hardware.
+
+Because neither of the works seemed solid enough for us to implement it, but because the learning improvements in Q-learning are very promising, we decided to create our own mixture of \ac{PER} and \ac{DDPG}.
+We decided to implement the Q-Learning part straightforward as in \cite{schaulPrioritizedExperienceReplay2016} and including importance sampling.
+We used the reformulated \ac{TD-Error} of \ac{DDPG} with the target actor and critic.
+To avoid oversampling bias in the policy, we only applied \ac{PER}-sampling to the training part of the critic and trained the actor on uniformly sampled data.
+It can be argued, that as the policy $\mu$ is anyways being trained on the state density function of another policy $\rho^{\beta}$, it can deal with that oversampling bias.
+However we doubt adding more bias is beneficial to the training, so we decided to sample uniformly for the policy.
+
+Because we saw performance problems storing the priorities alongside the experiences and directly sampling from a 100000-item weighted array, we followed the advice of \ac{PER} to use a sumtree datastructure, a binary weighted tree in which each node holds the sum of weights of the children.
+We implemented sampling as a tree walk starting at the root node.
+At every node, we normalized the priorities of the two child nodes to probabilities, did a random draw and recursively repeated until we reached a leaf.
+This recursive sampling technique retains the sample probabilities of the leaf nodes but is cheaper in complexity than to normalize the whole array each timestep and do a random draw over the whole probability distribution.
+All our operations on the replay buffer are in $O(log(n))$ complexity now.
 
@@
 We could already greatly improve our results by using huber-loss \cite{huber1964} for the critic instead of mean squared error loss as described in DDPG paper.
-However we still saw some loss explosions.
+In fact, the authors of \ac{DQN} have caused some confusion on this in the \ac{RL}-community.
+In their journal paper \cite{mnihHumanlevelControlDeep2015}, they write \"We also found it helpful to clip the error term from the update [...] to be between -1 and 1\", which could be interpreted as loss clipping.
+However they actually switched to absolute error instead of square error after those boundaries, which is the definition for huber loss.
+
+Although loss explosions gotten less frequent and smaller in magnitude with this update, we still saw some.
 
@@
 As in deep learning, batch normalization is helpful for increasing lr and thus to skip over local maxima, we tried to add batch normalization between the layers.
-The authors suspected problems with the noise in RL, however it looked good.
+We did get a performance boost when using this, however we ended up with NaN values after a while of running it, especially when experiencing q loss explosions.
+As these unexpected nan-values made training more complicated, we decided to not use batch normalization.
 
-\section{Automatized hyperparameter search}
+\subsection{Regarding death conditions}
 
-To test all the above things, we tried simulations with different hyperparameters.
+In \ac{DQN}, death conditions are accounted for by changing the Q target in the terminal step to be $y = r$ instead of $y = r + \gamma \max_a' Q(s', a')$.
+This way, behavior after a death will not be accounted for by the Q function.
+As we are seeing many death conditions, we thought it could be useful to include this into our DDPG algorithm.
+So now, instead of $y = r + \gamma Q'(s', \mu'(s'))$ we change it to be $y = r + \gamma Q'(s', \mu'(s')) * (1-d)$ where $d \in {0, 1}$ symbolize the death conditions.
+
+\chapter{Algorithm}
+
+After our experimentation phase, we will present our algorithm, on which we perform our evaluation.
+At first, we describe how we encapsulated the qblade environment.
+At second, we will present our replay buffer, our network structure, our exploration strategy and our update algorithm.
+
+\section{QBlade}
+
+As mentioned before, we set the yaw angle to always 0 degrees, as in reality controlling the orientation of the nacelle is trivial and in our simulations, wind will always come from the front.
+The generator torque on our NREL 5MW turbine has a sensible range of \num{0} to ca \num{4e6} Nm, we used a range of \num{0} to \num{8e6} as action space, as higher torque can be simulated by qblade and this way made it easier for the controller, as it can apply high torques to slow down quickly rotating blades.
+Because changing torque from zero to maximum in one timestep is not realistic for a real turbine, we restrict it to move a maximum of \num{3e5} Nm per one timestep of 0.1 seconds.
+If our agent chooses a value outside of that area, we set it to the closest value inside of that area.
+Blades can be pitched between 0 and 90 degrees, 0 being not pitched out at all and operating at maximum efficiency and 90 resulting in no aerodynamic torque from the rotor whatsoever.
+A sensible pitch motor can only turn the blades at a limited speed, we assume a limit of 5 degrees per second and again set controller inputs to a sensible value inside that.
+The simulation theoretically accepts a full pitch change in one timestep, however as inertia on the blade is so high for such a change, the blades break instantly and the rest of the simulation needs to be reset.
+
+QBlade simulates blade damage, also high rotational speeds cause blades to fall off.
+We however observed problems with our controller when reaching these extreme limits of the simulation In the course of our experiments, we decided to reset the simulation at an rotational speed of 3 rad/s, as our turbine normally operates at ca 0.8 rad/s.
+Also, high generator torque inputs cause the simulation to rotate the rotor backwards with a negative energy yield, effectively creating a multimillion-dollar leafblower.
+As this is neither a realistic scenario, we also reset the simulation at -0.5 rad/s.
+We tried to clip at 0, but this resulted in so frequent deaths that learning actions at this border was not possible.
+
+When resetting the simulation, we let it run for 100 timesteps with zero action and state, as the elastic structure needs a bit of time to move from simulation default to elastic equilibrium.
+In this time, simulation results do not make sense yet.
+Also, every 100000 timesteps or after seeing NaN values we reload the project file to maximize simulation stability.
+
+We hid all observations except for power and rotational speed from the algorithm and also merged all pitch controls into a single variable.
+
+\section{Core algorithm}
+
+As DDPG is capable of learning from a replay of experiences, we implemented a replay buffer.
+To enable \ac{PER}, we needed to sample according to priorities proportional to a metric, in our case \ac{TD-Error}.
+We implemented the replay buffer as a sum tree datastructure, a binary weighted tree in which each node holds the sum of the weights of the nodes below.
+The root node thus always held the sum of all metrics.
+Also, we did not store probabilities right away but kept the unnormalized experiences $p_i = (\abs{\sigma} + \epsilon)^{\alpha}$ with an epsilon of $\num{1e-6}$ and $\alpha = 0.5$.
+
+We implemented our actor model according to the \ac{DDPG} paper, but with less neurons per layer.
+Instead of 400 and 300 neurons, we use 32 and 16 neurons.
+Our critic is also according to the \ac{DDPG} paper but with less neurons.
+Instead of 400 and 300 we use 64 and 32 neurons.
+All states, rewards and actions seen by the network are normalized to be between $[-1, 1]$.
+
+We start our simulation with an expert exploration.
+We designed a P-controller which can keep the turbine at a constant rotational speed and added \ac{OU-Noise} to it to explore sensible operation range.
+For 3 epochs, we let our expert policy with noise control the turbine before we hand over to our \ac{DDPG}-Agent.
+During the handover, the policy is pretrained for 10k steps, lr 1e-4, adam optimizer, mse loss to directly predict actions from the expert controller.
+This way we started off early with better actions.
+Then, critic and actor training is done on the experiences from expert exploration.
+After this training, all priorities in the replay buffer are set to their maximum (so they all will be seen) and normal training starts.
+Each epoch lasts for 2000 steps and at the beginning of it, the simulation is reset to default state.
+We trained for 150 epochs.
+
+Critic updates are done similar to \ac{DQN} updates with \ac{PER}. We use importance sampling to correct for \ac{PER} overestimation bias and regard death conditions.
 
@@
 This result in itself is notable, because it needed a whole set of additions to plain \ac{DDPG} to get first sensible results on our windturbine environment.
-So apparently, our windturbine simulation is more complex to learn than the OpenAI gym examples, which are usually used as benchmarks for \ac{RL} performance, or we are having errors in our programme.
-As the final product we developped is close to \cite{fujimotoAddressingFunctionApproximation2018} and has improvements from a wide variety of papers, we conclude that tackling wind turbine control with \ac{RL} will need more research.
+So either, our windturbine simulation is more complex to learn than the OpenAI gym examples, which are usually used as benchmarks for \ac{RL} performance, or we are having errors in our programme.
+As the final product we developped has improvements from a wide variety of papers but still does not produce stable results, we conclude that tackling wind turbine control with \ac{RL} will need more research.
 Though we were not able to outperform industry controllers, the speed in which the area of \ac{RL} is advancing makes us hopeful to achieve good results in the near future.
 With our work on integrating qblade, we provided an open-source, accessible framework to develop future reinforcement learning controllers for wind turbines.
+
+To demonstrate the instability, we ran 5 training runs with the same hyperparameters and measured return, epoch return, killcount and time to q explosion (where we define q explosion as q loss > 1e4)
 

### 2019-12-03 python/automated_hparam_search.py

@@ -12,17 +12,6 @@ import os
   {"id": 5, "batch_size": 256},
 ]'''
 
-experiments = [
-  {"id": 30, "critic_lr": 1e-2},
-  {"id": 31, "critic_lr": 1e-3},
-  {"id": 32, "critic_lr": 1e-5},
-  {"id": 33, "critic_lr": 1e-6},
-  {"id": 34, "actor_lr": 1e-2},
-  {"id": 35, "actor_lr": 1e-3},
-  {"id": 36, "actor_lr": 1e-5},
-  {"id": 37, "actor_lr": 1e-6},
-]
-
 '''experiments = [
   {"id": 11, "twin_critics": True},
   {"id": 12, "twin_critics": True, "critic_loss": 'mse'},
@@ -43,7 +32,28 @@ experiments = [
   {"id": 27, "gamma": 0.8},
 ]'''
 
+'''experiments = [
+  {"id": 30, "critic_lr": 1e-2},
+  {"id": 31, "critic_lr": 1e-3},
+  {"id": 32, "critic_lr": 1e-5},
+  {"id": 33, "critic_lr": 1e-6},
+  {"id": 34, "actor_lr": 1e-2},
+  {"id": 35, "actor_lr": 1e-3},
+  {"id": 36, "actor_lr": 1e-5},
+  {"id": 37, "actor_lr": 1e-6},
+]'''
 
+
+experiments = [
+  {"id": 38},
+  {"id": 39},
+  {"id": 40},
+  {"id": 41},
+  {"id": 42},
+  {"id": 43},
+  {"id": 44},
+  {"id": 45},
+]
 
 
 def run_one(experiment):

### 2019-12-03 python/ddpg/ddpg.py

@@ -50,8 +50,8 @@ class DDPG:
             assert(not np.any(self.action_grad_denormalizer == 0))
         
         # initialize actor and critic networks
-        self.critic = Critic(self.obs_dim, self.act_dim, hparams.critic_sizes[0], hparams.critic_sizes[1], hparams.critic_sizes[2], hparams.init_weight_limit, hparams.critic_simple).to(self.device)
-        self.critic_target = Critic(self.obs_dim, self.act_dim, hparams.critic_sizes[0], hparams.critic_sizes[1], hparams.critic_sizes[2], hparams.init_weight_limit, hparams.critic_simple).to(self.device)
+        self.critic = Critic(self.obs_dim, self.act_dim, hparams.critic_sizes[0], hparams.critic_sizes[1], hparams.init_weight_limit, hparams.critic_simple).to(self.device)
+        self.critic_target = Critic(self.obs_dim, self.act_dim, hparams.critic_sizes[0], hparams.critic_sizes[1], hparams.init_weight_limit, hparams.critic_simple).to(self.device)
         
         self.actor = Actor(self.obs_dim, self.act_dim, hparams.actor_sizes[0], hparams.actor_sizes[1], hparams.init_weight_limit, hparams.actor_simple).to(self.device)
         self.actor_target = Actor(self.obs_dim, self.act_dim, hparams.actor_sizes[0], hparams.actor_sizes[1], hparams.init_weight_limit, hparams.actor_simple).to(self.device)
@@ -62,11 +62,11 @@ class DDPG:
         for target_param, param in zip(self.actor_target.parameters(), self.actor.parameters()):
             target_param.data.copy_(param.data)
         
-        if(self.hparams.clip_gradients):
-            for p in self.actor.parameters():
-                p.register_hook(lambda grad: torch.clamp(grad, -self.hparams.clip_gradients, self.hparams.clip_gradients))
-            for p in self.critic.parameters():
-                p.register_hook(lambda grad: torch.clamp(grad, -self.hparams.clip_gradients, self.hparams.clip_gradients))
+        #if(self.hparams.clip_gradients):
+        #    for p in self.actor.parameters():
+        #        p.register_hook(lambda grad: torch.clamp(grad, -self.hparams.clip_gradients, self.hparams.clip_gradients))
+        #    for p in self.critic.parameters():
[…]

### 2019-12-03 python/ddpg/models.py

@@ -3,7 +3,7 @@ import torch.nn as nn
 import torch.nn.functional as F
 
 class Critic(nn.Module):
-    def __init__(self, obs_dim, action_dim, hidden_size_1 = 1024, hidden_size_2 = 512, hidden_size_3 = 300, init_w = 0.1, simple = False, batch_normalization = False):
+    def __init__(self, obs_dim, action_dim, hidden_size_1 = 1024, hidden_size_2 = 512, init_w = 0.1, simple = False, batch_normalization = False):
         super(Critic, self).__init__()
 
         self.obs_dim = obs_dim
@@ -15,7 +15,6 @@ class Critic(nn.Module):
         if(batch_normalization):
             self.bn1 = nn.BatchNorm1d(hidden_size_1)
             self.bn2 = nn.BatchNorm1d(hidden_size_2)
-            self.bn3 = nn.BatchNorm1d(hidden_size_3)
         else:
             self.bn1 = lambda x: x
             self.bn2 = lambda x: x
@@ -28,12 +27,10 @@ class Critic(nn.Module):
         else:
             self.linear1 = nn.Linear(self.obs_dim, hidden_size_1)
             self.linear2 = nn.Linear(hidden_size_1 + self.action_dim, hidden_size_2)
-            self.linear3 = nn.Linear(hidden_size_2, hidden_size_3)
-            self.linear4 = nn.Linear(hidden_size_3, 1)
-            print('Initialized Critic with %d, %d, %d, %d, %d' % (self.obs_dim + self.action_dim,
-                                                              hidden_size_1,
+            self.linear3 = nn.Linear(hidden_size_2, 1)
+            print('Initialized Critic with %d, %d, %d, %d' % (self.obs_dim,
+                                                              hidden_size_1 + self.action_dim,
                                                               hidden_size_2,
-                                                              hidden_size_3,
                                                               1))
 
         self.init_weights(init_w)
@@ -44,7 +41,6 @@ class Critic(nn.Module):
 
         if(not self.simple):
             self.linear3.weight.data.uniform_(-init_w, init_w)
-            self.linear4.weight.data.uniform_(-init_w, init_w)
 
     def first_layer(self, x, a):
         if(self.simple):
@@ -56,8 +52,7 @@ class Critic(nn.Module):
         xa = F.relu(self.bn1(self.linear1(x)))
         xa = torch.cat([xa,a], 1)
[…]

### 2019-12-03 python/ddpg/utils.py

@@ -256,7 +256,7 @@ class BasicBuffer:
 
         done = 1 if done else 0
 
-        experience = (state, action, np.array([reward]), next_state, done)
+        experience = (state, action, np.array([reward]), next_state, np.array([done]))
 
         self.max_priority = max(self.max_priority, priority)
 

### 2019-12-03 python/env/qbladeadapter.py

@@ -27,15 +27,11 @@ class QBladeAdapter:
 
     print("loading qblade library")
     qbladeLib = ctypes.CDLL('../qblade_build/libQBlade.so')
-
-
     self._qbladeLib = qbladeLib
-
     self.map_functions(qbladeLib)
 
     print("creating qblade instance")
     self._qbladeLib._Z14createInstancev()
-
     self.load_default_project()
 
     print("qblade init done")
@@ -66,17 +62,21 @@ class QBladeAdapter:
     x = ctypes.c_char_p(b"../sample_projects/NREL_5MW_STR.wpa")
     print("loading qblade project")
     self._loadProject(x)
+    self.steps_since_reload = 0
 
   def reset(self):
     print('Resetting simulation')
+    if(self.steps_since_reload>100000):
+      self.load_default_project()
     self._initializeSimulation(ctypes.c_int(0))
     self.lastAction = np.zeros(5)
     self.storeAction(np.zeros(self.get_act_dim()))
 
     # Do some steps with zero action
-    #action = np.zeros(self.get_act_dim())
-    #for i in range(0, 5):
-    #  self.step(action)
+    action = np.zeros(self.get_act_dim())
+    self.storeAction(action)
+    for i in range(0, 100):
+      self._advanceSingleTimestep()
 
     print('reset done')
 
@@ -101,7 +101,7 @@ class QBladeAdapter:
     return [0, 0]
 
   def get_act_max_grad(self):
-    return [3e5, 0.5]
+    return [3e4, 0.5]
 
 
   def calc_reward(self, observation, action, death):
@@ -121,7 +121,9 @@ class QBladeAdapter:
     return 1-np.abs((observation[0]-rated_speed)/rated_speed)-death_penalty
 
   def calc_death(self, observation):
-    # If there are nan values, accumulate, after a few, die
+    # If there are nan values, accumulate, after a few, reload
+    self.steps_since_reload += np.sum(np.isnan(observation)) * 100
+
     if(np.any(np.isnan(observation))):
       return True
 
@@ -164,6 +166,7 @@ class QBladeAdapter:
     return np.array(obs[0:2])
 
   def step(self, action):
+    self.steps_since_reload += 1
 
     self.storeAction(action)
     self._advanceSingleTimestep()

### 2019-12-03 python/main.py

@@ -81,6 +81,7 @@ def main():
 
     if(reset):
       o = env.reset()
+      print('Reset at step %d' % t)
       a = agent.reset_finalize(o)
 
   # Run a test run after training if wanted

### 2019-12-04 paper/paper.tex
Finished background section, some improvements on the qblade adapter

@@
 
+\usepackage{fancyref}
 \usepackage{acro}
@@
 \usepackage{commath}
+\usepackage[ruled]{algorithm2e}
 \input{abbreviations}
@@
 
+\begin{figure}
+  \centering
+  \includegraphics{NREL-5MW.png}
+  \caption{Control curve of the NREL 5MW turbine, from \cite{jonkmanDefinition5MWReference2009}}
+  \label{fig:nrel5mw}
+\end{figure}
+
 As described in \cite[sec 8.3]{burtonWindEnergyHandbook2011}, the normal controller design for windturbines divides the operation of a turbine into two parts, below rated and above rated.
-Rated is a windspeed, in which the turbine gives maximum energy on highest aerodynamic effiency.
-At this point, the generator runs on its maximum torque and the blades are pitched to the optimum angle of 0 degrees.
-At this point, the tip-speed-ratio, which is a good indicator for loads on the blades is at maximum too.
+To explain this, we added reference plot of the windturbine model which we will use in our further testing, the NREL 5MW reference turbine \cite{jonkmanDefinition5MWReference2009}.
+Figure \ref{fig:nrel5mw} describes several properties of that turbine in relation to windspeed.
+Rated windspeed is a windspeed, in which the turbine gives maximum energy on highest aerodynamic effiency.
+In our example figure, this point is at 11.4m/s inflowing wind.
+At this point, the generator runs on its maximum torque (green graph) and the blades are pitched to the optimum angle of 0 degrees (red graph).
+At this point, the tip-speed-ratio (purple graph) is at its designed spot.
+When designing the turbine blades, this tip speed ratio plays an important role.
 Below that windspeed, the rotor generates less rotational torque than the maximum generator torque, thus the generator torque needs to be reduced to not slow the rotor down too much.
 Slowing the rotor down complicates the energy generation process, and though variable-speed wind turbines allow for some play, all windturbines are limited to a certain operational range.
-Fixed-speed wind turbines can only run on one single rotational speed and thus need to change the torque more agressively.
+Fixed-speed wind turbines can only run on one single rotational speed and thus need to change the torque more aggressively.
+In our reference turbine, you can see how the rotor speed is kept above 6.9 rpm, which is the cut-in speed of the NREL 5MW turbine.
+As the generated power of a windturbine is proportional to its rotor speed times generator torque, we produce less than our rated energy in this area.
 Above rated windspeed, the blades need to be pitched out to stop the rotor from spinning at faster speeds than what the turbine was designed for.
-The generator torque can be kept at maximum in this area, but to protect the turbine from damage, aerodynamic efficiency of the rotor is reduced by turning the blades along their long axis and thus moving their angle of attack against the incoming air into a less optimal range.
+See the red line in the reference plot - blades are kept at zero below rated and from then on control is done only through blade pitch change.
+The generator torque can be kept at maximum in this area and such will also the energy production stay at maximum.
+To protect the turbine from damage, aerodynamic efficiency of the rotor is reduced by turning the blades along their long axis and thus moving their angle of attack against the incoming air into a less optimal range.
+This is done in a way that the rotational speed of the rotor stays exactly constant.
 
@@
 Also, both generator torque and blade pitch can be used to counteract natural resonant frequencies of the structure, reducing material stress through extensive swinging of structural parts.
+In our 5MW reference turbine, only blade pitch is used to counteract resonances.
 
 Usually this control is implemented by two PID controllers, which are hand-designed as described in \cite[sec 8.4]{burtonWindEnergyHandbook2011}.
+PID-Controllers in general have one real-valued input over time and deliver an output.
+This is done by calculating an error term between the input and a desired input.
+The output is calculated as a mixture of the error itself (P - proportional to the error), integrating the error over the last few timesteps (I - integral of the error) and calculating a derivative to the last timestep (D - derivative).
+The output is the sum of these 3 terms each factorized with a constant factor.
+However back to windturbines.
+Usually, we implement two PID controllers to control a turbine, a torque and a pitch controller.
 Below rated windspeed, the controller for torque is active, above rated the one for pitch.
-The parameters to those controllers can be calculated according to the laws of control theory, and the resulting controllers work reasonably close to theoretical maximum.
-% TODO find reference
+The parameters to those controllers can be calculated according to the laws of control theory.
+Sometimes, this strong split into two operational ranges is relaxed a bit, as in the 5MW controller, the pitch part counteracts resonances also below rated speed.
+From this simple concept, the resulting controllers work reasonably close to theoretical maximum already.
+Our reference turbine has a peak power coefficient of 0.482 - meaning that at optimal wind speed, 48\% of the energy of the wind is converted into electricity.
+In fact, the theoretical maximum is given by the Betz limit of 59.3\% and if we account for electrical and frictional losses, the realistic maximum is even lower.
+So there is not much potential to be alleviated.
+Also, only a fraction of this potential lies in the responsibility of the control circuit.
+Blade design has so far been the major area for improvements.
+
 Still, in this work, we are trying to replace these hand-designed controllers with a reinforcement-learning algorithm.
+We are not expecting to exceed the performance of a industry controller.
+So why do we even do this?
+
+\subsection{Why we do what we do}
+As you might have seen before, all controllers follow a certain style.
+They take a single input variable, such as rotor speed, and adjust a single output such as torque.
+This type of controller is called \ac{SISO}.
+\ac{SISO} controllers are easy to design and work reliably.
+They can also be interconnected, such as with the pitch controller which gets a vibration input and a rotor speed input to combine them to one output.
+The main challenge with wind turbine control at the moment is however not to optimize for power, but to optimize for life-time \cite[Chapter 4]{vankuikLongtermResearchChallenges2016}.
+The controllers are dealing well with adjusting energy to a good level, but they are missing out on keeping the windturbine intact over a long life time.
+
+Also, the outputs get more complex.
+Back in the early days of wind turbines, you could control 3 variables - pitch, torque and yaw.
+Nowadays, we can control each of the blades individually and with a quick enough response time to wiggle the blades around within one single rotor rotation.
+Research on adding active control elements to the blades such as flaps is happening, which would add many more control parameters.
+You might have seen flaps on an aircraft wing, extending the length of the wing or adding gaps at takeoff and landing to account for the low wind speeds hitting the aircraft wing in these flight scenarios.
+This could also benefit wind turbines at low wind speeds.
+Tackling all this with \ac{SISO} is possible, but a lot of manual work.
+
+Next, the inputs also get more complex.
+It is getting cheaper to install high-precision sensors, and many of them.
+Wind measurement techniques such as LIDAR based aenometers are able to efficiently and precisely predict incoming wind situations.
+Load measurements can be done in real-time across a turbine blade and in theory, it would be possible to incorporate these into control systems.
+However, accomodating all these input parameters in \ac{SISO} systems is difficult as we would have to hand-model each of the inputs and how exactly it ends up in the output.
+
+We hope to lay a foundational step in solving these issues by using an end-to-end machine learning algorithm which is capable of learning a wind turbine control scenario with many inputs and outputs.
+Also, we must admit that we are generally interested in playing around with reinforcement learning, it's fucking cool.
 
@@
 
-This section should aim to give a brief overview into Reinforcement Learning assumptions and derive briefly a couple of methods until in the end, we will describe \ac{DDPG}, the method we utilized to control our virtual windturbine.
+So, why are we using reinforcement learning?
+We have posed the challenge earlier, that we want to learn a control system with a high number of inputs and outputs.
+Reinforcement learning has proven to be able to tackle high-complexity problems, both in input and output dimension.
+There have been successes in learning to play atari games based on pixel inputs \cite{mnihPlayingAtariDeep2013} and outperform humans, there have been successes in playing games with very long-term strategies such as go \cite{silverMasteringGameGo2016}, and research is continuing to improve.
+Because of all these successes, we try to automatically learn a windturbine controller.
+
+Unfortunately, reinforcement learning isn't possible without a good amount of maths, so the following sections will be mainly mathematical.
 
@@
 
+Additionally to the state and reward, an environment can optionally send out a done signal \(d\), which indicates a state in which further simulation is not possible and the environment wants to be reset to an initial state.
+In a chess game, this would be a loss or a win, and with a windturbine that could for example be fatal structural damage.
+
 In every timestep, the agent supplies an action \(a\).
@@
 In other words, it should steer the environment to achieve best performance.
-
-Additionally to the state and reward, an environment can optionally send out a done signal \(d\), which indicates a state in which further simulation is not possible and the environment wants to be reset to an initial state.
-In a chess game, this would be a loss or a win, and with a windturbine that could for example be fatal structural damage.
-As this is commonly ignored in reinforcement learning, we will also ignore it.
+Let's write down this mission.
+We want to maximize:
+
+\begin{equation}
+J = \mathbb{E}r
+\end{equation}
+
+This formula is still a bit incomplete - we didn't yet specify our expectation.
+And wasn't reward defined on states and actions?
+But that is because we are still missing some definitions.
+Luckily, there are some ahead.
 
 \subsection{Definitions}
+\label{sec:definitions}
 
@@
 These two frequently appear together, if that happens we are using an actor-critic algorithm.
+  \item \textbf{State-density} According to \cite{suttonPolicyGradientMethods}, $d^{\pi}(s) = \lim_{t -> \infty} p(s_t = s | s_0, \pi)$ describes how likely it is to be in state s when following a policy $\pi$ forever.
+If we for example have a policy which plays go as well as the authors of this paper, states with a lot of enemy stones and very few own stones might be very likely, and a state without enemy stones at all would be very unlikely.
+Also according to \cite{suttonPolicyGradientMethods} this distribution is independent of the starting state $s_0$ and only dependent on the policy and of course the environment.
+In later papers, commonly a discounted version of this is used.
+$\rho^{\pi}(s)$ weights future probabilities less and thus can be described as the probability of being in state s \textit{soon} when following policy $\pi$.
+We will omit a precise definition here, you can look it up at \cite[Section 2.1]{silverDeterministicPolicyGradient2013}.
 \end{itemize}
 
+Now, with all these definitions at hand, we can specify our mission a bit better. We want to maximize
+\begin{equation}
+J^{\pi} = \mathbb{E}_{s \sim \rho^{\pi}, a \sim \pi}r(s, a) = \int_S \rho^{\pi}(s) \int_A \pi(s) r(s, a)
+\label{equ:maximize_rewards}
+\end{equation}
+
+In other words, we want a very high probability of being in a state ($\rho^{\pi}$) which might give good rewards, and in that state we want a high probability of taking an action ($\pi$) which might give a good reward.
+In our wind turbine, rewards could for example be the all the costs and incomes summed together, so at every point in time we could count income by energy yield or costs because of maintenance (with our controller more likely costs for rebuilding the entire turbine).
+But before we choose $\pi$, there is a bit of a problem ahead.
+We can observe some rewards, but only very few of them.
+We might know that using 0.3MN of torque at rotational speed 11rpm is a good idea, but how about rotational speed 11.1rpm?
+We might have never seen that.
+Also, rotating the turbine at 15rpm might be a good strategy to immediately achieve high power outputs, but we might need to fix our turbine after running this policy for just a few days and incur a heavy penalty.
+Luckily we have defined Q-Values above.
+We could use them somehow, so let's discuss how to learn them.
+
 \subsection{Q-Learning}
 
-At first we will present a very basic algorithm for a machine learning agent, which is based on the basic Bellman-Equation \cite{bellmanTheoryDynamicProgramming1954}. It is proven, that
+At first we will present a very basic algorithm for a Q learning agent, which is based on the basic Bellman-Equation \cite{bellmanTheoryDynamicProgramming1954}. It is proven, that
 
@@
 So, for our work, we have to also learn a policy and not just a Q estimator.
-Luckily, our Q learning rule also works for other policies than the optimal policy, so let's try to derive a way to learn a policy.
+So let's try to derive a way to learn a policy.
 
@@
 The algorithm is called \ac{REINFORCE} (no joke) and was originally presented by \cite{williamsSimpleStatisticalGradientfollowing}.
-Then we will add a Q term to it according to \cite{suttonPolicyGradientMethods} and is called \ac{PG}.
-\cite{degrisOffPolicyActorCritic2013} generalized that term and made an off-policy version of policy gradients.
+Then we will add a Q term to it according to \cite{suttonPolicyGradientMethods} and call it \ac{PG}.
+\cite{degrisOffPolicyActorCritic2013} generalized that term and made an off-policy version of policy gradients, which we will not discuss in detail.
 That version was then used by \cite{silverDeterministicPolicyGradient2013} and the stochastic policy was replaced by a deterministic one, the algorithm was called \ac{DPG}.
-This algorithm finally lays the basis for \ac{DDPG} by \cite{lillicrapCONTINUOUSCONTROLDEEP2015}, which is what we used until finally optimizing it towards \ac{TD3} \cite{fujimotoAddressingFunctionApproximation2018}.
+This algorithm finally lays the basis for \ac{DDPG} by \cite{lillicrapCONTINUOUSCONTROLDEEP2015}, which is what we used to start our experimentation.
 
@@
 \theta_{i+1} = \theta_i + \alpha r(s,a) \Delta \pi_{\theta_i}(a|s)
+\label{equ:reinforce}
 \end{equation}
@@
 \theta_{i+1} = \theta_i + \alpha Q^{\pi}(s, a) \Delta \log \pi_{\theta_i}(a|s)
-\end{equation}
-
-This update equation is our basic policy gradient equation.
-
-Before, we ignored for brevity that we are dealing with expectation calculations and that we are doing these updates under a stochastic environment, so to write it down formally correct, we need to introduce the density function of a state under a policy.
-According to \cite{suttonPolicyGradientMethods}, $d^{\pi}(s) = \lim_{t -> \infty} p(s_t = s | s_0, \pi)$ describes how likely it is to be in state s when following a policy $\pi$ forever, and also according to them this distribution is independent of the starting state $s_0$ and only dependent on the policy and environment.
-\cite{silverDeterministicPolicyGradient2013} used a discounted version of this, $\rho^{\pi}(s)$, which weights future probabilities less and thus can be described as the probability of being in state s \textit{soon} when following policy $\pi$.
-Using the discounted version approximates the Monte-Carlo experiments we are doing better than the undiscounted version, as effectively we do not train forever on one constant policy, but change it frequently.
+\label{equ:pg_update}
+\end{equation}
+
+So, let's sum up what we have done before and formulate our goal again.
+Before, we ignored for brevity that we are dealing with expectation calculations and that we are doing these updates under a stochastic environment, so to write it down formally correct, we need to remember the density function of a state under a policy from section \ref{sec:definitions}.
+Using the discounted version approximates the Monte-Carlo experiments we are doing better than the undiscounted version, as effectively we do not train forever on one constant policy, but change it frequently, so we will use $\rho^{\pi}$ to approximate how likely we are in a state $s$ when using policy $\pi$.
 
@@
 \end{equation}
-
-We explained intuitively above, and \cite{suttonPolicyGradientMethods} proves that deriving this yields
+This looks similar to equation \ref{equ:maximize_rewards}, just that we use Q now.
+
+\cite{suttonPolicyGradientMethods} proves that deriving this yields
 \begin{equation}
 \Delta J(\pi) = \mathbb{E}_{s \sim \rho^{\pi}, a \sim \pi}[\Delta \log \pi(a|s)Q^{\pi}(s, a)]
-\end{equation}
+\label{equ:pg}
+\end{equation}
+This equation is very famous in reinforcement learning and is called policy gradients.
+And by coincidence, it is equivalent to what we intuitively derived in equation \ref{equ:pg_update}.
+We decided to skip the actual proof and use this intuitive explanation instead, and we used the intuitive explanation in \cite[Chapter 13, Section 3, after Equation 13.8]{suttonReinforcementLearningIntroduction2018} to explain the nature of this policy gradient equation.
 
@@
 
-We sketched an overview in \ref{fig:ddpgsetup}.
+We sketched an overview in figure \ref{fig:ddpgsetup}.
 Remember that actor was another name for policy and critic another name for Q estimator.
@@
 \begin{equation}
-\Delta_{\theta^{\mu}}J(\theta^{\mu}) = \mathbb{E}_{s \sim \rho^{\beta}}[\Delta_{\theta^{\mu}}Q(s, \mu(s;\theta^{\mu});\theta^Q)] = \mathbb{E}_{s \sim \rho^{\beta}}[\Delta_{\theta^{\mu}}Q(s, a;\theta^Q | a=\mu(s)) \Delta_{\theta^{\mu}}\mu(s;\theta^{\mu})]
+\Delta_{\theta^{\mu}}J(\theta^{\mu}) = \mathbb{E}_{s \sim \rho^{\beta}}[\Delta_{\theta^{\mu}}Q(s, \mu(s;\theta^{\mu});\theta^Q)] = \mathbb{E}_{s \sim \rho^{\beta}}[\Delta_a Q(s, a;\theta^Q | a=\mu(s)) \Delta_{\theta^{\mu}}\mu(s;\theta^{\mu})]
 \end{equation}
@@
 
-Furthermore, to aid exploration, they added noise in form of a Ornstein-Uhlenbeck process \cite{uhlenbeckTheoryBrownianMotion1930}
+Furthermore, to aid exploration, they added noise in form of a Ornstein-Uhlenbeck process \cite{uhlenbeckTheoryBrownianMotion1930}.
+
+For readability, we added a copy of their algorithm in Algorithm \ref{alg:vanilladdpg}, with minor adjustments to fit our notation.
+
+\begin{algorithm}
+\SetAlgoLined
+Randomly initialize critic network $Q(s, a;\theta^Q)$ and actor $\mu(s;\theta^{\mu})$ with weights $\theta^Q$ and $\theta^{\mu}$\\
+Initialize target networks $Q'$ and $\mu'$ with weights $\theta^{Q'} \leftarrow \theta^Q, \theta^{\mu'} \leftarrow \theta^{\mu}$\\
+Initialize replay buffer $\mathcal{R}$\\
+
+\For{episode = 1, M}{
+  Initialize a random process $\mathcal{N}$ for action exploration\\
+  Receive initial observation state $s_1$\\
+  \For{t = 1, T}{
+    Select action $a_t = \mu(s_t;\theta^{\mu}) + \mathcal{N}_t$ according to the current policy and exploration noise\\
+    Execute action $a_t$ and observe reward $r_t$ and observe new state $s_{t+1}$\\
+    Store transition $(s_t, a_t, r_t, s_{t+1})$ in $\mathcal{R}$\\
+    Sample a random minibatch of $N$ transitions $(s_i, a_i, r_i, s_{i+1})$ from $\mathcal{R}$\\
+    Set $y_i = r_i + \gamma Q'(s_{i+1}, \mu'(s_{i+1};\theta^{\mu'});\theta^{Q'})$\\
+    Update critic by minimizing the loss: $L = \frac{1}{N} \sum_i (y_i - Q(s_i, a_i;\theta^Q))^2$\\
+    Update the actor policy using the sampled policy gradient: $\Delta_{\theta^{\mu}} J \approx \frac{1}{N} \sum_{i} \Delta_a Q(s_i, a;\theta^Q | a=\mu(s_i)) \Delta_{\theta^{\mu}}\mu(s_i;\theta^{\mu})$\\
+    Update the target networks:\\
+    $\theta^{Q'} \leftarrow \tau \theta^{Q} + (1-\tau)\theta^{Q'}$\\
+    $\theta^{\mu'} \leftarrow \tau \theta^{\mu} + (1-\tau)\theta^{\mu'}$
+  }
+}
+\caption{Vanilla DDPG algorithm}
+\label{alg:vanilladdpg}
+\end{algorithm}
 
@@
 The team behind that paper built a miniature model of a windturbine and used a variation of the \ac{REINFORCE} algorithm to control it.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.5\linewidth]{Miniturbine.png}
+  \caption{The mini wind-turbine from \cite{kolterDesignAnalysisLearning2012}}
+  \label{fig:miniturbine}
+\end{figure}
+
+Their miniature model was built from cheap and low-scale parts, they simulated a wind-tunnel by attaching several fans in front of a wooden tube and at the end placed their turbine.
+Figure \ref{fig:miniturbine} illustrates this.
+Each of the blades of that miniature turbine is individually controllable and through a correction term, they are able to achieve independent pitch control which is able to pitch the motors to a value based on rotor position.
+Though this turbine rotates many times faster than a normal, full scale wind turbine, the underlying principles are the same (500rpm vs 11rpm).
+For us most importantly, they used reinforcement learning to control this turbine.
+
+As training input, they used a single observation in contrast to popular reinforcment learning which operates on high-dimensional observation space.
+This observation was power output of the turbine, and they treated it as reward to the system, not having any state observations.
+Their action space consisted of electrical load on the generator and blade pitch, which was uniformly applied to all 3 blades.
+In fact, they designed their policy as to return a gaussian distribution over these two parameters directly without any input: $\pi: \varnothing \rightarrow P(\mathbb{R}^2)$.
+To run the training, they kept the wind speed constant, let the turbine spool up to speed and then let it run with a set of parameters returned by the policy.
+They ran these parameters for 2 seconds on the turbine, averaged power output over that time into a reward and then ran an iteration of their algorithm to generate new parameters, starting with random parameters.
+They first tested the \ac{REINFORCE} algorithm, with operates without a Q network and trains a policy directly on rewards.
+This algorithm was discussed above and utilizes the following gradient equation:
+
+\begin{equation}
+\Delta_{\pi^{\theta}} J = \mathbb{E}_{a \sim \pi, r \sim \varepsilon | a} r \Delta \log \pi(a;\theta)
+\end{equation}
+
+Note, how they denoted J in their paper as rewards, while we are using r for rewards and J for objective functions.
+Also, they used $\omega$ as a symbol for the action taken, while we use $a$.
+Also note how their policy is not dependent on a state but directly predicts actions.
+They did not explicitely note what they formulated their expectation term on, but we added it here for clarity.
+Note the similarity to what we discussed in equation \ref{equ:reinforce} and \ref{equ:pg}.
+In addition to this, they came up with an algorithm which they call Trust Region Policy Search, and which is astonishingly different from Trust Region Policy Optimization.
+With a few additions, they incorporated off-policy learning and second order error approximation into their algorithm, and showed later that it performs nicely.
+They were able to find optimal pitch and load settings as soon as 15 timesteps after the start of the simulation.
+
+Having described this paper, we want to also summarize what we do differently and why.
+First, they had a simpler environment.
+While a full scale windturbine has a strong tendency to destroy itself when outside of their operational margin, their turbine did not fly apart when not applying braking.
+Also, they were setting electrical resistance, not rotor torque.
+This system applies a braking force constant to rotational speed times resistance parameter, and thus already acts as a P controller.
+We were setting torque, so we had to adjust it regularly to rotor speed.
+In general, their turbine behaved more concavely opposed to our turbine, which with a constant parameter set, would settle to either rotating quickly backwards, forwards or not at all.
+Because of this, they were able to give the turbine some time at the beginning to stabilize before applying reinforcement learning while we needed to start with the standing turbine.
+Lastly and most importantly, they aimed to find optimal control settings for a constant wind speed.
+To create a full controller setup from this, they would have to repeat this process for every windspeed and use the resulting look-up table to derive a PID-controller.
+We however want to create an end-to-end system and want to experiment with the possibly higher expressive nature of predicting actions based on a full state input.
+We want to create a version as general as possible.
+Also, as \ac{DDPG} is a direct successor on \ac{REINFORCE} and has proven to outperform it in many areas, we do not reimplement their paper.
 
@@
 As we found that both random policies rarely explore a sensible operational range, we later added an expert PID controller, which implements torque control, and combined that with our Ornstein-Uhlenbeck process.
-We set the parameters to that controller by experimentation and came up with the following equation, observing rotational speed $rot$ from the environment: $(torque, pitch) = (\num{2e7}*rot-\num{12e6}, 128*rot-103)$.
-These actions were clipped to our action bounds.
+We set the parameters to that controller by experimentation and came up with the following equation, only dependent on rotational speed $\mathrm{rot}$: $\mathcal{C}:(\mathrm{rot})\rightarrow(\mathrm{torque}, \mathrm{pitch}), \mathcal{C}(\mathrm{rot}) = (\num{2e7}*\mathrm{rot}-\num{12e6}, 128*\mathrm{rot}-103)$.
+These actions were clipped as in section \ref{sec:clip_act_grad}.
 So effectively, we designed a P controller, and only using this, we could already achieve relatively stable control.
@@
 The PID controller was easily able to keep the windturbine in a sensible operation range, and without adding noise it resulted in a smooth and constant operation.
+We summarize this in algorithm \ref{alg:expertexplo}
+
+\begin{algorithm}
+\SetAlgoLined
+Use replay buffer $\mathcal{R}$ from normal training\\
+
+\For{episode = 1, E}{
+  Initialize expert controller $\mathcal{C}$ and random process $\mathcal{N}$ for expert exploration\\
+  Receive initial observation state $s_1$\\
+  \For{t = 1, T}{
+    Select action $a_t = \mathrm{clip}(\mathcal{C}(s_t) + \mathcal{N}_t)$
+    Execute action $a_t$ and observe reward $r_t$, death condition $d_t$ and new state $s_{t+1}$\\
+    Store transition $(s_t, a_t, r_t, s_{t+1}, d_t)$ in $\mathcal{R}$\\
+    On death condition, reset the environment.
+  }
+}
+\caption{Expert exploration}
+\label{alg:expertexplo}
+\end{algorithm}
 
@@
 Reducing complexity partially tackled the problem of zero gradients, we could then observe small gradients which however vanished over time.
+Later, we found that two layers of 64 and 32 in the critic and 32 and 16 neurons in the actor also still yielded good gradients with higher generalization potential.
 
@@
 More effectively, we added data normalization.
-We normalized the data so states, actions and reward always stayed between 1 and -1.
-This is a normal technique in machine learning, and we are irritated by the fact that this is not part of the default algorithm.
-\cite{vanhasseltLearningValuesMany2016} propose an integration mechanism into \ac{RL}, but we decided to implement a more straight-forward variant.
-At first, we used observed states and rewards from the random exploration phase to calculate maximum/minimum values.
-As sometimes after random exploration, no extreme values were observed, we started normalising on the min/max of replay data of a previous run.
-We could also imagine manually setting normalization bounds on low-dimensional problems, as sometimes human knowledge about the possible state space is usually present.
-This finally fixed our problem of zero gradients and we could observe normal learning.
-We recommend this technique as a default technique for every reinforcement learning algorithm.
-As in our simulation, where our state reflects measurements from very different parts of the simulation, some values in the state-array were 8 orders of magnitude different from others.
+As in our simulation, where our state reflects measurements from very different parts of the simulation, some values in the state-array were 8 orders of magnitude different from others as they were measured in very different units.
 As most of the high-magnitude values are vibrations, the net could not create a link between these and the state-action, while low-magnitude but insightful observations like rotational speed were not considered due to their small absolute values.
+
+We normalized the data so states, actions and reward usually stayed between 1 and -1.
+This is a normal technique in machine learning, and we are irritated by the fact that the reinforcement learning community largely made no mention of normalization.
+At times, rewards were clipped \cite{mnihPlayingAtariDeep2013} and observations were sometimes, especially when working on pixel inputs, cropped to be of square dimensions.
+Only in 2016, \cite{vanhasseltLearningValuesMany2016} brought up the topic.
+They propose an adaptive normalization mechanism which integrates into a generic \ac{RL} algorithm and which can normalize all inputs adaptively.
+We decided to implement a more straight-forward variant.
+At first, we used observed states and rewards from the random exploration phase to calculate the 5\%, 95\% quartile values.
+We calculated normalization constants from it to normalize those quartiles to $[-1, 1]$.
+As sometimes during random exploration, no extreme values were observed and later extreme values lay far outside of $[-1, 1]$, we instead normalized on replay data of a previous run.
+Also, we clipped inputs which lay outside of $[-3, 3]$, as sometimes in boundary conditions, qblade would produce observations far outside normal operational ranges.
+We could also imagine manually setting normalization bounds on low-dimensional problems, as human knowledge about the possible state space is usually present.
+In case when human knowledge about input magnitudes is not present, we recommend the mechanism from \cite{vanhasseltLearningValuesMany2016}.
+
+This finally fixed our problem of zero gradients and we could observe normal learning. We recommend this technique as a default technique for every reinforcement learning algorithm.
 
@@
 \subsection{Clipping action gradients}
-
+\label{sec:clip_act_grad}
 We eventually solved the problem with the high action gradients by still letting the controller output absolute actions across the entire action spaces, but if gradients exceed a certain threshold, the actual action taken is set to the nearest value with sensible gradients.
+We defined our sensible gradient vector $g = (\num{3e4}, 0.1)$ for pitch and torque.
+Clipping action $a$ and previos action $a'$ can be defined as $\mathrm{clip}(a_i, a'_i) = \min(\max(a_i, a'_i -g_i), a'_i + g_i)$.
+The results of this are again clipped to be in the action space.
 Though the actions returned from the policy are still wildly unstable, they at least didn't result in death of the turbine anymore and still we didn't hinder convergence through our gradient actionspace.
@@
 We hoped for a more sensible default policy at the beginning and indeed observed reaching sensible behavior a little earlier.
+We summarize this in algorithm \ref{alg:pretrain}, parametrized with training time $T=10000$ and batch size $N=32$
+
+\begin{algorithm}
+\SetAlgoLined
+Use prefilled replay buffer $\mathcal{R}$ from expert exploration\\
+Use expert policy $\mathcal{C}$ from expert exploration\\
+\For{t = 1, T}{
+  Sample a random minibatch of $N$ states $(s_i)$\\
+  Calculate noise-free expert actions $a_i = \mathcal{C}(s_i)$\\
+  Update the actor by minimizing the loss: $L = \frac{1}{N}\sum_i(\mu(s_i) - a_i)^2$\\
+}
+\caption{Policy pretraining}
+\label{alg:pretrain}
+\end{algorithm}
 
@@
 \cite{schaulPrioritizedExperienceReplay2016} proposed a method to sample experiences from the replay buffer according to how much they benefit training, and not just uniformly.
-This method is called \ac{PER} According to how much they benefit means that we add a priority to the experiences, which we somehow derive from the performance of the algorithm on that samples.
+This method is called \ac{PER}.
+According to how much they benefit training means that we add a priority to the experiences, which we somehow derive from the performance of the algorithm on that samples.
 In their paper, they showed how especially cliffwalk problems, but effectively all problems, benefit from using prioritized experience replay.
@@
 Seeing this short high reward might not be enough for the controller to learn that being there is good.
+Furthermore, the technique is generally promising as in Rainbow-\ac{DQN} \cite{hesselRainbowCombiningImprovements2017}, \ac{PER} made up for most of the gain on the algorithm.
 In the original case, they use it on \ac{DQN}, but we will apply it to \ac{DDPG}.
@@
 \subsection{Batch normalization}
-As in deep learning, batch normalization is helpful for increasing lr and thus to skip over local maxima, we tried to add batch normalization between the layers.
+As in deep learning, batch normalization is found helpful for increasing lr and thus to skip over local maxima \cite{bjorckUnderstandingBatchNormalization}.
+It works by normalizing the outputs between the layers to be of zero mean and unit standart deviation, and for not losing generalization adding an extra learnt parameter per output dimension.
 We did get a performance boost when using this, however we ended up with NaN values after a while of running it, especially when experiencing q loss explosions.
+We can explain this because our prioritized experience replay most likely sampled only one item at a timee.
+To circumvent this, we clipped individual priorities.
 As these unexpected nan-values made training more complicated, we decided to not use batch normalization.
@@
 
-Critic updates are done similar to \ac{DQN} updates with \ac{PER}. We use importance sampling to correct for \ac{PER} overestimation bias and regard death conditions.
+Critic updates are done similar to \ac{DQN} updates with \ac{PER}.
+We use importance sampling to correct for \ac{PER} overestimation bias and regard death conditions.
+All in all, algorithm \ref{alg:ourddpg} summarizes all improvements:
+
+\begin{algorithm}
+\SetAlgoLined
+Randomly initialize critic network $Q(s, a;\theta^Q)$ and actor $\mu(s;\theta^{\mu})$ with weights $\theta^Q$ and $\theta^{\mu}$\\
+Initialize target networks $Q'$ and $\mu'$ with weights $\theta^{Q'} \leftarrow \theta^Q, \theta^{\mu'} \leftarrow \theta^{\mu}$\\
+Initialize replay buffer $\mathcal{R}$\\
+
+Execute algorithm \ref{alg:expertexplo} for expert exploration \\
+
+Execute algorithm \ref{alg:pretrain} for policy pretraining \\
+
+\For{episode = 1, M}{
+  Initialize a random process $\mathcal{N}$ for action exploration\\
+  Receive initial observation state $s_1$\\
+  \For{t = 1, T}{
+    Select action $a_t = \mathrm{clip}(\mu(s_t;\theta^{\mu}) + \mathcal{N}_t)$ according to the current policy and exploration noise\\
+    Execute action $a_t$ and observe reward $r_t$, death condition $d_t$ and new state $s_{t+1}$\\
+    Store transition with highest priority $(s_t, a_t, r_t, s_{t+1}, d_t, \max_i(p_i))$ in $\mathcal{R}$\\
+    Calculate sampling probabilities $P(i) = \frac{p_i}{\sum_j p_j}$\\
+    Sample a random minibatch of $N$ transitions $(s_i, a_i, r_i, s_{i+1}, d_i)$ from $\mathcal{R}$ according to $P$\\
+    Calculate importance sampling weights $w_i = (N \times P(i))^{-\beta}$ and normalize $ w_i = \frac{w_i}{\max_j w_j}$\\
+    Set $y_i = r_i + \gamma Q'(s_{i+1}, \mu'(s_{i+1};\theta^{\mu'});\theta^{Q'}) (1 - d_i)$\\
+    Calculate TD-Error $\delta_i = y_i - Q(s_i, a_i;\theta^Q)$\\
+    Store TD-Errors in the replay buffer: $p_i = (\abs{\delta_i}+\epsilon)^{\alpha}$\\
+    Apply importance sampling $\delta_i = \delta_i w_i$\\
+    Update critic by minimizing the huber loss: $L = \frac{1}{N} \sum_i \begin{cases} (\delta_i)^2 & \mathrm{if} \delta < 1\\ \abs{\delta_i} & \mathrm{else}\end{cases}$\\
+    Sample a random minibatch of $N$ states $(s_i)$ from $\mathcal{R}$ uniformly\\
+    Update the actor policy using the sampled policy gradient: $\Delta_{\theta^{\mu}} J \approx \frac{1}{N} \sum_{i} \Delta_a Q(s_i, a;\theta^Q | a=\mu(s_i)) \Delta_{\theta^{\mu}}\mu(s_i;\theta^{\mu})$\\
+    Update the target networks:\\
+    $\theta^{Q'} \leftarrow \tau \theta^{Q} + (1-\tau)\theta^{Q'}$\\
+    $\theta^{\mu'} \leftarrow \tau \theta^{\mu} + (1-\tau)\theta^{\mu'}$\\
+    On death condition, reset the environment
+  }
+}
+\caption{Our DDPG algorithm}
+\label{alg:ourddpg}
+\end{algorithm}
 
@@
 
+\section{Train PID inputs}
+
+As PID controllers have proven well in controlling a turbine and neural nets directly performed poorly, we could imagine to predict PID controller parameters instead of directly building the controller.
+\cite{kolterDesignAnalysisLearning2012} had successes with this approach, and maybe using a more advanced algorithm such as DDPG would result in even better performance.
+
 \section{Reward functions}

### 2019-12-04 paper/watch.sh

@@ -1,0 +1,6 @@
+make watch
+while inotifywait -r -e close_write,modify,move,attrib,delete,create . 
+do
+	echo "updating"
+ 	make watch
+done

### 2019-12-04 python/automated_hparam_search.py

@@ -45,14 +45,14 @@ import os
 
 
 experiments = [
-  {"id": 38},
-  {"id": 39},
-  {"id": 40},
-  {"id": 41},
-  {"id": 42},
-  {"id": 43},
-  {"id": 44},
-  {"id": 45},
+  {"id": 48},
+  {"id": 49},
+  {"id": 50},
+  {"id": 51},
+  {"id": 52},
+  {"id": 53},
+  {"id": 54},
+  {"id": 55},
 ]
 
 
@@ -85,7 +85,7 @@ def run_one(experiment):
     return (id, -1, -1)
 
 # Set the number of parallel runs here
-pool = Pool(2)
+pool = Pool(1)
 results = pool.map(run_one, experiments)
 
 for (id, reward, deaths) in results:

### 2019-12-04 python/ddpg/ddpg.py

@@ -61,12 +61,6 @@ class DDPG:
             target_param.data.copy_(param.data)
         for target_param, param in zip(self.actor_target.parameters(), self.actor.parameters()):
             target_param.data.copy_(param.data)
-        
-        #if(self.hparams.clip_gradients):
-        #    for p in self.actor.parameters():
-        #        p.register_hook(lambda grad: torch.clamp(grad, -self.hparams.clip_gradients, self.hparams.clip_gradients))
-        #    for p in self.critic.parameters():
-        #        p.register_hook(lambda grad: torch.clamp(grad, -self.hparams.clip_gradients, self.hparams.clip_gradients))
 
         # optimizers
         if(self.hparams.optimizer == 'adam'):
@@ -83,10 +77,6 @@ class DDPG:
             for target_param, param in zip(self.critic2_target.parameters(), self.critic2.parameters()):
                 target_param.data.copy_(param.data)
 
-            #if(self.hparams.clip_gradients):
-            #    for p in self.critic2.parameters():
-            #        p.register_hook(lambda grad: torch.clamp(grad, -self.hparams.clip_gradients, self.hparams.clip_gradients))
-
             self.critic2_optimizer = self.optimizer(self.critic2.parameters(), lr=hparams.critic_lr)
 
         if(self.hparams.critic_loss == 'huber'):
@@ -101,7 +91,7 @@ class DDPG:
         self.critic_loss_function = lambda target, pred, weights: generic_loss_function(target, pred, weights)
 
         # Buffor for experience replay
-        self.replay_buffer = BasicBuffer(hparams.buffer_maxlen)
+        self.replay_buffer = BasicBuffer(hparams.buffer_maxlen, self.device, self.obs_dim, self.act_dim, hparams.prioritized_experience_replay_alpha)
 
         # Noises
         # Random exploration
@@ -265,7 +255,6 @@ class DDPG:
             print([(x, y) for (x, y) in self.actor.cpu().named_parameters()])
             assert(False)
 
-        assert(not np.any(np.isnan(action)))
         self.logger.logAction(self.time, action, 'act-unclipped')
 
         action = self.clip_action(action)
@@ -297,7 +286,7 @@ class DDPG:
     def calc_normalizations(self):
         # Load additional data if wanted
         if(self.hparams.normalization_extradata):
-            tmpbuffer = BasicBuffer(100000)
+            tmpbuffer = BasicBuffer(100000, torch.device('cpu'), self.obs_dim, self.act_dim)
[…]

### 2019-12-04 python/ddpg/utils.py

@@ -1,6 +1,7 @@
 import random
 import numpy as np
 import pickle
+import torch
 from torch.utils.tensorboard import SummaryWriter
 from datetime import datetime
 from collections import deque
@@ -195,7 +196,7 @@ class SumTree:
     # Update a leaf and propagate the sums back through the net
     def update_entry(self, leaf_idx, weight):
         self.leaf_weights[leaf_idx] = float(weight)
-        weight = float(weight)**self.alpha
+        weight = np.clip(float(weight)**self.alpha, 0, 1.5)
 
         trajectory = self.leaf_trajectory(leaf_idx)
         # Get the previous weight at this position
@@ -237,13 +238,34 @@ class SumTree:
 # A "Basic" Buffer which stores entries and can sample them by priority
 class BasicBuffer:
 
-    def __init__(self, max_size, alpha=1):
+    def __init__(self, max_size, device, state_dim, action_dim, alpha=1):
+        self.device = device
         self.max_size = max_size
         self.iterator = 0 # Next position to write to
         self.full = False # Whether we are full
+
+        # CPU buffer
         self.buffer = np.array([(None, None, None, None, None) for _ in range(max_size)])
+
+        self.dtype = torch.float
+
+        # GPU buffers
+        self.buffer_s = torch.zeros(max_size, state_dim, dtype=self.dtype, device=device)
+        self.buffer_a = torch.zeros(max_size, action_dim, dtype=self.dtype, device=device)
+        self.buffer_snext = torch.zeros(max_size, state_dim, dtype=self.dtype, device=device)
+        self.buffer_r = torch.zeros(max_size, 1, dtype=self.dtype, device=device)
+        self.buffer_d = torch.zeros(max_size, 1, dtype=self.dtype, device=device)
+
         self.priorities = SumTree(max_size, alpha)
         self.max_priority = 1
+
+    def copy_buffer_to_device(self):
+        for i in range(0, len(self)):
+            self.buffer_s[i] = torch.tensor(self.buffer[i][0])
+            self.buffer_a[i] = torch.tensor(self.buffer[i][1])
+            self.buffer_r[i] = torch.tensor(self.buffer[i][2])
+            self.buffer_snext[i] = torch.tensor(self.buffer[i][3])
+            self.buffer_d[i] = torch.tensor(self.buffer[i][4])
 
     def push(self, state, action, reward, next_state, done, priority=None):
         # If not given a priority, make sure it's getting sampled
[…]

### 2019-12-04 python/env/qbladeadapter.py

@@ -90,10 +90,9 @@ class QBladeAdapter:
     #return 5
     return 2
 
-  # TODO get actual values for this
   def get_act_high(self):
     #return [10, 10, 10, 10, 10]
-    return [8e6, 90]
+    return [474029.1, 90]
 
   # TODO get actual values for this
   def get_act_low(self):
@@ -101,7 +100,8 @@ class QBladeAdapter:
     return [0, 0]
 
   def get_act_max_grad(self):
-    return [3e4, 0.5]
+    timestep = 0.1
+    return [150000*timestep, 8*timestep]
 
 
   def calc_reward(self, observation, action, death):
@@ -118,11 +118,11 @@ class QBladeAdapter:
     #return np.clip(5 
     #  - np.abs((observation[1]-rated_power)/rated_power)
     #  - 5e-2*(np.abs(observation[16]) + np.abs(observation[17]) + np.abs(observation[18])), -10, 10)
-    return 1-np.abs((observation[0]-rated_speed)/rated_speed)-death_penalty
+    return np.clip(1-np.abs((observation[0]-rated_speed)/rated_speed)-death_penalty, -4, 4)
 
   def calc_death(self, observation):
-    # If there are nan values, accumulate, after a few, reload
-    self.steps_since_reload += np.sum(np.isnan(observation)) * 100
+    # If there are nan values, reload completely
+    self.steps_since_reload += 100000
 
     if(np.any(np.isnan(observation))):
       return True

### 2019-12-04 paper/paper.tex
Added hyperparameter class

@@
 
-After a while of experimenting with the unrestricted simulation, we added some bounds to our control parameters. In our final version, we had the following restrictions:
-As mentioned before, we set the yaw angle to always 0 degrees, as in reality controlling the orientation of the nacelle is trivial and in our simulations, wind will always come from the front.
-The generator torque on our NREL 5MW turbine has a sensible range of \num{0} to \num{4e6} Nm.
-Because changing torque from zero to maximum in one timestep is not realistic for a real turbine, we restrict it to move a maximum of \num{3e5} Nm per one timestep of 0.1 seconds.
-If our agent chooses a value outside of that area, we set it to the closest value inside of that area.
-Blades can be pitched between 0 and 90 degrees, 0 being not pitched out at all and operating at maximum efficiency and 90 resulting in no aerodynamic torque from the rotor whatsoever.
-A sensible pitch motor can only turn the blades at a limited speed, we assume a limit of 5 degrees per second and again set controller inputs to a sensible value inside that.
-The simulation theoretically accepts a full pitch change in one timestep, however as inertia on the blade is so high for such a change, the blades break instantly and the rest of the simulation needs to be reset.
-
-QBlade simulates blade damage, also high rotational speeds cause blades to fall off.
-QBlade is not validated to realistically simulate these extreme conditions and thus posed a problem to our controllers.
-In the course of our experiments, we decided to reset the simulation at an rotational speed of 4 rad/s, as our turbine operates at ca 0.8 rad/s.
-Also, high generator torque inputs cause the simulation to rotate the rotor backwards with a negative energy yield, effectively creating a multimillion-dollar leafblower.
-As this is neither a realistic scenario, we also reset the simulation at -0.5 rad/s.
+After a while of experimenting with the unrestricted simulation, we added some bounds to our control parameters. Restrictions to our final version are described in section \ref{sec:alg_qblade}.
 
@@
 In our wind turbine, rewards could for example be the all the costs and incomes summed together, so at every point in time we could count income by energy yield or costs because of maintenance (with our controller more likely costs for rebuilding the entire turbine).
-But before we choose $\pi$, there is a bit of a problem ahead.
-We can observe some rewards, but only very few of them.
-We might know that using 0.3MN of torque at rotational speed 11rpm is a good idea, but how about rotational speed 11.1rpm?
+This work would be easy if we could directly calculate this equation but clearly, there is a bit of a problem ahead.
+Firstly, we don't know $\rho$ as we don't exactly know the environment we are modelling.
+We can observe some rewards, states and actions, but only very few of them.
+We might then see that using 0.3MN of torque at rotational speed 11rpm yields good energy yield, but how about rotational speed 11.1rpm?
 We might have never seen that.
 Also, rotating the turbine at 15rpm might be a good strategy to immediately achieve high power outputs, but we might need to fix our turbine after running this policy for just a few days and incur a heavy penalty.
+So we don't know how rewards will spread out in the future.
 Luckily we have defined Q-Values above.
@@
 \section{QBlade}
+
+\label{sec:alg_qblade}
 
@@
 
+\section{More computational power}
+
+As we weren't able to solve the problem by our reasoning, we could just implement a huge hyperparameter search.
+Limiting factor in this is currently the simulation, which currently uses 95\% of the computation time, while we only use 5\% to train our networks.
+
 \section{Train PID inputs}
 
-As PID controllers have proven well in controlling a turbine and neural nets directly performed poorly, we could imagine to predict PID controller parameters instead of directly building the controller.
-\cite{kolterDesignAnalysisLearning2012} had successes with this approach, and maybe using a more advanced algorithm such as DDPG would result in even better performance.
+As PID controllers have proven well in controlling a turbine and neural nets directly performed poorly, we could imagine to predict PID controller parameters and error targets instead of directly building the controller.
 

### 2019-12-04 python/ddpg/ddpg.py

@@ -9,8 +9,7 @@ from datetime import datetime
 from collections import deque
 
 # For hparams
-import tensorflow as tf
-
+from .hparams import HParams
 from .models import Critic, Actor
 from .utils import *
 
@@ -704,7 +703,7 @@ class DDPG:
             "prefix": prefix,
             "time": str(datetime.now()),
             "killcount": self.killcount,
-            "hparams": self.hparams.values(),
+            "hparams": self.hparams.get_dict(),
             "real_last_action": self.real_last_action.tolist() if self.hparams.action_gradients else None,
         }
         with open('%s/%s_metadata' % (directory, prefix), 'w') as f:
@@ -736,7 +735,7 @@ class DDPG:
 
 
     def get_default_hparams():
-        return tf.contrib.training.HParams(
+        return HParams(
             # The name of the run to make identification easier
             run_name = '',
 

### 2019-12-04 python/ddpg/hparams.py

@@ -1,0 +1,44 @@
+import json
+import re
+
+class HParams:
+  def __init__(self, **first_data):
+    self.data = first_data
+
+  def __str__(self):
+    return str(self.data)
+
+  def get_dict(self):
+    return self.data
+
+  def items(self):
+    return self.data.items()
+
+  def values(self):
+    return self.get_dict()
+
+  def parse_json(self, args):
+    if(not args):
+      return
+
+    data = json.loads(args)
+    for key, value in data:
+      self.data[key] = value
+
+  # parses in the form of a=1,b=2,
+  def parse(self, args):
+    if(not args):
+      return
+
+    for line in re.compile("(\\w*?=(?:\\[.*?\\]|{.*?}|[^,])*)").split(args):
+      if(not line or line == ','):
+        continue
+      [key, value] = line.split('=')
+      try:
+        value = eval(value) # unsafe as fuck
+      except:
+        pass
+      self.data[key] = value
+
+  def __getattr__(self, name):
+    return self.data[name]

### 2019-12-04 python/main.py

@@ -11,6 +11,7 @@ import tensorflow as tf
 import argparse
 import os
 import json
+import time
 
 def main():
 
@@ -68,6 +69,8 @@ def main():
   o = env.reset()
   a = agent.prepare(o)
 
+  execution_time_env = 0
+  execution_time_agent = 0
 
   for t in range(start_time, hparams.steps_per_epoch * hparams.epochs):
     # do a checkpoint if required
@@ -75,9 +78,18 @@ def main():
       file = agent.save_checkpoint(checkpoint_dir)
       print("Checkpoint written to %s" % file)
 
+    start = time.time()
     o, r, d = env.step(a)
-
+    step = time.time()
     a, reset = agent.step(o, r, d)
+    end = time.time()
+    execution_time_env += step - start
+    execution_time_agent += end - step
+
+    if(t%100 == 0):
+      print("Step %d, Execution times env: %f, agent: %f" % (t, execution_time_env, execution_time_agent))
+      execution_time_agent = 0
+      execution_time_env = 0
 
     if(reset):
       o = env.reset()

### 2019-12-05 paper/paper.tex
Wrote TD3 section

@@
 As sometimes during random exploration, no extreme values were observed and later extreme values lay far outside of $[-1, 1]$, we instead normalized on replay data of a previous run.
-Also, we clipped inputs which lay outside of $[-3, 3]$, as sometimes in boundary conditions, qblade would produce observations far outside normal operational ranges.
 We could also imagine manually setting normalization bounds on low-dimensional problems, as human knowledge about the possible state space is usually present.
@@
 
-\subsection{Double critic}
-Adding twin critic networks such as in \cite{fujimotoAddressingFunctionApproximation2018} eventually solved our problem with loss explosions completely, however the thing didn't reach sensible states anymore.
-To use it in conjunction with \ac{PER}, we needed to decide for a priority formula and decided to take the minimum of the two q values, after the maximum showed biasing towards one of the networks.
+\subsection{Double critics}
+
+We also had a look at \cite{fujimotoAddressingFunctionApproximation2018}.
+They propose a set of three mechanisms that address function approximation error in \ac{DDPG}.
+At first, they show in section 4.1 that Q learning with a deterministic policy tends to overestimate true Q values.
+This has been observed in normal Q learning as well, but hasn't been generalized to \ac{DDPG} before.
+They show, that the interaction of the policy and the Q network leads to a residual error accumulating over the recursive nature of Q updates.
+To mitigate this, they at first proposed to use two Q networks and choose the minimum of the two to compute the target $y$.
+They combine this with the target network idea, so effectively we end up with 6 networks now.
+They called this technique Clipped Double Q learning.
+We illustrate this in figure \ref{fig:clippedq}.
+The Q target is computed as
+
+\begin{equation}
+y = r(s, a) + \gamma \min_{i \in {1,2}} Q_i'(s', \mu'(s';\theta^{\mu'});\theta^{Q_i'})
+\end{equation}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{TD3-Setup.png}
+  \caption{Clipped double Q learning}
+  \label{fig:clippedq}
+\end{figure}
+
+As they observed that a quickly changing policy introduces instability into the Q learning process, they proposed to update the policy less often than the actor.
+They do this simply by updating the policy only when $t\mod d = 0$ where $t$ is the timestep and $d$ describes how many timesteps to wait before an actor update.
+They call this \textit{delaying} policy updates.
+Lastly they proposed adding noise in the target policy calculations to achieve smoothing around policy results.
+So they changed the target updates again to:
+\begin{equation}
+\tilde{a} = \mu'(s';\theta^{\mu'}) + \text{clip}(\mathcal{N}(0, \tilde{\sigma}), -\tilde{c}, \tilde{c})\\
+\end{equation}
+\begin{equation}
+y = r(s, a) + \gamma \min_{i \in {1,2}} Q_i'(s', \tilde{a});\theta^{Q_i'})
+\end{equation}
+
+Where $\tilde{a}$ is the target action with added noise on it. In their appendix they describe that they furthermore clip this target action to sensible action range.
+
+When implementing this, we found that we now have two \ac{TD-Error}s which left us with a choice how to implement \ac{PER} in conjunction with this.
+We could either use one of the two errors.
+When implementing this, we found that the critic whose error we chose for \ac{PER} performed way better than the other, effectively rendering it useless to have two critics.
+We also tried to use the maximum and minimum of the two, which yielded more stable results, where minimum seemed better than maximum.
+Finally, we ended up with adding the two errors for prioritisation.
+
+Using the techniques from \cite{fujimotoAddressingFunctionApproximation2018} still didn't solve
 
@@
 So now, instead of $y = r + \gamma Q'(s', \mu'(s'))$ we change it to be $y = r + \gamma Q'(s', \mu'(s')) * (1-d)$ where $d \in {0, 1}$ symbolize the death conditions.
+
+\subsection{Clipping observations}
+
+Finally, we found that the solution to our problem lay in qblade.
+In certain conditions, qblade returned observations 20 orders of magnitude away from normal operation.
+As we based our reward calculation on the observations, the reward was also far off from normal reward.
+This condition happened only every 200k steps or so, and as we did not log every timestep, these spikes didn't end up in our logs.
+It was up to chance, that one of these spikes finally did, and we then decided to clip all normalized values to $[-3, 3]$.
 
@@
 Execute algorithm \ref{alg:expertexplo} for expert exploration \\
-
+Calculate normalization factors based on exploraton buffer and data from a previous run \\
 Execute algorithm \ref{alg:pretrain} for policy pretraining \\
@@
     Select action $a_t = \mathrm{clip}(\mu(s_t;\theta^{\mu}) + \mathcal{N}_t)$ according to the current policy and exploration noise\\
-    Execute action $a_t$ and observe reward $r_t$, death condition $d_t$ and new state $s_{t+1}$\\
+    Execute action $a_t$ in $\varepsilon$ and observe $r_t$, $d_t$ and $s_{t+1}$ and normalize these\\
     Store transition with highest priority $(s_t, a_t, r_t, s_{t+1}, d_t, \max_i(p_i))$ in $\mathcal{R}$\\
@@
 As we weren't able to solve the problem by our reasoning, we could just implement a huge hyperparameter search.
-Limiting factor in this is currently the simulation, which currently uses 95\% of the computation time, while we only use 5\% to train our networks.
+Limiting factor in this is currently the simulation, which currently uses 70\% of the computation time, while we only use 30\% to train our networks.
+As on windows, gpu support in qblade is available, it might be interesting to run qblade on another machine than the training.
 

### 2019-12-05 python/automated_hparam_search.py

@@ -85,7 +85,7 @@ def run_one(experiment):
     return (id, -1, -1)
 
 # Set the number of parallel runs here
-pool = Pool(1)
+pool = Pool(2)
 results = pool.map(run_one, experiments)
 
 for (id, reward, deaths) in results:

### 2019-12-05 python/ddpg/ddpg.py

@@ -12,967 +12,872 @@ from collections import deque
 from .hparams import HParams
 from .models import Critic, Actor
 from .utils import *
-
+from .normalizer import Normalizer
 
 class DDPG:
+  
+  def __init__(self, hparams):
+    self.device = torch.device("cuda" if torch.cuda.is_available() and hparams.use_gpu else "cpu")
     
-    def __init__(self, hparams):
-        self.device = torch.device("cuda" if torch.cuda.is_available() else "cpu")
-        
-        self.hparams = hparams
-        self.obs_dim = hparams.obs_dim
-        self.act_dim = hparams.act_dim
-        self.act_high = np.array(hparams.act_high)
-        self.act_low = np.array(hparams.act_low)
-        self.batch_size = hparams.batch_size
-        self.gamma = hparams.gamma
-        self.tau = hparams.tau
+    self.hparams = hparams
+    self.obs_dim = hparams.obs_dim
+    self.act_dim = hparams.act_dim
+    self.act_high = np.array(hparams.act_high)
+    self.act_low = np.array(hparams.act_low)
+    self.batch_size = hparams.batch_size
+    self.gamma = hparams.gamma
+    self.tau = hparams.tau
 
-        # If we should feed in past data, multiply obs dim by that
-        if(self.hparams.feed_past):
-            self.obs_dim *= self.hparams.feed_past + 1
-            self.past_obs = CapacityBuffer(self.hparams.feed_past)
+    # If we should feed in past data, multiply obs dim by that
+    if(self.hparams.feed_past):
+      self.obs_dim *= self.hparams.feed_past + 1
+      self.past_obs = CapacityBuffer(self.hparams.feed_past)
 
-        # In case of action gradients, reduce action high and low and extend the observation space
-        if(self.hparams.action_gradients):
-            self.real_obs_dim = self.obs_dim
-            self.obs_dim += self.act_dim
-            self.real_act_high = self.act_high
-            self.real_act_low = self.act_low
-            self.act_high = np.ones(self.act_dim)
-            self.act_low = -np.ones(self.act_dim)
-            self.real_last_action = self.hparams.starting_action
-            self.action_grad_denormalizer = (self.real_act_high - self.real_act_low) * self.hparams.action_gradient_stepsize
+    # In case of action gradients, reduce action high and low and extend the observation space
+    if(self.hparams.action_gradients):
+      self.real_obs_dim = self.obs_dim
+      self.obs_dim += self.act_dim
[…]

### 2019-12-05 python/ddpg/hparams.py

@@ -1,6 +1,7 @@
 import json
 import re
 
+# This class is partially compatible to tensorflows training.contrib.HParams class
 class HParams:
   def __init__(self, **first_data):
     self.data = first_data
@@ -11,11 +12,33 @@ class HParams:
   def get_dict(self):
     return self.data
 
+  def override_from_dict(self, new_data):
+    for k,v in new_data:
+      self.data[k] = v
+
   def items(self):
     return self.data.items()
 
   def values(self):
-    return self.get_dict()
+    data = {}
+
+    def is_jsonable(x):
+      try:
+        json.dumps(x)
+        return True
+      except:
+        return False
+
+    for k,v in self.data.items():
+      if(not is_jsonable):
+        data[k] = str(v)
+      else:
+        data[k] = v
+
+    return data
+
+  def to_json(self):
+    return json.dumps(self.values())
 
   def parse_json(self, args):
     if(not args):

### 2019-12-05 python/ddpg/normalizer.py

@@ -1,0 +1,155 @@
+import torch
+import numpy as np
+
+from .utils import BasicBuffer
+
+class Normalizer:
+  def __init__(self, obs_dim, act_dim, act_low, act_high, device):
+    self.device = device
+    self.obs_dim = obs_dim
+    self.act_dim = act_dim
+    self.act_high = np.array(act_high)
+    self.act_low = np.array(act_low)
+
+    # Calculate action normalization
+    m = (self.act_low - self.act_high) / (-1 - 1)
+    c = self.act_low - m * (-1)
+    assert(not np.any(m == 0))
+    self.action_normalizer_params = (m, c)
+    self.action_normalizer_params_gpu = (torch.FloatTensor(m).to(self.device),
+                                         torch.FloatTensor(c).to(self.device))
+
+    # We just set the rest to identity projection
+    self.state_normalizer_params = (np.ones(obs_dim),
+                                    np.zeros(obs_dim))
+    self.state_normalizer_params_gpu = (torch.FloatTensor(np.ones(obs_dim)).to(self.device), 
+                                        torch.FloatTensor(np.zeros(obs_dim)).to(self.device))
+
+    self.reward_normalizer_params = (np.ones(1),
+                                     np.zeros(1))
+    self.reward_normalizer_params_gpu = (torch.FloatTensor(np.ones(1)).to(self.device), 
+                                         torch.FloatTensor(np.zeros(1)).to(self.device))
+
+  def to_dict(self):
+    return {
+      "state_m": self.state_normalizer_params[0].tolist(),
+      "state_c": self.state_normalizer_params[1].tolist(),
+      "action_m": self.action_normalizer_params[0].tolist(),
+      "action_c": self.action_normalizer_params[1].tolist(),
+      "reward_m": self.reward_normalizer_params[0].tolist(),
+      "reward_c": self.reward_normalizer_params[1].tolist(),
+    }
+
+  def from_dict(self, data):
+    self.state_normalizer_params = (np.array(data["state_m"]), np.array(data["state_c"]))
+    self.action_normalizer_params = (np.array(data["action_m"]), np.array(data["action_c"]))
+    self.reward_normalizer_params = (np.array(data["reward_m"]), np.array(data["reward_c"]))
+
+    self.state_normalizer_params_gpu =  (torch.FloatTensor(data["state_m"]).to(self.device),
[…]

### 2019-12-05 python/env/qbladeadapter.py

@@ -66,7 +66,7 @@ class QBladeAdapter:
 
   def reset(self):
     print('Resetting simulation')
-    if(self.steps_since_reload>100000):
+    if(self.steps_since_reload>50000):
       self.load_default_project()
     self._initializeSimulation(ctypes.c_int(0))
     self.lastAction = np.zeros(5)

### 2019-12-05 python/tcp_env/main.py

@@ -1,0 +1,28 @@
+from .env.qbladeadapter import QBladeAdapter
+from .ddpg.ddpg import DDPG
+
+import torch
+
+def main():
+
+  environment = 'qblade'
+  policy = 'actor'
+
+  hparams = DDPG.get_default_hparams()
+
+  if(environment == 'qblade'):
+    env = QBladeAdapter()
+    # Set environment-specific hparams
+    hparams.obs_dim = env.get_obs_dim()
+    hparams.act_dim = env.get_act_dim()
+    hparams.act_high = env.get_act_high()
+    hparams.act_low = env.get_act_low()
+    hparams.act_max_grad = env.get_act_max_grad()
+
+  if(policy == 'actor'):
+    pol = Actor(hparams.obs_dim, hparams.act_dim, hparams.actor_sizes[0], hparams.actor_sizes[1], hparams.init_weight_limit, hparams.actor_simple)
+
+  for t in range(0, hparams.steps_per_epoch * hparams.epochs):
+
+
+main()

### 2019-12-06 paper/paper.tex
Added first evaluation parts

@@
 \usepackage[ruled]{algorithm2e}
+\usepackage{multicol}
 \input{abbreviations}
@@
   \centering
-  \includegraphics{NREL-5MW.png}
+  \includegraphics{images/NREL-5MW.png}
   \caption{Control curve of the NREL 5MW turbine, from \cite{jonkmanDefinition5MWReference2009}}
@@
 The observations returned in getControlVars match those that are visible to standart industry controllers, which in turn are modelled after what can be measured on a real windturbine.
-Concretely, we measure:
-\begin{verbatim}
-* rotational speed [rad/s]
-* power [kW]
-* wind velocity [m/s]
-* yaw angle [deg]
-* pitch blade 1 [deg]
-* pitch blade 2 [deg]
-* pitch blade 3 [deg]
-* tower top bending local x [Nm]
-* tower top bending local y [Nm]
-* tower top bending local z [Nm]
-* out-of-plane bending blade 1 [Nm]
-* out-of-plane bending blade 2 [Nm]
-* out-of-plane bending blade 3 [Nm]
-* in-plane bending blade 1 [Nm]
-* in-plane bending blade 2 [Nm]
-* in-plane bending blade 3 [Nm]
-* out-of-plane tip deflection blade 1 [m]
-* out-of-plane tip deflection blade 2 [m]
-* out-of-plane tip deflection blade 3 [m]
-* in-plane tip deflection blade 1 [m]
-* in-plane tip deflection blade 2 [m]
-* in-plane tip deflection blade 3 [m]
-* current time [s]
-\end{verbatim}
-
-Because it has no semantic meaning, we mask away current time from our reinforcement learning agent. The inputs we can give to the turbine are:
-\begin{verbatim}
-* generator torque [Nm]
-* yaw angle [deg]
-* pitch blade 1 [deg]
-* pitch blade 2 [deg]
-* pitch blade 3 [deg]
-\end{verbatim}
-
-After a while of experimenting with the unrestricted simulation, we added some bounds to our control parameters. Restrictions to our final version are described in section \ref{sec:alg_qblade}.
+We have 23 observations and 5 actions.
+
+After a while of experimenting with the unrestricted simulation, we added some bounds to our control parameters and hid some observations and actions.
+Restrictions to our final version are described in section \ref{sec:alg_qblade}.
 
@@
   \centering
-  \includegraphics[width=0.6\linewidth]{DDPG_Setup.png}
+  \includegraphics[width=0.6\linewidth]{images/DDPG_Setup.png}
   \caption{DDPG Network}
@@
   \centering
-  \includegraphics[width=0.5\linewidth]{Miniturbine.png}
+  \includegraphics[width=0.5\linewidth]{images/Miniturbine.png}
   \caption{The mini wind-turbine from \cite{kolterDesignAnalysisLearning2012}}
@@
 At first, we used observed states and rewards from the random exploration phase to calculate the 5\%, 95\% quartile values.
-We calculated normalization constants from it to normalize those quartiles to $[-1, 1]$.
+We calculated normalization constants from it to linearly normalize those quartiles to $[-1, 1]$.
+Plain min/max normalization worked less well.
 As sometimes during random exploration, no extreme values were observed and later extreme values lay far outside of $[-1, 1]$, we instead normalized on replay data of a previous run.
 We could also imagine manually setting normalization bounds on low-dimensional problems, as human knowledge about the possible state space is usually present.
-In case when human knowledge about input magnitudes is not present, we recommend the mechanism from \cite{vanhasseltLearningValuesMany2016}.
+In case when human knowledge about input magnitudes is not present, we recommend the mechanism from \cite{vanhasseltLearningValuesMany2016}, or if you are willing to accept the extra computational effort of running once without normalization, our version.
 
@@
 We hoped that this way, the agent could derive its own gradients internally.
-The result of this were high and very noisy q-losses and no sensible results still.
-Though gradients looked sensible, we suspect that the high Q-losses resulted in too high steps along the gradient and thus diverging behavior.
+Also, a PID controller has a view of the past, with this extra information, the countroller would be able to generalize further.
+The result of this however were high and very noisy q-losses and q loss explosions, so we disabled this again.
 
@@
 
+\cite{fujimotoAddressingFunctionApproximation2018} addressed something similar on actions generated by the target policy, see section \ref{sec:double_critics} for this.
+
 \section{Exploding Q-Loss}
@@
 In their journal paper \cite{mnihHumanlevelControlDeep2015}, they write \"We also found it helpful to clip the error term from the update [...] to be between -1 and 1\", which could be interpreted as loss clipping.
-However they actually switched to absolute error instead of square error after those boundaries, which is the definition for huber loss.
-
-Although loss explosions gotten less frequent and smaller in magnitude with this update, we still saw some.
+However they actually switched to absolute error instead of square error outside those boundaries, which is the definition for huber loss.
+
+Although loss explosions gotten less frequent and smaller in magnitude with this update, we still saw some. In general we could observe an improvement in training, so we kept this active.
 
 \subsection{Double critics}
+\label{sec:double_critics}
 
 We also had a look at \cite{fujimotoAddressingFunctionApproximation2018}.
-They propose a set of three mechanisms that address function approximation error in \ac{DDPG}.
+They propose a set of three mechanisms that address function approximation error in \ac{DDPG}, and they call it \ac{TD3}.
 At first, they show in section 4.1 that Q learning with a deterministic policy tends to overestimate true Q values.
@@
   \centering
-  \includegraphics[width=\linewidth]{TD3-Setup.png}
+  \includegraphics[width=\linewidth]{images/TD3-Setup.png}
   \caption{Clipped double Q learning}
@@
 
-Using the techniques from \cite{fujimotoAddressingFunctionApproximation2018} still didn't solve
+However, we could not get \ac{TD3} to work properly, with or without \ac{PER}, and didn't see any convergence.
 
@@
 This condition happened only every 200k steps or so, and as we did not log every timestep, these spikes didn't end up in our logs.
-It was up to chance, that one of these spikes finally did, and we then decided to clip all normalized values to $[-3, 3]$.
+We decided to filter out these spikes by clipping rewards and observations after normalization to $[-3, 3]$.
 
@@
 
-It must be stated that we did not achieve a single run which stayed within sensible operational range for longer than 30K steps.
-Even our handcrafted PID controller for which we took the first set of parameters that somehow works widely outperforms our learnt policies, an industry controller is even better.
-This result in itself is notable, because it needed a whole set of additions to plain \ac{DDPG} to get first sensible results on our windturbine environment.
-So either, our windturbine simulation is more complex to learn than the OpenAI gym examples, which are usually used as benchmarks for \ac{RL} performance, or we are having errors in our programme.
-As the final product we developped has improvements from a wide variety of papers but still does not produce stable results, we conclude that tackling wind turbine control with \ac{RL} will need more research.
-Though we were not able to outperform industry controllers, the speed in which the area of \ac{RL} is advancing makes us hopeful to achieve good results in the near future.
-With our work on integrating qblade, we provided an open-source, accessible framework to develop future reinforcement learning controllers for wind turbines.
-
-To demonstrate the instability, we ran 5 training runs with the same hyperparameters and measured return, epoch return, killcount and time to q explosion (where we define q explosion as q loss > 1e4)
-
-\section{First working version}
-
-Our model consists of the DDPG algorithm with following additions:
-We integrated prioritized experience replay to
-
-How we achieved first successes, hparams, etc
+For our evaluation, we ran every experiment 4 times for 150 epochs, each lasting 2000 steps.
+We used a machine with an i7 4-core processor, 16GB RAM and a 1080 GTX with 8GB as accelerator.
+With these specs, we were able to run 1 experiment concurrently due to RAM limitations
+
+\section{Learning rates}
+
+At first we regard the algorithm under different learning rates.
+We vary the critic learning rate between 1e-3 (left), 1e-4 (middle, recommended by DDPG authors) and 1e-5 (right) in figure \ref{fig:critic_lr}.
+The task to solve was to hold a rotational speed of 0.8 rad/s over the time of 2000 steps, starting with a simulation that has artificially been stabilized at the desired rotational speed with 0 degrees pitch and 0 generator torque.
+The agent gets a reward of 1 per step if it exactly matches rotational speed, deviation from it is linearly punished, 0 rad/s would yield a reward of 0.
+Thus, the theoretically possible maximum reward would be 2000, holding the turbine at perfect speed for the entire length of the simulation.
+However, as we start with 0 pitch and torque, action gradient limitations will naturally incur a small penalty at the beginning as these values need to be adjusted to sensible values and the resulting rotational speed deviation needs to be recovered.
+A perfect policy would thus reach a score of a litte bit below 2000.
+
+Each of the controllers reaches this theoretical optimum and is able to hold it for a number of timesteps.
+With learning rate 1e-3, we reach it somewhere around episode 10, with learning rate 1e-4 a bit later and with 1e-5 even later.
+The time we are holding it before divergence also raises with learning rate.
+
+Interesting to see is the strategy which the controller found.
+As described before, both increasing pitch and increasing torque reduces rotor speed, so through both it would be possible to control the rotational speed.
+However, all of our runs decided to implement pitch-only control.
+A possible explanation for this decision could be that with high pitches, lower forces act on the blades, thus rotational speed is easier to control.
+However, this insight is above our human knowledge of windturbines.
+
+With the aggressive learning rate, we can clearly see how quickly the algorithm is diverging.
+After a short period of convergence around epoch 10, the high learning rate steps the critic out of the minimum and causes instabilities.
+With even higher learning rates, we observed not reaching convergence at all.
+Interestingly, we can not see such divergence in the loss.
+In normal machine learning, a too high learning rate would cause a phenomenon of jumping across a minimum, taking gradient steps greater than the width of the local minimum.
+This would reduce in an increase in loss.
+Here, however, the q loss stays small on such divergence, as plotted in figure \ref{fig:critic_qloss} for the diverging run of critic lr=1e-3.
+This is because we are generating our prediction targets based on delayed versions of the same net, as we do not know a ground truth.
+
+In general, it is difficult to measure convergence or divergence in reinforcement learning, \cite{cobbeQuantifyingGeneralizationReinforcement2019} have made some efforts.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.5\linewidth]{images/plot_critic_1e-3.png}
+  \caption{Critic learning rate 1e-3 reward over epoch}
+  \label{fig:critic_lr}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.5\linewidth]{images/plot_critic_1e-3_qloss.png}
+  \caption{Critic learning rate 1e-3 q loss over step}
+  \label{fig:critic_qloss}
+\end{figure}
 

### 2019-12-06 python/automated_hparam_search.py

@@ -44,15 +44,22 @@ import os
 ]'''
 
 
-experiments = [
-  {"id": 48},
-  {"id": 49},
-  {"id": 50},
-  {"id": 51},
-  {"id": 52},
+'''experiments = [
+  #{"id": 48}, # Twin critic, target policy smoothing on, results shitty
+  #{"id": 49},
+  #{"id": 50},
+  #{"id": 51},
+  {"id": 52}, # Single critic, critic-lr 1e-3, PER on, converged at first then diverged
   {"id": 53},
   {"id": 54},
   {"id": 55},
+]'''
+
+experiments = [
+  {"id": 56}, # Single critic, PER on, critic lr = 1e-4
+  {"id": 57},
+  {"id": 58},
+  {"id": 59},
 ]
 
 
@@ -85,7 +92,7 @@ def run_one(experiment):
     return (id, -1, -1)
 
 # Set the number of parallel runs here
-pool = Pool(2)
+pool = Pool(4)
 results = pool.map(run_one, experiments)
 
 for (id, reward, deaths) in results:

### 2019-12-06 python/ddpg/ddpg.py

@@ -18,7 +18,8 @@ class DDPG:
   
   def __init__(self, hparams):
     self.device = torch.device("cuda" if torch.cuda.is_available() and hparams.use_gpu else "cpu")
-    
+    self.dtype = torch.float
+
     self.hparams = hparams
     self.obs_dim = hparams.obs_dim
     self.act_dim = hparams.act_dim
@@ -42,6 +43,7 @@ class DDPG:
       self.act_high = np.ones(self.act_dim)
       self.act_low = -np.ones(self.act_dim)
       self.real_last_action = self.hparams.starting_action
+      assert(len(self.real_last_action) == self.act_dim)
       self.action_grad_denormalizer = (self.real_act_high - self.real_act_low) * self.hparams.action_gradient_stepsize
 
       assert(np.all(self.real_last_action <= self.real_act_high))
@@ -66,7 +68,7 @@ class DDPG:
       self.optimizer = optim.Adam
     else:
       self.optimizer = optim.SGD
-    self.critic_optimizer = self.optimizer(self.critic.parameters(), lr=hparams.critic_lr)
+    self.critic_optimizer = self.optimizer(self.critic.parameters(), lr=hparams.critic_lr, weight_decay=hparams.critic_weight_decay)
     self.actor_optimizer  = self.optimizer(self.actor.parameters(), lr=hparams.actor_lr)
 
     # If in twin critic mode, add another critic
@@ -95,6 +97,7 @@ class DDPG:
     # Noises
     # Random exploration
     if(self.hparams.random_exploration_type == 'correlated'):
+      assert(len(self.hparams.random_exploration_mu) == self.act_dim)
       self.random_exploration_noise = OUNoise(
         self.act_dim,
         self.hparams.random_exploration_theta,
@@ -145,7 +148,7 @@ class DDPG:
                                  self.hparams.act_high,
                                  self.device)
 
-    self.logger = QBladeLogger(self.hparams.logdir, self.hparams.log_steps, self.hparams.run_name)
+    self.logger = QBladeLogger(self.hparams.logdir, self.hparams.log_steps, self.hparams.run_name, self.hparams.obs_labels, self.hparams.act_labels)
     for key, value in self.hparams.values().items():
       self.logger.writer.add_text(key, str(value), 0)
 
@@ -177,7 +180,7 @@ class DDPG:
     # Send the observation to the device, normalize it
     # Then calculate the action and denormalize it
     assert(not np.any(np.isnan(obs)))
[…]

### 2019-12-06 python/ddpg/hparams.py

@@ -4,20 +4,20 @@ import re
 # This class is partially compatible to tensorflows training.contrib.HParams class
 class HParams:
   def __init__(self, **first_data):
-    self.data = first_data
+    self.__dict__ = first_data
 
   def __str__(self):
-    return str(self.data)
+    return str(self.__dict__)
 
   def get_dict(self):
-    return self.data
+    return self.__dict__
 
   def override_from_dict(self, new_data):
     for k,v in new_data:
-      self.data[k] = v
+      self.__dict__[k] = v
 
   def items(self):
-    return self.data.items()
+    return self.__dict__.items()
 
   def values(self):
     data = {}
@@ -29,7 +29,7 @@ class HParams:
       except:
         return False
 
-    for k,v in self.data.items():
+    for k,v in self.__dict__.items():
       if(not is_jsonable):
         data[k] = str(v)
       else:
@@ -46,7 +46,7 @@ class HParams:
 
     data = json.loads(args)
     for key, value in data:
-      self.data[key] = value
+      self.__dict__[key] = value
 
   # parses in the form of a=1,b=2,
   def parse(self, args):
@@ -61,7 +61,10 @@ class HParams:
         value = eval(value) # unsafe as fuck
       except:
         pass
-      self.data[key] = value
+      self.__dict__[key] = value
 
-  def __getattr__(self, name):
-    return self.data[name]
+  #def __getattr__(self, name):
+  #  return self.__dict__[name]
+#
+#  def __setattr__(self, name, val):
+#    self.__dict__[name] = val

### 2019-12-06 python/ddpg/normalizer.py

@@ -6,6 +6,7 @@ from .utils import BasicBuffer
 class Normalizer:
   def __init__(self, obs_dim, act_dim, act_low, act_high, device):
     self.device = device
+    self.dtype = torch.float
     self.obs_dim = obs_dim
     self.act_dim = act_dim
     self.act_high = np.array(act_high)
@@ -16,19 +17,19 @@ class Normalizer:
     c = self.act_low - m * (-1)
     assert(not np.any(m == 0))
     self.action_normalizer_params = (m, c)
-    self.action_normalizer_params_gpu = (torch.FloatTensor(m).to(self.device),
-                                         torch.FloatTensor(c).to(self.device))
+    self.action_normalizer_params_gpu = (torch.tensor(m, dtype=self.dtype).to(self.device),
+                                         torch.tensor(c, dtype=self.dtype).to(self.device))
 
     # We just set the rest to identity projection
     self.state_normalizer_params = (np.ones(obs_dim),
                                     np.zeros(obs_dim))
-    self.state_normalizer_params_gpu = (torch.FloatTensor(np.ones(obs_dim)).to(self.device), 
-                                        torch.FloatTensor(np.zeros(obs_dim)).to(self.device))
+    self.state_normalizer_params_gpu = (torch.tensor(np.ones(obs_dim), dtype=self.dtype).to(self.device), 
+                                        torch.tensor(np.zeros(obs_dim), dtype=self.dtype).to(self.device))
 
     self.reward_normalizer_params = (np.ones(1),
                                      np.zeros(1))
-    self.reward_normalizer_params_gpu = (torch.FloatTensor(np.ones(1)).to(self.device), 
-                                         torch.FloatTensor(np.zeros(1)).to(self.device))
+    self.reward_normalizer_params_gpu = (torch.tensor(np.ones(1), dtype=self.dtype).to(self.device), 
+                                         torch.tensor(np.zeros(1), dtype=self.dtype).to(self.device))
 
   def to_dict(self):
     return {
@@ -45,12 +46,12 @@ class Normalizer:
     self.action_normalizer_params = (np.array(data["action_m"]), np.array(data["action_c"]))
     self.reward_normalizer_params = (np.array(data["reward_m"]), np.array(data["reward_c"]))
 
-    self.state_normalizer_params_gpu =  (torch.FloatTensor(data["state_m"]).to(self.device),
-                                         torch.FloatTensor(data["state_c"]).to(self.device))
[…]

### 2019-12-06 python/ddpg/utils.py

@@ -100,11 +100,11 @@ class UniformNoiseDec:
 # Helper class for not having noise at all
 class NoNoise:
     def __init__(self, dim, default_action=None):
-        self.default_action = default_action if default_action is not None else np.zeros(self.dim)
         self.dim = dim
+        self.default_action = default_action if default_action is not None else np.zeros(self.dim)
 
     def get_noise(self, _t=0):
-        return np.zeros(dim)
+        return self.default_action
 
 class CapacityBuffer:
     def __init__(self, max_size):
@@ -113,10 +113,11 @@ class CapacityBuffer:
 
     def append(self, data):
         self.push(data)
+
     def push(self, data):
         self.buffer.append(data)
         while(len(self.buffer) > self.max_size):
-            del self.buffer[-1]
+            self.buffer.pop(0)
 
     def get_buffer(self):
         return self.buffer
@@ -204,36 +205,8 @@ class SumTree:
         # Add the difference to all the nodes in the trajectory
         self.weights[trajectory] += weight_diff
 
-    # Samples an index from the tree
-    # Takes into account weights
-    def sample(self):
-        # Assert not empty
-        assert(self.weights[0] != 0)
+        return weight
 
-        return self.sample_rec(2) - 1 - self.tree_size
-
-    # Make one branch decision
-    # Index is a 1-terminated index in the tree, pointing to the left node of the two between which to decide
-    # Regularization is the weight of the parent node
-    # It returns the 1-terminated index in the weights
-    def sample_rec(self, index):
-        # Make a decision between the two weights
-        prob_left = self.weights[index-1]
-        prob_right = self.weights[index]
-        regularization = prob_left + prob_right
-        assert(regularization != 0)
-
-        go_left = np.random.uniform() < (prob_left/regularization)
-
-        if(not go_left):
-            index += 1
-
-        # Check whether we reached the leaf already
-        if(index > self.tree_size):
-            return index
-
-        # If not, continue recursively
-        return self.sample_rec(index * 2)
 
 # A "Basic" Buffer which stores entries and can sample them by priority
 class BasicBuffer:
@@ -250,22 +223,22 @@ class BasicBuffer:
         self.dtype = torch.float
 
         # GPU buffers
[…]

### 2019-12-06 python/env/gymadapter.py

@@ -1,8 +1,8 @@
 import gym
 
 class GymAdapter:
-  def __init__(self, hparams):
-    self.env = gym.make(hparams.env_name)
+  def __init__(self, env_name):
+    self.env = gym.make(env_name)
 
   def reset(self):
     self.observation = self.env.reset()
@@ -22,6 +22,15 @@ class GymAdapter:
 
   def get_act_low(self):
     return self.env.action_space.low
+
+  def get_act_max_grad(self):
+    return self.env.action_space.high - self.env.action_space.low
+
+  def get_obs_labels(self):
+    return {}
+
+  def get_act_labels(self):
+    return {}
 
   def step(self, action):
     self.observation, self.reward, self.death, _ = self.env.step(action)

### 2019-12-06 python/env/qbladeadapter.py

@@ -22,7 +22,8 @@ def unloadLib(handle, path):
 
 
 class QBladeAdapter:
-  def __init__(self):
+  def __init__(self, project="../sample_projects/NREL_5MW_STR.wpa"):
+    self.project = project
     self._qbladeLib = None
 
     print("loading qblade library")
@@ -59,7 +60,7 @@ class QBladeAdapter:
     del self._advanceSingleTimestep
 
   def load_default_project(self):
-    x = ctypes.c_char_p(b"../sample_projects/NREL_5MW_STR.wpa")
+    x = ctypes.c_char_p(bytes(self.project, 'utf-8'))
     print("loading qblade project")
     self._loadProject(x)
     self.steps_since_reload = 0
@@ -75,8 +76,8 @@ class QBladeAdapter:
     # Do some steps with zero action
     action = np.zeros(self.get_act_dim())
     self.storeAction(action)
-    for i in range(0, 100):
-      self._advanceSingleTimestep()
+    #for i in range(0, 100):
+    #  self._advanceSingleTimestep()
 
     print('reset done')
 
@@ -103,6 +104,42 @@ class QBladeAdapter:
     timestep = 0.1
     return [150000*timestep, 8*timestep]
 
+  def get_obs_labels(self):
+    return {
+      0: 'rotational speed [rad/s]',
+      1: 'power [kW]',
+      2: 'HH wind velocity [m/s]',
+      3: 'yaw angle [deg]',
+      4: 'pitch blade 1 [deg]',
+      5: 'pitch blade 2 [deg]',
+      6: 'pitch blade 3 [deg]',
+      7: 'tower top bending local x [Nm]',
+      8: 'tower top bending local y [Nm]',
+      9: 'tower top bending local z [Nm]',
+      10: 'oop bending blade 1 [Nm]',
+      11: 'oop bending blade 2 [Nm]',
+      12: 'oop bending blade 3 [Nm]',
+      13: 'ip bending blade 1 [Nm]',
+      14: 'ip bending blade 2 [Nm]',
+      15: 'ip bending blade 3 [Nm]',
+      16: 'oop tip deflection blade 1 [m]',
+      17: 'oop tip deflection blade 2 [m]',
+      18: 'oop tip deflection blade 3 [m]',
+      19: 'ip tip deflection blade 1 [m]',
+      20: 'ip tip deflection blade 2 [m]',
+      21: 'ip tip deflection blade 3 [m]',
+      22: 'current time [s]'
+    }
+
[…]

### 2019-12-06 python/main.py

@@ -13,6 +13,7 @@ import os
 import json
 import time
 
+
 def main():
 
   import argparse
@@ -23,7 +24,8 @@ def main():
   parser.add_argument('--load_checkpoint_hparams', action="store_true", help="also load hparams from checkpoint file")
   args = parser.parse_args()
 
-  env = QBladeAdapter()
+  env = QBladeAdapter("../sample_projects/NREL_5MW_STR.wpa")
+  #env = GymAdapter('Pendulum-v0')
 
   # Get default hparams
   hparams = DDPG.get_default_hparams()
@@ -47,6 +49,8 @@ def main():
   hparams.act_high = env.get_act_high()
   hparams.act_low = env.get_act_low()
   hparams.act_max_grad = env.get_act_max_grad()
+  hparams.act_labels = env.get_act_labels()
+  hparams.obs_labels = env.get_obs_labels()
 
   checkpoint_steps = hparams.checkpoint_steps
   checkpoint_dir = hparams.checkpoint_dir
@@ -80,13 +84,14 @@ def main():
 
     start = time.time()
     o, r, d = env.step(a)
+    reset = d
     step = time.time()
     a, reset = agent.step(o, r, d)
     end = time.time()
     execution_time_env += step - start
     execution_time_agent += end - step
 
-    if(t%100 == 0):
+    if(t%300 == 0):
       print("Step %d, Execution times env: %f, agent: %f" % (t, execution_time_env, execution_time_agent))
       execution_time_agent = 0
       execution_time_env = 0
@@ -108,6 +113,10 @@ def main():
       o, r, d = env.step(a)
 
       total_reward += r
+
+      agent.logger.logAction(t, a, 'act_test')
+      agent.logger.logObservation(t, o, 'obs_test')
+      agent.logger.add_scalar('Reward/test', r, t)
 
       if(d):
         total_deaths += 1

### 2019-12-06 python/tcp_env/actor_policy.py

@@ -1,0 +1,7 @@
+from .ddpg.models import Actor
+
+class ActorPolicy:
+  def __init__(self, hparams):
+    self.actor = Actor(hparams.obs_dim, hparams.act_dim, hparams.actor_sizes[0], hparams.actor_sizes[1], hparams.init_weight_limit, hparams.actor_simple)
+
+  def prepare(self, obs):

### 2019-12-07 paper/paper.tex
Almost done with the paper, missing more eval and intro

@@
-\documentclass[hyperref,german,beleg]{cgvpub}
+\documentclass[hyperref,beleg]{cgvpub}
 
@@
 \usepackage{multicol}
+\usepackage{subcaption}
 \input{abbreviations}
@@
 
-\copyrighterklaerung{Hier soll jeder Autor die von ihm eingeholten
-Zustimmungen der Copyright-Besitzer angeben bzw. die in Web Press
-Rooms angegebenen generellen Konditionen seiner Text- und
-Bild"ubernahmen zitieren.}
 \acknowledgments{At first thanks to my advisors Mattew Lennie and Dmitrij Shlezinger for their support.
@@
 Thanks to my laptop for almost dying, but then surviving.
-And big thanks to the mighty god of windturbines for letting me crush $\num{12345}$ windturbines for science - yes I counted.}
+And big thanks to the mighty god of windturbines for letting me crush several thousand windturbines for science.}
 
@@
 \chapter{Abstract}
-Small
-
-\chapter{Intro}
 
@@
 The range of topics, RL is applicable to, increases.
-However, so far, there hasn't been an effort to control a windturbine with a reinforcement learning algorithm.
+However, so far, these advancements were only partially applied to the area of windturbine pitch and torque control.
 Current state of the art linear controllers are performing well in maintaining turbine control, and the optimization margin for these is small, as they achieve the theoretical maximum energy output up to rated windspeed and manage to keep the turbine intact above rated windspeeds.
-
-Trying it nontheless can still provide for some foundational knowledge for future developments.
-Park control becomes a bigger topic in wind-energy, where controlling all windturbines of a park together can increase overall energy yield or protect windturbines further down the windstream from wakes by turbines up front.
+However, these simple controllers can only deal with a limited amount of data inputs and are preventing turbine damages.
+We are evaluating the use of an actor-critic reinforcement learning algorithm to solve continous control of a windturbine, hoping to lay a foundational work for later improvements.
+We find that utilizing a wildly proven continous control algorithm does not solve our windturbine problem.
+Finally, we demonstrate that this control is possible, however we do not achieve results outperforming industry controllers.
+
+\chapter{Introduction}
+
+With the issue of global warming, we are facing maybe the biggest challenge to our generation.
+Though wheather models have predicted the effect since the 1980s, only in the recent years this topic has been brought into societal focus.
+CO2 is seen as the main cause for a tendency in the global weather to heat up.
+Since our industrial age, humankind has been releasing CO2 into the athmosphere in large scales.
+If however we continue doing so the way we are doing now, we will likely increase global temperature by a degree which will make living on this planet difficult.
+With the Paris agreement, the United Nations decided to limit global warming to 1.5 degrees, for which massive reductions in CO2 emissions are necessary.
+As energy production holds one of the biggest parts in the global CO2 emissions, we need to replace CO2 emitting power plants.
+For this, renewable energies have played a major role, and as a big part of this also windpower.
+
+Winds are a direct result of sunshine.
+The sun heats different parts of the earth to different temperatures, as the surface of the earth isn't uniform.
+The air over hotter regions such as close to the equator tends to rise, while on mountains and the poles it tends to fall.
+Combine this with a vast and complex interplay of humidities, Coriolis effects, mountains, cloud formation and enough effects to fill an entire branch of science and you will get wind.
+Windturbines operate by drawing energy from this weather phenomenon.
+Large rotors placed in windy regions which are connected to a generator yield electrical power.
+As weather is a complex phenomenon, wind tends to come from different directions with different speeds.
+This makes drawing energy from it more complex, or less efficient.
+If we were to draw energy from both storms and light breezes with the same turbine, we would not be efficient.
+Thus, modern windturbines are built to be adjustable.
+With these adjustments, a turbine can operate sensibly over a broader range of wind conditions.
+However, somehow, these adjustments need to be automatically adapted to the wind condition.
+This is the task of the turbine controller.
+We are attempting to learn such a controller.
+
+Machine learning has developped immensly over the last years.
+Benefitting from Moore's law, huge amounts of computational power can be used to train more and more complex models.
+Machine learning is generally split into three big branches.
+Supervised learning, unsupervised learning and reinforcement learning.
+In supervised learning, some form of fitting is done on a dataset.
 
@@
 
-This section aims at facilitating the background knowledge necessary for understanding this paper. We will first introduce windturbines and common terms around this area of research.
+This section aims at facilitating the background knowledge necessary for understanding this work.
+We will first introduce windturbines and common terms around this area of research.
+Then we will motivate our work precisely.
+We will give insights on the simulation tool we used and then derive the reinforcement learning algorithm we decided to use.
+Lastly we will present another paper which tried to combine reinforcement learning and windturbines.
 
@@
 
-\subsection{Why we do what we do}
+\subsection{Motivation}
+
 As you might have seen before, all controllers follow a certain style.
@@
 You might have seen flaps on an aircraft wing, extending the length of the wing or adding gaps at takeoff and landing to account for the low wind speeds hitting the aircraft wing in these flight scenarios.
-This could also benefit wind turbines at low wind speeds.
+This could also benefit wind turbines at low wind speeds, however building a good controller for them is challenging.
 Tackling all this with \ac{SISO} is possible, but a lot of manual work.
@@
 However, \ac{DPG} is not definitely proven to converge for a non-linear function approximator, as this transfer was, to our knowledge, never fully formulated.
-Observations show that it still does converge.
+However, in practice, it is still trained off-policy.
+In our own experiments, we found that using a wildly different policy for exploration than for training did in fact result in worse behavior, and \cite{fujimotoOffPolicyDeepReinforcement2019} observed the same behavior.
+Although there is some evidence against this being possible, we will from now on just assume that DDPG is trainable off-policy because it delivered good results in the past and because the paper stating it is possible has been cited more than 100 times more often than the one saying it's not.
+Science!
 
@@
 
+Further works on reinforcement learning on windturbine are also \cite{saenz-aguirreArtificialNeuralNetwork2019}.
+This work uses traditional supervised learning to fit a neural network of 10 neurons to the straight line that describes the relation between yaw angle and power output and because what they got looks similar to a Q function they call it reinforcement learning.
+\cite{fernandez-gaunaVariableSpeedWind2016} published in a proprietary journal.
+
+%we didn't get \cite{abouheafModelfreeAdaptiveLearning2018a}
+
 \chapter{Experimentation}
@@
 
+\item \textbf{Penalize death conditions} Penalize when a death condition is reached.
+With $r(s)$ being another reward function, $d(s) \in {0,1}$ being our indicator for deaths and $c$ a constant penalty, calculate $r'(s) = r(s) - c * d(s)$
+
 \end{itemize}
@@
 We used this reward for our first working version.
+
 \textbf{Rated power} is a bit more difficult.
@@
 However especially if setting an artificial rated power lower than what the turbine was built for, there is a certain play in how strongly to use blade pitch and how much needs to be done through torque.
+
 \textbf{Maximal power} is an intuitive reward, but might yield unrealistic results.
@@
 So this reward makes more sense when combined with a stress penalty.
+
 \textbf{Penalize current stress} is the easiest stress penalty, where a penalty proportional to the current bending of the structure can be used.
@@
 This is a simple variant which leaves out bending in the middle of the blades, bending in the rotational plane, vibration or bending of the tower and torsional stress on the shaft.
+
 \textbf{Penalize accumulated stress} is a more complete, but time dependent variant.
@@
 We suspect this penalty to be hard to learn, as it reacts slowly to changes and a once given penalty will never be lifted again.
+
 \textbf{Penalize action gradients} is an option which is useful in contexts where high action gradients are impossible on the real system.
@@
 
-Most of our experimentation we did with rewarding rated speed, as we expected this reward function to be the easiest to learn.
+\textbf{Penalize deaths} is simple to implement but yields a not differentiable loss function at deaths.
+Especially because we are learning a continuous value estimation, it will most likely smooth out along a death and likely underestimate the death penalty and underestimate values close to the death.
+Alternatively we could imagine penalizing getting close to a death condition continously.
+
+Most of our experimentation we did with rewarding rated speed and penalize deaths, as we expected this reward function to be the easiest to learn.
+In a real application, crafting a good reward function will be a bigger challenge, but this is out of the scope of this work.
 
@@
 
+We did however not really observe better results with expert exploration enabled.
+We tried combining it with the algorithm from section \ref{sec:pretrain}, because we were suspecting that the large difference between the actor distribution and the expert distribution lead to problems in learning the respective Q distribution.
+We argued before that we suspect DDPG not to be able to learn completely off-policy but just a bit off-policy.
+Together with the pretraining algorithm, we had some slight improvements, but these didn't justify the extra computation time needed for this.
+Also, though an expert controller which yields somewhat sensible result is easy to build for each turbine, this is a step of human interaction and expert knowledge, which machine learning generally tries to minimize in algorithm design.
+
 \subsection{Parameter noise}
@@
 \subsection{Feeding past timesteps}
+\label{sec:feedpast}
 
@@
 \subsection{Pretraining the policy}
+\label{sec:pretrain}
 
 We added a small period of direct pretraining of the policy on the actions taken by the expert policy during random exploration.
-We hoped for a more sensible default policy at the beginning and indeed observed reaching sensible behavior a little earlier.
+We hoped for a more sensible default policy at the beginning.
 We summarize this in algorithm \ref{alg:pretrain}, parametrized with training time $T=10000$ and batch size $N=32$
@@
 
+The training is relatively straightforward, however because we decided to abolish expert exploration, we also abolished this.
+
 \section{Other improvements}
@@
 
-As DDPG is capable of learning from a replay of experiences, we implemented a replay buffer.
-To enable \ac{PER}, we needed to sample according to priorities proportional to a metric, in our case \ac{TD-Error}.
+As DDPG is capable of learning from a replay of experiences, we implemented a replay buffer with capacity 1e6.
+To enable \ac{PER}, we needed to sample according to priorities proportional to a metric, in our case absolute \ac{TD-Error}.
 We implemented the replay buffer as a sum tree datastructure, a binary weighted tree in which each node holds the sum of the weights of the nodes below.
-The root node thus always held the sum of all metrics.
 Also, we did not store probabilities right away but kept the unnormalized experiences $p_i = (\abs{\sigma} + \epsilon)^{\alpha}$ with an epsilon of $\num{1e-6}$ and $\alpha = 0.5$.
@@
 Instead of 400 and 300 we use 64 and 32 neurons.
-All states, rewards and actions seen by the network are normalized to be between $[-1, 1]$.
-
-We start our simulation with an expert exploration.
-We designed a P-controller which can keep the turbine at a constant rotational speed and added \ac{OU-Noise} to it to explore sensible operation range.
-For 3 epochs, we let our expert policy with noise control the turbine before we hand over to our \ac{DDPG}-Agent.
-During the handover, the policy is pretrained for 10k steps, lr 1e-4, adam optimizer, mse loss to directly predict actions from the expert controller.
-This way we started off early with better actions.
-Then, critic and actor training is done on the experiences from expert exploration.
-After this training, all priorities in the replay buffer are set to their maximum (so they all will be seen) and normal training starts.
-Each epoch lasts for 2000 steps and at the beginning of it, the simulation is reset to default state.
-We trained for 150 epochs.
+We use batch normalization between the layers.
+All states and rewards seen by the network are normalized to be between $[-1, 1]$ 90\% of the time and are clipped to $[-3, 3]$.
+Actions are normalized to $[-1, 1]$.
+Learning rate for both actor and critic are 1e-4 and we use adam optimization.
+We train with batch size 128, discounting factor $\gamma$ = 0.01 and target update rate $\tau$ = 0.05.
+Weights are initialized by uniform random in $[-0.5, 0.5]$.
 
 Critic updates are done similar to \ac{DQN} updates with \ac{PER}.
-We use importance sampling to correct for \ac{PER} overestimation bias and regard death conditions.
+We use importance sampling to correct for \ac{PER} overestimation bias in the critic and redraw samples uniformly for policy training.
+We regard death conditions in the Q update as in \ac{DQN}.
+Also, we use huber loss instead of MSE.
 All in all, algorithm \ref{alg:ourddpg} summarizes all improvements:
@@
 Initialize replay buffer $\mathcal{R}$\\
-
-Execute algorithm \ref{alg:expertexplo} for expert exploration \\
-Calculate normalization factors based on exploraton buffer and data from a previous run \\
-Execute algorithm \ref{alg:pretrain} for policy pretraining \\
+Calculate normalization factors based on data from a previous run \\
 
@@
 
-For our evaluation, we ran every experiment 4 times for 150 epochs, each lasting 2000 steps.
-We used a machine with an i7 4-core processor, 16GB RAM and a 1080 GTX with 8GB as accelerator.
-With these specs, we were able to run 1 experiment concurrently due to RAM limitations
-
-\section{Learning rates}
-
-At first we regard the algorithm under different learning rates.
-We vary the critic learning rate between 1e-3 (left), 1e-4 (middle, recommended by DDPG authors) and 1e-5 (right) in figure \ref{fig:critic_lr}.
+For our evaluation, we used a machine with an Intel\textsuperscript{\textregistered} i7 4-core processor, 16GB RAM and a Nvidia 1080 GTX with 8GB as accelerator.
+On this system, a fully realistic run with 300k steps would take roughly 200 hours, of which almost everything would be used by qblade.
+Partly this is because we didn't manage to enable hardware acceleration because of compilation problems, but computing a fluid simulation is always time-consuming.
+We could bring the computation time down to 12 hours per run by deactivating wake calculations and using a different structural computation method in qblade.
+Wake calculations are responsible for calculating the effects of the rotor on the surrounding air.
+Without this effect, the rotor can draw infinite energy from the air as it can generate lift without slowing down the air.
+Effectively, this increased instability of the system, as now the rotor does not get slowed down at high rotational speeds, but otherwise we would not have been able to evaluate our work.
+
+We ran every experiment 4 times except stated otherwise.
+We visualize our measures as confidence plots, for which we calculated the mean and a 90\% confidence interval over the results.
+The mean is displayed as a solid line, while the confidence is plotted as a colored background.
+
+\section{Hold speed}
+
+We start off with the simplest task.
 The task to solve was to hold a rotational speed of 0.8 rad/s over the time of 2000 steps, starting with a simulation that has artificially been stabilized at the desired rotational speed with 0 degrees pitch and 0 generator torque.
+Wind-speed is kept constant at 11m/s.
 The agent gets a reward of 1 per step if it exactly matches rotational speed, deviation from it is linearly punished, 0 rad/s would yield a reward of 0.
@@
 A perfect policy would thus reach a score of a litte bit below 2000.
-
-Each of the controllers reaches this theoretical optimum and is able to hold it for a number of timesteps.
-With learning rate 1e-3, we reach it somewhere around episode 10, with learning rate 1e-4 a bit later and with 1e-5 even later.
+This task is simple because it can be achieved either through pitching out, increasing generator torque or a combination of both.
+This task is not realistic, as it can be solved without generating energy whatsoever.
+
+We look at two different learning rates for the critic network, 1e-3 (a) and 1e-4 (b). The latter is what was recommended in the DDPG paper.
+
+\begin{figure}
+\centering
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.8\linewidth]{images/plot_critic_1e-3.png}
+  \caption{critic lr=1e-3}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.8\linewidth]{images/plot_critic_1e-4.png}
+  \caption{critic lr=1e-4}
+\end{subfigure}
+\caption{Learning rate comparison for the critic network}
+\label{fig:critic_lr}
+\end{figure}
+
+Each of the controllers reaches close to theoretical optimum and is able to hold it for a number of epochs.
+With learning rate 1e-3, we reach it somewhere around episode 10, with learning rate 1e-4 a bit later around epoch 25.
 The time we are holding it before divergence also raises with learning rate.
-
-Interesting to see is the strategy which the controller found.
+The most aggressive learning rate reached 1828.8 average epoch reward between epoch 7 and 11 and a single maximum of 1917.5 points.
+On the lower learning rate, it held an average of 1896 points from episode 30 to 56 and reached a maximum of 1968.4 points.
+The maximum we could reach with our hand-designed PID controller was 1981 points, where the 19 points were lost through the swing at the beginning and then rotational speed was held perfectly.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.5\linewidth]{images/epoch40.png}
+  \caption{Pitch, torque and rotational speed from a sample epoch}
+  \label{fig:strategy}
+\end{figure}
+
+Interesting to see is the strategy which the agent found.
 As described before, both increasing pitch and increasing torque reduces rotor speed, so through both it would be possible to control the rotational speed.
-However, all of our runs decided to implement pitch-only control.
+However, all of our runs decided to implement pitch-only control and kept torque constant at zero, as possible to see in figure \ref{fig:strategy}.
 A possible explanation for this decision could be that with high pitches, lower forces act on the blades, thus rotational speed is easier to control.
 However, this insight is above our human knowledge of windturbines.
+Furthermore should be noted that rotational speed is still varying between 0.75 and 0.85 and the pitch is ranging from 15 to 25 degrees (note the scaling of the pitch and torque).
+In a realistic scenario, oscillations of this magnitude would likely destroy the turbine very soon, as pitching the blades around this quickly creates strong material stress.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.5\linewidth]{images/plot_critic_1e-3_qloss.png}
+  \caption{Critic q loss during divergence}
+  \label{fig:critic_qloss}
+\end{figure}
 
@@
 This is because we are generating our prediction targets based on delayed versions of the same net, as we do not know a ground truth.
-
-In general, it is difficult to measure convergence or divergence in reinforcement learning, \cite{cobbeQuantifyingGeneralizationReinforcement2019} have made some efforts.
+As this delayed target net will diverge with the original net if divergence happens slow enough, we do not see it.
+
+\section{Hold rated power}
+A more difficult task to solve, but also a more realistic one it to hold a rated power.
+Analog to our hold speed task, the agent gets a simulation which was artificially stabilized at 0.8rpm with pitch and torque 0, and should generate exactly 5MW of power during an epoch of 2000 steps.
+This power is the reference output of the NREL 5MW turbine, and in our wind scenario it should theoretically be able to reach this power.
+Wind speed was kept constant at 11m/s.
+Again, the agent gets a reward of 1 for hitting 5MW perfectly, and 0 for having no power at all, so a perfect policy would get something a little below 2000.
+Unfortunately, we did not get any policy to achieve something close to optimal, the best score we achieved was 1500 points with an average of 5 kills per episode.
+Unfortunately, this result wasn't very stable, so we can not plot an average over many runs.
+An example for a relatively successful epoch is figure \ref{fig:crashing_turbines}.
+Here, the controller maxed out torque and tried to hold the turbine stable with pitch.
+Power output for these insane regions was up to 9MW, which would have instantly roasted any electronics in a real wind turbine.
+Also, blades would have taken serious damage on these rotational speeds.
 
@@
   \centering
-  \includegraphics[width=0.5\linewidth]{images/plot_critic_1e-3.png}
-  \caption{Critic learning rate 1e-3 reward over epoch}
-  \label{fig:critic_lr}
+  \includegraphics[width=0.5\linewidth]{images/crashing_turbines.png}
+  \caption{An epoch from the hold power task}
+  \label{fig:crashing_turbines}
 \end{figure}
 
-\begin{figure}
-  \centering
-  \includegraphics[width=0.5\linewidth]{images/plot_critic_1e-3_qloss.png}
-  \caption{Critic learning rate 1e-3 q loss over step}
-  \label{fig:critic_qloss}
-\end{figure}
-
-\section{HParam tuning}
-We tried different hyper parameters.
-To evaluate our controller, we measure the time the simulation stayed within a sensible range.
-We define sensible as rotational speeds between 0 and 1.2 rad/s.
-We let each simulation run 1 million steps.
-As during our random exploration, the simulation was always in sensible range, we remove the random exploration phase from our evaluation.
-
-We tried
-\begin{itemize} 
-  \item increasing/decreasing both learning rates
-  \item increasing/decreasing tau and gamma
-  \item increasing/decreasing batch size
-  \item increasing/decreasing replay noise
-\end{itemize}
-
-\section{Comparison to industry controllers}
-
-% What should I put here?
-
-% Holds power, doesn't diverge, way to many actions, super noisy, we are at least somewhere close to sensible
-% Different reward function
-% Different wind speed
+At the end, most controllers converged to leaving pitch at zero and left the turbine to die regularily with maximum generator torque.
+
+\section{Hold power}
+The next task
+
+\section{Pendulum}
+To exclude the possibility that our implementation is buggy, we decided to run our DDPG algorithm on the Pendulum task.
+With our hyperparameters, we could not observe convergence and only switching back to the algorithm of \ac{DDPG}, the agent managed to balance out the pendulum after 11 epochs.
+Especially the network size hyperparameter was important for this.
+
+\section{Discussion}
+
+Commonly, algorithms are evaluated over many different problems.
+The OpenAI Gym suite is the common suite for evaluating reinforcement learning algorithms, and typical problems in this are balance problems (all cartpole* and pendulum* tasks) and actuation tasks (gripper*, reacher*, cheetah*).
+Balance tasks could be argued to resemble our windturbine in a way, as we want to balance an output aswell.
+However a windturbine is vastly more complex than a pendulum.
+It exhibits resonant frequencies in addition to the optimization problem which the controller needs to keep under control.
+Our experiments showed that it didn't.
+We were able to solve easy tasks, in which resonances of high magnitude still lead to acceptable results, however as soon as it got more difficult, high swinging lead to our control task not being accomplished.
+Also, noisy sensor measurements were a more difficult task to learn.
+We could not implement the same size of network and thus not achieve the same learning potential as the authors in \ac{DDPG}.
 
 \chapter{Future work}
-% Reworking reward functions
-% Trying different algorithms for better stability?
-% Pretraining the policy with a reference controller
-% Expert policy in instable conditions
-% Whatever comes to our mind
-% Punishment for high action gradients
-% Switch reward functions during training?
-% Interest from TU Berlin from control elements guy
-
-\section{More computational power}
-
-As we weren't able to solve the problem by our reasoning, we could just implement a huge hyperparameter search.
-Limiting factor in this is currently the simulation, which currently uses 70\% of the computation time, while we only use 30\% to train our networks.
-As on windows, gpu support in qblade is available, it might be interesting to run qblade on another machine than the training.
+
+Just because we didn't manage doesn't means it's impossible.
+We could imagine that with some additional work, learning stable policies would be possible.
+We have some ideas how we could achieve this result.
+
+\section{Scale up}
+
+It was limiting to perform all our experiment on the tiny machine.
+Limiting factor in this is currently the simulation, even with the lowered accuracy, which currently uses 70\% of the computation time, while we only use 30\% to train our networks.
+As currently, linux hardware acceleration is broken in qblade, a nice idea could be to fix this and to run future experiments on a cluster.
+This would also allow efficient hyperparameter sweeps, as we saw that hyperparameter choice greatly affects the outcome of the simulation.
+
+This would also allow to evaluate different wind scenarios and simulations with enabled wakes.
+
+\section{Validate \ac{PER}}
+
+We did not validate our combination of \ac{PER} and \ac{DDPG}.
+For our task, it gave better results, but it would be interesting to know whether our algorithm might also yield better results on general reinforcement learning tasks.
+This might also yield some insights into what makes our case special in comparison to general reinforcement learning.
 
@@
 As PID controllers have proven well in controlling a turbine and neural nets directly performed poorly, we could imagine to predict PID controller parameters and error targets instead of directly building the controller.
+Alternatively, feeding the algorithm with derivative and integrative inputs similar to what we described in section \ref{sec:feedpast} could enable the policy to build more resonant-robust results.
 
@@
 
-We did all our training under the simple hold-speed reward function.
-A realistic reward function however would incorporate power generated and damage incurred.
-As soon as a stable agent under the hold speed reward function was created, using different rewards could bring the performance of the controller closer to that of industry controllers.
-A complex reward function like what we proposed in section \ref{rewardfunctions} however also makes learning more difficult, and thus would need more work in refining the algorithm.
-
-\section{Switch reward functions during training}
+We did all our training under the simple hold-speed or hold-power reward function.
+A realistic reward function however would incorporate power generated and long-term damages occured.
+Especially the last part is tricky, as it would somehow have to be included into the scalar reward function.
+An interesting approach could be to try to train under multi-dimensional reward functions.
+To our knowledge, there haven't been efforts to accomodate this into Q-learning.
+Also, we could imagine interpolating between different reward functions during training.
+
 To combine the learning ease of a simple reward function with the performance of a complex reward, we could imagine fading over between a simple and complex reward function.
-In early stages of training, the network is conditioned to only output anything sane, while in later training it could optimize power or damage prevention more.
+In early stages of training, the network is conditioned to only output anything sane, while in later training it could optimize complex goals such as damage prevention more.
+This could be achieved by increasing the magnitude of the reward function further down in training, like after focussing on the big chunky goals zoom in on more fine-grained problems.
 
@@
 This would result in a safe controller, that upon insane policies would fall back to sane behavior.
-It would degrade exploration beyond what we doom safe, so a controller with this fallback method would never explore a death condition.
+It would degrade exploration to what we deem safe, so a controller with this fallback method would never explore a death condition.
 This would not necessarily hinder learning, as a complete death condition might not be necessary to explore to see the decreasing loss gradient leading to it.
@@
 
-\chapter{Conclusion}
-
-% It's pretty hard to create a working controller for a windturbine, apparently RL needs some more foundational research or we are just not able to
-
 \end{document}

### 2019-12-07 python/automated_hparam_search.py

@@ -55,13 +55,40 @@ import os
   {"id": 55},
 ]'''
 
-experiments = [
-  {"id": 56}, # Single critic, PER on, critic lr = 1e-4
+'''experiments = [
+  {"id": 56}, # Single critic, PER on, critic lr = 1e-4, hold speed
   {"id": 57},
   {"id": 58},
   {"id": 59},
-]
+]'''
 
+
+'''experiments = [
+  {"id": 60}, # Single critic, PER on, critic lr = 1e-4, hold power
+  {"id": 61, "critic_lr": 5e-5},
+  {"id": 62, "critic_lr": 1e-5},
+  {"id": 63, "actor_lr": 1e-5},  # Got closest
+  {"id": 64, "prioritized_experience_replay": False}, 
+  {"id": 65, "batch_size": 64},
+  {"id": 66, "batch_size": 32},
+  {"id": 67, "prioritized_experience_replay_alpha": 0.8, "prioritized_experience_replay_beta": 0.8},
+  {"id": 68, "prioritized_experience_replay_alpha": 0.3},
+  {"id": 69, "prioritized_experience_replay_alpha": 1, "prioritized_experience_replay_beta": 0.8},
+]'''
+
+# Hold lower power, actor lr 5e-5
+experiments = [
+  # {"id": 70}, # Disaster
+  # {"id": 71, "critic_lr": 5e-5}, # Disaster
+  # {"id": 72, "critic_lr": 1e-5}, # Disaster
+  # {"id": 73, "actor_lr": 1e-5}, $ Disaster
+  {"id": 74, "prioritized_experience_replay": False}, 
+  {"id": 75, "batch_size": 64},
+  {"id": 76, "batch_size": 32},
+  {"id": 77, "prioritized_experience_replay_alpha": 0.8, "prioritized_experience_replay_beta": 0.8},
+  {"id": 78, "prioritized_experience_replay_alpha": 0.3},
+  {"id": 79, "prioritized_experience_replay_alpha": 1, "prioritized_experience_replay_beta": 0.8},
+]
 
 def run_one(experiment):
   id = experiment['id']

### 2019-12-07 python/ddpg/ddpg.py

@@ -92,7 +92,7 @@ class DDPG:
     self.critic_loss_function = lambda target, pred, weights: generic_loss_function(target, pred, weights)
 
     # Buffor for experience replay
-    self.replay_buffer = BasicBuffer(hparams.buffer_maxlen, self.device, self.obs_dim, self.act_dim, hparams.prioritized_experience_replay_alpha)
+    self.replay_buffer = BasicBuffer(hparams.buffer_maxlen, self.device, self.obs_dim, self.act_dim, hparams.prioritized_experience_replay_alpha, self.dtype)
 
     # Noises
     # Random exploration
@@ -146,7 +146,8 @@ class DDPG:
                                  self.act_dim,
                                  self.hparams.act_low,
                                  self.hparams.act_high,
-                                 self.device)
+                                 self.device,
+                                 self.dtype)
 
     self.logger = QBladeLogger(self.hparams.logdir, self.hparams.log_steps, self.hparams.run_name, self.hparams.obs_labels, self.hparams.act_labels)
     for key, value in self.hparams.values().items():
@@ -214,7 +215,7 @@ class DDPG:
   # Returns an unnormalized expert action from the expert controller
   def get_expert_action(self, state):
     if(self.act_dim == 2):
-      return [2e7*state[0]-12e6,128.6*state[0]-102.9]
+      return [2e7*state[0]-12e6,128.6*state[0]-128.6]
     else:
       return np.zeros(self.act_dim)
 
@@ -338,7 +339,6 @@ class DDPG:
 
 
     # When using PER, calculate IS-weights
-    weights = torch.ones(batch_size)
     if(self.hparams.prioritized_experience_replay):
       weights = self.replay_buffer.get_probabilities(indices)
       weights = (weights*batch_size) ** (-self.hparams.prioritized_experience_replay_beta)
@@ -346,6 +346,8 @@ class DDPG:
       assert(not np.any(np.isnan(weights)))
       assert(not np.any(weights > 1))
       weights = torch.tensor(weights, dtype=self.dtype).to(self.device)
+    else:
+      weights = torch.ones(batch_size, dtype=self.dtype, device=self.device)
 
     # Run the critic
     next_actions = self.actor_target.forward(next_state_batch).detach()
@@ -597,7 +599,7 @@ class DDPG:
       "prefix": prefix,
       "time": str(datetime.now()),
       "killcount": self.killcount,
-      "hparams": self.hparams.values(),
[…]

### 2019-12-07 python/ddpg/normalizer.py

@@ -4,9 +4,9 @@ import numpy as np
 from .utils import BasicBuffer
 
 class Normalizer:
-  def __init__(self, obs_dim, act_dim, act_low, act_high, device):
+  def __init__(self, obs_dim, act_dim, act_low, act_high, device, dtype=torch.float):
     self.device = device
-    self.dtype = torch.float
+    self.dtype = dtype
     self.obs_dim = obs_dim
     self.act_dim = act_dim
     self.act_high = np.array(act_high)

### 2019-12-07 python/ddpg/utils.py

@@ -211,7 +211,7 @@ class SumTree:
 # A "Basic" Buffer which stores entries and can sample them by priority
 class BasicBuffer:
 
-    def __init__(self, max_size, device, state_dim, action_dim, alpha=1):
+    def __init__(self, max_size, device, state_dim, action_dim, alpha=1, dtype=torch.float):
         self.device = device
         self.max_size = max_size
         self.iterator = 0 # Next position to write to
@@ -220,7 +220,7 @@ class BasicBuffer:
         # CPU buffer
         self.buffer = np.array([(None, None, None, None, None) for _ in range(max_size)])
 
-        self.dtype = torch.float
+        self.dtype = dtype
 
         # GPU buffers
         self.buffer_s = torch.zeros(max_size, state_dim, dtype=self.dtype, device=device, requires_grad=False)

### 2019-12-07 python/env/qbladeadapter.py

@@ -92,17 +92,22 @@ class QBladeAdapter:
     return 2
 
   def get_act_high(self):
-    #return [10, 10, 10, 10, 10]
-    return [474029.1, 90]
+    gearbox_ratio = 97
+    generator_torque = 47402.91
+    return [generator_torque*gearbox_ratio, 90]
 
   # TODO get actual values for this
   def get_act_low(self):
     #return [0, 0, 0, 0, 0]
-    return [0, 0]
+    return [0,0]
 
   def get_act_max_grad(self):
     timestep = 0.1
-    return [150000*timestep, 8*timestep]
+    gearbox_ratio = 97
+    torque_step = 15000
+    pitch_step = 8
+
+    return [torque_step*gearbox_ratio*timestep, pitch_step*timestep]
 
   def get_obs_labels(self):
     return {
@@ -147,15 +152,15 @@ class QBladeAdapter:
     else:
       death_penalty = 0
 
-    rated_power = 3200
+    rated_power = 3500
     rated_speed = 0.8
     #return -np.abs(observation[1]-rated_power) - 1e3*(np.abs(observation[16]) + np.abs(observation[17]) + np.abs(observation[18]))
-    #return 1-np.abs((observation[1]-rated_power)/rated_power)-death_penalty
+    return np.clip(1-np.abs((observation[1]-rated_power)/rated_power)-death_penalty, -4, 4)
     #return np.clip(observation[1], 0, None) - 1e4*(np.abs(observation[16]) + np.abs(observation[17]) + np.abs(observation[18]))
     #return np.clip(5 
     #  - np.abs((observation[1]-rated_power)/rated_power)
     #  - 5e-2*(np.abs(observation[16]) + np.abs(observation[17]) + np.abs(observation[18])), -10, 10)
-    return np.clip(1-np.abs((observation[0]-rated_speed)/rated_speed)-death_penalty, -4, 4)
+    #return np.clip(1-np.abs((observation[0]-rated_speed)/rated_speed)-death_penalty, -4, 4)
 
   def calc_death(self, observation):
     # If there are nan values, reload completely
@@ -168,11 +173,15 @@ class QBladeAdapter:
 
     # 63 is the rotor size, so if anything bends further than that, it's broken off.
     broken_state = 20
+    max_rotorspeed = 3
+    min_rotorspeed = -0.05
+    max_power = 8000
     return np.abs(observation[16]) > broken_state or \
            np.abs(observation[17]) > broken_state or \
[…]

### 2019-12-07 paper/paper.tex
Reworked death conditions part

@@
 Supervised learning, unsupervised learning and reinforcement learning.
-In supervised learning, some form of fitting is done on a dataset.
+In supervised learning, a function approximator is fitted to model properties of a dataset.
+In unsupervised learning, properties in a dataset are measured, compared or grouped.
+In reinforcement learning, the dataset is explored interactively.
+We are trying to learn a controller for a windturbine, so in theory we could use both supervised an reinforcement learning.
+If we were to use supervised learning, we would need some precomputed data, which could only be generated by already running a windturbine with a controller and measuring the data.
+As we however want to learn this controller, it is not present at the start of supervised learning, and imitating known controllers could at best get exactly as good as the known controller.
+So we are using reinforcement learning.
+
+With the combination of both, we aim to achieve insights into the applicability of reinforcement learning to wind turbine control tasks.
+We are aiming at laying a foundational work, not necessarily yet outperforming state of the art control systems.
+Our hope is that the rising performance and more sophisticated algorithms are able to solve pending problems in wind energy now or in the near future.
 
@@
 \section{Gym Experiments}
-
+\label{sec:gymexperiments}
 At the start of our experimentation, we took off with a well-known problem and a proven implementation.
@@
 
+The authors of \ac{DDPG} trained their networks up to 2.5 million steps, which we can't afford with our computational power, so a simpler network architecture will hopefully also converge faster.
+
 \subsection{Normalization}
@@
 As sometimes during random exploration, no extreme values were observed and later extreme values lay far outside of $[-1, 1]$, we instead normalized on replay data of a previous run.
+Additionally, we added a constant of 3 after normalization because of the phenomenon described in section \ref{sec:death}.
 We could also imagine manually setting normalization bounds on low-dimensional problems, as human knowledge about the possible state space is usually present.
@@
 \subsection{Regarding death conditions}
-
+\label{sec:death}
+
+We were observing many exploding turbines in our experiments, so we had a look at how \ac{DDPG} handles death conditions.
 In \ac{DQN}, death conditions are accounted for by changing the Q target in the terminal step to be $y = r$ instead of $y = r + \gamma \max_a' Q(s', a')$.
-This way, behavior after a death will not be accounted for by the Q function.
-As we are seeing many death conditions, we thought it could be useful to include this into our DDPG algorithm.
+In fact, we regard the lack of this in \ac{DDPG} as a bug.
+To explain why, let's have a look at our Q-function again:
+\begin{equation}
+Q(s,a) = r(s,a) + \gamma Q(s', a')
+\end{equation}
+
+The right term of this equation $\gamma Q(s', a')$ is likely several magnitudes higher than the left term $r(s,a)$.
+With our $\gamma$ = 0.99, circa 99 times as high as $\int_1^{\infty} 0.99^x \approx 98.5$.
+Though we punish death conditions severely, a death condition will not exceed 3 times the rewards of normal operations due to our clipping.
+So if, after a death condition, a good condition is reached, our algorithm would gladly take that penalty.
+For this reason we introduced the updated death notion
+
+\begin{equation}
+Q(s,a) = r(s,a) + \begin{cases} 0& \text{if death}\\ Q(s', a') & \text{else}\end{cases}
+\end{equation}
+
+However, now it could be advantageous to die in case of negative Q values.
+Let's suppose we have a terrible policy which under no circumstances produces positive rewards.
+Thus, our Q term would be negative.
+As discussed before, it is likely of high magnitude, so elimiating the high negative Q for a 0 would be a good strategy.
+Thus we need to make sure the Q term is never negative.
+We do this by adding the lower clipping bound to rewards after clipping them.
 So now, instead of $y = r + \gamma Q'(s', \mu'(s'))$ we change it to be $y = r + \gamma Q'(s', \mu'(s')) * (1-d)$ where $d \in {0, 1}$ symbolize the death conditions.
+We haven't found any literature which does a similar thing with \ac{DDPG}
 
@@
     Apply importance sampling $\delta_i = \delta_i w_i$\\
-    Update critic by minimizing the huber loss: $L = \frac{1}{N} \sum_i \begin{cases} (\delta_i)^2 & \mathrm{if} \delta < 1\\ \abs{\delta_i} & \mathrm{else}\end{cases}$\\
+    Update critic by minimizing the huber loss: $L = \frac{1}{N} \sum_i \begin{cases} (\delta_i)^2 & \text{if } \delta < 1\\ \abs{\delta_i} & \text{else}\end{cases}$\\
     Sample a random minibatch of $N$ states $(s_i)$ from $\mathcal{R}$ uniformly\\
@@
 With learning rate 1e-3, we reach it somewhere around episode 10, with learning rate 1e-4 a bit later around epoch 25.
+We also tried 1e-5, but we could not see any convergence with this, so we ommitted the plot.
 The time we are holding it before divergence also raises with learning rate.
@@
 As this delayed target net will diverge with the original net if divergence happens slow enough, we do not see it.
+
+From this we see that higher learning rates cause a more stable learning, but reach convergence later. The results suffer from strong oscillations both with the low and high learning rate.
 
@@
 At the end, most controllers converged to leaving pitch at zero and left the turbine to die regularily with maximum generator torque.
-
-\section{Hold power}
-The next task
+Though this is not an optimum in the global reward function, apparently this became an optimum in the Q estimate.
+Looking back at our explanation from section \ref{sec:death} where we gave some options on why an agent could favor a death over returning to a sensible region, we can exclude the case that good Q values after a death (because after a reset the agent returns to sensible area) propagate.
+So we looked at the case of negative Q estimates.
+When plotting Q estimates, we found that indeed Q is sometimes predicting negative values, especially in early stages of training.
+
+A possible explanation for favouring deaths could be that our death penalties are spikes in the reward functions.
+As explained in section \ref{sec:rewardfunctions}, we add a constant penalty upon death but not in the steps before.
+As our Q function approximator approximates smooth functions, it will likely have a high TD-Error on these samples.
+This in turn will cause \ac{PER} to replay these samples relatively often in comparison to other samples.
+As we didn't use a \ac{PER} beta of 1 to fully correct against importance sampling bias but 0.5 as proposed in the paper, we could imagine that still some oversampling happened along these points and they actually ended up being largely overestimated.
+
+Also, we could imagine that the Q function gets deformed before deaths and maybe actually exposes a local maximum right before it.
+The spikey nature could lead to weird situations in a high-dimensional regression, which could be similar to overfitting.
+As the regression wants to fit to the outlier, it might create a strong curvature before this.
+To illustrate this, let's plot a typical overfit scenario in figure \ref{fig:overfit}.
+The dot on the bottom right is our spike in death.
+Deriving along the linear regression (blue line) would lead us away from the death.
+The orange line however exposes a local maximum shortly before the death.
+If our policy for some reason got stuck in this local maximum, it would be trained to stay close to a death condition, which would, because of system instability, lead to many death conditions happening.
+Oversampling effects might then even increase this phenomenon.
+
+We are unsure why the controllers showed such a bad performance, and couldn't get to fix this problem anymore.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.5\linewidth]{images/overfit.png}
+  \caption{Overfitting a 5 point regression}
+  \label{fig:overfit}
+\end{figure}
 
@@
 To exclude the possibility that our implementation is buggy, we decided to run our DDPG algorithm on the Pendulum task.
-With our hyperparameters, we could not observe convergence and only switching back to the algorithm of \ac{DDPG}, the agent managed to balance out the pendulum after 11 epochs.
-Especially the network size hyperparameter was important for this.
+With our hyperparameters, we could not observe convergence and only switching back to the algorithm of \ac{DDPG} which we already tried in \ref{sec:gymexperiments}, the agent managed to balance out the pendulum.
+Thus we can not exclude misimplementation as a bug might only occur with our set of hyperparameters, however we suspect that windturbines are just not very similar to pendulums.
 
 \section{Discussion}
+
+To summarize it, the level of control we have achieved does not outperform any previous benchmark controller.
+In fact, it would destroy most windturbines relatively quickly and does not yield sensible energy.
+Still we showed that it is possible, that existing algorithms however exhibit some significant problems and need reworking.
+Also we showed that we are not able to perform this reworking in the given amount of time.
 
@@
 Balance tasks could be argued to resemble our windturbine in a way, as we want to balance an output aswell.
-However a windturbine is vastly more complex than a pendulum.
+However a windturbine is apparently different enough so we can not optimize it with the same algorithm which optimized the other games.
 It exhibits resonant frequencies in addition to the optimization problem which the controller needs to keep under control.
@@
 This would also allow efficient hyperparameter sweeps, as we saw that hyperparameter choice greatly affects the outcome of the simulation.
-
-This would also allow to evaluate different wind scenarios and simulations with enabled wakes.
-
-\section{Validate \ac{PER}}
+Also, it would be interesting to run simulations for a longer time.
+The authors of \ac{TD3} ran their simulations for up to 1 million steps, \ac{DDPG} 2.5 million steps and \ac{DQN} 50 million.
+On easier tasks, TD3 could reach first sensible results after 0.2 million steps.
+With our computational power we couldn't run anything above 0.3 million steps.
+But we also chose a less complex architecture which should be faster to train.
+
+With high computational power, we could also evaluate many different wind scenarios and turbine models, in contrast to our evaluation which was done on a single wind speed and turbine type.
+
+\section{Validate against OpenAI-Gyms}
 
 We did not validate our combination of \ac{PER} and \ac{DDPG}.
+Neither did we validate our addition of death conditions.
 For our task, it gave better results, but it would be interesting to know whether our algorithm might also yield better results on general reinforcement learning tasks.
-This might also yield some insights into what makes our case special in comparison to general reinforcement learning.
+Running through the suite of OpenAI Gyms would be an interesting work and could show whether our critique on \ac{DDPG} was justified.
+Also, we might be able to get \ac{TD3} to work there.
+This comparison might also yield some insights into what makes our case special in comparison to general reinforcement learning.
 
@@
 A difficulty incurred in that research is the design of a controller, as the high-dimensional action space makes it difficult for a human to engineer a controller.
-Using our automatically learning agent could enable a first working version to that research, providing a first direction or even lead to performance increases over not using active control elements in a turbine.
+As soon as we could provide a working version of an automatically learning, stable agent, we could assist the efforts of the people working on flap design to automatically learn controllers for them.
 

### 2019-12-07 python/ddpg/ddpg.py

@@ -393,6 +393,7 @@ class DDPG:
     self.critic_optimizer.step()
     self.logger.add_scalar('Loss/q', q_loss.detach(), time)
     self.logger.add_scalar('Loss/next_Q_mean', next_Q.mean().detach(), time)
+    self.logger.add_scalar('critic/deaths_seen', (1 - masks).sum(), time)
 
     # Also, for correction of oversampling in later policy replays, redraw samples
     if(self.hparams.prioritized_experience_replay):
@@ -707,7 +708,7 @@ class DDPG:
       buffer_maxlen = 100000,
 
       # Learning rate of the Q approximator
-      critic_lr = 1e-5,
+      critic_lr = 1e-4,
 
       # Neural network sizes of the critic
       critic_sizes = [128, 64],
@@ -731,7 +732,7 @@ class DDPG:
       twin_critics = False,
 
       # Learning rate of the policy
-      actor_lr = 1e-5,
+      actor_lr = 5e-5,
 
       # Network sizes of the policy
       actor_sizes = [32, 16],
@@ -760,7 +761,7 @@ class DDPG:
       optimizer = 'adam',
 
       # How many training steps to do per environment iteration
-      training_steps_per_env_iteration = 1,
+      training_steps_per_env_iteration = 2,
 
       # Whether to clip gradients
       # If yes, to which values to clip them to

### 2019-12-07 python/ddpg/normalizer.py

@@ -60,7 +60,7 @@ class Normalizer:
       action = torch.clamp((action - c) / m, -3, 3)
     else:
       m, c = self.action_normalizer_params
-      action = np.clip((action - c) / m, -3, 3)
+      action = np.clip((action - c) / m, -3, 3) 
     
     return action
 
@@ -87,16 +87,16 @@ class Normalizer:
 
     if(gpu):
       m, c = self.reward_normalizer_params_gpu
-      reward = torch.clamp((reward - c) / m, -3, 3)
+      reward = torch.clamp((reward - c) / m, -3, 3) + 3
     else:
       m, c = self.reward_normalizer_params
-      reward = np.clip((reward - c) / m, -3, 3)
+      reward = np.clip((reward - c) / m, -3, 3) + 3
     
     return reward
 
   def denormalize_reward(self, reward, gpu=False):
     m, c = self.reward_normalizer_params_gpu if gpu else self.reward_normalizer_params
-    return reward * m + c
+    return (reward - 3) * m + c
 
   def calc_normalizations(self, replay_buffer, normalization_extradata=None):
     # Load additional data if wanted

### 2019-12-07 python/env/qbladeadapter.py

@@ -175,7 +175,7 @@ class QBladeAdapter:
     broken_state = 20
     max_rotorspeed = 3
     min_rotorspeed = -0.05
-    max_power = 8000
+    max_power = 10000
     return np.abs(observation[16]) > broken_state or \
            np.abs(observation[17]) > broken_state or \
            np.abs(observation[18]) > broken_state or \

### 2019-12-09 paper/paper.tex
Abgabe

@@
-\documentclass[hyperref,beleg]{cgvpub}
+\documentclass[hyperref,final,beleg]{cgvpub}
 
@@
 
-\betreuer{Dr. Dmitrij Schlesinger}
+\betreuer{Dr. Dmitrij Schlesinger, PhD. Matthew Lennie}
 \bibfiles{literatur}
 \problem{
-  \section{Background:}
+  The following problem formulation is an exact copy from what M. Lennie drafted at the beginning of the work and the only part of this extract which is not formuluated by N. Westerbeck.
+
+  \textbf{Problem}
+
   Here at the HFI Experimental fluid mechanics group, we have developed an
@@
 
-  \section{Tasks}
+  \textbf{Tasks}
+
   The major tasks of the project are as follows:
@@
 
-\acknowledgments{At first thanks to my advisors Mattew Lennie and Dmitrij Shlezinger for their support.
+\acknowledgments{At first thanks to my advisors Matthew Lennie and Dmitrij Shlezinger for all the support they gave in the process of writing this thesis.
+Thanks to David Marten for helping with QBlade.
 Also, I guess, thanks David Silver - 7 of my most important papers were from you.
@@
 
-Recent advancements in reinforcment learning have managed to tackle more and more complex problems, like StarCraft or Go.
-The range of topics, RL is applicable to, increases.
+Recent advancements in \ac{RL} have managed to tackle more and more complex problems, like StarCraft or Go.
+The range of topics, \ac{RL} is applicable to, increases.
 However, so far, these advancements were only partially applied to the area of windturbine pitch and torque control.
-Current state of the art linear controllers are performing well in maintaining turbine control, and the optimization margin for these is small, as they achieve the theoretical maximum energy output up to rated windspeed and manage to keep the turbine intact above rated windspeeds.
-However, these simple controllers can only deal with a limited amount of data inputs and are preventing turbine damages.
-We are evaluating the use of an actor-critic reinforcement learning algorithm to solve continous control of a windturbine, hoping to lay a foundational work for later improvements.
-We find that utilizing a wildly proven continous control algorithm does not solve our windturbine problem.
-Finally, we demonstrate that this control is possible, however we do not achieve results outperforming industry controllers.
+Current state of the art linear controllers are performing well in maintaining turbine control, and the optimization margin for these is small, as they achieve near theoretical maximum energy output.
+However, these simple controllers can only deal with a limited amount of data inputs and are struggling to prevent long-term turbine damages.
+We are evaluating the use of DDPG as an actor-critic reinforcement learning algorithm to solve continuous control of a windturbine, hoping to lay a foundational work for later improvements.
+We find that utilizing the unmodified algorithm does not solve our windturbine problem, thus propose to extend DDPG with a range of known and novel methods, including prioritized experience replay, death handling and normalization.
+Finally, we demonstrate that windturbine control is possible, however we do not achieve results outperforming industry controllers.
 
@@
 With the issue of global warming, we are facing maybe the biggest challenge to our generation.
-Though wheather models have predicted the effect since the 1980s, only in the recent years this topic has been brought into societal focus.
-CO2 is seen as the main cause for a tendency in the global weather to heat up.
-Since our industrial age, humankind has been releasing CO2 into the athmosphere in large scales.
-If however we continue doing so the way we are doing now, we will likely increase global temperature by a degree which will make living on this planet difficult.
+Though weather models have predicted the effect since the 1980s, only in the recent years this topic has been brought into societal focus.
+CO2 is seen as one of the main cause for a tendency in the global weather to heat up.
+Since our industrial age, humankind has been releasing CO2 into the atmosphere in large scales.
+If however we continue doing so the way we are doing now, we will likely increase global temperatures by roughly 3 degrees which will make living on this planet difficult.
 With the Paris agreement, the United Nations decided to limit global warming to 1.5 degrees, for which massive reductions in CO2 emissions are necessary.
@@
 The sun heats different parts of the earth to different temperatures, as the surface of the earth isn't uniform.
-The air over hotter regions such as close to the equator tends to rise, while on mountains and the poles it tends to fall.
-Combine this with a vast and complex interplay of humidities, Coriolis effects, mountains, cloud formation and enough effects to fill an entire branch of science and you will get wind.
+The air over hotter regions such as close to the equator tends to rise, while over oceans and the poles it tends to fall.
+Combine this with a vast and complex interplay of humidities, Coriolis effects, mountains, cloud formation and enough co-dependence to fill an entire branch of science and you will get wind.
 Windturbines operate by drawing energy from this weather phenomenon.
@@
 This makes drawing energy from it more complex, or less efficient.
-If we were to draw energy from both storms and light breezes with the same turbine, we would not be efficient.
+If we were to draw energy from both storms and light breezes with the same turbine, we would not be efficient as the turbine can only draw so much energy that it survives the storm, which projected on the light breeze is very little energy yield.
 Thus, modern windturbines are built to be adjustable.
 With these adjustments, a turbine can operate sensibly over a broader range of wind conditions.
-However, somehow, these adjustments need to be automatically adapted to the wind condition.
+However, somehow, these adjustments need to be automatically adapted to the inflowing wind.
 This is the task of the turbine controller.
@@
 
-Machine learning has developped immensly over the last years.
-Benefitting from Moore's law, huge amounts of computational power can be used to train more and more complex models.
+Machine learning has developed immensely over the last years.
+Benefiting from Moore's law, huge amounts of computational power can be used to train more and more complex models.
 Machine learning is generally split into three big branches.
@@
 In unsupervised learning, properties in a dataset are measured, compared or grouped.
-In reinforcement learning, the dataset is explored interactively.
+In reinforcement learning, the dataset is an environment and it is explored interactively.
 We are trying to learn a controller for a windturbine, so in theory we could use both supervised an reinforcement learning.
@@
 % Quote wind-energy handbook, talk about how currently wind turbines are controlled.
-There are two mayor types of wind turbine designs, \ac{HAWT} and \ac{VAWT}, which differ by their rotation axis.
-In this work, we will only look at \ac{HAWT} and all mentions of windturbines mean \ac{HAWT} type turbines.
+There are two major types of wind turbine designs, \ac{HAWT} and \ac{VAWT}, which differ by their rotation axis.
+In this work, we will only look at \ac{HAWT} as these are more commonly used, in our work all mentions of windturbines mean \ac{HAWT} type turbines.
 Such a turbine is made up by 3 big components, a tower on which a nacelle is rested which itself has a rotor in front.
-The joint between tower and nacelle allows for rotation to turn the rotor into the wind.
+The joint between tower and nacelle allows for rotation to turn the rotor into the wind, and the angle between inflowing wind and nacelle orientation is called yaw angle.
 Except for recent ideas \cite{howlandWindFarmPower2019}, the best value for the rotational direction was to always yaw the rotor directly into the wind, and for the sake of simplicity we will omit this control parameter from our simulation and leave it at 0 degrees with wind facing straight onto the rotor.
 In the nacelle there is, most importantly, an electrical generator, which can be given a specific torque.
-It will then slow down the rotor with that torque and at the same time generate energy proportional to rotor speed and torque.
+Torque is a word for the amount of rotational force, here in opposite direction to rotation.
+This force will slow down the rotor and at the same time generate energy.
+If generator torque is approximately equal to the torque created by wind, the two cancel out and leave the rotor at constant rotational speed, which is what we want to achieve.
 This torque is one of the two more important control parameters, together with blade pitch.
-The blades are designed to operate on maximal aerodynamic efficiency when they are pitched at 0 degrees, increasing the pitch angle will (except for possible slight improvements in the first few degrees) reduce aerodynamic efficiency of the rotor.
-
+The blades are designed to operate on maximal aerodynamic efficiency when they are pitched close to 0 degrees, increasing the pitch angle will (except for possible slight improvements in the first few degrees) reduce aerodynamic efficiency of the rotor.
+The main use case of blade pitch is to protect the turbine from damage, as high wind speeds can generate higher loads than the turbine could handle.
+Our generator has a maximum torque limit, and if the inflowing wind generates more than that torque, the turbine would spool up faster than what it can handle.
+We want to pitch the blades as little as possible so we can maintain maximum generator torque.
 \begin{figure}
@@
 PID-Controllers in general have one real-valued input over time and deliver an output.
-This is done by calculating an error term between the input and a desired input.
-The output is calculated as a mixture of the error itself (P - proportional to the error), integrating the error over the last few timesteps (I - integral of the error) and calculating a derivative to the last timestep (D - derivative).
+This is done by calculating an error term between the measured input and a desired input.
+The output is calculated as a mixture of the error itself (P - proportional to the error), integrating the error over the last few time-steps (I - integral of the error) and calculating a derivative to the last time-step (D - derivative).
 The output is the sum of these 3 terms each factorized with a constant factor.
+There are well proven formulas how to choose these parameters and also some adaptive methods, but control theory usually chooses these parameters by hand.
 However back to windturbines.
@@
 Our reference turbine has a peak power coefficient of 0.482 - meaning that at optimal wind speed, 48\% of the energy of the wind is converted into electricity.
-In fact, the theoretical maximum is given by the Betz limit of 59.3\% and if we account for electrical and frictional losses, the realistic maximum is even lower.
+In fact, the theoretical maximum is given by the Betz limit of 59.3\% and if we account for unavoidable electrical and frictional losses, the realistic maximum is even lower.
 So there is not much potential to be alleviated.
@@
 You might have seen flaps on an aircraft wing, extending the length of the wing or adding gaps at takeoff and landing to account for the low wind speeds hitting the aircraft wing in these flight scenarios.
-This could also benefit wind turbines at low wind speeds, however building a good controller for them is challenging.
+This could also benefit wind turbines either for higher efficiency on low wind speeds or to furthermore reduce blade loads.
+However building a good controller for them is challenging due to the high number of interplaying parameters.
 Tackling all this with \ac{SISO} is possible, but a lot of manual work.
@@
 It is getting cheaper to install high-precision sensors, and many of them.
-Wind measurement techniques such as LIDAR based aenometers are able to efficiently and precisely predict incoming wind situations.
+Wind measurement techniques such as LIDAR based anemometers are able to efficiently and precisely predict incoming wind situations.
 Load measurements can be done in real-time across a turbine blade and in theory, it would be possible to incorporate these into control systems.
-However, accomodating all these input parameters in \ac{SISO} systems is difficult as we would have to hand-model each of the inputs and how exactly it ends up in the output.
-
-We hope to lay a foundational step in solving these issues by using an end-to-end machine learning algorithm which is capable of learning a wind turbine control scenario with many inputs and outputs.
+However, accommodating all these input parameters in \ac{SISO} systems is difficult as we would have to hand-model each of the inputs and how exactly it ends up in the output.
+
+We hope to lay a foundational step in solving these issues by using an end-to-end reinforcement learning algorithm which is capable of learning a wind turbine control scenario with many inputs and outputs.
 Also, we must admit that we are generally interested in playing around with reinforcement learning, it's fucking cool.
@@
 \section{QBlade}
-% Quickly introduce qblade as a simulation tool
+% Quickly introduce QBlade as a simulation tool
 Our data source in this work is the open-source simulation tool QBlade \cite{martenQBLADEOPENSOURCE2013} developed at TU Berlin.
@@
 It uses an algorithm which approximates the 3D blade structures with 2D structures and corrects the results through several error terms.
-Though a full CFD simulation would yield higher accuracy, our computational resources don't allow for that.
+Though a full 3D CFD simulation would yield higher accuracy, the increase in computational power needed doesn't justify the accuracy improvement yet.
 To be used in reinforcement learning, we designed an interface, over which the QBlade simulation can be embedded into an external environment.
 Most machine-learning frameworks are written in python, so we decided to compile QBlade into library format and expose the most fundamental functions to allow it to communicate with any programming language that can load libraries.
+During this work, David Marten of TU Berlin was of great help, implementing all necessary changes to QBlade so it can be compiled as a library, thank you for this!
 As python has a ctypes interface to load c-code, we could link a python agent to the QBlade C++ environment.
@@
 QBlade allows different simulation scenarios, for our testing we decided to only use the NREL 5MW \cite{jonkmanDefinition5MWReference2009} turbine with the default structural model and using all implemented correction mechanisms to achieve the most realistic data possible.
-As reference data to this turbine is easily available and publicly published by NREL, this allows us for good cross-validation.
-
-The interface is made of 7 functions, which allow for loading a project, resetting the simulation, setting controller inputs and getting environment measurements and advancing the simulation a timestep.
-The observations returned in getControlVars match those that are visible to standart industry controllers, which in turn are modelled after what can be measured on a real windturbine.
+As reference data to this turbine is easily available and publicly published by NREL, this allows for good cross-validation.
+
+The interface is made of 7 functions, which allow for loading a project, resetting the simulation, setting controller inputs and getting environment measurements and advancing the simulation a time-step.
+The observations returned in getControlVars match those that are visible to standart industry controllers, which in turn are modeled after what can be measured on a real windturbine.
 We have 23 observations and 5 actions.
@@
 After a while of experimenting with the unrestricted simulation, we added some bounds to our control parameters and hid some observations and actions.
-Restrictions to our final version are described in section \ref{sec:alg_qblade}.
+Restrictions to our final version are described in section \ref{sec:alg_QBlade}.
 
@@
 This could be an actual cyber-physical system or a simulation.
-The first assumption to this environment is that this system operates over time and we can discretize time into timesteps.
+The first assumption to this environment is that this system operates over time and we can discretize time into time-steps.
 Our environment here is the simulated wind turbine, but it could be a computer game \cite{lillicrapCONTINUOUSCONTROLDEEP2015}, a physics simulation\cite{brockmanOpenAIGym2016} or even an actual existing wind turbine \cite{kolterDesignAnalysisLearning2012}.
-In every timestep, the environment supplies us with an observation \(x\) and a scalar reward \(r\) for taking an action \(a\) in that state.
-In theory, the full trajectory \((x_0, a_0, ..., x_t)\) could be needed to describe the full state of the simulation at timestep t \(s_t\).
-However, the algorithms we looked at assume that the state progressions can be modelled as a \ac{MDP} of observations, as in that what will happen at timestep t+1 is only dependent on the observation at timestep t, not also on for example t-10.
-In \ac{MDP}, the state transition probability \(p(x_{t+1}|x_t, a_t)\) is fully descriptive, as in it is equal to \(p(x_{t+1}|x_0, a_0, ..., x_t, a_t\).
+In every time-step, the environment supplies us with an observation \(x\) and a scalar reward \(r\) for taking an action \(a\) in that state.
+In theory, the full trajectory \((x_0, a_0, ..., x_t)\) could be needed to describe the full state of the simulation at time-step t \(s_t\).
+However, the algorithms we looked at assume that the state progressions can be modeled as a \ac{MDP} of observations, as in what will happen at time-step t+1 is only dependent on the observation at time-step t, not also on for example t-10.
+In other words, in \ac{MDP}, the state transition probability \(p(x_{t+1}|x_t, a_t)\) is fully descriptive, as in it is equal to \(p(x_{t+1}|x_0, a_0, ..., x_t, a_t)\).
 Because we assume the observation \(x\) to be fully descriptive, we will use the term state as equivalent to observation \(s_t = x_t\).
@@
 
-In every timestep, the agent supplies an action \(a\).
+In every time-step, the agent supplies an action \(a\).
 This action could be button presses in an arcade game or blade pitch in a windturbine simulation.
@@
 \begin{equation}
-J = \mathbb{E}r
+J = \mathbb{E}[r]
 \end{equation}
@@
 Additionally to the terms defined by the environment, we introduce some terms which are common to reinforcement learning
+
 \begin{itemize}
   \item \textbf{Policy} A function that maps from states to actions.
-There are deterministic policies $\mu: S -> A$ and stochastic policies $\pi: S -> P(A)$.
+There are deterministic policies $\mu: S \rightarrow A$ and stochastic policies $\pi: S \rightarrow P(A)$.
 If we approximate a policy by a function approximator parametrized with $\theta$, we write $\pi(s;\theta)$ or omit the parameters when clear.
+We sometimes call a learnt policy \textit{actor}
+  \item \textbf{State-density} We define a function similar to \cite{suttonPolicyGradientMethods} which describes how likely it is to be in state $s$ when following a policy $\pi$ forever.
+$d^{\pi}(s) = \lim_{t -> \infty} p(s_t = s | s_0, \pi)$.
+If we for example have a policy which plays Go as well as the author of this paper, states with a lot of enemy stones and very few own stones might happen more often than others, and a state without any enemy stones would be very unlikely be seen at all.
+Also according to \cite{suttonPolicyGradientMethods} this distribution is independent of the starting state $s_0$ and only dependent on the policy and of course the environment.
+In later papers, commonly a discounted version of this is used.
+$\rho^{\pi}(s)$ weights future probabilities less and thus can be described as the probability of being in state s \textit{soon} when following policy $\pi$.
+We will omit a precise definition here, you can look it up at \cite[Section 2.1]{silverDeterministicPolicyGradient2013}.
   \item \textbf{Value} A value with respect to a policy and a starting state s gives the accumulative reward of following that policy from state s until infinitely in the future.
-$V^{\pi}(s_t) = \mathbb{E}_{s_i \sim \varepsilon, a_i \sim \pi, i>t} r(s_i, a_i) $.
+$V^{\pi}(s_t) = \mathbb{E}_{s_i \sim \rho^{\pi}, a_i \sim \pi, i>t} r(s_i, a_i) = \int_{s_i, i>t} \rho^{\pi}(s_i) \int_{a_i, i>t} \pi(a_i) r(s_i, a_i) $.
 As we have a stochastic environment and policy, we have to add an expectation term over actions taken and states resulting from those actions.
@@
   \item \textbf{Return} If we measured a value function, that would be the return.
-Obviously we can't measure infinitely, so a return is only defined on a finite trajectory of state-transitions, but otherwise is equal to a value.
-  \item \textbf{Optimal policy} The theoretical construct of a policy that always takes the best action with respect to accumulative reward.
-Though we rarely know this policy, there must always be at least one policy which gives the highest value of all possible policies in a finite \ac{MDP}
-  \item \textbf{Q-Value} The reward of taking action a in state s plus the value following a policy after that $Q^{\pi}(s,a) = r(s, a) + \mathbb{E}_{s' \sim \varepsilon} V^{\pi}(s')$.
+Obviously we can't measure infinitely, so a return is only defined on a finite trajectory of state-transitions that we can actually observe, but otherwise is equal to a value.
+  \item \textbf{Optimal policy} The theoretical construct of a policy that always takes the best action with respect to accumulative reward is called optimal policy.
+Though we rarely know this policy, we usually want to know it.
+In a finite \ac{MDP}, there must always be at least one policy which gives this highest value of all possible policies.
+  \item \textbf{Q-Value} The reward of taking action a in state s plus the value following a policy after that $Q^{\pi}(s,a) = r(s, a) + \mathbb{E}_{s' \sim \varepsilon} V^{\pi}(s')$ is called Q-Value.
 When following the optimal policy, we simply write $Q^*(s,a)$.
-Usually, our Q functions are discounted.
-Note that we can also define Values on Q-Values: $V^{\pi}(s) = \mathbb{E}_{a \sim \pi} Q(s, a)$.
+Usually, our Q functions are discounted similar to how we discount values.
+Note that we can define Values and Q-Values on each other: $V^{\pi}(s) = \mathbb{E}_{a \sim \pi} Q(s, a)$.
 If we approximate Q by a function approximator parametrized with $\theta$, we write $Q(s, a;\theta)$ or omit it in case it is clear.
-  \item \textbf{Actor-Critic} Actor is just another name for a learnt policy and critic just another name for an estimated Q-Value function.
-These two frequently appear together, if that happens we are using an actor-critic algorithm.
-  \item \textbf{State-density} According to \cite{suttonPolicyGradientMethods}, $d^{\pi}(s) = \lim_{t -> \infty} p(s_t = s | s_0, \pi)$ describes how likely it is to be in state s when following a policy $\pi$ forever.
-If we for example have a policy which plays go as well as the authors of this paper, states with a lot of enemy stones and very few own stones might be very likely, and a state without enemy stones at all would be very unlikely.
-Also according to \cite{suttonPolicyGradientMethods} this distribution is independent of the starting state $s_0$ and only dependent on the policy and of course the environment.
-In later papers, commonly a discounted version of this is used.
-$\rho^{\pi}(s)$ weights future probabilities less and thus can be described as the probability of being in state s \textit{soon} when following policy $\pi$.
-We will omit a precise definition here, you can look it up at \cite[Section 2.1]{silverDeterministicPolicyGradient2013}.
+We sometimes call a learnt value estimation \textit{critic}.
+  \item \textbf{Actor-Critic} Actor is just another name for a learnt policy and critic just another name for an estimated Q-Value or Value function.
+These two frequently appear together, if that happens we are using an actor-critic type algorithm.
+
 \end{itemize}
 
-Now, with all these definitions at hand, we can specify our mission a bit better. We want to maximize
+Now, with all these definitions at hand, we can specify our mission a bit better. We want to find a policy $\pi = \mathrm{argmax}_{\pi} J$ which maximizes expected value
 \begin{equation}
@@
 In other words, we want a very high probability of being in a state ($\rho^{\pi}$) which might give good rewards, and in that state we want a high probability of taking an action ($\pi$) which might give a good reward.
-In our wind turbine, rewards could for example be the all the costs and incomes summed together, so at every point in time we could count income by energy yield or costs because of maintenance (with our controller more likely costs for rebuilding the entire turbine).
+In our wind turbine, rewards could for example be the all the costs and incomes summed together, so at every point in time we could count income by energy yield or costs because of maintenance (with not so optimal policies costs for rebuilding the entire turbine).
 This work would be easy if we could directly calculate this equation but clearly, there is a bit of a problem ahead.
-Firstly, we don't know $\rho$ as we don't exactly know the environment we are modelling.
+Firstly, we don't know $\rho$ as we don't exactly know the environment we are modeling.
 We can observe some rewards, states and actions, but only very few of them.
-We might then see that using 0.3MN of torque at rotational speed 11rpm yields good energy yield, but how about rotational speed 11.1rpm?
+We might then see that using 0.3MN of torque at rotational speed 0.8 rad/s yields good energy yield, but how about rotational speed 0.9 rad/s?
 We might have never seen that.
-Also, rotating the turbine at 15rpm might be a good strategy to immediately achieve high power outputs, but we might need to fix our turbine after running this policy for just a few days and incur a heavy penalty.
+Also, rotating the turbine at 2 rad/s might be a good strategy to immediately achieve high power outputs, but we might need to fix our turbine after running this policy for just a few days and incur a heavy penalty.
 So we don't know how rewards will spread out in the future.
@@
 We could use them somehow, so let's discuss how to learn them.
+If you already know about Q-Learning, you can skip the next section.
 
@@
 We could create a lookup table for all s and a, initialize it all zero and whenever observing a state transition, we could update the table at that state and action after the above notation.
+This iterative Q update is proven to converge to the optimal solution.
 
@@
 Thus we just derived a method to train a Q function under the assumption of an optimal policy.
-In \cite{mnihPlayingAtariDeep2013}, the authors assumed a finite state space and could use the greedy policy $argmax_a Q(s', a)$ by searching the entire action space.
+In \cite{mnihPlayingAtariDeep2013}, the authors worked on a finite action space and could use the greedy policy $argmax_a Q(s', a)$, coming reasonably close to an optimal policy.
+They call their algorithm \ac{DQN}, and with some improvements which we omitted here, they are able to play atari games above human level.
 We are presented with an infinite action space though, which is why we can't implement a greedy policy without high computational expense.
@@
 
-We will at first give an intuitive explanation of a basic algorithm which forms the basis for many modern reinforcement learning algorithms, including the one we chose.
-The algorithm is called \ac{REINFORCE} (no joke) and was originally presented by \cite{williamsSimpleStatisticalGradientfollowing}.
+In this section, we will at first give an intuitive explanation of a basic algorithm which forms the basis for many modern reinforcement learning algorithms, including the one we chose.
+The algorithm is called \ac{REINFORCE} (yes that is the name) and was originally presented by \cite{williamsSimpleStatisticalGradientfollowing}.
 Then we will add a Q term to it according to \cite{suttonPolicyGradientMethods} and call it \ac{PG}.
 \cite{degrisOffPolicyActorCritic2013} generalized that term and made an off-policy version of policy gradients, which we will not discuss in detail.
-That version was then used by \cite{silverDeterministicPolicyGradient2013} and the stochastic policy was replaced by a deterministic one, the algorithm was called \ac{DPG}.
-This algorithm finally lays the basis for \ac{DDPG} by \cite{lillicrapCONTINUOUSCONTROLDEEP2015}, which is what we used to start our experimentation.
-
-Let's assume we have a probablistic policy $\pi: S \rightarrow P(A)$ which is parametrized by $\theta$.
+That version was then used by \cite{silverDeterministicPolicyGradient2013} and the stochastic policy was replaced by a deterministic one, the algorithm was called \ac{DPG} and we explain it in section \ref{sec:dpg}.
+This algorithm finally lays the basis for \ac{DDPG} by \cite{lillicrapCONTINUOUSCONTROLDEEP2015}, which is what we used in our work and which we explain in \ref{sec:ddpg}.
+
+Let's assume we have a probabilistic policy $\pi: S \rightarrow P(A)$ which is parametrized by $\theta$.
 An intuitive way of improving this policy could be to increase the gradient proportional to the reward an action yielded:
@@
 \end{equation}
-We could do this for all states we see and would end up with an iterative way of improving our policy.
-In fact, this is similar to the \ac{REINFORCE} algorithm (if we add our importance sampling trick from below).
+We could do this every time we see a state transition and we would end up with an iterative way of improving our policy.
+In fact, this is the basic idea of the \ac{REINFORCE} algorithm, with some additions.
 However, we might not know rewards for any action, state or have noisy rewards.
-Luckily, we derived Q-Learning in the section before and thus have an estimation which action will give which return in this step.
+We derived Q-Learning in the section before and it would be nice to use the advantage of being able to look into the future for our policies.
+We can simply replace r with Q, thus have an estimation which action will give which return in this step.
 So, with our newly gained Q, for all actions and states, do
 \begin{equation}
-\theta_{i+1} = \theta_i + \alpha Q(s, a)\Delta \pi_{\theta_i}(a|s)
+\theta_{i+1} = \theta_i + \alpha Q^*(s, a)\Delta \pi_{\theta_i}(a|s)
 \end{equation}
@@
 Problematic is though, that the Q term we used, always took the best possible actions, which could be vastly different from the actions under our policy.
+So for the optimal policy, jumping down a bridge is the best possible strategy because it also knows how to swim, but we might not know how to swim yet and should better not jump down a bridge.
 Thus, we need to reformulate Q to account for our policy instead of the optimal policy
@@
 Now, in theory, we can not regard Q as constant with respect to $\theta$ anymore, because it is dependent on the policy.
-\cite{suttonPolicyGradientMethods} claim though that it can be ommitted and still regarded as constant.
+\cite{suttonPolicyGradientMethods} claim though that it can be omitted and still regarded as constant.
 However there is still a problem if we integrate this new Q into our policy updates.
 We assume that we use the same $\pi$ for exploration while doing training, so this policy is responsible for taking actions in the environment and also responsible for which experiences we see during our training.
-So let's assume the policy is poorly initialized at the beginning of our training and gives action 1 a probability 4 times as high as action 2, though taking action 2 would yield a 10\% better return in our Q function.
-Now, as we are taking action 1 4 times as often as action 2, we will also perform gradient updates on action 1 4 times as often in our Monte-Carlo expectation draw.
-Though each of the gradient update steps is smaller due to the lower Q factor, the higher frequency of updating action 1 will lead to action 1 getting an advantage over action 2 in this update strategy.
+So let's assume the policy is poorly initialized at the beginning of our training and gives action jump a probability many times as high as action walk on, though taking action walk would yield a better return in our (already perfectly trained) Q function.
+Now, as we are taking action jump many times more often than action walk on, we will also perform gradient updates on action jump more often in our Monte-Carlo draw.
+Though with each update step, we only get a tiny value in comparison to the value of the better action, the higher frequency of updating that action will compensate for it and summing all the tiny values together leads to jumping getting an advantage over walking in this update strategy.
 To correct for this oversampling bias, we could employ a trick which is called importance sampling.
@@
 \theta_{i+1} = \theta_i + \alpha Q^{\pi}(s, a) \frac{ \Delta \pi_{\theta_i}(a|s)}{\pi_{\theta_i}(a|s)}
+\label{equ:pg_importancesampling}
 \end{equation}
@@
 \end{equation}
-This equation is very famous in reinforcement learning and is called policy gradients.
+This equation is very famous in reinforcement learning and is called \textit{policy gradient}.
 And by coincidence, it is equivalent to what we intuitively derived in equation \ref{equ:pg_update}.
@@
 In fact, we were already able to use a replay buffer in the Q learning part.
-If we wanted to do that on our policy however, we would need to correct for the bias of the other policy aswell with importance sampling.
+If we wanted to do that on our policy however, we would need to correct for the bias of the other policy as well with importance sampling.
 However we will not derive stochastic off-policy updates, because using the \ac{DPG} algorithm, this problem is solved.
@@
 We needed this expectation term, because we were dealing with stochastic policies.
-If instead learning deterministic policies, we can exclude policy expectation from the equation.
+If instead learning deterministic policies, we can exclude the policy expectation from the equation.
 But before going through all the advantages of deterministic policies, let's first explain it.
@@
 \subsection{Deterministic Policy Gradients}
+\label{sec:dpg}
 
 Before \cite{silverDeterministicPolicyGradient2013}, stochastic policies were preferred because the stochastic nature aided exploration and because there was simply no theory on how to derive deterministic policy gradients.
-The idea behind deterministic policy gradients is to replace the stochastic policy $\pi: S -> P(A)$ with a deterministic one $\mu: S -> A$.
-Also, we are switching from small finite action spaces to continous action spaces.
+The idea behind deterministic policy gradients is to replace the stochastic policy $\pi: S \rightarrow P(A)$ with a deterministic one $\mu: S \rightarrow A$.
+Also, we are switching from small finite action spaces to continuous action spaces.
 Now, updating this policy according to the equations above does not work anymore due to two problems.
 At first and most significantly, above we needed all $a$ or at least some form of maximization to calculate the full gradient of $\pi(a|s)$.
-This is difficult with a continous action space now.
+This is difficult with a continuous action space now.
 Furthermore our importance sampling trick, our correction for overestimation bias based on the bias of the policy, does not work anymore.
@@
 This is an advantage, because computing only one expectation requires less samples and we do not have to worry for any bias from the policy in Q-updates.
-When we do our Monte-Carlo experiments to update it, we can take our generated samples from the environment and assume they are modelling the underlying environment.
+When we do our Monte-Carlo experiments to update it, we can take our generated samples from the environment and assume they are modeling the underlying environment.
 
@@
 If we derive this surface along the action dimension, we get a gradient which points in the direction of a better action for this state.
-We could direcly follow this gradient and calculate our policy update from it:
+We could directly follow this gradient and calculate our policy update from it:
 
@@
 They proved, that a linear function approximator $Q^{\omega}$ which minimizes the MSE between $Q^{\omega}$ and $Q^{\mu}$ will retain the gradient, even when trained on samples generated by $\beta$, and called this Q \textit{compatible}.
-However, in practice we are not minimizing that loss and neither are we using linear function approximators.
+However, in practice we are not explicitly minimizing that loss and neither are we using linear function approximators but usually nonlinear multi-layer perceptrons.
 \cite{bhatnagarConvergentTemporalDifferenceLearning2009} proved that using a non-linear function approximator to predict the value function $V(s)$ will converge towards a local optimum.
@@
 In our own experiments, we found that using a wildly different policy for exploration than for training did in fact result in worse behavior, and \cite{fujimotoOffPolicyDeepReinforcement2019} observed the same behavior.
-Although there is some evidence against this being possible, we will from now on just assume that DDPG is trainable off-policy because it delivered good results in the past and because the paper stating it is possible has been cited more than 100 times more often than the one saying it's not.
-Science!
+They proved DDPG to not function truly off-policy due to a phenomenon which they call extrapolation error.
+This phenomenon describes inaccuracies due to poor generalization in the Q function on unseen values.
+When a target policy predicts such an unseen action, the Q function is likely to overestimate that action.
+Although there is some evidence against training \ac{DDPG} off-policy in general, we will still do so because we are not training that far off-policy.
+We utilize samples from our own, slowly changing policy in the last 100k steps.
+Also, empirically it delivered good results in the past.
 
@@
 
+$Q^{\omega}$ being a nonlinear function approximator trained under $\beta$.
+
 \subsection{DDPG}
+\label{sec:ddpg}
 
@@
 
-\ac{DDPG} is a combination of \ac{DPG} and \ac{DQN}.
-\cite{lillicrapCONTINUOUSCONTROLDEEP2015} tackle several problems in the two algorithms to combine them.
-At first, traditional Q learning updates the Q network parameters based on a target $y$ calculated on the Q network parameters themselves.
-This introduces instability and to tackle this, \cite{mnihPlayingAtariDeep2013} introduced a second Q network $Q'$ which they call target network, for which the parameters are held constant over a period of updates.
-They calculate the $y$ based on the second network $Q'$.
+\ac{DDPG}, first presented by \cite{lillicrapCONTINUOUSCONTROLDEEP2015}, is a combination of \ac{DPG} and \ac{DQN}.
+The authors tackle several problems in the two algorithms to combine them.
+At first, in traditional Q learning, updates to the Q network parameters are based on a target $y$ calculated on the Q network parameters themselves $\text{loss}(Q) = (f(Q)-Q)^2$.
+This introduces instability and to tackle this, \cite{mnihPlayingAtariDeep2013} introduced a second Q network $Q'$, for which the parameters are held constant over a period of updates.
+They calculate the learning target $y$ based on the second network $Q'$, which is why they call $Q'$ the \textit{target network}.
 After a number of steps, the parameters from the actual Q network are then copied over to the target.
-The stability gain in their paper depends greatly on the number of steps after which to perform an update.
+The stability gain in their algorithm depends greatly on the number of steps after which to perform an update.
 In \ac{DDPG}, they instead use smooth target updates, where every step, the parameters of the actual network fade over to the target: $\theta_{target} = \tau \theta + (1 - \tau) \theta_{target}$.
-The fade-over parameter $\tau$ was recommended to be set to a small number like 0.001.
+The fade-over parameter $\tau$ was recommended to be set to a small number like 0.01.
 Additionally to the target Q-network, they also found that a target policy $\mu'$ improves stability, and they update it the same way as the main policy.
@@
 \begin{equation}
-L(\theta^Q) = \mathbb{E}_{s \sim \rho^{\beta}, a \sim \beta, s' \sim \varepsilon}[(Q(s, a;\theta^Q) - y)^2]
+L(\theta^Q) = \mathbb{E}_{s \sim \rho^{\beta}, a \sim \beta, s' \sim \varepsilon}[(y - Q(s, a;\theta^Q))^2]
 \end{equation}
@@
 Also, this Q is approximated by a non-linear function approximator instead of the linear one from compatible \ac{DPG}.
-\cite{lillicrapCONTINUOUSCONTROLDEEP2015} do not provide a proof of convergence for this new, mixed Q update.
-Still, we have an off-policy way of improving the critic now.
-The actor is directly copied from \ac{DPG} even though it does not fulfil the \textit{compatible} property.
+\cite{lillicrapCONTINUOUSCONTROLDEEP2015} do not provide a proof of convergence for this new, mixed Q update, but instead show empiric proof that it works.
+So, we have an off-policy way of improving the critic now.
 
@@
 
-And the Q update:
-\begin{equation}
-\Delta_{\theta^Q} L(\theta^Q) = \mathbb{E}_{s \sim \rho^{\beta}, a \sim \beta, s' \sim \varepsilon}[(Q(s, a;\theta^Q) - y)\Delta_{\theta^Q}Q(s, a;\theta^Q)]
+Intuitively, this new policy gradient equation follows the curvature of the Q function directly.
+Policy gradients point in the direction where the Q function sees the highest values.
+Our new Q update looks like
+
+\begin{equation}
+\Delta_{\theta^Q} L(\theta^Q) = \mathbb{E}_{s \sim \rho^{\beta}, a \sim \beta, s' \sim \varepsilon}[(y - Q(s, a;\theta^Q))\Delta_{\theta^Q}Q(s, a;\theta^Q)]
 \end{equation}
@@
 Similar to other reinforcement learning algorithms, we will again not compute the full gradient but use stochastic gradient ascend with batch learning.
-To rephrase gradient ascend to gradient descend, they multiply the policy loss with $-1$.
+To rephrase gradient ascend to gradient descend, which is commonly implemented in machine learning frameworks, they multiply the equation with $-1$.
 In fact, in the original \ac{DDPG} algorithm, they used Adam \cite{kingmaAdamMethodStochastic2017} optimization instead of stochastic gradient descend.
@@
 Their network architecture for the actor consisted of two fully connected layers with 400 and 300 units and instead of relu they used tanh activations in the last layer to bound actions.
-For the critic, they had a version which learnt upon pixel outputs and used convolutional layers, while we only look at the low-dimensional version.
-For this, they also used a two layer variant with 400 and 300 units.
+They had a version which learnt upon pixel outputs and used convolutional layers, while we only look at the low-dimensional version.
+For the critic, they also used a two layer variant with 400 and 300 units.
 State inputs were fed through the whole network, while actions were concatenated in onto the second layer, skipping the first.
-
-Furthermore, to aid exploration, they added noise in form of a Ornstein-Uhlenbeck process \cite{uhlenbeckTheoryBrownianMotion1930}.
-
-For readability, we added a copy of their algorithm in Algorithm \ref{alg:vanilladdpg}, with minor adjustments to fit our notation.
+They used batch normalization between the layers and L2 regularization on the weights.
+
+Furthermore, to aid exploration, they added noise in form of a \ac{OU-Noise} \cite{uhlenbeckTheoryBrownianMotion1930}.
+This noise is a correlated noise which is sampled over time.
+Instead of the uncorrelated version, this correlated noise can step further away from the noise center (parametrized by $\mu$ in their paper).
+It consists of a mixture of going towards this center ($\theta$) and going into a direction drawn from a normal distribution ($\sigma$).
+As we use the same symbols for other things in our work, we will not adapt their notation.
+
+For readability, we added a copy of the entire \ac{DDPG} algorithm in Algorithm \ref{alg:vanilladdpg}, with minor adjustments to fit our notation.
 
@@
 
-\For{episode = 1, M}{
+\For{epoch = 1, M}{
   Initialize a random process $\mathcal{N}$ for action exploration\\
@@
 \section{RL on windturbines}
-We are aware of one paper \cite{kolterDesignAnalysisLearning2012} which tried RL on windturbines already, we want to dedicate a section to it.
+We are aware of one high-quality paper \cite{kolterDesignAnalysisLearning2012} which tried RL on windturbines already, we want to dedicate a section to it.
+This section is not necessary for understanding the rest of the work, and we have no references to it outside of this section.
+If you want to understand why we chose a different approach, you can still read this section.
 The team behind that paper built a miniature model of a windturbine and used a variation of the \ac{REINFORCE} algorithm to control it.
@@
 
-As training input, they used a single observation in contrast to popular reinforcment learning which operates on high-dimensional observation space.
+As training input, they used a single observation in contrast to popular reinforcement learning which operates on high-dimensional observation space.
 This observation was power output of the turbine, and they treated it as reward to the system, not having any state observations.
@@
 They ran these parameters for 2 seconds on the turbine, averaged power output over that time into a reward and then ran an iteration of their algorithm to generate new parameters, starting with random parameters.
-They first tested the \ac{REINFORCE} algorithm, with operates without a Q network and trains a policy directly on rewards.
+They first tested the \ac{REINFORCE} algorithm, which operates without a Q network and trains a policy directly on rewards.
 This algorithm was discussed above and utilizes the following gradient equation:
@@
 
-Note, how they denoted J in their paper as rewards, while we are using r for rewards and J for objective functions.
+When reading their work, you will find that they denoted J in their paper as rewards, while we are using r for rewards and J for objective functions.
 Also, they used $\omega$ as a symbol for the action taken, while we use $a$.
 Also note how their policy is not dependent on a state but directly predicts actions.
-They did not explicitely note what they formulated their expectation term on, but we added it here for clarity.
+They did not explicitly note what they formulated their expectation term on, but we added it here for clarity.
 Note the similarity to what we discussed in equation \ref{equ:reinforce} and \ref{equ:pg}.
-In addition to this, they came up with an algorithm which they call Trust Region Policy Search, and which is astonishingly different from Trust Region Policy Optimization.
+In addition to this, they came up with an algorithm which they call Trust Region Policy Search, and which is astonishingly different from Trust Region Policy Optimization \cite{schulmanTrustRegionPolicy2017}.
 With a few additions, they incorporated off-policy learning and second order error approximation into their algorithm, and showed later that it performs nicely.
-They were able to find optimal pitch and load settings as soon as 15 timesteps after the start of the simulation.
+They were able to find optimal pitch and load settings as soon as 15 time-steps after the start of the simulation.
 
@@
 In general, their turbine behaved more concavely opposed to our turbine, which with a constant parameter set, would settle to either rotating quickly backwards, forwards or not at all.
-Because of this, they were able to give the turbine some time at the beginning to stabilize before applying reinforcement learning while we needed to start with the standing turbine.
+Because of this, they were able to give the turbine some time at the beginning to stabilize before applying reinforcement learning while we needed to start with an artificially stabilized turbine.
 Lastly and most importantly, they aimed to find optimal control settings for a constant wind speed.
@@
 We want to create a version as general as possible.
-Also, as \ac{DDPG} is a direct successor on \ac{REINFORCE} and has proven to outperform it in many areas, we do not reimplement their paper.
+Also, as \ac{DDPG} is a direct successor to \ac{REINFORCE} it has proven to outperform it in many areas.
 
@@
 This work uses traditional supervised learning to fit a neural network of 10 neurons to the straight line that describes the relation between yaw angle and power output and because what they got looks similar to a Q function they call it reinforcement learning.
-\cite{fernandez-gaunaVariableSpeedWind2016} published in a proprietary journal.
+\cite{fernandez-gaunaVariableSpeedWind2016} published in a proprietary journal, so we didn't have access to their work.
 
@@
 We managed to find a solution with our DDPG implementation.
-We hoped that the implementation might yield results on our windturbine environment as well and exchanged the OpenAI gym for qblade.
-
-The first runs looked marvellous, perfectly holding rated speed, until we found out that those perfect values were prescribed by a setting in the qblade project file, and disabling this setting marked the beginning of our actual experimentation phase.
-
-\section{Starting with qblade}
-
-On our first experimentation steps, we did not yet have experience with how the qblade simulation behaves with a random or not well designed controller.
+We hoped that the implementation might yield results on our windturbine environment as well and exchanged the OpenAI gym for QBlade.
+
+The first runs looked marvelous, perfectly holding rated speed, until we found out that those perfect values were prescribed by a setting in the QBlade project file, and disabling this setting marked the beginning of our actual experimentation phase.
+
+\section{Starting with QBlade}
+
+On our first experimentation steps, we did not yet have experience with how the QBlade simulation behaves with a random or not well designed controller.
 Our observation was, that the simulation was resting between two different states, which we want to call \ac{FMM} and \ac{BMM}.
@@
 \ac{BMM} is characterized by a negative rotational speed, up to -3.8 rad/s.
-This can be reached by setting a higher generator torque than the rotor torque which is produced by aerodynamic lift on the blades.
+This can be reached by setting a higher generator torque than the aerodynamic torque induced by wind flowing through the turbine and producing lift on the blades.
 Also in this state, extreme vibrations and sometimes blades flying off can be observed, and like \ac{FMM} this state is stable until a blade falls off.
 
-In neither of these modes, the simulation is within realistic bounds and the observations gotten from the simulation at this part are not credible.
+In neither of these modes, the simulation is within realistic bounds and the observations received from the simulation at this part are not credible.
 After we observed that the controller itself wasn't able to avoid these states on its own, we inhibited reaching these states by outputting a death condition when being in negative rotational areas or high positive.
@@
 \section{Designing reward functions}
-\label{rewardfunctions}
-
-As qblade doesn't provide a built-in reward function, we needed to craft our own reward function based on the observed state.
+\label{sec:rewardfunctions}
+
+As QBlade doesn't provide a built-in reward function, we needed to craft our own reward function based on the observed state.
 We considered several possible variants.
-We will first introduce all of them and give a more detailed explanation later
+We will first introduce all of them and give a more detailed explanation later.\\
 
@@
 
-\item \textbf{Maximal power} Generate the maximum power possible. Observe current power \(s_{pow}\) and with given normalization constant \(c\) calculate \(r(s) = s_{pow} * c\)
+  \item \textbf{Maximal power} Generate the maximum power possible. Observe current power \(s_{pow}\) and with given normalization constant \(c\) calculate \(r(s) = s_{pow} c\)
 
   \item \textbf{Penalize current stress} Penalize high blade bending.
-Observe current bending \(s_{bend}\) and with given normalization constant \(c\) and another reward function \(r(s)\) calculate new reward \(r'(s) = r(s) - c * s_{bend} \)
+Observe current bending \(s_{bend}\) and with given normalization constant \(c\) and another reward function \(r(s)\) calculate new reward \(r'(s) = r(s) - s_{bend}c \)
 
   \item \textbf{Penalize accumulated stress} Penalize accumulated structural stress.
-With \(x = rainflow(s_0...s_t\) being the rainflow table and \(p(x)\) being a function that maps from rainflow outputs to a scalar penalty, calculate \(r'(s_0...s_t) = r(s_t) - p(rainflow(s_0...s_t))\)
+With \(x = rainflow(s_0...s_t)\) being the rainflow table and \(p(x)\) being a function that maps from rainflow outputs to a scalar penalty, calculate \(r'(s_0...s_t) = r(s_t) - p(rainflow(s_0...s_t))\)
 
   \item \textbf{Penalize action gradients} Penalize high action gradients.
-With \(a_t\) being the current action, \(a_{t-1}\) the last action, \(c\) a penalty constant and \(r(s)\) another reward function, calculate: \(r'(s, a) = r(s) - c * (a_t - a_{t-1})\)
+With \(a_t\) being the current action, \(a_{t-1}\) the last action, \(c\) a penalty constant and \(r(s)\) another reward function, calculate: \(r'(s, a) = r(s) - c  (a_t - a_{t-1})\)
 
   \item \textbf{Penalize death conditions} Penalize when a death condition is reached.
-With $r(s)$ being another reward function, $d(s) \in {0,1}$ being our indicator for deaths and $c$ a constant penalty, calculate $r'(s) = r(s) - c * d(s)$
+With $r(s)$ being another reward function, $d(s) \in {0,1}$ being our indicator for deaths and $c$ a constant penalty, calculate $r'(s) = r(s) - c  d(s)$
 
@@
 We suspect this penalty to be hard to learn, as it reacts slowly to changes and a once given penalty will never be lifted again.
+Alternatively, we could imagine a moving-average like solution, in which rainflow fatigue is calculated on a small trajectory of past states and not on the entirety of the simulation.
 
 \textbf{Penalize action gradients} is an option which is useful in contexts where high action gradients are impossible on the real system.
+High action gradients mean for example turning a blade 90 degrees in a single timestep, which would rupture a normal blade.
 Instead of clipping them, penalizing them could include action gradients into the learning process.
@@
 Especially because we are learning a continuous value estimation, it will most likely smooth out along a death and likely underestimate the death penalty and underestimate values close to the death.
-Alternatively we could imagine penalizing getting close to a death condition continously.
+Alternatively we could imagine penalizing getting close to a death condition continuously.
 
@@
 
-We suspected the first problem to happen, as neither of the two modes had good rewards, and going out of these modes continously improves rewards.
+We suspected the first problem to happen, as neither of the two modes had good rewards, and going out of these modes continuously improves rewards.
 To feed a network with new values closer to an optimum, a common practice in reinforcement learning is to add noise.
@@
 
-We first tried a gaussian noise instead of the recommended \ac{OU-Noise} \cite{uhlenbeckTheoryBrownianMotion1930} from \ac{DDPG}: \(n \leftarrow \mathcal{N}(\mu, \sigma^2) \).
+We first tried a Gaussian noise instead of the recommended \ac{OU-Noise} \cite{uhlenbeckTheoryBrownianMotion1930} from \ac{DDPG}: \(n \leftarrow \mathcal{N}(\mu, \sigma^2) \).
 We didn't see any improvements over our old behavior, so we switched to \ac{OU-Noise}.
@@
 Our decay is parameterized by a start sigma, end sigma and end step and then held constant at the end sigma.
+We are unsure whether the decay actually benefited our training, but we kept it in nonetheless as it also didn't seem to negatively impact it.
+We started it 0.1 above and let it decay to 0.1 below the values from the \ac{DDPG} paper ($\sigma=0.2, \theta=0.15$) within the first 50k training steps.
 
@@
 The TD3-paper \cite{fujimotoAddressingFunctionApproximation2018} utilizes a random exploration phase, in which a completely random policy \cite{mozerDiscoveringStructureReactive1990} initially explores the environment, before the actual agent switches in.
-We implemented the random walk at first through gaussian noise and then through an Ornstein-Uhlenbeck process.
+We implemented the random walk at first through Gaussian noise and then through an Ornstein-Uhlenbeck process.
 The later version creates less vibration through less excessive changes in control parameters.
 As we found that both random policies rarely explore a sensible operational range, we later added an expert PID controller, which implements torque control, and combined that with our Ornstein-Uhlenbeck process.
-We set the parameters to that controller by experimentation and came up with the following equation, only dependent on rotational speed $\mathrm{rot}$: $\mathcal{C}:(\mathrm{rot})\rightarrow(\mathrm{torque}, \mathrm{pitch}), \mathcal{C}(\mathrm{rot}) = (\num{2e7}*\mathrm{rot}-\num{12e6}, 128*\mathrm{rot}-103)$.
+We set the parameters to that controller by experimentation and came up with the following equation, only dependent on rotational speed $s_{\mathrm{rot}}$: $\mathcal{C}:(s_{\mathrm{rot}})\rightarrow(a_{\mathrm{torque}}, a_{\mathrm{pitch}}), \mathcal{C}(s_{\mathrm{rot}}) = (\num{2e7} s_{\mathrm{rot}}-\num{12e6}, 128s_{\mathrm{rot}}-103)$.
 These actions were clipped as in section \ref{sec:clip_act_grad}.
-So effectively, we designed a P controller, and only using this, we could already achieve relatively stable control.
+So, effectively, we designed a P controller, and only using this, we could already achieve relatively stable control.
 As further work, it could be imagined to use an established controller here, but we did not need this for our purposes.
@@
 
-\For{episode = 1, E}{
+\For{epoch = 1, E}{
   Initialize expert controller $\mathcal{C}$ and random process $\mathcal{N}$ for expert exploration\\
@@
 Together with the pretraining algorithm, we had some slight improvements, but these didn't justify the extra computation time needed for this.
-Also, though an expert controller which yields somewhat sensible result is easy to build for each turbine, this is a step of human interaction and expert knowledge, which machine learning generally tries to minimize in algorithm design.
+Also, though an expert controller which yields somewhat sensible results is easy to build for each turbine, this is a step of human interaction and expert knowledge, which machine learning generally tries to minimize in algorithm design.
 
@@
 The authors of \ac{DDPG} trained their networks up to 2.5 million steps, which we can't afford with our computational power, so a simpler network architecture will hopefully also converge faster.
+However we loose generalization potential with the smaller network.
+We will not be able to solve the same complexity of tasks as \ac{DDPG} with our smaller networks.
+We estimate windturbine control to be a complex problem though.
+So effectively, we had to look for other solutions.
 
@@
 As in our simulation, where our state reflects measurements from very different parts of the simulation, some values in the state-array were 8 orders of magnitude different from others as they were measured in very different units.
-As most of the high-magnitude values are vibrations, the net could not create a link between these and the state-action, while low-magnitude but insightful observations like rotational speed were not considered due to their small absolute values.
+As most of the high-magnitude values are vibrations, the net could not create a link between these and the state-action, while low-magnitude but insightful observations like rotational speed were likely not considered due to their small absolute values.
 
 We normalized the data so states, actions and reward usually stayed between 1 and -1.
-This is a normal technique in machine learning, and we are irritated by the fact that the reinforcement learning community largely made no mention of normalization.
+This is a normal technique in most of machine learning, but for some reason, in reinforcement learning this is not so common.
 At times, rewards were clipped \cite{mnihPlayingAtariDeep2013} and observations were sometimes, especially when working on pixel inputs, cropped to be of square dimensions.
@@
 They propose an adaptive normalization mechanism which integrates into a generic \ac{RL} algorithm and which can normalize all inputs adaptively.
-We decided to implement a more straight-forward variant.
+Because of the extra implementation effort however, we decided to implement a more straight-forward variant.
 At first, we used observed states and rewards from the random exploration phase to calculate the 5\%, 95\% quartile values.
@@
 As sometimes during random exploration, no extreme values were observed and later extreme values lay far outside of $[-1, 1]$, we instead normalized on replay data of a previous run.
-Additionally, we added a constant of 3 after normalization because of the phenomenon described in section \ref{sec:death}.
-We could also imagine manually setting normalization bounds on low-dimensional problems, as human knowledge about the possible state space is usually present.
+Additionally, we added a constant of 3 after normalization because of the phenomenon described in section \ref{sec:death}, but only in later evaluations.
+We reference to normalization by $\text{norm}(s)$, $\text{norm}(a)$ and $\text{norm}(r)$
+
+Alternatively to calculating normalization bounds, it would be possible to set them with human expert knowledge for low-dimensional problems, as human knowledge about the possible state space is usually present.
 In case when human knowledge about input magnitudes is not present, we recommend the mechanism from \cite{vanhasseltLearningValuesMany2016}, or if you are willing to accept the extra computational effort of running once without normalization, our version.
 
-This finally fixed our problem of zero gradients and we could observe normal learning. We recommend this technique as a default technique for every reinforcement learning algorithm.
+Normalizing our data drastically helped with our problem of zero gradients and we could observe normal learning.
+We took the penalty of requiring data from a previous run for our algorithm, but in simulation environments it is possible to acquire this data easily.
+For a real world environment, it is less easy and our method would not be applicable to these scenarios.
+We however recommend using any normalization when dealing with different magnitude sensor data.
+
+\subsection{Last-layer actor activations}
+\label{sec:activations}
+
+In the \ac{DDPG} architecture, the last layer of the actor is activated with a tanh activation.
+Like sigmoid, this activation returns close to 1 or -1 for high or low activations.
+These activations commonly produce vanishing gradient problems when their activations are driven close to this range.
+A big change in input at these borders results in a very small change in outputs, for example $\text{tanh}(2)=0.9640$ and $\text{tanh}(4)=0.9993$.
+We duplicated the input, and the output barely changed.
+Thus, also the gradients in this activation range are minimal.
+In our windturbine control scenario, maximum actions however are beneficial in many scenarios, so we expect actions of 1 and -1 quite often.
+This trains our net towards the flat outside areas of the tanh activation, and increases our vanishing gradient problem.
+We solve this by multiplying a constant factor to our activations, moving 1 and -1 further to the inside of the activation.
+If for example we chose 1.5 as a factor, our last layer activation was $a = 1.5tanh(x)$ where x is the weighted and biased output from the last layer.
+This allows our actor to choose actions slightly outside of the action range, which has a disadvantage: Q will not see any actions outside of $[-1, 1]$ during Q training, as the training samples to Q are clipped.
+Thus, Q has to provide estimates for values outside of the area it has seen.
+As the authors in \cite{fujimotoOffPolicyDeepReinforcement2019} have described, Q tends to overestimate values it has not seen.
+In fact, when implementing it as described, we did observe the policy to very likely predict actions outside of the action space.
+We did not manage to fix this problem in the scope of this work, but we could imagine a follow up work where we change the policy gradient equation to account for this overestimation.
+Alternatively, we could imagine as well to train the Q network on actions slightly outside of the action space with negative estimates to prevent this effect from happening.
+
+In the end, we could not find a definite solution to our vanishing gradient problem and thus had to stick to the smaller architecture, loosing the ability to learn more complex tasks.
 
@@
 
-We observed that our controller delivers high action gradients and reacts strongly between timesteps, sometimes jumping from no pitch/torque all the way to the maximum.
+We observed that our controller delivers high action gradients and reacts strongly between time-steps, sometimes jumping from no pitch/torque all the way to the maximum.
+This happened especially in the beginning of a training run, as the relations between input and output were still very random and sudden spikes not yet trained away.
 This creates vibrations in the structure and especially high blade pitch changes lead to blades breaking or falling off.
+Even worse, QBlade was not able to handle this type of structural failure and completely crashed in these scenarios.
 We tried different methods to circumvent this.
@@
 \subsection{Gradient actionspace}
+\label{sec:grad_actionspace}
 
@@
 For aiding training, we added the actual actions taken to the observation space of the next step.
-This way, we could clip the action gradients by reducing the gradient action space, so that in each timestep, only a fraction of the actual action space could be traversed.
+This way, we could clip the action gradients by reducing the gradient action space, so that in each time-step, only a fraction of the actual action space could be traversed.
 If the controller would output a positive change at maximum pitch/torque or a negative change at minimum pitch, this would be ignored.
@@
 
-\subsection{Feeding past timesteps}
+\subsection{Feeding past time-steps}
 \label{sec:feedpast}
 
-Another idea was that we feed the last n timesteps concatenated to the current observations.
+Another idea was to feed the last n time-steps concatenated to the current observations.
 We hoped that this way, the agent could derive its own gradients internally.
-Also, a PID controller has a view of the past, with this extra information, the countroller would be able to generalize further.
-The result of this however were high and very noisy q-losses and q loss explosions, so we disabled this again.
+Also, a PID controller has a view of the past.
+With this extra information, the controller could be able to generalize further and perform better especially counteracting resonant vibrations.
+The result of this however were high and very noisy Q-losses and Q loss explosions, so we disabled this again.
 
@@
 We eventually solved the problem with the high action gradients by still letting the controller output absolute actions across the entire action spaces, but if gradients exceed a certain threshold, the actual action taken is set to the nearest value with sensible gradients.
-We defined our sensible gradient vector $g = (\num{3e4}, 0.1)$ for pitch and torque.
-Clipping action $a$ and previos action $a'$ can be defined as $\mathrm{clip}(a_i, a'_i) = \min(\max(a_i, a'_i -g_i), a'_i + g_i)$.
+We defined the sensible gradient vector $g$ according to \cite[Table 7-2]{jonkmanDefinition5MWReference2009}.
+The clipping function on action $a$ and previous action $a'$ can be defined as $\mathrm{clip}_a(a, a') = \min(\max(a, a' - g), a' + g)$.
 The results of this are again clipped to be in the action space.
-Though the actions returned from the policy are still wildly unstable, they at least didn't result in death of the turbine anymore and still we didn't hinder convergence through our gradient actionspace.
-This resulted in our first ever controller keeping the turbine within sensible operational range for at least a part of the training, though diverging quickly after.
+Though the actions returned from the policy are still wildly unstable, they at least didn't result in death of the turbine anymore and still we didn't hinder convergence through gradient actionspace such as in section \ref{sec:grad_actionspace}.
+With this, we could completely exclude QBlade crashing due to unrealistic pitch changes.
 
 We at first hid this clipping from the replay buffer, but then decided to store the actual values taken, as the models then don't have to also learn the clipping effect.
-We observed better learning when not hiding the clipping from the replay buffer.
+Before we reasoned, hiding this clipping effect would result in a more efficient exploration of the action space, as the policy can predict all actions without restriction and thus easily sweep over the action space.
+However, we observed better learning when not hiding the clipping from the replay buffer.
 
@@
 
-The training is relatively straightforward, however because we decided to abolish expert exploration, we also abolished this.
+The training is relatively straightforward.
+We could see a better performance of the policy in the first steps in contrast to not using pretraining, but the effects were minimal.
+So when we decided to abolish expert exploration, we also abolished this.
 
@@
 
-Because still generally not performing well, we added some techniques which we thought would aid general performance.
+Because still generally not performing well, we added some techniques which we thought could aid general performance.
 
@@
 There hasn't been much work on how to use \ac{PER} in conjunction with \ac{DDPG}, in fact we only found two papers: \cite{houImprovingDDPGPrioritized} and \cite{zhaExperienceReplayOptimization2019}.
-We will at first present the general prioritized experience replay algorithm, then we will discuss the papers and then present how we incorporated \ac{PER}.
+In the rest of this section, we will at first present the general prioritized experience replay algorithm, then we will discuss the papers and then present how we incorporated \ac{PER} ourselves.
 
@@
 This difference, which in Q learning is comparable to the loss of the approximator, tells us how far off the prediction was in this step.
-So, alongside with the experiences, a sampling priority for each item $p_i$ is stored in the replay buffer.
+So, alongside with the experiences, a sampling priority for each item $p_i = \abs{\delta}$ is stored in the replay buffer.
 In theory, we would have to calculate this \ac{TD-Error} every time we update our critic.
 This would mean to feed our entire replay buffer through the critic on every update step.
-Clearly, this is a huge performance hit, which is why in the algorithm, they approximated by only updating the error on samples that are anyways being used for training.
-We calculate the \ac{TD-Error} anyways for the critic update, so we can store this without performance penalty.
+Clearly, this is a huge performance hit, which is why in the algorithm, they approximated the real errors by only updating the error on samples that are anyway being used for training.
+We calculate the \ac{TD-Error} anyway for the critic update, so we can store this without performance penalty.
+There is a risk though, that older experiences which used to produce low errors but under a new critic produce high ones are not being fed soon with this approximation.
 New samples are always stored with maximum priority to make sure they are sampled at least once.
@@
 When replaying, the sampling probability of a sample is set to $P(j) = \frac{p_j^{\alpha}}{\sum_i{p_i^{\alpha}}}$, $\alpha$ being a hyperparameter allowing to fade between uniform sampling ($\alpha=0$) and pure priority based sampling ($\alpha=1$).
-Because we introduced oversampling bias by sampling some experiences more often than others, a correction term is applied to the \ac{TD-Error}: $w_j = \frac{(N*P(j))^{-\beta}}{\max_i{w_i}}$.
+Additionally, usually a small $\epsilon$ is added to avoid zero priorities: $p_i = \abs{\delta} + \epsilon$.
+Because we introduce oversampling bias by sampling some experiences more often than others, a correction term is applied to the \ac{TD-Error}: $w_j = \frac{(N*P(j))^{-\beta}}{\max_i{w_i}}$.
 It is simply multiplied to the real \ac{TD-Error}.
 $\beta$ is a parameter describing how strongly to correct oversampling, where $\beta=1$ is full correction and $\beta=0$ no correction at all.
-This importance sampling is a common technique in reinforcement learning and we have addressed it before in our background part.
+This importance sampling is a common technique in reinforcement learning and we have used it before in our background part in equation \ref{equ:pg_importancesampling}.
 
 In addition to sampling directly based on $\abs{\delta}$, \cite{schaulPrioritizedExperienceReplay2016} propose a variant where they sample according to the rank of an experience: $p_i = \frac{1}{\text{rank}(i)}$ where rank is defined as the position in the replay memory when sorted according to $\abs{\delta}$.
-This method is more robust to outliers and slightly increases performance, but because of the extra computational effort of having to calculate the rank every time, we decided to go for the first method.
+This method is more robust to outliers and slightly increases performance, but because of the extra computational effort of having to calculate the rank every time, we decided to implement the first method.
 Also, \cite{zhaExperienceReplayOptimization2019} measured worse performance of the rank based version in contrast to even plain \ac{DDPG}.
@@
 To avoid oversampling bias in the policy, we only applied \ac{PER}-sampling to the training part of the critic and trained the actor on uniformly sampled data.
-It can be argued, that as the policy $\mu$ is anyways being trained on the state density function of another policy $\rho^{\beta}$, it can deal with that oversampling bias.
+It can be argued, that as the policy $\mu$ is anyway being trained on the state density function of another policy $\rho^{\beta}$, it can deal with that oversampling bias.
 However we doubt adding more bias is beneficial to the training, so we decided to sample uniformly for the policy.
@@
 Because we saw performance problems storing the priorities alongside the experiences and directly sampling from a 100000-item weighted array, we followed the advice of \ac{PER} to use a sumtree datastructure, a binary weighted tree in which each node holds the sum of weights of the children.
-We implemented sampling as a tree walk starting at the root node.
-At every node, we normalized the priorities of the two child nodes to probabilities, did a random draw and recursively repeated until we reached a leaf.
-This recursive sampling technique retains the sample probabilities of the leaf nodes but is cheaper in complexity than to normalize the whole array each timestep and do a random draw over the whole probability distribution.
-All our operations on the replay buffer are in $O(log(n))$ complexity now.
+At first we implemented sampling from this as a tree walk, however found that this way we are in danger of sampling the same item several times.
+In edge-cases, we ended up with a batch consisting of only one item repeatedly in a Q loss explosion, which resulted in batch normalization variation compensation to perform a divide by zero, storing NaN in our weights and effectively ruining the network.
+We replaced it with a draw without duplicates.
+Also, we did not store probabilities right away but kept the unnormalized experiences $p_i = (\abs{\sigma} + \epsilon)^{\alpha}$ and only normalized them to probabilities right before sampling.
 
@@
 There have been some experiments with data augmentation in reinforcement learning \cite{cobbeQuantifyingGeneralizationReinforcement2019}, but it is not often mentioned on tutorials or blog posts.
-We implemented by adding a small noise term to the sampled states, actions and next states \((s, a, s')\) when performing experience replay.
-We didn't really see any indication of bad generalization, but as implementing it wasn't a big issue and normally doesn't bring any disadvantages, we still added this feature and set the noise level to a very small value.
+We implemented adding a small noise term to the sampled states, actions and next states \((s, a, s')\) when performing experience replay.
+We didn't really see any indication of better or worse generalization, but as implementing it wasn't a big issue and we neither saw nor could imagine any disadvantages, we still added this feature and set the noise level to a very small value.
 
 \cite{fujimotoAddressingFunctionApproximation2018} addressed something similar on actions generated by the target policy, see section \ref{sec:double_critics} for this.
+They argued that noise added to the target action smoothes out the critic.
+With the same argumentation, also our replay noise would smooth out Q estimates.
 
@@
 We observed that our Q-Loss, especially after a death, explodes into e15 magnitude values, while most of the times it stayed around e2 magnitude.
-Instable Q-Losses are a general problem with DQN and DDPG, and some literature \cite{fujimotoAddressingFunctionApproximation2018} proposes to use two Q networks, of which only the minimum is used for training.
+Instable Q-Losses are a general problem with DQN and DDPG, however literature normally doesn't speak about Q loss explosions, or if yes than in the form of divergence.
+It is possible that a high learning rate makes the Q function jump away from the target, stepping over the minimum of predicting target Q values and instead predicting higher, then lower, then even higher values.
+This process however is different to what we observed, as during our Q loss explosions mean Q estimates did not jump as well.
+Also, the loss usually jumped up within 1 to 10 training steps and then decreased again, in contrast to a divergence which would somehow constantly and smoothly escalate.
+So we tried to figure out what caused our loss explosions.
 
 \subsection{Huber loss}
-We could already greatly improve our results by using huber-loss \cite{huber1964} for the critic instead of mean squared error loss as described in DDPG paper.
+The \ac{DQN} authors are using Huber-loss \cite{Huber1964} for the critic instead of mean squared error loss as described in DDPG paper.
 In fact, the authors of \ac{DQN} have caused some confusion on this in the \ac{RL}-community.
-In their journal paper \cite{mnihHumanlevelControlDeep2015}, they write \"We also found it helpful to clip the error term from the update [...] to be between -1 and 1\", which could be interpreted as loss clipping.
-However they actually switched to absolute error instead of square error outside those boundaries, which is the definition for huber loss.
-
-Although loss explosions gotten less frequent and smaller in magnitude with this update, we still saw some. In general we could observe an improvement in training, so we kept this active.
+In their journal paper \cite{mnihHumanlevelControlDeep2015}, they write
+\begin{quotation}
+We also found it helpful to clip the error term from the update [...] to be between -1 and 1
+\end{quotation}
+which could be interpreted as loss clipping.
+However later they explained that they actually switched to mean absolute error instead of mean square error outside those boundaries.
+This is the exact definition for Huber loss.
+
+\begin{equation}
+\text{loss} = \frac{1}{N}\sum_{i \in N}\begin{cases} \delta_i^2& \text{if } \delta_i <= 1\\ \abs{\delta_i} & \text{else}\end{cases}
+\end{equation}
+
+Although loss explosions got less frequent and smaller in magnitude with this update, we still saw some. In general we could observe an improvement in training, so we kept this active.
+
+\subsection{Large batches}
+We also had to increase batch size up to 128 to counteract noisy observations.
+A large batch size results in less noisy losses, as the loss is averaged across a high number of samples.
+However, large batch sizes can also limit learning performance, as the network is smoothing across a high number of samples.
+Generally, batch sizes of 32 or 64 have established in deep learning, but we could reduce our Q loss problem slightly by taking a bigger batch size.
 
@@
 They propose a set of three mechanisms that address function approximation error in \ac{DDPG}, and they call it \ac{TD3}.
-At first, they show in section 4.1 that Q learning with a deterministic policy tends to overestimate true Q values.
+At first, they show in their section 4.1 that Q learning with a deterministic policy tends to overestimate true Q values.
 This has been observed in normal Q learning as well, but hasn't been generalized to \ac{DDPG} before.
@@
 They combine this with the target network idea, so effectively we end up with 6 networks now.
-They called this technique Clipped Double Q learning.
+They called this technique \textit{Clipped Double Q learning}.
 We illustrate this in figure \ref{fig:clippedq}.
@@
   \centering
-  \includegraphics[width=\linewidth]{images/TD3-Setup.png}
+  \includegraphics[width=0.7\linewidth]{images/TD3-Setup.png}
   \caption{Clipped double Q learning}
@@
 As they observed that a quickly changing policy introduces instability into the Q learning process, they proposed to update the policy less often than the actor.
-They do this simply by updating the policy only when $t\mod d = 0$ where $t$ is the timestep and $d$ describes how many timesteps to wait before an actor update.
-They call this \textit{delaying} policy updates.
-Lastly they proposed adding noise in the target policy calculations to achieve smoothing around policy results.
+They do this simply by updating the policy only when $t\mod d = 0$ where $t$ is the time-step and $d$ describes how many time-steps to wait before an actor update.
+They call this \textit{delaying policy updates}.
+Lastly they proposed adding noise in the target policy calculations to achieve smoothing around action value estimates.
 So they changed the target updates again to:
@@
 
-Where $\tilde{a}$ is the target action with added noise on it. In their appendix they describe that they furthermore clip this target action to sensible action range.
+Where $\tilde{a}$ is the target action with added noise on it.
+In their appendix they describe that they furthermore clip this target action to sensible action range.
+This is called \textit{target policy smoothing}.
+Summarizing, \ac{TD3} consists of three mechanisms, Clipped double Q learning, delayed policy updates and target policy smoothing.
 
@@
 We also tried to use the maximum and minimum of the two, which yielded more stable results, where minimum seemed better than maximum.
-Finally, we ended up with adding the two errors for prioritisation.
+Finally, we ended up with adding the two errors together for prioritization.
 
 However, we could not get \ac{TD3} to work properly, with or without \ac{PER}, and didn't see any convergence.
-
-\subsection{Batch normalization}
-As in deep learning, batch normalization is found helpful for increasing lr and thus to skip over local maxima \cite{bjorckUnderstandingBatchNormalization}.
-It works by normalizing the outputs between the layers to be of zero mean and unit standart deviation, and for not losing generalization adding an extra learnt parameter per output dimension.
-We did get a performance boost when using this, however we ended up with NaN values after a while of running it, especially when experiencing q loss explosions.
-We can explain this because our prioritized experience replay most likely sampled only one item at a timee.
-To circumvent this, we clipped individual priorities.
-As these unexpected nan-values made training more complicated, we decided to not use batch normalization.
+We thus disabled all three features, but found that using a lower learning rate for the actor has a similar stabilizing effect as actor delays.
+Also, we later found that target policy smoothing helps with a problem in death conditions, but we didn't have it active for most of the training.
 
@@
 
-We were observing many exploding turbines in our experiments, so we had a look at how \ac{DDPG} handles death conditions.
+We were observing many burning turbines and flying-off blades in our experiments, so we had a look at how \ac{DDPG} handles death conditions.
 In \ac{DQN}, death conditions are accounted for by changing the Q target in the terminal step to be $y = r$ instead of $y = r + \gamma \max_a' Q(s', a')$.
@@
 With our $\gamma$ = 0.99, circa 99 times as high as $\int_1^{\infty} 0.99^x \approx 98.5$.
-Though we punish death conditions severely, a death condition will not exceed 3 times the rewards of normal operations due to our clipping.
-So if, after a death condition, a good condition is reached, our algorithm would gladly take that penalty.
+Though we punish death conditions severely, a death condition will not exceed 99 times the rewards of normal operations without causing a Q loss explosion.
+So if, after a death condition, a good condition is reached due to resetting the environment, our algorithm would gladly take that penalty.
 For this reason we introduced the updated death notion
@@
 
-However, now it could be advantageous to die in case of negative Q values.
-Let's suppose we have a terrible policy which under no circumstances produces positive rewards.
-Thus, our Q term would be negative.
-As discussed before, it is likely of high magnitude, so elimiating the high negative Q for a 0 would be a good strategy.
-Thus we need to make sure the Q term is never negative.
-We do this by adding the lower clipping bound to rewards after clipping them.
+This update notion is similar to the update notion from \ac{DQN}.
 So now, instead of $y = r + \gamma Q'(s', \mu'(s'))$ we change it to be $y = r + \gamma Q'(s', \mu'(s')) * (1-d)$ where $d \in {0, 1}$ symbolize the death conditions.
-We haven't found any literature which does a similar thing with \ac{DDPG}
+We believe this addition to \ac{DDPG} to be novel.
+Also, in section \ref{sec:hold_rated_power}, we discuss why it is necessary to also make sure the algorithm sees mainly positive rewards.
+Unfortunately, we didn't know this at the beginning of our experimentation phase so this is not included in our original algorithm.
 
@@
 
-Finally, we found that the solution to our problem lay in qblade.
-In certain conditions, qblade returned observations 20 orders of magnitude away from normal operation.
+Finally, we found that the solution to our problem lay in QBlade.
+In certain conditions, QBlade returned observations 20 orders of magnitude away from normal operation for a single timestep.
 As we based our reward calculation on the observations, the reward was also far off from normal reward.
-This condition happened only every 200k steps or so, and as we did not log every timestep, these spikes didn't end up in our logs.
+This condition happened only every 200k steps or so, and as we did not log every time-step, these spikes didn't end up in our logs.
 We decided to filter out these spikes by clipping rewards and observations after normalization to $[-3, 3]$.
+We reference this clipping by $\text{clip}_{s}$.
+
+\section{Concluding all changes}
+
+We want to quickly summarize all the changes we made.
+We tried several things and abolished some of them because they didn't show results.
+Among the ones abolished, we had \textit{parameter-space noise} due to it not giving a visible advantage.
+We abolished \textit{expert exploration} and \textit{policy pretraining}, because adding human knowledge to the training process was what we wanted to avoid in the first place.
+The advantage of having them was small anyway.
+We tried predicting \textit{action gradients} instead of actions directly, and stopped because results were poor.
+\textit{Feeding past timesteps} caused even more instable Q losses so we didn't use this either.
+Also, we couldn't get \textit{TD3} to run as a whole.
+\textit{Last-layer activations} drifted our policy out of the action space so we deactivated this as well.
+
+We were unsure about the effects of some mechanisms.
+\textit{Data augmentation} has neither proven helpful nor limiting, but because we think it is a good thing, we left it in with small magnitude.
+Same goes for \textit{decaying action noise}.
+We could get \textit{target policy smoothing} to work but only used it on our last experiments.
+
+Other mechanisms have proven helpful.
+\textit{Clipping action gradients} was necessary to stop the agent from destroying blades by too quick pitching movements.
+\textit{Clipping observations} helped removing spikes from faulty simulation results.
+\textit{Large batches} and \textit{Huber loss} added stability to the Q function.
+A \textit{simpler architecture} prevented zero gradients.
+
+Lastly, we have three mechanisms which are of higher complexity.
+\textit{\ac{PER}} is uncommon to \ac{DDPG}, but it has delivered better results for us.
+\textit{Normalization} was critical for learning anything at all and removed a lot of our problems.
+And lastly, we argued why \textit{death conditions} are not properly handled in \ac{DDPG} and proposed a solution.
 
@@
 After our experimentation phase, we will present our algorithm, on which we perform our evaluation.
-At first, we describe how we encapsulated the qblade environment.
+At first, we describe how we encapsulated the QBlade environment.
 At second, we will present our replay buffer, our network structure, our exploration strategy and our update algorithm.
@@
 
-\label{sec:alg_qblade}
+\label{sec:alg_QBlade}
 
 As mentioned before, we set the yaw angle to always 0 degrees, as in reality controlling the orientation of the nacelle is trivial and in our simulations, wind will always come from the front.
-The generator torque on our NREL 5MW turbine has a sensible range of \num{0} to ca \num{4e6} Nm, we used a range of \num{0} to \num{8e6} as action space, as higher torque can be simulated by qblade and this way made it easier for the controller, as it can apply high torques to slow down quickly rotating blades.
-Because changing torque from zero to maximum in one timestep is not realistic for a real turbine, we restrict it to move a maximum of \num{3e5} Nm per one timestep of 0.1 seconds.
+The generator torque on our NREL 5MW turbine has a sensible range of \num{0} to ca \num{4.6e6} Nm, we copied exact value from \cite[Table 7-2]{jonkmanDefinition5MWReference2009}.
+For this it should be noted that the torque in the specification sheet is measured after a gearbox, while qblade needs torque before the gearbox.
+Thus the gearbox ratio of 97 has to be multiplied to the values in the specification.
+Because changing torque from zero to maximum in one time-step is not realistic for a real turbine, we restrict it to move a maximum of ca \num{1.5e6} Nm per one time-step of 0.1 seconds.
+We again took exact values from \cite[Table 7-2]{jonkmanDefinition5MWReference2009}.
 If our agent chooses a value outside of that area, we set it to the closest value inside of that area.
 Blades can be pitched between 0 and 90 degrees, 0 being not pitched out at all and operating at maximum efficiency and 90 resulting in no aerodynamic torque from the rotor whatsoever.
-A sensible pitch motor can only turn the blades at a limited speed, we assume a limit of 5 degrees per second and again set controller inputs to a sensible value inside that.
-The simulation theoretically accepts a full pitch change in one timestep, however as inertia on the blade is so high for such a change, the blades break instantly and the rest of the simulation needs to be reset.
-
-QBlade simulates blade damage, also high rotational speeds cause blades to fall off.
-We however observed problems with our controller when reaching these extreme limits of the simulation In the course of our experiments, we decided to reset the simulation at an rotational speed of 3 rad/s, as our turbine normally operates at ca 0.8 rad/s.
+A sensible pitch motor can only turn the blades at a limited speed, and the NREL specs give a maximum change of 8 degrees per second.
+We again clipped controller inputs to a sensible value inside that.
+The simulation theoretically accepts a full pitch change in one time-step, however as inertia on the blade is so high for such a change, the blades break instantly and the rest of the simulation needs to be reset.
+
+We observed problems with our controller when reaching extreme limits of the simulation in the course of our experiments, thus we decided to reset the simulation at a rotational speed of 3 rad/s, as our turbine normally operates at ca 0.8 rad/s.
 Also, high generator torque inputs cause the simulation to rotate the rotor backwards with a negative energy yield, effectively creating a multimillion-dollar leafblower.
-As this is neither a realistic scenario, we also reset the simulation at -0.5 rad/s.
-We tried to clip at 0, but this resulted in so frequent deaths that learning actions at this border was not possible.
-
-When resetting the simulation, we let it run for 100 timesteps with zero action and state, as the elastic structure needs a bit of time to move from simulation default to elastic equilibrium.
-In this time, simulation results do not make sense yet.
-Also, every 100000 timesteps or after seeing NaN values we reload the project file to maximize simulation stability.
-
-We hid all observations except for power and rotational speed from the algorithm and also merged all pitch controls into a single variable.
+As this is neither a realistic scenario, we also reset the simulation at negative rotational speeds.
+Also, power outputs over 10MW and excessive swinging more than 2 magnitudes higher than normal vibrations are considered a death condition.
+
+When resetting the simulation, we artificially stabilized it to 0.8rad/s rotational speed with the precomp option in QBlade.
+During this stabilizing phase, torque and pitch are fixed to 0.
+After vibrations at simulation start were overcome, we handed over control to our agent and ended the precomp phase.
+Starting with an uninitialized simulation with standing rotor very likely enters small negative rotational speeds at the beginning and thus results in constant resetting.
+
+We hid all observations except for power and rotational speed from the algorithm and also merged all pitch controls into a single variable, implementing collective pitch control.
 
@@
 
-As DDPG is capable of learning from a replay of experiences, we implemented a replay buffer with capacity 1e6.
+We implemented a replay buffer with capacity 1e6.
 To enable \ac{PER}, we needed to sample according to priorities proportional to a metric, in our case absolute \ac{TD-Error}.
-We implemented the replay buffer as a sum tree datastructure, a binary weighted tree in which each node holds the sum of the weights of the nodes below.
-Also, we did not store probabilities right away but kept the unnormalized experiences $p_i = (\abs{\sigma} + \epsilon)^{\alpha}$ with an epsilon of $\num{1e-6}$ and $\alpha = 0.5$.
+We used \ac{PER} with minimum priority $\epsilon = \num{1e-6}$, prioritization intensity $\alpha = 0.5$ and importance sampling correction $\beta = 0.5$.
 
@@
 All states and rewards seen by the network are normalized to be between $[-1, 1]$ 90\% of the time and are clipped to $[-3, 3]$.
-Actions are normalized to $[-1, 1]$.
+Actions are normalized to $[-1, 1]$ and clipped to not extend sensible action gradients.
 Learning rate for both actor and critic are 1e-4 and we use adam optimization.
@@
 We regard death conditions in the Q update as in \ac{DQN}.
-Also, we use huber loss instead of MSE.
+Also, we use Huber loss instead of MSE.
 All in all, algorithm \ref{alg:ourddpg} summarizes all improvements:
@@
   \For{t = 1, T}{
-    Select action $a_t = \mathrm{clip}(\mu(s_t;\theta^{\mu}) + \mathcal{N}_t)$ according to the current policy and exploration noise\\
-    Execute action $a_t$ in $\varepsilon$ and observe $r_t$, $d_t$ and $s_{t+1}$ and normalize these\\
+    Select action $a_t = \mathrm{norm}(\mathrm{clip}_a(\mu(s_t;\theta^{\mu})) + \mathcal{N}_t)$ according to the current policy and exploration noise\\
+    Execute action $a_t$ in $\varepsilon$ and observe $r_t$, $d_t$ and $s_{t+1}$.
+    Clip and normalize rewards and states according to $\mathrm{clip}_s$ and $\mathrm{norm}$ \\
     Store transition with highest priority $(s_t, a_t, r_t, s_{t+1}, d_t, \max_i(p_i))$ in $\mathcal{R}$\\
@@
     Apply importance sampling $\delta_i = \delta_i w_i$\\
-    Update critic by minimizing the huber loss: $L = \frac{1}{N} \sum_i \begin{cases} (\delta_i)^2 & \text{if } \delta < 1\\ \abs{\delta_i} & \text{else}\end{cases}$\\
+    Update critic by minimizing the Huber loss: $L = \frac{1}{N} \sum_i \begin{cases} (\delta_i)^2 & \text{if } \delta < 1\\ \abs{\delta_i} & \text{else}\end{cases}$\\
     Sample a random minibatch of $N$ states $(s_i)$ from $\mathcal{R}$ uniformly\\
@@
 
-For our evaluation, we used a machine with an Intel\textsuperscript{\textregistered} i7 4-core processor, 16GB RAM and a Nvidia 1080 GTX with 8GB as accelerator.
-On this system, a fully realistic run with 300k steps would take roughly 200 hours, of which almost everything would be used by qblade.
+For the evaluation, we used a machine with an Intel\textsuperscript{\textregistered} i7 4-core processor, 16GB RAM and a Nvidia 1080 GTX with 8GB as accelerator.
+On this system, a fully realistic run with 300k steps would take roughly 200 hours, of which almost everything would be used by QBlade.
 Partly this is because we didn't manage to enable hardware acceleration because of compilation problems, but computing a fluid simulation is always time-consuming.
-We could bring the computation time down to 12 hours per run by deactivating wake calculations and using a different structural computation method in qblade.
+We could bring the computation time down to 12 hours per run by deactivating wake calculations and using a different structural computation method in QBlade.
 Wake calculations are responsible for calculating the effects of the rotor on the surrounding air.
@@
 Effectively, this increased instability of the system, as now the rotor does not get slowed down at high rotational speeds, but otherwise we would not have been able to evaluate our work.
+Also, during most of our training we were able to fit a single run into the 16 GB RAM of the machine, but after optimizing our code for memory usage, we could accommodate 3 easily or 4 with heavy swapping penalties on one of the runs.
 
@@
 The mean is displayed as a solid line, while the confidence is plotted as a colored background.
+
+Unfortunately we couldn't evaluate all our improvements in a proper grid search and measure their impact.
+We will limit us to very few scenarios, but hope to discuss them thorough enough to provide valuable insights.
 
@@
 
-We look at two different learning rates for the critic network, 1e-3 (a) and 1e-4 (b). The latter is what was recommended in the DDPG paper.
+We look at two different learning rates for the critic network, 1e-3 and 1e-4. The latter is what was recommended in the DDPG paper.
 
@@
 
+Our results are plotted in figure \ref{fig:critic_lr}.
 Each of the controllers reaches close to theoretical optimum and is able to hold it for a number of epochs.
-With learning rate 1e-3, we reach it somewhere around episode 10, with learning rate 1e-4 a bit later around epoch 25.
-We also tried 1e-5, but we could not see any convergence with this, so we ommitted the plot.
-The time we are holding it before divergence also raises with learning rate.
-The most aggressive learning rate reached 1828.8 average epoch reward between epoch 7 and 11 and a single maximum of 1917.5 points.
-On the lower learning rate, it held an average of 1896 points from episode 30 to 56 and reached a maximum of 1968.4 points.
+With learning rate 1e-3, we reach it somewhere around epoch 10, with learning rate 1e-4 a bit later around epoch 25.
+We also tried 1e-5, but we could not see any convergence with this, so we omitted the plot.
+The time we are holding it also raises with learning rate.
+The most aggressive learning rate reached 1828.8 average epoch reward between epoch 7 and 11 and a single maximum of 1917.5 epoch reward.
+On the lower learning rate, it held an average of 1896 points from epoch 30 to 56 and reached a maximum of 1968.4 points.
 The maximum we could reach with our hand-designed PID controller was 1981 points, where the 19 points were lost through the swing at the beginning and then rotational speed was held perfectly.
+A totally random policy reached 440 points on average.
 
@@
 
+So, we were able to reach sensible control for a number of timesteps.
+4 epochs on the high learning rate and 26 epochs on the low one ranged in an area which is close to optimal.
+We widely outperformed the random policy and also we needed quite some time of trial and error until we were able to beat our learnt policy in this task with our handmade controller.
+
 Interesting to see is the strategy which the agent found.
@@
 However, this insight is above our human knowledge of windturbines.
-Furthermore should be noted that rotational speed is still varying between 0.75 and 0.85 and the pitch is ranging from 15 to 25 degrees (note the scaling of the pitch and torque).
+Furthermore should be noted that rotational speed is still oscillating between 0.75 and 0.85 and the pitch is ranging from 15 to 25 degrees (note the scaling of the pitch and torque).
 In a realistic scenario, oscillations of this magnitude would likely destroy the turbine very soon, as pitching the blades around this quickly creates strong material stress.
+This makes the algorithm not usable as it is.
 
@@
   \includegraphics[width=0.5\linewidth]{images/plot_critic_1e-3_qloss.png}
-  \caption{Critic q loss during divergence}
+  \caption{Critic Q loss for lr=1e-3}
   \label{fig:critic_qloss}
@@
 
-With the aggressive learning rate, we can clearly see how quickly the algorithm is diverging.
-After a short period of convergence around epoch 10, the high learning rate steps the critic out of the minimum and causes instabilities.
-With even higher learning rates, we observed not reaching convergence at all.
-Interestingly, we can not see such divergence in the loss.
-In normal machine learning, a too high learning rate would cause a phenomenon of jumping across a minimum, taking gradient steps greater than the width of the local minimum.
-This would reduce in an increase in loss.
-Here, however, the q loss stays small on such divergence, as plotted in figure \ref{fig:critic_qloss} for the diverging run of critic lr=1e-3.
-This is because we are generating our prediction targets based on delayed versions of the same net, as we do not know a ground truth.
-As this delayed target net will diverge with the original net if divergence happens slow enough, we do not see it.
-
-From this we see that higher learning rates cause a more stable learning, but reach convergence later. The results suffer from strong oscillations both with the low and high learning rate.
+Both algorithms converge to doing nothing after a while, which indicates a divergence in the algorithm.
+Interestingly, we can not see our divergence in the loss.
+A divergence caused by a too high learning rate would cause a phenomenon of jumping across a minimum, taking gradient steps greater than the width of the local minimum and skyrocketing both the loss and Q estimates, and we observed plenty of these scenarios during our experimentation phase.
+Here, however, the Q loss stays constant on such divergence, as plotted in figure \ref{fig:critic_qloss} for the diverging run (a) with the high learning rate.
+So, what we have is not a traditional Q divergence.
+Instead, we must have learned some maximum in the Q function, towards which the policy now optimizes.
+This minimum is doing nothing and leaving the turbine still.
+In this minimum, death conditions are strongly reduced in number, as turbines tend to not die when they don't rotate.
+This minimum could be learned after learning the maximum towards which the policy converged before but on samples generated before the policy was able to reach this maximum.
+Learning about the negative nature of death conditions is a slower process, as death conditions in our dataset are relatively sparse, so learning the penalty happens after optimizing the policy towards a good behavior.
+With the lower learning rate 1e-4, this happens even later.
+With even lower learning rates however, we have accumulated enough death condition for Q to already learn them before optimizing the policy towards a sensible state, so a learning rate of 1e-5 does not bring us to any convergence (not plotted here).
+A solution for this would be to train the network for a longer time until the death conditions from the beginning ended up outside of the replay buffer.
+When that happened, movement towards higher rotational speeds could be rated as actually viable by Q without Q already punishing the near certain deaths of the policy.
+Training for longer however alleviates our problem of vanishing gradients, as the policy is pushed towards actions at the borders of the action space when bringing the turbine to a standstill.
+Training for a longer time, we observed very small gradients.
+We don't know whether it would recover from the low gradients, and trying is outside of our computational possibilities.
+We will call the process of converging to doing nothing and strictly avoiding any deaths \textit{panic mode} in the rest of the evaluation.
+
+From this we see that higher learning rates cause a more stable learning, but reach convergence later.
+The results suffer from strong oscillations with both the low and high learning rate.
+We do not expect the Q function to learn resonant behavior of the windturbine in the few training steps which we allowed it to run, so we are not surprised that the policy can not yet counteract this.
+In a production systems, oscillations of the magnitude which we observed would wear the turbine strongly and would likely destroy it within a few days.
 
 \section{Hold rated power}
+\label{sec:hold_rated_power}
 A more difficult task to solve, but also a more realistic one it to hold a rated power.
@@
 Again, the agent gets a reward of 1 for hitting 5MW perfectly, and 0 for having no power at all, so a perfect policy would get something a little below 2000.
-Unfortunately, we did not get any policy to achieve something close to optimal, the best score we achieved was 1500 points with an average of 5 kills per episode.
+Unfortunately, we did not get any policy to achieve something close to optimal, the best score we achieved was 1500 points with an average of 5 kills per epoch.
 Unfortunately, this result wasn't very stable, so we can not plot an average over many runs.
-An example for a relatively successful epoch is figure \ref{fig:crashing_turbines}.
+An example for an epoch with high reward is figure \ref{fig:crashing_turbines}.
 Here, the controller maxed out torque and tried to hold the turbine stable with pitch.
+It reaches rotational speeds above 2rad/s though.
 Power output for these insane regions was up to 9MW, which would have instantly roasted any electronics in a real wind turbine.
@@
 
-At the end, most controllers converged to leaving pitch at zero and left the turbine to die regularily with maximum generator torque.
-Though this is not an optimum in the global reward function, apparently this became an optimum in the Q estimate.
-Looking back at our explanation from section \ref{sec:death} where we gave some options on why an agent could favor a death over returning to a sensible region, we can exclude the case that good Q values after a death (because after a reset the agent returns to sensible area) propagate.
-So we looked at the case of negative Q estimates.
-When plotting Q estimates, we found that indeed Q is sometimes predicting negative values, especially in early stages of training.
-
-A possible explanation for favouring deaths could be that our death penalties are spikes in the reward functions.
+At the end, most controllers converged to leaving pitch at zero and left the turbine to die regularly with maximum generator torque.
+So again, the controller has learned something which obviously isn't an optimum in the global reward function.
+We again don't see divergence indicators, so we need to ask again how this relatively poor strategy became an optimum in the Q estimate.
+
+A possible explanation for favoring deaths could be that our death penalties are spikes in the reward functions.
 As explained in section \ref{sec:rewardfunctions}, we add a constant penalty upon death but not in the steps before.
@@
 
-We are unsure why the controllers showed such a bad performance, and couldn't get to fix this problem anymore.
-
 \begin{figure}
@@
 
+However, there is another, more credible explanation to the problem.
+If you remember section \ref{sec:death}, we explained how in the Q update, the advantage of a single result is roughly 100 times lower than accumulated future rewards in the update notion to Q: $Q(s, a) = r + \gamma Q(s', a')$.
+There is a case in which it could be advantageous to die.
+Let's suppose we have a terrible policy which very likely produces rewards below zero, or an environment in which it is difficult to obtain rewards greater than zero.
+Thus, our Q term would likely be negative, as however we act, we will incur negative rewards.
+As discussed before, Q is likely of high magnitude, so eliminating the high negative Q for a 0 would be a good strategy.
+A policy might learn this strategy and result in a behavior similar to what we observed.
+Thus we need to make sure the Q term is never negative.
+We propose a simple method, we circumvent this by adding the lower clipping bound to rewards after clipping them.
+This results in the policy seeing only positive rewards, which would cause Q estimates to also be positive.
+
+Our environment is in fact an environment in which is is difficult to achieve positive results.
+In contrast to the hold speed task, a small change in rotational speed does not end up in a small change in reward, but the turbine is required to first spin the turbine up to cut-in speed and then torque the generator up.
+This results in the average Q estimate to be below zero, as executing this chain of actions happens far less likely than to for example kill a turbine.
+When plotting average Q estimates (not shown here), we in fact see negative Q estimates.
+
+We evaluate whether our explanation with the negative Q parts holds by running once with rewards in the range of $[0, 6]$ and once rewards between $[-6, 0]$.
+We see the result of the direct comparison in figure \ref{fig:death_norm}.
+Unfortunately, we can only perform a single run of 20 epochs for evaluating these theories due to time constraints.
+To improve performance, we performed 2 training operations per step, so epoch 20 is comparable to epoch 40 from our previous setting.
+In our results, we can see that the controller which gets only positive rewards performs widely better, and nearly without killed turbines.
+Both controllers fail and converge to doing nothing then.
+Because of our low number of runs and short evaluation time, we only see this as a hint to support our thesis, not a full proof.
+
+\begin{figure}
+\centering
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.8\linewidth]{images/death_killcount.png}
+  \caption{Number of deaths per epoch}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.8\linewidth]{images/death_rewards.png}
+  \caption{Cumulative reward per epoch}
+\end{subfigure}
+\caption{Comparing adding and subtracting a constant factor to rewards}
+\label{fig:death_norm}
+\end{figure}
+
+We do not evaluate our spike theory again but instead run a training on the hold power task with our new normalization addition.
+
+\section{Hold rated power with death conditions}
+
+We did a last evaluation run similar to section \ref{sec:hold_rated_power}, however with the aim of holding 3.5MW instead of 5MW, as we deemed this task slightly easier.
+Also we changed the critic size to 128 and 64 neurons because in previous runs we observed going to panic mode quickly.
+Using a bigger and thus slower to train network made us hope to achieve sensible results later but with more stability.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.5\linewidth]{images/hold_power.png}
+  \caption{Results of the hold power task}
+  \label{fig:holdpower}
+\end{figure}
+
+In figure \ref{fig:holdpower} we can see at least some convergence happening.
+Though no agent reached anything near perfect, we are at least seeing some improvements over a totally random policy, which gets an average reward of 410 points.
+If we look in detail at actions the controller took, as displayed in figure \ref{fig:holdpowerepoch}, we see that it even managed to smooth out vibrations.
+However, it stabilized at 4 MW instead of 3.5MW.
+We would again explain this miss to the early stage in training we are looking at.
+Also only one of our 4 runs managed to stabilize this well, the others still had severe oscillations occur.
+This instability in training could be due to the bigger network size, however our better results could also be due to the bigger network size.
+In general machine learning, training bigger networks is usually more difficult but provides the ability to tackle more complex problems.
+In supervised learning, overfit is a clear indication of a too large network and can be measured directly by validation performance.
+As we do not have access to such measures in reinforcement learning, rating generalization is less trivial.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.5\linewidth]{images/hold_power_epoch.png}
+  \caption{An epoch from the hold power task}
+  \label{fig:holdpowerepoch}
+\end{figure}
+
 \section{Pendulum}
-To exclude the possibility that our implementation is buggy, we decided to run our DDPG algorithm on the Pendulum task.
-With our hyperparameters, we could not observe convergence and only switching back to the algorithm of \ac{DDPG} which we already tried in \ref{sec:gymexperiments}, the agent managed to balance out the pendulum.
-Thus we can not exclude misimplementation as a bug might only occur with our set of hyperparameters, however we suspect that windturbines are just not very similar to pendulums.
+To exclude the possibility that our implementation is buggy, we decided to run our DDPG algorithm on the Pendulum task again.
+With our hyperparameters, we could not observe sensible results.
+Only switching back to the original algorithm of \ac{DDPG} which we already tried in \ref{sec:gymexperiments}, the agent managed to balance out the pendulum.
+Thus we can not exclude mis-implementation as a bug might only occur with our set of hyperparameters, however we suspect that windturbines are just not very similar to pendulums.
 
@@
 In fact, it would destroy most windturbines relatively quickly and does not yield sensible energy.
-Still we showed that it is possible, that existing algorithms however exhibit some significant problems and need reworking.
-Also we showed that we are not able to perform this reworking in the given amount of time.
+We have poor training stability, as runs are doing very different things.
+The best words for our results are that at least our agents learned something.
+We can outperform random policies easily until panic mode sets in, however this is a very low milestone.
+We conclude that we are not able to reach sensible windturbine control in the scope of this work.
 
@@
 The OpenAI Gym suite is the common suite for evaluating reinforcement learning algorithms, and typical problems in this are balance problems (all cartpole* and pendulum* tasks) and actuation tasks (gripper*, reacher*, cheetah*).
-Balance tasks could be argued to resemble our windturbine in a way, as we want to balance an output aswell.
+Balance tasks could be argued to resemble our windturbine in a way, as we want to balance an output as well.
 However a windturbine is apparently different enough so we can not optimize it with the same algorithm which optimized the other games.
 It exhibits resonant frequencies in addition to the optimization problem which the controller needs to keep under control.
-Our experiments showed that it didn't.
-We were able to solve easy tasks, in which resonances of high magnitude still lead to acceptable results, however as soon as it got more difficult, high swinging lead to our control task not being accomplished.
-Also, noisy sensor measurements were a more difficult task to learn.
-We could not implement the same size of network and thus not achieve the same learning potential as the authors in \ac{DDPG}.
+The multitude of resonating elements on a windturbine results in observations being noisy.
+And also, wind turbines shatter when exceeding operational ranges, which none of the OpenAI gym examples does.
+This difference made it difficult to train our networks, and without reformulation of the original algorithm wasn't possible.
+
+We did not fix our panic mode after some epochs neither could we get stable convergence on the more difficult task.
+We were not able to create a definite solution to our vanishing gradient problem and thus had to train with smaller networks than the original authors.
+We did not properly evaluate all our changes in a grid search and comparison of the individual effects of our changes, so we can not say which change has which effect.
+In fact, our entire evaluation shows a total of 14 runs, so none of our claims are statistically significant.
+
+However, we at least achieved some things.
+We could achieve partially sensible control and could outperform a random policy despite the problems we were facing.
+When solving especially the vanishing gradient and performance problems, we believe that achieving high quality results is possible.
 
@@
 
-Just because we didn't manage doesn't means it's impossible.
-We could imagine that with some additional work, learning stable policies would be possible.
-We have some ideas how we could achieve this result.
+Just because we didn't manage doesn't mean it's impossible.
+We can imagine that with some additional work, learning stable policies is possible.
+We have some ideas how we could achieve this result and what was missing in our work.
 
@@
 
-It was limiting to perform all our experiment on the tiny machine.
-Limiting factor in this is currently the simulation, even with the lowered accuracy, which currently uses 70\% of the computation time, while we only use 30\% to train our networks.
-As currently, linux hardware acceleration is broken in qblade, a nice idea could be to fix this and to run future experiments on a cluster.
+It was limiting to perform all our experiment on the tiny machine, not being able to calculate even by far as many iterations as the authors of prominent papers and having very limited ability to try different parameters in parallel.
+Limiting factor in this is currently the simulation, even with the lowered accuracy, it currently uses 90\% of the computation time, while we only use 10\% to train our networks.
+As currently, linux hardware acceleration is broken in QBlade, we recommend any future researcher to fix this and to run future experiments on a more powerful machine.
 This would also allow efficient hyperparameter sweeps, as we saw that hyperparameter choice greatly affects the outcome of the simulation.
@@
 With our computational power we couldn't run anything above 0.3 million steps.
-But we also chose a less complex architecture which should be faster to train.
+But we also chose a less complex architecture which should be faster to train, and did reach first results as early as 20k steps.
 
@@
 
+\section{Activation problems}
+
+We mentioned some ideas in section \ref{sec:activations} we discussed why changing the policy gradient equation or preventing the Q network from overestimating unseen values in combination with the constant factor on the activations could solve our problem with vanishing gradients.
+Future work could try one of these methods.
+
 \section{Validate against OpenAI-Gyms}
 
-We did not validate our combination of \ac{PER} and \ac{DDPG}.
-Neither did we validate our addition of death conditions.
-For our task, it gave better results, but it would be interesting to know whether our algorithm might also yield better results on general reinforcement learning tasks.
+We did not validate our combination of \ac{PER} and \ac{DDPG} through an excessive evaluation.
+Neither did we validate our addition of death conditions that way.
+For our task, it gave better results in the first stages of training, but it would be interesting to know whether these results are reproducible and whether our algorithm might even yield better results on general reinforcement learning tasks.
 Running through the suite of OpenAI Gyms would be an interesting work and could show whether our critique on \ac{DDPG} was justified.
@@
 We did all our training under the simple hold-speed or hold-power reward function.
-A realistic reward function however would incorporate power generated and long-term damages occured.
+A realistic reward function however would incorporate power generated and long-term damages occurred.
 Especially the last part is tricky, as it would somehow have to be included into the scalar reward function.
 An interesting approach could be to try to train under multi-dimensional reward functions.
-To our knowledge, there haven't been efforts to accomodate this into Q-learning.
+To our knowledge, there haven't been efforts to accommodate this into Q-learning.
 Also, we could imagine interpolating between different reward functions during training.
@@
 In early stages of training, the network is conditioned to only output anything sane, while in later training it could optimize complex goals such as damage prevention more.
-This could be achieved by increasing the magnitude of the reward function further down in training, like after focussing on the big chunky goals zoom in on more fine-grained problems.
+This could be achieved by increasing the magnitude of the reward function further down in training, like after focusing on the big chunky goals zoom in on more fine-grained problems.
 
@@
 
+\chapter{Conclusion}
+
+We showed that it is not trivially possible to extend existing continuous control reinforcement learning algorithms to provide windturbine control.
+However, we were able to reach near sensible control in two simplistic control tasks.
+To achieve this, we needed to modify existing algorithms to the pelicularities of windturbine control.
+Though our main task is not solved yet, we provided a promising strategy to handle death conditions in \ac{DDPG}, which can alleviate the algorithm to scenarios in which the environment might step out of sensible operational ranges.
+We integrated \ac{PER} into \ac{DDPG} and at least gave a theoretical explanation why it works.
+We showed how to integrate QBlade into a reinforcement scenario.
+Though our results can not outperform an industry controller, we made some progress towards solving reinforcement learning on wind turbines.
+
 \end{document}

### 2019-12-09 python/automated_hparam_search.py

@@ -77,7 +77,7 @@ import os
 ]'''
 
 # Hold lower power, actor lr 5e-5
-experiments = [
+'''experiments = [
   # {"id": 70}, # Disaster
   # {"id": 71, "critic_lr": 5e-5}, # Disaster
   # {"id": 72, "critic_lr": 1e-5}, # Disaster
@@ -88,6 +88,13 @@ experiments = [
   {"id": 77, "prioritized_experience_replay_alpha": 0.8, "prioritized_experience_replay_beta": 0.8},
   {"id": 78, "prioritized_experience_replay_alpha": 0.3},
   {"id": 79, "prioritized_experience_replay_alpha": 1, "prioritized_experience_replay_beta": 0.8},
+]'''
+
+experiments = [
+  {"id": 81},
+  {"id": 82},
+  {"id": 83},
+  {"id": 84},
 ]
 
 def run_one(experiment):

### 2019-12-09 python/ddpg/ddpg.py

@@ -20,6 +20,9 @@ class DDPG:
     self.device = torch.device("cuda" if torch.cuda.is_available() and hparams.use_gpu else "cpu")
     self.dtype = torch.float
 
+    if(hparams.target_policy_smoothing):
+      assert(hparams.normalize_actions)
+
     self.hparams = hparams
     self.obs_dim = hparams.obs_dim
     self.act_dim = hparams.act_dim
@@ -92,7 +95,12 @@ class DDPG:
     self.critic_loss_function = lambda target, pred, weights: generic_loss_function(target, pred, weights)
 
     # Buffor for experience replay
-    self.replay_buffer = BasicBuffer(hparams.buffer_maxlen, self.device, self.obs_dim, self.act_dim, hparams.prioritized_experience_replay_alpha, self.dtype)
+    self.replay_buffer = BasicBuffer(hparams.buffer_maxlen,
+                                     self.device,
+                                     self.obs_dim,
+                                     self.act_dim,
+                                     hparams.prioritized_experience_replay_alpha,
+                                     self.dtype)
 
     # Noises
     # Random exploration
@@ -106,8 +114,8 @@ class DDPG:
     elif(self.hparams.random_exploration_type == 'uncorrelated'):
       self.random_exploration_noise = GaussianNoise(
         self.act_dim,
-        self.hparams.random_exploration_mean,
-        self.hparams.random_exploration_variance)
+        self.hparams.random_exploration_mu,
+        self.hparams.random_exploration_sigma)
     else:
       self.random_exploration_noise = NoNoise(
         self.act_dim)
@@ -147,9 +155,16 @@ class DDPG:
                                  self.hparams.act_low,
                                  self.hparams.act_high,
                                  self.device,
-                                 self.dtype)
+                                 self.dtype,
+                                 self.hparams.reward_offset)
 
-    self.logger = QBladeLogger(self.hparams.logdir, self.hparams.log_steps, self.hparams.run_name, self.hparams.obs_labels, self.hparams.act_labels)
+    self.logger = QBladeLogger(self.hparams.logdir,
+                               self.hparams.log_steps,
+                               self.hparams.run_name,
+                               self.hparams.obs_labels,
+                               self.hparams.act_labels,
+                               self.hparams.feed_past)
+
     for key, value in self.hparams.values().items():
       self.logger.writer.add_text(key, str(value), 0)
 
[…]

### 2019-12-09 python/ddpg/models.py

@@ -77,6 +77,9 @@ class Actor(nn.Module):
         self.action_dim = action_dim
         self.simple = simple
 
+        self.f_inner = torch.nn.ReLU()
+        self.f_last = torch.nn.Tanh()
+
         if(batch_normalization):
             self.bn1 = nn.BatchNorm1d(hidden_size_1)
             self.bn2 = nn.BatchNorm1d(hidden_size_2)

### 2019-12-09 python/ddpg/normalizer.py

@@ -4,13 +4,14 @@ import numpy as np
 from .utils import BasicBuffer
 
 class Normalizer:
-  def __init__(self, obs_dim, act_dim, act_low, act_high, device, dtype=torch.float):
+  def __init__(self, obs_dim, act_dim, act_low, act_high, device, dtype=torch.float, reward_offset=0):
     self.device = device
     self.dtype = dtype
     self.obs_dim = obs_dim
     self.act_dim = act_dim
     self.act_high = np.array(act_high)
     self.act_low = np.array(act_low)
+    self.reward_offset = reward_offset
 
     # Calculate action normalization
     m = (self.act_low - self.act_high) / (-1 - 1)
@@ -87,29 +88,42 @@ class Normalizer:
 
     if(gpu):
       m, c = self.reward_normalizer_params_gpu
-      reward = torch.clamp((reward - c) / m, -3, 3) + 3
+      reward = torch.clamp((reward - c) / m, -3, 3) + self.reward_offset
     else:
       m, c = self.reward_normalizer_params
-      reward = np.clip((reward - c) / m, -3, 3) + 3
+      reward = np.clip((reward - c) / m, -3, 3) + self.reward_offset
     
     return reward
 
   def denormalize_reward(self, reward, gpu=False):
     m, c = self.reward_normalizer_params_gpu if gpu else self.reward_normalizer_params
-    return (reward - 3) * m + c
+    return (reward - self.reward_offset) * m + c
 
-  def calc_normalizations(self, replay_buffer, normalization_extradata=None):
+  def calc_normalizations(self, replay_buffer, normalization_extradata=None, feed_past=0):
     # Load additional data if wanted
     if(normalization_extradata):
-      tmpbuffer = BasicBuffer(100000, torch.device('cpu'), self.obs_dim, self.act_dim)
+      
+      if(feed_past):
+        obs_dim = int(self.obs_dim/(feed_past+1))
+      else:
+        obs_dim = self.obs_dim
+
+      tmpbuffer = BasicBuffer(100000, torch.device('cpu'), obs_dim, self.act_dim)
       tmpbuffer.load(normalization_extradata)
       extrastates = [state for state, _, _, _, _ in tmpbuffer.get_buffer()]
       extrarewards = [reward for _, _, reward, _, _ in tmpbuffer.get_buffer()]
+
+      if(feed_past):
+        extrastates = [np.array([s for i in range(0, feed_past+1)]).flatten() for s in extrastates]
 
[…]

### 2019-12-09 python/ddpg/utils.py

@@ -64,7 +64,7 @@ class GaussianNoise:
         self.variance = variance
 
     def get_noise(self, _t=0):
-        return np.random.randn(dim) * variance + mean
+        return np.random.randn(self.dim) * self.variance + self.mean
 
 
 class UniformNoise:
@@ -360,16 +360,20 @@ class BasicBuffer:
 
 class QBladeLogger:
 
-    def __init__(self, logdir, log_steps, run_name, obs_labels = {}, act_labels = {}):
+    def __init__(self, logdir, log_steps, run_name, obs_labels = {}, act_labels = {}, feed_past=0):
         self.logdir = logdir
         self.log_steps = log_steps
         run_name = run_name if run_name else str(datetime.now())
         self.writer = SummaryWriter('%s/%s' % (logdir, run_name))
         self.obs_labels = obs_labels
         self.act_labels = act_labels
+        self.feed_past = feed_past
 
     def logObservation(self, step, observation, prefix='obs'):
-        for i in range(0, len(observation)):
+        length = len(observation)
+        if(self.feed_past):
+            length = int(length/(self.feed_past+1))
+        for i in range(0, length):
             if i in self.obs_labels:
                 label = self.obs_labels[i]
             else:
